---
title: "Evaluate annotations and their effect on document scores"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Evaluate annotations and their effect on document scores}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ldfreq)
```

## Start with the analysis that uses the annotations

Suppose a study compares the diversity of nouns across texts. A POS tagger
decides which words enter each noun sequence. Two taggers may select the same
number of nouns while selecting different words. A count comparison alone
cannot reveal whether those selections change the diversity score.

`lexdiv_evaluate_annotations()` connects three checks: agreement with an
explicit reference at each source position; label-specific detection errors
and missingness; and document coverage needed to interpret downstream metrics.
The example below then uses the existing `lexdiv_metrics()` to compare noun
TTR. The question determines that selection and metric; the evaluator does
not choose them for you.

This design follows the distinction between feature-level annotation
evaluation and effects on text indices in Kyle and Eguchi (2024),
[doi:10.1016/j.rmal.2024.100120](https://doi.org/10.1016/j.rmal.2024.100120).
It does not reproduce their dependency analysis or establish a universal
accuracy threshold. Our sentences and labels are authored illustrations,
including deliberately wrong and unavailable labels. They are not independent
human annotations or a performance benchmark for English or Japanese.

## Run the complete example offline

The installed script requires only ldfreq and its ordinary dependencies.
It does not invoke Python, download a model, load a corpus, or write files.

```{r}
source(system.file("examples", "annotation-evaluation.R", package = "ldfreq"))
example <- annotation_evaluation_example
evaluated <- example$evaluation
evaluated$documents[, c("document_id", "tokens", "reference_available",
  "prediction_available", "agreement", "disagreement", "reference_only", "prediction_only")]
```

The five documents include English and Japanese examples, an unresolved
reference label, an unavailable prediction, and an empty document. An empty
document stays in the roster and has undefined coverage, not zero coverage.

## Supply your own predictions and reference

Each input comes from `lexdiv_import_annotations()`. Supply all token surfaces,
including punctuation, with original segment text and consecutive token IDs.
Unavailable labels are `NA_character_`; do not delete the corresponding rows.
Extra columns can hold separate reviewer judgments, unresolved/unreviewed
statuses and prediction-failure reasons. The evaluator preserves those columns
in its complete inputs but does not interpret their names or adjudicate them.

```{r}
head(example$inputs$reference$tokens)
example$inputs$reference$segments
example$inputs$reference_info
```

In real work, replace the authored source, predictions and labels with your
local data. Import them separately, recording the actual analyzer and dictionary
versions, token unit and normalization. Assign the reference only after following
your annotation protocol. Retain original individual judgments separately from
adjudication, and record who made them. `model_exposure = "not_shown"` and
`evaluation_role = "held_out"` are declarations, not checks that establish
blinding, independence or absence of training overlap.

Evaluate one categorical column at a time with a complete label inventory:

```{r}
evaluated_again <- lexdiv_evaluate_annotations(
  predicted = example$inputs$predicted,
  reference = example$inputs$reference,
  column = "upos",
  labels = example$inputs$labels,
  reference_info = example$inputs$reference_info,
  context_chars = 12L
)
stopifnot(identical(evaluated, evaluated_again))
```

For an actual UPOS analysis, use your full UPOS inventory; for Penn XPOS or a
Japanese POS scheme, supply that scheme's labels and definitions. `labels`
is the allowed inventory, not a filter of the feature you want to report.
Unexpected labels cause an error instead of disappearing from the evaluation.
Both tables must use comparable label meanings. The function does not map
different tagsets, and matching label strings alone does not establish equivalence.

The two imports must describe the same document/segment IDs, exact segment
text and token surfaces/positions. Different document or segment order is
allowed and matched by IDs; a changed token or segment boundary is rejected.
Do not force a split/merged word into a one-to-one match by reassigning row
numbers. That requires a separate boundary-alignment analysis. See
[Japanese annotations](japanese-annotations.html) for an external morphology
input example.

## Interpret label-specific errors with their denominators

```{r}
evaluated$labels
evaluated$summary[, c("tokens", "reference_coverage", "prediction_coverage",
  "agreement_among_paired", "matches_among_reference_available")]
```

For a label such as `NOUN`, a true positive (TP) is a noun in both inputs.
A false positive (FP) is a predicted noun with an available non-noun reference.
A false negative (FN) is a reference noun with a different **or missing**
prediction. Precision is `TP / (TP + FP)`, recall is `TP / (TP + FN)`, and
F1 is `2 * TP / (2 * TP + FP + FN)`.

A missing reference is not a non-noun. Its prediction contributes to
`predictions_without_reference`, not FP. A missing prediction on a known noun
contributes to `missing_predictions` and FN. Thus `agreement_among_paired`
describes only positions with both labels, while
`matches_among_reference_available` also includes reference positions the
predictor failed to label. All zero-denominator rates remain `NA`. Punctuation
is included in the overall rates; report the specific classes relevant to
your question rather than relying on one pooled agreement value.

These are unweighted occurrence counts for the supplied sample. If references
were checked only on predicted nouns, noun precision is conditional on that
reviewed subset and false negatives elsewhere cannot be estimated. Sample
complete segments, including non-candidates, to assess detection failures.
Keep any deliberately oversampled rare-feature strata separate from a
population estimate unless their sampling probabilities are accounted for.

## Return from a false positive or negative to the source

All occurrences have context, including those the predictor did not mark as
the target. The following extracts noun errors while excluding unavailable
references from the FP calculation:

```{r}
pairs <- evaluated$pairs
known <- !is.na(pairs$reference_label)
false_positives <- known & pairs$predicted_label %in% "NOUN" &
  !pairs$reference_label %in% "NOUN"
false_negatives <- pairs$reference_label %in% "NOUN" &
  !pairs$predicted_label %in% "NOUN"
shown <- c("document_id", "token_index", "pre", "keyword", "post",
  "reference_label", "predicted_label")
pairs[false_positives | false_negatives, shown]
```

`pre`, `keyword` and `post` use the original text. The window is measured in
Unicode codepoints and stays inside the segment, so it works without an
additional tokenizer for either language. Its edge may cut a combining
sequence; it is not a grapheme count or quanteda token window. Read the full
source segment when more context is needed. The document/segment/token keys
link back to the preserved annotation tables, including reviewer or failure
reasons; source positions are segment-local, 1-based and inclusive.

`review_queue` contains disagreements and all unavailable comparisons. A row
in that queue is a request for examination, not a declaration that the model
is wrong. Use a separate decision table and reimport a new reference snapshot
when correcting labels. Re-evaluation of those corrected development labels
does not become independent held-out validation.

## Check the effect on each document

The installed script fixes the analysis to **exact supplied surfaces of nouns**
and TTR. No lowercasing, lemmatization or normalization is performed. It requires
complete labels in a condition before computing its document score: an unknown
label could change the selected noun sequence.

```{r}
example$differences
example$document_scores[, c("document_id", "condition", "missing_labels",
  "observed_selected", "selected_total", "ttr", "status")]
```

In both the English and Japanese example, each condition selects two nouns.
The reference selects the same surface twice (`cats` or `猫`), giving TTR
`1 / 2 = 0.5`. The prediction instead includes `run` or `見る`, giving TTR
`2 / 2 = 1`. The difference, predicted minus reference, is `+0.5` in each
document. The unchanged count concealed a false positive and a false negative.
These deterministic examples demonstrate the mechanism, not its prevalence
or size in a corpus, and not equivalence of English and Japanese measurement.

For the reference-gap document, a reference-derived noun total and TTR are
`NA`, even though an observed noun is present. The prediction-gap document
has the converse limitation. `observed_selected` remains useful as a partial
count, while `selected_total` is unknown. The empty document has zero selected
tokens and an undefined TTR. None of these documents silently disappears from
the comparison.

Choose other metrics and parameters according to your research question using
`lexdiv_metrics()` or the profile APIs. Keep selection, lexical form, window
size and reference resources fixed across paired conditions, and retain metric
failure reasons. An incomplete reference supports evaluation on reviewed
positions, not an unqualified full-document reference score. For larger studies,
link participant/task/register metadata by document ID; repeated texts and
tokens from one participant are not independent observations for uncertainty
estimation. The function provides descriptive counts, not confidence intervals
or a universal F1 requirement.

## Save the inputs, policy and results

The example object contains both complete inputs, reference declarations,
the evaluation and document-score settings. Keep these together:

```{r, eval=FALSE}
saveRDS(example, "annotation-evaluation.rds", version = 2)
restored <- readRDS("annotation-evaluation.rds")
replayed <- do.call(lexdiv_evaluate_annotations, restored$inputs)
stopifnot(identical(replayed, restored$evaluation))
```

Keep the installed script, package version and `sessionInfo()` with the result
to reproduce the document-score calculation too. CSV exports are convenient
for review, but alone do not preserve complete nested metadata or the types
of all-missing columns. Stored hashes identify snapshots; they do not authenticate
human judgments. When using restricted corpora, keep the saved source-containing
objects under the corresponding access conditions rather than including them
in a distributable package.

This evaluator concerns one categorical label on fixed token boundaries.
It does not compute dependency-head attachment accuracy, identify word senses,
or establish psychological validity. The separate
[annotation sensitivity guide](auditing-vocabulary-profiles.html) compares
lemma/UPOS/flemma changes without designating either side as a reference.
