---
title: "From corpus annotations to quanteda and lexical profiles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From corpus annotations to quanteda and lexical profiles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ldfreq)
has_xml <- requireNamespace("xml2", quietly = TRUE)
has_quanteda <- has_xml && requireNamespace("quanteda", quietly = TRUE)
```

Suppose a corpus already provides token boundaries, lemmas and part-of-speech
tags. Retokenizing the text can change the units being counted. This guide
imports those annotations, uses **quanteda** for n-grams and concordances,
and keeps the IDs and positions needed to inspect the original evidence.

The example is an original, MIT-licensed text with authored annotations,
included with ldfreq. It runs offline. It is not a sample of MASC and is not
a frequency norm. Install the optional `xml2` and `quanteda` packages to run
the corresponding steps; no Python or AntConc installation is needed.

## Read an annotated document

```{r read, eval=has_xml}
header <- system.file("extdata", "masc-example", "example.anc", package = "ldfreq")
corpus <- lexdiv_read_masc(header, resource_version = "authored-example-1")
corpus$tokens[c("annotation_id", "segment_id", "token_index",
                "surface", "lemma", "pos", "start", "end")]
```

The text is `We re-use words. We go.` The supplied annotation treats `re-use`
as one token, assembled from three adjacent source regions. The reader keeps
that decision and returns `words` with the supplied lemma `word`. It does not
correct the tags, invent missing lemmas, or translate Penn tags into UPOS.

`start` and `end` are 1-based inclusive Unicode-codepoint positions in the
unchanged primary text. Original zero-based, end-exclusive anchors remain in
`start_anchor` and `end_anchor`. For a source whose anchors count UTF-16 code
units, specify `offset_unit = "utf16"`. UTF-8 file encoding alone does not tell
you which coordinate convention an annotation uses.

The document table retains original text and available header metadata. The
segment table retains the supplied sentence/utterance regions, including
regions without tokens. A completely empty document may have no segment rows;
its existence remains in `corpus$documents` and must be retained when setting
document-level denominators.

## Transfer selected tokens without closing gaps

Choose exclusions explicitly. Here punctuation is identified by the authored
Penn tag `.`. This small example is not a complete punctuation policy for
all corpora or tagsets.

```{r adapter, eval=has_quanteda}
words <- corpus$tokens[corpus$tokens$pos != ".", ]
q <- lexdiv_as_quanteda(words, corpus$segments)
as.list(q$tokens)
quanteda::docvars(q$tokens)[c("document_id", "segment_id", "token_count")]
quanteda::ntoken(q$tokens, remove_padding = TRUE)
```

Every quanteda document represents one original segment. `docvars` connects
it to the original document. Removed token positions remain empty padding,
including at segment ends. Keep `token_index` and the segment's original
`token_count` unchanged when filtering. Renumbering after removing `a` from
`make a decision` would incorrectly create the adjacent bigram `make decision`.
The adapter preserves gaps; it cannot reconstruct positions already discarded.

For supplied lemmas, use `term_col = "lemma"`. Missing or multiword values
require an explicit decision and are rejected by the adapter. To compare
surface and lemma analyses, retain both sets of settings and results rather
than overwriting one analysis.

For non-ASCII terms, run the quanteda step in an R session with a UTF-8
`LC_CTYPE` locale (`l10n_info()[["UTF-8"]]` should be `TRUE`). The adapter rejects
other locales because the public quanteda import can alter such terms. It
does not change the session locale. This restriction does not apply to the
reader's UTF-8 text and source-coordinate conversion.

This is a setting of the running R session, not a restriction on Macs. Check
the session used for your analysis:

```{r locale}
l10n_info()[["UTF-8"]]
```

If this returns `TRUE`, no locale change is needed. `C.UTF-8` is UTF-8-capable;
it is different from the plain `C` locale used in an encoding stress test.
RStudio, R.app and terminal R can start with different environment settings,
so check the session where you run the analysis. The package does not require
changing your Mac's display language. See [R's locale documentation](https://stat.ethz.ch/R-manual/R-devel/library/base/html/locales.html).

## Use quanteda for n-grams and original-document frequencies

```{r counts, eval=has_quanteda}
ng <- quanteda::tokens_ngrams(q$tokens, n = 2:4)
segment_counts <- quanteda::dfm(ng, tolower = FALSE)
document_counts <- quanteda::dfm_group(
  segment_counts, groups = quanteda::docvars(ng, "document_id")
)
quanteda::topfeatures(document_counts)
quanteda::docfreq(document_counts)
```

Grouping matters: counting how many segments contain a phrase does not give
the number of original documents containing it. This example has one original
document and two segments. The returned tables do not establish population
prevalence, proficiency, or lexical employability.

`tolower = FALSE` preserves the case of the supplied tokens. The default of
`dfm()` lowercases features, which changes case-sensitive frequency tables.
Choose the same case policy for targets and references and record it.

Quanteda already supports four-word and longer n-grams. The ldfreq adjacent
reference API currently supports only bigrams and trigrams, so a quanteda
four-gram table is not yet interchangeable with that reference format.
Also retain component keys when preparing reference tables: splitting a
concatenated feature on `_` is ambiguous if a component already contains `_`.

## Return from a result to a source annotation

Search the original token object, not the n-gram token object whose positions
refer to a transformed sequence.

```{r kwic, eval=has_quanteda}
hits <- as.data.frame(quanteda::kwic(q$tokens, "re-use", valuetype = "fixed"))
locations <- q$positions[c("quanteda_docname", "token_index", "document_id",
                           "annotation_id", "start", "end")]
evidence <- merge(hits, locations,
  by.x = c("docname", "from"), by.y = c("quanteda_docname", "token_index"))
evidence[c("keyword", "document_id", "annotation_id", "start", "end")]
```

This joins the start of a hit. For a multi-token phrase, also join its `to`
position to obtain the final token's character endpoint. The mapping applies
to the returned tokens and transformations that preserve positions. Collapsing
padding, compounding tokens, or rebuilding tokenization requires a new mapping.

## Connect to the existing ldfreq reference workflow

```{r adjacent, eval=has_quanteda}
prepared <- q$positions
prepared$term <- prepared$surface
sequences <- lexdiv_ngrams(prepared,
  preprocessing_id = "authored-penn-surface-exclude-period-tag-v1",
  documents = corpus$documents$document_id)
sequences$counts
```

Follow the **Adjacent n-gram frequency profiles** guide to build a local
reference with explicit source metadata and denominators. A preprocessing ID
is a declaration of comparable rules, not proof that an unrelated tokenizer
is compatible. MASC Penn tokens, quanteda word4 tokens and the bundled TUBELEX
tokenization must not be silently treated as the same tokenization.

Save source evidence with the analysis:

```{r saving, eval=has_quanteda}
path <- tempfile(fileext = ".rds")
saveRDS(list(corpus = corpus, quanteda = q, sequences = sequences), path)
restored <- readRDS(path)
stopifnot(identical(restored$corpus$provenance, corpus$provenance),
          identical(as.list(restored$quanteda$tokens), as.list(q$tokens)))
unlink(path)
```

## Use your local MASC files

The reader is verified against the **Penn layer of Mini-MASC 1.0**, using its
GrAF 1.0 `cesHeader` layout. Supply local `.anc` document headers and the actual
resource version. No corpus is downloaded by this function.
In the verified eight-document Mini-MASC release, 851 of 7,263 Penn tokens
have no `base` feature. They retain `NA` lemmas; a lemma analysis therefore
needs an explicit missingness policy even when surface tokens are available.

```{r local-corpus, eval=FALSE}
headers <- list.files("/path/to/Mini-MASC/data", pattern = "\\.anc$",
                      recursive = TRUE, full.names = TRUE)
masc <- lexdiv_read_masc(headers, resource_version = "Mini-MASC-1.0")
```

The source format includes separate primary text, segmentation, sentence or
utterance, and Penn feature files. Header references must resolve within the
header's directory. The reader rejects missing links, overlapping spans,
tokens crossing segment boundaries, and discontinuous tokens. It does not
silently repair those cases or switch to another annotation layer.

Reader interface 0.2.0 also accepts MASC 3.0.0's GrAF `documentHeader`
layout: `.hdr` files use `f.id="f.penn"` etc. and `loc` attributes. This is
format support with a restricted validated scope. A local audit of all 392
documents in the [official MASC 3.0.0 archive](https://anc.org/data/masc/downloads/data-download/)
accepted **122 documents** under the existing strict rules. First failures
among the other 270 were: 130 documents with tokens outside a single supplied
sentence/utterance region, 126 with overlapping sentence/utterance regions,
nine with invalid or incompatible anchors, four with overlapping tokens and
one with a feature structure outside the supported flat name/value format.
These are reader diagnostics, not a comprehensive correction of the corpus.
Multiple issues may exist in one file. No rejected annotations were repaired
or silently replaced by new sentence/token boundaries.

Consequently, do not catch errors, omit these files and label the remaining
sample "full MASC". Such exclusions can change register composition and
reference frequencies. Record accepted and rejected document IDs, reasons,
source hashes and preparation choices before interpreting any subset. The
accepted subset is useful for exercising the software; it is not supplied
as a representative English norm. The inspected archive's SHA-256 was
`0b300d96016233c7c79b43559b477a6bb0ed16bb416d5c517eccb108a6228493`.

The inspected [OANC GrAF archive](https://anc.org/data/oanc/download/) mainly
uses a `hepple` annotation layer with token regions in the same graph. It is
not the MASC Penn/seg layout. This reader does not provide a validated OANC
importer; shared GrAF namespaces alone do not establish compatibility.

[MASC](https://anc.org/data/masc/) is distributed under CC BY 3.0 US. Retain
its attribution, licence and changes when redistributing material. The
reader records file hashes and the declared release but does not determine
the licence of arbitrary input files. Corpus rights differ from the licence
of this package and its authored example.

MASC and [OANC](https://anc.org/data/oanc/) can support different reference
sizes and registers. Their sample composition, overlapping source material,
annotation choices and sparse long n-grams matter. A reference's usefulness
for your question requires more than successful file import. These functions
do not yet provide a validated standard MASC/OANC frequency resource.

Once inputs have been validated, `lexdiv_ngram_reference_build()` can accumulate
references in whole-document chunks. This avoids retaining every occurrence
table, while counts and the document roster remain in memory. See the
[n-gram guide](ngram-profiles.html#accumulate-a-reference-in-whole-document-chunks)
for the callback, denominator and exclusion-record requirements.
