---
title: "From corpus annotations to quanteda and lexical profiles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From corpus annotations to quanteda and lexical profiles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ldfreq)
has_xml <- requireNamespace("xml2", quietly = TRUE)
has_quanteda <- has_xml && requireNamespace("quanteda", quietly = TRUE)
has_phrase_example <- requireNamespace("quanteda", quietly = TRUE) &&
  utils::packageVersion("quanteda") >= "4.5.0" &&
  isTRUE(l10n_info()[["UTF-8"]])
knitr::read_chunk(system.file("examples", "phrase-review-demo.R",
  package = "ldfreq", mustWork = TRUE))
```

Suppose a corpus already provides token boundaries, lemmas and part-of-speech
tags. Retokenizing the text can change the units being counted. This guide
imports those annotations, uses **quanteda** for n-grams and concordances,
and keeps the IDs and positions needed to inspect the original evidence.

The example is an original, MIT-licensed text with authored annotations,
included with ldfreq. It runs offline. It is not a sample of MASC and is not
a frequency norm. Install the optional `xml2` and `quanteda` packages to run
the corresponding steps; no Python or AntConc installation is needed.

## Read an annotated document

```{r read, eval=has_xml}
header <- system.file("extdata", "masc-example", "example.anc", package = "ldfreq")
corpus <- lexdiv_read_masc(header, resource_version = "authored-example-1")
corpus$tokens[c("annotation_id", "segment_id", "token_index",
                "surface", "lemma", "pos", "start", "end")]
```

The text is `We re-use words. We go.` The supplied annotation treats `re-use`
as one token, assembled from three adjacent source regions. The reader keeps
that decision and returns `words` with the supplied lemma `word`. It does not
correct the tags, invent missing lemmas, or translate Penn tags into UPOS.

`start` and `end` are 1-based inclusive Unicode-codepoint positions in the
unchanged primary text. Original zero-based, end-exclusive anchors remain in
`start_anchor` and `end_anchor`. For a source whose anchors count UTF-16 code
units, specify `offset_unit = "utf16"`. UTF-8 file encoding alone does not tell
you which coordinate convention an annotation uses.

The document table retains original text and available header metadata. The
segment table retains the supplied sentence/utterance regions, including
regions without tokens. A completely empty document may have no segment rows;
its existence remains in `corpus$documents` and must be retained when setting
document-level denominators.

## Transfer selected tokens without closing gaps

Choose exclusions explicitly. Here punctuation is identified by the authored
Penn tag `.`. This small example is not a complete punctuation policy for
all corpora or tagsets.

```{r adapter, eval=has_quanteda}
words <- corpus$tokens[corpus$tokens$pos != ".", ]
q <- lexdiv_as_quanteda(words, corpus$segments)
as.list(q$tokens)
quanteda::docvars(q$tokens)[c("document_id", "segment_id", "token_count")]
quanteda::ntoken(q$tokens, remove_padding = TRUE)
```

Every quanteda document represents one original segment. `docvars` connects
it to the original document. Removed token positions remain empty padding,
including at segment ends. Keep `token_index` and the segment's original
`token_count` unchanged when filtering. Renumbering after removing `a` from
`make a decision` would incorrectly create the adjacent bigram `make decision`.
The adapter preserves gaps; it cannot reconstruct positions already discarded.

For supplied lemmas, use `term_col = "lemma"`. Missing or multiword values
require an explicit decision and are rejected by the adapter. To compare
surface and lemma analyses, retain both sets of settings and results rather
than overwriting one analysis.

For non-ASCII terms, run the quanteda step in an R session with a UTF-8
`LC_CTYPE` locale (`l10n_info()[["UTF-8"]]` should be `TRUE`). The adapter rejects
other locales because the public quanteda import can alter such terms. It
does not change the session locale. This restriction does not apply to the
reader's UTF-8 text and source-coordinate conversion.

This is a setting of the running R session, not a restriction on Macs. Check
the session used for your analysis:

```{r locale}
l10n_info()[["UTF-8"]]
```

If this returns `TRUE`, no locale change is needed. `C.UTF-8` is UTF-8-capable;
it is different from the plain `C` locale used in an encoding stress test.
RStudio, R.app and terminal R can start with different environment settings,
so check the session where you run the analysis. The package does not require
changing your Mac's display language. See [R's locale documentation](https://stat.ethz.ch/R-manual/R-devel/library/base/html/locales.html).

## Use quanteda for n-grams and original-document frequencies

```{r counts, eval=has_quanteda}
ng <- quanteda::tokens_ngrams(q$tokens, n = 2:4)
segment_counts <- quanteda::dfm(ng, tolower = FALSE)
document_counts <- quanteda::dfm_group(
  segment_counts, groups = quanteda::docvars(ng, "document_id")
)
quanteda::topfeatures(document_counts)
quanteda::docfreq(document_counts)
```

Grouping matters: counting how many segments contain a phrase does not give
the number of original documents containing it. This example has one original
document and two segments. The returned tables do not establish population
prevalence, proficiency, or lexical employability.

`tolower = FALSE` preserves the case of the supplied tokens. The default of
`dfm()` lowercases features, which changes case-sensitive frequency tables.
Choose the same case policy for targets and references and record it.

Quanteda already supports four-word and longer n-grams. The ldfreq adjacent
reference API currently supports only bigrams and trigrams, so a quanteda
four-gram table is not yet interchangeable with that reference format.
Also retain component keys when preparing reference tables: splitting a
concatenated feature on `_` is ambiguous if a component already contains `_`.

## Return from a result to a source annotation

Search the original token object, not the n-gram token object whose positions
refer to a transformed sequence.

```{r kwic, eval=has_quanteda}
hits <- as.data.frame(quanteda::kwic(q$tokens, "re-use", valuetype = "fixed"))
locations <- q$positions[c("quanteda_docname", "token_index", "document_id",
                           "annotation_id", "start", "end")]
evidence <- merge(hits, locations,
  by.x = c("docname", "from"), by.y = c("quanteda_docname", "token_index"))
evidence[c("keyword", "document_id", "annotation_id", "start", "end")]
```

This joins the start of a hit. For a multi-token phrase, also join its `to`
position to obtain the final token's character endpoint. The mapping applies
to the returned tokens and transformations that preserve positions. Collapsing
padding, compounding tokens, or rebuilding tokenization requires a new mapping.

For single-token candidate decisions, the [ambiguity review guide](ambiguity-review.html)
shows a source-verified KWIC workflow with complete external imports, original
context, occurrence IDs and explicit unresolved decisions. That interface
accepts `lexdiv_import_annotations()` results; do not pass the MASC reader's
document-relative coordinates as segment-relative positions without conversion.

## Match your phrase list and inspect its coverage

Suppose you want to find a supplied list of expressions in essays, articles,
or conversations, then check whether each occurrence is relevant in context.
The installed `phrase-list-kwic.R` helper connects existing quanteda phrase
search to original source spans and document summaries. It is a **copyable
example**, sourced explicitly, rather than a new exported ldfreq function.
It does not download phrase lists, run a parser, or infer idiomatic meanings.

Run the complete English/Japanese illustration:

```{r phrase-demo, eval=has_phrase_example}
example_env <- new.env()
sys.source(system.file("examples", "phrase-list-demo.R", package = "ldfreq"),
           envir = example_env)
phrase_result <- example_env$result
```

All text, annotations and list entries in this example were authored for
ldfreq and carry its MIT licence. They are not extracts from published phrase
inventories or empirical frequency norms. The example requires only optional
quanteda and a UTF-8 R session; the XML reader is not needed for this step.

For your own analysis, first import **complete** token annotations using
`lexdiv_import_annotations()`, as in the [Japanese annotation guide](japanese-annotations.html).
The helper rechecks this unchanged object before any exclusions. It accepts
segment-local positions from that importer; it does not accept the MASC
reader's document-relative positions directly. Token and segment boundaries
must already be specified. The helper never guesses where sentences end.

Provide a named list of explicit component vectors:

```{r phrase-input, eval=has_phrase_example}
phrases <- list(
  long = c("In", "the", "end", "of", "the", "day"),
  nested = c("the", "end"),
  short = c("in", "the", "end"),
  japanese = c("国際", "協力", "を", "学ぶ"),
  absent = c("not", "present")
)
```

IDs must be unique and each entry needs at least two tokens. There is no
bigram/trigram restriction. Japanese components follow the supplied analyzer's
segmentation: `c("国際", "協力")` will not match a single token `"国際協力"`.
Underscores within components are literal. Searches are case sensitive and
surface based; no lemmatization, Unicode normalization, variant expansion or
discontinuous matching occurs. `In` and `in` are different. Canonically
equivalent Unicode strings are checked against their original components
after quanteda search, so they are not silently merged.

To prepare the list in a spreadsheet, use long columns `phrase_id`,
`component_index`, `term`, with one component per row. Read IDs and terms as
character strings; validate unique consecutive component indices starting at
1 within each ID, sort by those indices, and convert each group to a vector.
Keep citation, version and permission information with the list. Merely
putting spaces into a Japanese expression does not establish token boundaries.

```{r phrase-run, eval=has_phrase_example}
source(system.file("examples", "phrase-list-kwic.R", package = "ldfreq"), local = TRUE)
a <- example_env$annotations
keep <- !a$tokens$surface %in% c(",", ".", "。", "very")
result <- phrase_list_kwic(a, phrases, example_env$resource, keep, window = 2L)
result$occurrences[c("phrase_id", "document_id", "segment_id", "from", "to",
                     "start", "end", "keyword")]
result$counts
result$documents
```

Exclusions are an explicit logical mask over the original token table, not
deletions followed by renumbering. Kept surfaces must be non-whitespace tokens.
If your annotations include whitespace tokens, exclude their slots explicitly;
those slots remain barriers to matching. Punctuation is retained unless you
exclude it and can be part of a listed sequence. In the illustration,
removing `very` from `in the very end` does **not** create `in the end`.
No match crosses a removed slot, segment boundary, or document boundary.

`from` and `to` count original token slots within a segment. `start` and `end`
are **1-based inclusive Unicode-codepoint positions in `segment_text`**, not
byte offsets or whole-document offsets. `keyword` is the exact substring
between those endpoints, including original spacing. `pre` and `post` retain
original text up to `window` token slots either side, within that segment;
excluded punctuation/words can therefore remain visible as context. A window
of zero produces empty `pre`/`post`. This source display differs from quanteda's
space-joined display, particularly for Japanese. Save the full segment for
interpretation when a short window is insufficient.

Three outputs answer distinct questions:

| Output | Interpretation |
|---|---|
| `occurrences` | Every matched phrase ID and span, including nested and overlapping matches |
| `counts` / `summary` | Per-document/per-ID occurrences, and occurrence/document totals for each ID; absent phrases and empty documents stay visible |
| `documents` | All original token slots, retained tokens, occurrence counts, and the union of covered tokens divided by retained tokens |

For the English example, four occurrences cover nine distinct tokens out of
12 retained tokens: coverage is `9/12`, not the sum of phrase lengths divided
by 12. The longer expression contains a shorter listed expression, and both
are recorded. Repeated identical sequences under different IDs also remain
separate list entries; they increase per-ID occurrence totals, but not union
coverage. For an empty or fully excluded document, coverage is `NA`, with
zero counts. A nonempty retained document with no match has coverage zero.
The denominator includes every retained token, including punctuation if kept;
it is not automatically a word-only or eligible-window denominator.

Keep a full result for replay and a separate review table for judgments:

```{r phrase-save, eval=has_phrase_example}
path <- tempfile(fileext = ".rds")
saveRDS(result, path)
restored <- readRDS(path)
replayed <- phrase_list_kwic(restored$source, restored$phrases, restored$resource,
  restored$settings$keep, restored$settings$window)
stopifnot(identical(replayed, restored))
unlink(path)
```

The saved object includes source annotations, phrase components, resource
declarations, exclusions, settings, input hash and package versions. The next
section adds a checked worksheet for occurrence-specific decisions. Record
literal/idiomatic or other functional judgments separately, with reviewer,
criteria and unresolved cases. Phrase matches do not
establish collocational strength, formulaicity, proficiency, or a learner's
knowledge. The current single-token ambiguity API does not accept these phrase
rows; this example does not create a phrase-sense classifier.

This helper is intended for inspectable, modest lists: it calls quanteda once
per ID and retains source text plus every document-by-ID count, including zeros.
It is not a large dictionary index or a streaming corpus search. The resource
record documents caller declarations, not verified redistribution rights;
saving or sharing the result also copies your source text and phrase list.
Existing [quanteda phrase patterns](https://quanteda.io/reference/phrase.html)
and [KWIC](https://quanteda.io/reference/kwic.html) perform the search. The
addition here is the reproducible connection to source spans, exclusions,
overlap-aware denominators and subsequent human inspection.

## Review phrase matches and return decisions to document counts {#review-phrase-matches}

A match for `take off` can describe a departing plane or removing a coat.
Whether it belongs in an analysis depends on the declared question and its
context. Similarly, the authored Japanese examples distinguish hindering a
plan from physically pulling a doll's leg. The search finds both; it does not
make that contextual judgment.

From development version **0.3.0.9003**, the explicitly sourced
`phrase_list_review()` helper connects the existing phrase search to a complete
decision worksheet. It is not an exported package API. Install that version
or later using the [installation instructions](https://ryuya-dot-com.github.io/ldfreq/#installation),
and use optional quanteda >= 4.5.0 in a UTF-8 R session. The standalone example
runs offline, without a corpus, model or API key:

```{r phrase-review-quick-start, eval=FALSE}
source(system.file("examples", "phrase-review-demo.R",
  package = "ldfreq", mustWork = TRUE), local = TRUE)
phrase_after$documents
phrase_after$counts
```

All texts, phrase entries and decisions are authored teaching material, not
learner observations, an established phrase inventory or an independent human
reference. Six document IDs include complete English/Japanese reviews, pending
reviews, a nonempty no-match document and an empty document.

```{r phrase-review-prepare, eval=has_phrase_example, include=FALSE}
```

```{r phrase-review-context, eval=has_phrase_example, echo=FALSE}
knitr::kable(phrase_before$occurrences[c("document_id", "segment_id", "phrase_id",
  "keyword", "segment_text")], row.names = FALSE)
```

The declared criterion accepts the English departure interpretation and the
Japanese hindrance interpretation. A longer `will take off` entry deliberately
overlaps `take off`, so both entries remain inspectable. The Japanese entry
uses the supplied authored components `足 / を / 引っ張っ / た`; other
tokenizations require correspondingly specified entries. No automatic
lemmatization, alternative-form expansion or discontinuous search is implied.

For your own unchanged search result, start with:

```{r phrase-review-own-input, eval=FALSE}
source(system.file("examples", "phrase-list-kwic.R",
  package = "ldfreq", mustWork = TRUE), local = TRUE)
before <- phrase_list_review(my_search,
  criterion = "Write the study's interpretation and inclusion rule here.")
```

### Export context separately from the editable worksheet

```{r phrase-review-export, eval=has_phrase_example}
```

`context.csv` is for reading the KWIC and complete original segment.
`decisions.csv` is the editable worksheet. It has one row per **phrase ID and
source occurrence**, including nested spans and identical forms listed under
different IDs. Edit only `status`, `reviewer` and `reason`; keep all other cells
and every row, although you may reorder rows. Read all columns as character
when importing the CSV so IDs and coordinates survive unchanged.

| Status | Meaning and required fields |
|---|---|
| `accepted` | The occurrence meets the declared criterion; supply reviewer and reason |
| `rejected` | The match is outside that criterion; supply reviewer and reason. This does not label the writer's expression incorrect |
| `unresolved` | It was examined but the interpretation remains unsettled; supply reviewer and reason |
| `unreviewed` | No decision has been submitted; leave reviewer and reason missing |

Here `<MISSING>` is the CSV missing-value marker. Do not use that literal string
as a reviewer name or reason. To resume an existing review, export its latest
`$worksheet`; starting again from the initial blank worksheet replaces the
earlier decisions. Save each stage or rater separately when that history matters.

The following edits encode the teaching interpretations and deliberately leave
some rows unresolved or unreviewed. They are not automated semantic analysis.

```{r phrase-review-decide, eval=has_phrase_example}
```

### Reapply by identity, then inspect counts and coverage

```{r phrase-review-apply, eval=has_phrase_example}
```

The helper replays the original search before applying the worksheet. It
checks the full row set, source anchors, review fingerprint and required
decision fields. A removed/duplicated row, edited fixed cell or partly filled
unreviewed row fails. Changed source, phrase entries, resource declaration,
search settings **including the context window**, or criterion require a new
worksheet. Do not replace IDs to force old decisions onto changed material.

Each document and document-by-phrase cell satisfies
`occurrences = accepted + rejected + unresolved + unreviewed`.
`covered_tokens` describes all exact matches; `accepted_covered_tokens`
describes only the **union** of accepted spans. The English complete review
accepts two nested occurrences of lengths three and two, but their union is
three tokens, so accepted coverage is `3/11`, not `5/11`. The Japanese complete
review gives `4/18`. These authored ratios are not a language or ability comparison.

`accepted_coverage` retains the coverage of decisions accepted so far.
`reportable_accepted_coverage` is missing whenever a document still has an
unresolved or unreviewed match. A zero in the former column for a pending
review is therefore not reported as an observed absence of relevant uses.
The no-match document has zero coverage; the empty document has missing
coverage because its denominator is zero.

`review_complete` means that every **matched entry** has an accepted/rejected
decision. It does not establish that the inventory is exhaustive, that the
annotations are correct, or that the decisions are reliable. Different list
entries at the same span retain separate decisions; union coverage prevents
double counting but does not adjudicate inconsistent judgments. Original
tokens remain unchanged: this workflow does not compound accepted expressions
into single words or alter TTR, MATTR or MTLD.

### Save the complete review and reproduce its tables

```{r phrase-review-save, eval=has_phrase_example}
```

The temporary directory makes the example runnable. For a study, use a
persistent project directory and separate files for successive review stages;
writing the same filename overwrites it. The RDS retains original text, search
settings, the declared criterion, every judgment and the imported CSV record.
The two result CSVs are flat tables for later analysis. Reproduction uses the
same software versions; retain the original files and saved environment.
These records contain source text and the phrase inventory, so their sharing
conditions follow those materials.

A methods description should identify the phrase inventory/version, analyzer
and word unit, case/normalization policy, segment boundaries, exclusions,
context and decision criterion, reviewers and handling of disagreements.
Report accepted/rejected/unresolved/unreviewed counts and the retained-token
denominator. State that overlapping accepted spans contribute their union to
coverage, and report how many documents remain incomplete. Search and worksheet
checks verify correspondence, not inter-rater agreement or psycholinguistic validity.

## Compare phrase judgments before discussion {#phrase-agreement}

Two researchers may read the same `take off` match differently. The question
here is **whether they make the same inclusion decision, and whether their
decisions change the document profile**. It is not a test of phrase-detection
recall: expressions absent from the list have not been sampled for review.
The following guide-local code runs with the existing 0.3.0.9003 installation.

The design follows these distinctions in the annotation literature:

| Research basis | Consequence for this workflow |
|---|---|
| [Carletta (1996)](https://aclanthology.org/J96-2004/), pp. 250--253: chance agreement and the choice of coding unit matter | Declare one phrase ID at one original span as the item; retain the category counts and all sampled items |
| [Di Eugenio and Glass (2004)](https://aclanthology.org/J04-1005/), pp. 96--100: kappa variants and marginal distributions can change the interpretation | Name unweighted **Cohen's kappa**, show both raters' margins, and do not label a fixed cutoff as a universal pass criterion |
| [Artstein and Poesio (2008)](https://aclanthology.org/J08-4004/), sections 2.1--2.4: agreement, validity and chance models are different questions | Preserve independent decisions before discussion; a high agreement coefficient or a consensus is not evidence that the interpretation is correct |

For this binary inclusion task, `accepted` and `rejected` are nominal labels,
not ordered scores. `unresolved` records an examined but unsettled case;
`unreviewed` records no submitted decision. **Our analysis policy** retains
all four states, but estimates binary agreement only where both raters made
an accepted/rejected decision. This conditional subset may omit the hardest
cases. Its agreement is not an estimate for all occurrences without further
assumptions. If uncertainty itself is a substantive response category, define
and evaluate that different coding scheme explicitly.

### Pair two unchanged reviews and expose the denominators

The two review columns must each belong to one declared person, with different
IDs. Different names cannot verify independent work: in a study, assign the
same sampled material and written criterion, collect decisions separately,
and retain the pre-discussion files. Treat a repeated pass by one person as
intra-rater comparison instead. The helper below is local to this guide.

```{r phrase-agreement-function, eval=has_phrase_example}
compare_phrase_reviews <- function(a, b, reviewers) {
  if (!is.character(reviewers) || length(reviewers) != 2L ||
      anyNA(reviewers) || anyDuplicated(reviewers) ||
      any(!validUTF8(reviewers)) ||
      any(!nzchar(stringi::stri_trim_both(reviewers))))
    stop("Declare two distinct, nonblank reviewer IDs.")
  reviews <- list(a, b)
  for (i in seq_along(reviews)) {
    x <- reviews[[i]]
    checked <- phrase_list_review(x$source, x$criterion, x$worksheet)
    if (!identical(x, checked)) stop("Review changed; reapply its worksheet first.")
    submitted <- x$worksheet$status != "unreviewed"
    if (any(x$worksheet$reviewer[submitted] != reviewers[i]))
      stop("Each review must contain decisions from its declared reviewer only.")
  }
  if (!identical(a$provenance$review_id, b$provenance$review_id))
    stop("Reviews must share source, inventory, settings and criterion.")
  p <- a$occurrences[c("review_id", "occurrence_id", "phrase_id", "document_id",
    "segment_id", "start", "end", "pre", "keyword", "post", "segment_text")]
  j <- match(p$occurrence_id, b$occurrences$occurrence_id)
  if (anyNA(j) || anyDuplicated(j) || length(j) != nrow(b$occurrences))
    stop("Keep the same complete occurrence set in both reviews.")
  for (field in c("status", "reviewer", "reason")) {
    p[[paste0("a_", field)]] <- a$occurrences[[field]]
    p[[paste0("b_", field)]] <- b$occurrences[[field]][j]
  }
  labels <- c("accepted", "rejected")
  states <- c(labels, "unresolved", "unreviewed")
  p$both_decided <- p$a_status %in% labels & p$b_status %in% labels
  p$outcome <- ifelse(p$both_decided,
    ifelse(p$a_status == p$b_status, "agreement", "disagreement"), "incomplete")
  ids <- names(a$source$phrases)
  status_counts <- as.data.frame(table(
    phrase_id = factor(p$phrase_id, levels = ids),
    a_status = factor(p$a_status, levels = states),
    b_status = factor(p$b_status, levels = states)), responseName = "occurrences")
  ratings <- setNames(lapply(ids, function(id) {
    rows <- p$phrase_id == id & p$both_decided
    data.frame(a = factor(p$a_status[rows], levels = labels),
      b = factor(p$b_status[rows], levels = labels))
  }), ids)
  summary <- do.call(rbind, lapply(ids, function(id) {
    rows <- p$phrase_id == id
    r <- ratings[[id]]; n <- sum(rows); paired <- nrow(r)
    agreement <- sum(r$a == r$b)
    data.frame(phrase_id = id, occurrences = n,
      documents = length(unique(p$document_id[rows])), both_decided = paired,
      paired_documents = length(unique(p$document_id[rows & p$both_decided])),
      incomplete = n - paired, agreement = agreement, disagreement = paired - agreement,
      paired_proportion = if (n) paired / n else NA_real_,
      observed_agreement = if (paired) agreement / paired else NA_real_,
      a_accepted = sum(r$a == "accepted"), b_accepted = sum(r$b == "accepted"))
  }))
  list(summary = summary, status_counts = status_counts, ratings = ratings,
    pairs = p, review_queue = p[p$outcome != "agreement", , drop = FALSE],
    reviews = reviews, reviewers = reviewers,
    policy = "Per phrase ID; binary agreement conditional on both decisive; all states retained")
}
```

For your files, use `readRDS()` to restore the separately saved reviews, then
call `compare_phrase_reviews(a, b, reviewers = c("rater-1", "rater-2"))` with
the actual recorded IDs. It reuses the installed worksheet checks before
pairing by occurrence identity; altered source or criterion cannot be joined
silently. No-match phrase IDs remain in the summary with a zero denominator.

### Run the authored example and inspect open cases

Continue the example above, or first run `phrase-review-demo.R`. The following
second worksheet deliberately introduces disagreements and different completion
states. **It is simulated teaching material, not an independent human rating.**

```{r phrase-agreement-example, eval=has_phrase_example}
review_a <- phrase_after
sheet_b <- phrase_before$worksheet
complete <- sheet_b$document_id %in% c("en", "ja")
sheet_b$status[complete & sheet_b$segment_id == "s1"] <- "accepted"
sheet_b$status[complete & sheet_b$segment_id == "s2"] <- "rejected"
sheet_b$status[sheet_b$document_id == "en" & sheet_b$segment_id == "s2"] <- "accepted"
sheet_b$status[sheet_b$document_id == "en" & sheet_b$phrase_id == "departure_long"] <- "unresolved"
sheet_b$status[sheet_b$document_id == "en_pending" & sheet_b$segment_id == "s1"] <- "rejected"
sheet_b$status[sheet_b$document_id == "ja_pending" & sheet_b$segment_id == "s1"] <- "unresolved"
sheet_b$status[sheet_b$document_id == "ja_pending" & sheet_b$segment_id == "s2"] <- "accepted"
submitted <- sheet_b$status != "unreviewed"
sheet_b$reviewer[submitted] <- "authored-B"
sheet_b$reason[submitted] <- "Deliberately varied teaching decision; not a human observation."
review_b <- phrase_list_review(phrase_search, phrase_criterion,
  sheet_b[rev(seq_len(nrow(sheet_b))), ])
phrase_comparison <- compare_phrase_reviews(review_a, review_b,
  reviewers = c("authored-demo-reviewer", "authored-B"))
phrase_comparison$summary
subset(phrase_comparison$status_counts, occurrences > 0)
phrase_comparison$review_queue[c("document_id", "phrase_id", "keyword",
  "a_status", "b_status", "a_reason", "b_reason", "outcome")]
```

There are ten list matches. Four have binary decisions from both sides: three
agree and one disagrees. The other six remain incomplete, including two
unresolved judgments on the same occurrence. The queue therefore has seven
rows; `$pairs` also retains the original KWIC and full segment text.
The per-phrase summaries keep the denominators separate: `departure` has
one agreement in two paired decisions out of four matches; `hindrance` has
two in two out of four. Neither `departure_long` nor `absent` has any paired
binary decision, for different reasons visible in their counts.

### Optional Cohen kappa with an existing R package

For two fixed raters applying these binary labels, use the optional
[`irr::kappa2()`](https://search.r-project.org/CRAN/refmans/irr/html/kappa2.html)
with `weight = "unweighted"`. Its missing rows are removed listwise; here the
paired rows have already been selected explicitly and their excluded states
remain in the comparison. Install `irr` separately to run this block.

```{r phrase-agreement-kappa, eval=FALSE}
# install.packages("irr")  # once, when this optional analysis is needed
stopifnot(requireNamespace("irr", quietly = TRUE))
phrase_kappa <- do.call(rbind, lapply(names(phrase_comparison$ratings), function(id) {
  r <- phrase_comparison$ratings[[id]]
  status <- if (!nrow(r)) "no_paired_decisions" else if
    (length(unique(c(as.character(r$a), as.character(r$b)))) == 1L)
      "constant_single_category" else "computed_descriptive"
  value <- if (status == "computed_descriptive")
    irr::kappa2(r, weight = "unweighted")$value else NA_real_
  data.frame(phrase_id = id, both_decided = nrow(r), kappa = value, status = status)
}))
phrase_kappa
saveRDS(list(comparison = phrase_comparison, estimates = phrase_kappa,
  irr_version = as.character(packageVersion("irr"))),
  file.path(phrase_review_dir, "kappa.rds"), version = 2)
```

The authored `departure` and `hindrance` values are 0 and 1; their tiny paired
samples are not evidence of real annotator reliability. No pooled kappa across
different phrase criteria/languages is calculated. Two raters assigning only
one identical category have observed agreement 1 but undefined kappa, not a
kappa of 1. Ordinal ratings or more than two raters require a different declared
model; ICC and a correlation between document scores do not replace this
nominal-label comparison.

To check the external implementation against a published calculation, this
block reconstructs [Artstein and Poesio's Tables 1--2](https://aclanthology.org/J08-4004/)
(pp. 558, 560). The counts describe dialogue-act categories, not learner data.

```{r phrase-agreement-published-check, eval=FALSE}
published_ratings <- data.frame(
  a = rep(c("STAT", "IREQ", "STAT", "IREQ"), c(20, 20, 10, 50)),
  b = rep(c("STAT", "STAT", "IREQ", "IREQ"), c(20, 20, 10, 50)))
published_kappa <- irr::kappa2(published_ratings, weight = "unweighted")$value
stopifnot(isTRUE(all.equal(published_kappa, 8/23, tolerance = 1e-12)))
published_kappa  # 0.3478261; observed agreement is 70/100
```

The verification also reproduced the prevalence examples in Di Eugenio and
Glass (2004), Figure 3 (p. 99), using `irr` 0.85: both tables have 90% observed
agreement, but counts `90, 5; 5, 0` give kappa -0.052632, whereas
`45, 5; 5, 45` give 0.8. The former figure prints -0.048; its displayed counts
and expected agreement 0.905 instead imply `(0.9 - 0.905)/(1 - 0.905)`.
The check follows those counts. Retaining the table and margins makes this
distinction inspectable; a single threshold would conceal it.
In Figure 4, the Example 5 counts (`40, 15; 20, 25`) imply expected agreement
0.51 and kappa 0.285714, rather than the printed 0.52 and 0.27. Example 6
(`40, 35; 0, 25`) gives kappa 0.363636, rather than 0.418. Independent exact
fraction arithmetic confirms the recomputed values. These are count-versus-text
discrepancies, not differences between definitions of kappa.

A document can contribute several related or overlapping occurrences. This
recipe reports descriptive point estimates, not the default test statistic or
p-value returned by `irr`. Uncertainty analysis needs the study's sampling
design and dependence structure, for example retaining related occurrences
together when resampling writers/documents. A different coefficient alone
does not resolve selective missing judgments or non-independent sampling.

### Record discussion separately and compare document profiles

After inspecting the queue, record discussion decisions separately. These
three authored edits resolve the coat/flight examples and decide one Japanese
context; other open cases remain open. Starting from A here is an explicit
working-copy choice, not automatic acceptance of A whenever the raters differ.

```{r phrase-agreement-discussion, eval=has_phrase_example}
discussion <- review_a$worksheet
edit <- (discussion$document_id == "en" & discussion$segment_id == "s2") |
  (discussion$document_id == "en" & discussion$phrase_id == "departure_long") |
  (discussion$document_id == "ja_pending" & discussion$segment_id == "s2")
discussion_log <- discussion[edit, ]
discussion_log$status <- ifelse(discussion_log$document_id == "en" &
  discussion_log$segment_id == "s2", "rejected", "accepted")
discussion_log$reviewer <- "authored-discussion"
discussion_log$reason <- ifelse(discussion_log$status == "rejected",
  "Authored discussion: removing a coat is outside the departure criterion.",
  "Authored discussion: flight departure or hindrance in the meeting fits the criterion.")
j <- match(discussion_log$occurrence_id, discussion$occurrence_id)
discussion[j, c("status", "reviewer", "reason")] <-
  discussion_log[c("status", "reviewer", "reason")]
after_discussion <- phrase_list_review(phrase_search, phrase_criterion, discussion)
stages <- list(A = review_a, B = review_b, partial_discussion = after_discussion)
phrase_profiles <- do.call(rbind, lapply(names(stages), function(stage) {
  data.frame(stage = stage, stages[[stage]]$documents, row.names = NULL)
}))
rownames(phrase_profiles) <- NULL
phrase_profiles[c("stage", "document_id", "accepted", "unresolved", "unreviewed",
  "retained_tokens", "accepted_covered_tokens", "reportable_accepted_coverage")]
```

All six documents remain in each stage. For the English document, A accepts
three original token slots, B accepts four, and the partial discussion returns
three. B still has an unresolved match there, so its complete-review coverage
is unavailable. Empty documents keep a zero denominator and missing coverage.
Discussion changes neither the original tokens nor the stored pre-discussion
agreement. Report these original profiles side by side, with their completion
status; do not turn a discussion result into an independent third rating.

```{r phrase-agreement-save, eval=has_phrase_example}
agreement_record <- list(comparison = phrase_comparison,
  discussion_log = discussion_log, after_discussion = after_discussion,
  profiles = phrase_profiles, session = sessionInfo())
saveRDS(agreement_record, file.path(phrase_review_dir, "agreement.rds"), version = 2)
write.csv(phrase_comparison$summary, file.path(phrase_review_dir, "agreement-summary.csv"),
  row.names = FALSE, na = "<MISSING>", fileEncoding = "UTF-8")
write.csv(phrase_profiles, file.path(phrase_review_dir, "review-stage-profiles.csv"),
  row.names = FALSE, na = "<MISSING>", fileEncoding = "UTF-8")
agreement_restored <- readRDS(file.path(phrase_review_dir, "agreement.rds"))
stopifnot(identical(agreement_restored, agreement_record), identical(
  compare_phrase_reviews(agreement_restored$comparison$reviews[[1]],
    agreement_restored$comparison$reviews[[2]], agreement_restored$comparison$reviewers),
  agreement_restored$comparison))
```

The optional block saves `kappa.rds` with its comparison, estimates and `irr`
version. For a study, retain the original submitted CSVs and the sampling record.
The RDS above retains original text, both reviews, comparison and discussion;
use a persistent study directory and respect the source materials' sharing terms.

A methods paragraph can follow this structure, replacing bracketed fields with
the actual design and numbers:

> Two raters independently applied [criterion/version] to [sampling rule]
> occurrences from [documents/writers]. Each item was one phrase ID at a fixed
> source span under [inventory/version and segmentation]. We retained unresolved
> and unreviewed states. For each phrase, binary observed agreement and unweighted
> Cohen's kappa (irr [version]) used the [n] jointly decisive occurrences of [N]
> sampled matches; category cross-tables and each rater's margins were reported.
> Discussion followed the independent ratings and was recorded separately.
> We compared document profiles under both ratings and subsequent decisions,
> counting overlapping accepted spans by their token union and withholding
> complete-review coverage for unfinished documents.

Use the independence sentence only for genuinely independent human work. A pilot
should test the written rules on separate practice material before a held-out
agreement sample; revised rules need a new recorded evaluation, not a relabeled
post-discussion score. This executable teaching example validates the workflow,
not human reliability, phrase-inventory completeness or learner proficiency.

## Connect to the existing ldfreq reference workflow

```{r adjacent, eval=has_quanteda}
prepared <- q$positions
prepared$term <- prepared$surface
sequences <- lexdiv_ngrams(prepared,
  preprocessing_id = "authored-penn-surface-exclude-period-tag-v1",
  documents = corpus$documents$document_id)
sequences$counts
```

Follow the **Adjacent n-gram frequency profiles** guide to build a local
reference with explicit source metadata and denominators. A preprocessing ID
is a declaration of comparable rules, not proof that an unrelated tokenizer
is compatible. MASC Penn tokens, quanteda word4 tokens and the bundled TUBELEX
tokenization must not be silently treated as the same tokenization.

Save source evidence with the analysis:

```{r saving, eval=has_quanteda}
path <- tempfile(fileext = ".rds")
saveRDS(list(corpus = corpus, quanteda = q, sequences = sequences), path)
restored <- readRDS(path)
stopifnot(identical(restored$corpus$provenance, corpus$provenance),
          identical(as.list(restored$quanteda$tokens), as.list(q$tokens)))
unlink(path)
```

## Use your local MASC files

The reader is verified against the **Penn layer of Mini-MASC 1.0**, using its
GrAF 1.0 `cesHeader` layout. Supply local `.anc` document headers and the actual
resource version. No corpus is downloaded by this function.
In the verified eight-document Mini-MASC release, 851 of 7,263 Penn tokens
have no `base` feature. They retain `NA` lemmas; a lemma analysis therefore
needs an explicit missingness policy even when surface tokens are available.

```{r local-corpus, eval=FALSE}
headers <- list.files("/path/to/Mini-MASC/data", pattern = "\\.anc$",
                      recursive = TRUE, full.names = TRUE)
masc <- lexdiv_read_masc(headers, resource_version = "Mini-MASC-1.0")
```

The source format includes separate primary text, segmentation, sentence or
utterance, and Penn feature files. Header references must resolve within the
header's directory. The reader rejects missing links, overlapping spans,
tokens crossing segment boundaries, and discontinuous tokens. It does not
silently repair those cases or switch to another annotation layer.

Reader interface 0.2.0 also accepts MASC 3.0.0's GrAF `documentHeader`
layout: `.hdr` files use `f.id="f.penn"` etc. and `loc` attributes. This is
format support with a restricted validated scope. A local audit of all 392
documents in the [official MASC 3.0.0 archive](https://anc.org/data/masc/downloads/data-download/)
accepted **122 documents** under the existing strict rules. First failures
among the other 270 were: 130 documents with tokens outside a single supplied
sentence/utterance region, 126 with overlapping sentence/utterance regions,
nine with invalid or incompatible anchors, four with overlapping tokens and
one with a feature structure outside the supported flat name/value format.
These are reader diagnostics, not a comprehensive correction of the corpus.
Multiple issues may exist in one file. No rejected annotations were repaired
or silently replaced by new sentence/token boundaries.

Consequently, do not catch errors, omit these files and label the remaining
sample "full MASC". Such exclusions can change register composition and
reference frequencies. Record accepted and rejected document IDs, reasons,
source hashes and preparation choices before interpreting any subset. The
accepted subset is useful for exercising the software; it is not supplied
as a representative English norm. The inspected archive's SHA-256 was
`0b300d96016233c7c79b43559b477a6bb0ed16bb416d5c517eccb108a6228493`.

The inspected [OANC GrAF archive](https://anc.org/data/oanc/download/) mainly
uses a `hepple` annotation layer with token regions in the same graph. It is
not the MASC Penn/seg layout. This reader does not provide a validated OANC
importer; shared GrAF namespaces alone do not establish compatibility.

[MASC](https://anc.org/data/masc/) is distributed under CC BY 3.0 US. Retain
its attribution, licence and changes when redistributing material. The
reader records file hashes and the declared release but does not determine
the licence of arbitrary input files. Corpus rights differ from the licence
of this package and its authored example.

MASC and [OANC](https://anc.org/data/oanc/) can support different reference
sizes and registers. Their sample composition, overlapping source material,
annotation choices and sparse long n-grams matter. A reference's usefulness
for your question requires more than successful file import. These functions
do not yet provide a validated standard MASC/OANC frequency resource.

Once inputs have been validated, `lexdiv_ngram_reference_build()` can accumulate
references in whole-document chunks. This avoids retaining every occurrence
table, while counts and the document roster remain in memory. See the
[n-gram guide](ngram-profiles.html#accumulate-a-reference-in-whole-document-chunks)
for the callback, denominator and exclusion-record requirements.
