---
title: "Compare token boundaries before comparing labels and document scores"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compare token boundaries before comparing labels and document scores}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ldfreq)
```

## Why token numbers are insufficient

One analyzer keeps `can't` as one token; another supplies `ca` and `n't`.
Every later token index shifts, although the original text is unchanged.
Japanese compounds create the same practical problem: `国際連合` and
`国際 | 連合` require different word units. Comparing rows would pair unrelated
occurrences. Comparing only the tokens that match could hide most of the change.

`lexdiv_align_annotations()` compares original source positions and connects
three distinct quantities: segmentation correspondence across all tokens,
optional label agreement on exactly corresponding spans, and the complete
sequences needed for document metrics. Its reference is a comparison baseline;
neither segmentation is automatically the correct linguistic unit.

## Import the complete annotations

Use `lexdiv_import_annotations()` separately for each analyzer's output, with
the same original document/segment IDs and text. Keep all non-whitespace source
text, including punctuation, before applying analytical exclusions. The importer
checks every surface and records segment-local, one-based, inclusive Unicode
codepoint positions. No text normalization or retokenization occurs here.
Byte offsets, UTF-16 offsets and grapheme clusters are different coordinate
systems. Changed segment rosters are rejected rather than guessed into alignment.

The installed, authored English/Japanese example runs entirely offline:

```{r}
env <- new.env(parent = baseenv())
sys.source(system.file("examples", "annotation-alignment.R",
  package = "ldfreq", mustWork = TRUE), env)
example <- env$annotation_alignment_example
aligned <- example$alignment
aligned$groups[c("document_id", "keyword", "reference_n",
                 "predicted_n", "relation")]
```

The relation is directional: `split` means one reference token and multiple
predicted tokens; `merge` means the reverse. `complex` means overlapping
many-to-many spans, such as `AB | C` versus `A | BC`. A single token on each
side with different endpoints is `changed_span`. Optional explicit whitespace
tokens can produce `reference_only` or `prediction_only` groups. No labels are
propagated from a whole compound to its parts.

`aligned$review_queue` contains all non-exact groups with `pre`, `keyword`
and `post` for reading the original context. Join its `alignment_id` to
`aligned$members` to recover every token on both sides. The original compound
IDs and both complete inputs preserve other annotations and reviewer notes.
These group IDs identify this result, not a permanent registry of occurrences.

## Report correspondence and conditional label agreement separately

```{r}
aligned$summary
aligned$annotation_evaluation$summary[
  c("tokens", "reference_available", "paired", "agreement_among_paired")]
```

Only four pairs have identical spans out of nine reference tokens and fifteen
predicted tokens: reference-side coverage is `4 / 9`, and predicted-side
coverage is `4 / 15`. All four paired labels agree. Reporting only their 100%
agreement would obscure the segmentation differences. Punctuation participates
in these counts; do not describe them as content-word accuracy.

The `token_span` row's recall and precision give these respective coverage
rates. The `internal_boundary` row compares junctions defined by **both** the
left token's end and the right token's start. This preserves whitespace gaps
and ownership; compulsory segment edges are excluded. Zero- and one-token
segments therefore have no boundary denominator, yielding `NA`, not a perfect
score. This declared junction convention may differ from an external benchmark.

To request label evaluation, supply all of `column`, `labels`, and
`reference_info`, using the same inventory and reference declarations as
[`lexdiv_evaluate_annotations()`](annotation-evaluation.html). Omit all three
for geometry alone. The conditional label evaluator retains both original
token indices and counts missing predictions on available references as false
negatives. Non-exact spans are outside its label denominators, with their
exclusion reasons visible in `members$relation`. Labels absent on exact spans
remain explicit too. Interpret errors only against a suitable reference standard;
these authored labels provide no accuracy evidence for a real analyzer.

## Calculate document metrics from each complete sequence

```{r}
example$differences
```

The Japanese example changes from four tokens and three types to six tokens
and four types. TTR changes from `3 / 4 = 0.75` to `4 / 6`, a difference of
`-1 / 12`. The English example changes from four to eight tokens while TTR
stays `0.75`: unchanged TTR does not imply unchanged segmentation. The example
also calculates MATTR with a fixed four-token window. One-token and empty
documents remain in the roster with their metric-specific missing reasons;
the requested window is not silently reduced.

This comparison uses **all supplied surface tokens**, including punctuation,
with no POS selection or normalization. That explicit policy isolates the
segmentation contrast. For a content-word or lemma analysis, declare the
selection and lexical-form rules, retain missing-label coverage, and do not
infer a full-document score from an incomplete reference. Metrics across
different units show sensitivity to those units; they do not establish that
one analyzer yields a more valid psychological measure. Use the whole input
for each condition, rather than computing diversity on the matched subset.

## Save and replay without rerunning an analyzer

```{r, eval=FALSE}
saveRDS(example, "annotation-alignment.rds", version = 2)
restored <- readRDS("annotation-alignment.rds")
replayed <- do.call(lexdiv_align_annotations, restored$inputs)
stopifnot(identical(replayed, restored$alignment))
```

The saved object includes the original inputs, reference declarations, metric
settings and results. Keep the installed script, package version and
`sessionInfo()` too. Source-containing RDS files remain subject to the original
data's access conditions. This workflow needs no bundled corpus or model;
it does not perform word-sense disambiguation or dependency-head evaluation.
Independent reference construction and evaluation on the actual research
population are separate from this reproducible demonstration.
