---
title: "Understanding unmatched words and lexical-unit choices"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Understanding unmatched words and lexical-unit choices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ldfreq)
```

## Begin with the question

A teacher wants to describe how much vocabulary in student essays matches
New JACET 8000. Several frequent words fail to match. Before interpreting this
as advanced vocabulary, check which lexical units were compared with the list.
NJ8 contains representative headwords; surface forms and dictionary lemmas can
produce different coverage. These authored sentences make that choice visible.

```{r}
texts <- c(a = "Students have jobs.", b = "Students enjoy jobs.", blank = "")
prepared <- lexdiv_tokenize_batch(texts, tokenizer = "english", case = "lower")
surface <- nj8_profile_batch(prepared, unit = "surface")
audit <- nj8_diagnostics(surface)
audit$unmatched_terms
audit$coverage[, c("document_id", "eligible_tokens", "off_list_tokens", "token_coverage")]
```

`token_count` counts occurrences; `document_count` counts distinct documents
containing the term. Repeating a word in one essay raises only the former.
The empty document remains in coverage with an undefined rate. Do not silently
drop it when describing the sample, or replace its undefined coverage with zero.

## Compare explicit annotations on the same input

Here we supply lemmas ourselves so that the example runs without an optional
dictionary or model. In real work use reviewed annotations, or explicitly call
`lexdiv_lemmatize(x, method = "textstem")` after installing `textstem`.
That dictionary backend does not provide context-sensitive POS disambiguation.

```{r}
lemmas <- list(a = c("student", "have", "job"),
               b = c("student", "enjoy", "job"), blank = character())
annotated <- Map(function(x, lemma) lexdiv_lemmatize(x, lemmas = lemma,
  backend_id = "authored-example", backend_version = "1"), prepared, lemmas)
lemma <- nj8_profile_batch(annotated, unit = "lemma")
comparison <- data.frame(document_id = surface$coverage$document_id,
  surface_coverage = surface$coverage$token_coverage,
  lemma_coverage = lemma$coverage$token_coverage)
comparison
audit_lemma <- nj8_diagnostics(lemma)
subset(audit_lemma$unit_mappings, unit_changed)
```

The change is an answer about matching different units to the same resource.
It is not evidence that a writer learned more words or that lemmatization is
universally preferable. Choose a lexical unit for the research question and
keep it consistent across documents. For a supplied character vector, NJ8
cannot reconstruct surface forms that the caller already replaced upstream.

## Compare the same documents under explicit conditions

Lemmatization and word inclusion can change lexical findings
([Caltabellotta et al., 2026](https://doi.org/10.1016/j.asw.2026.101039)).
Separate the choice of lexical unit from the choice of MATTR window. This
example crosses two units with two windows using the same annotated documents.
These very small windows illustrate the workflow; they are not recommended
settings for learner essays or a replication of the cited study.

```{r}
conditions <- expand.grid(unit = c("surface", "lemma"),
  window_length = c(2L, 4L), stringsAsFactors = FALSE)
conditions$condition_id <- paste0(conditions$unit, "_w", conditions$window_length)

condition_runs <- lapply(seq_len(nrow(conditions)), function(i) {
  lexdiv_metrics_text_batch(annotated, unit = conditions$unit[i],
    word_inclusion = "all", metrics = "mattr",
    window_length = conditions$window_length[i])
})
names(condition_runs) <- conditions$condition_id

condition_results <- do.call(rbind, lapply(seq_along(condition_runs), function(i) {
  rows <- condition_runs[[i]]$results
  stopifnot(identical(rows$document_id, names(annotated)))
  data.frame(condition_id = conditions$condition_id[i],
    unit = conditions$unit[i], window_length = conditions$window_length[i],
    rows[, c("document_id", "N", "V", "value", "status", "missing_reason")])
}))
row.names(condition_results) <- NULL
condition_results
```

All three documents remain in each condition. The four-token window is too
large for these sentences; the output retains the missing result and its reason.
The empty document also remains. For computable rows, these toy sentences have
no within-sentence repetition in either unit, so MATTR is unchanged even though
NJ8 coverage changes. Coverage and diversity answer different questions.

Keep `condition_runs` as well as this display table: the full objects retain
method identities, requested/effective parameters, token audits and preprocessing.
For a paired comparison, match document IDs within a fixed specification and
report the number of usable pairs; do not silently omit failed conditions.

An all-word/content-word comparison additionally needs aligned UPOS annotations
and their backend identity on every document. Once these exist, repeat the same
explicit workflow with `word_inclusion = "content"`. The retained token count
then changes, and a window spans selected words rather than the original token
positions. A surface/lemma contrast is also different from a default/custom
lemmatizer contrast: the latter requires two separately recorded annotation
sets on the same original texts. Do not treat them as interchangeable replications.

## Audit alternative annotation schemes

The custom scheme in Caltabellotta et al. (2026, Section 4.3 and Table 1)
distinguished verbal mood and tense. It was not a dictionary-correction exercise.
Here we supply a small authored example with explicit POS-sensitive labels and
a refined verbal scheme. The package counts supplied strings as given;
`unit = "lemma"` does not automatically append POS, mood or tense.

```{r}
scheme_tokens <- lexdiv_tokenize(
  "Students see students and saw teachers.", case = "lower")
scheme_pos <- c("NOUN", "VERB", "NOUN", "CCONJ", "VERB", "NOUN")
base_labels <- c("student_NOUN", "see_VERB", "student_NOUN",
  "and_CCONJ", "see_VERB", "teacher_NOUN")
refined_labels <- base_labels
refined_labels[c(2, 5)] <- c("see_VERB_Ind_Pres", "see_VERB_Ind_Past")
make_scheme <- function(labels, version) {
  lexdiv_lemmatize(scheme_tokens, lemmas = labels, upos = scheme_pos,
    backend_id = "authored-counting-scheme", backend_version = version,
    upos_backend_id = "authored-upos", upos_backend_version = "1")
}
base_scheme <- make_scheme(base_labels, "base-v1")
refined_scheme <- make_scheme(refined_labels, "tense-v1")
annotation_audit <- lexdiv_compare_annotations(base_scheme, refined_scheme)
annotation_audit$documents
annotation_audit$changes[, c("token_index", "surface", "before_lemma", "after_lemma")]

schemes <- list(base = base_scheme, refined = refined_scheme)
scheme_conditions <- expand.grid(scheme = names(schemes),
  inclusion = c("all", "content"), stringsAsFactors = FALSE)
scheme_runs <- lapply(seq_len(nrow(scheme_conditions)), function(i) {
  lexdiv_metrics_text_batch(list(essay = schemes[[scheme_conditions$scheme[i]]]),
    unit = "lemma", word_inclusion = scheme_conditions$inclusion[i],
    metrics = "mattr", window_length = 3)
})
names(scheme_runs) <- with(scheme_conditions, paste(scheme, inclusion, sep = "_"))
lapply(scheme_runs, function(x) x$results[, c("document_id", "N", "V", "value")])
```

Two lemma labels changed; no UPOS labels changed. The conjunction remains in the
all-word condition and is excluded in the content-word condition. The window
length is measured in the selected tokens. These authored annotations and the
three-token window illustrate the workflow, not a replication or a recommended
setting. The published study used spaCy, a 50-token window, and the same
processing for learner texts and its reference corpora.

Do not submit these POS/morphology keys directly to NJ8 or an ordinary lemma
frequency table: their lexical units differ. Keep separate prepared objects
for headword lookup and scheme-sensitive diversity, with explicit identities.
For frequency comparisons, the reference counts need a corresponding scheme.
An improved match rate alone cannot establish that correspondence.

`lexdiv_compare_annotations()` also accepts named batches. It matches IDs even
if their order differs, retains empty documents, and rejects differences in
original text or tokenization. Missing-to-value and value-to-missing transitions
are changes; two missing values are unchanged. Backend-only differences remain
visible in `$provenance` even when `$changes` is empty. Save both prepared inputs
and the audit: `saveRDS(list(before = base_scheme, after = refined_scheme,
audit = annotation_audit, conditions = scheme_conditions, runs = scheme_runs), path)`.
Backend labels alone are not proof of dictionary content identity. The next
example records the dictionary content fingerprint and saves the actual table.

## Choose and save the actual lemma dictionary

A backend version alone does not identify its dictionary. With optional
`textstem` installed, pass a two-column character data frame and give it an
explicit identity and version. Lookup terms must be unique and lowercase under
the current R locale. Unmatched tokens retain their surface forms. This tiny
authored dictionary illustrates lookup; it is not a general English dictionary.

```{r}
if (requireNamespace("textstem", quietly = TRUE)) {
  dictionary <- data.frame(term = c("cats", "ran", "third"),
    lemma = c("cat", "run", "3"))
  input <- lexdiv_tokenize("Cats ran. The third group ate one third.",
    tokenizer = "english", case = "lower")
  baseline <- lexdiv_lemmatize(input, method = "textstem",
    dictionary = dictionary, dictionary_id = "authored-demo",
    dictionary_version = "1")
  baseline$tokens[, c("token_index", "surface", "lemma")]
  baseline$provenance$annotation$dictionary
}
```

The record includes a content fingerprint, entry count and lookup locale.
Reordering rows leaves the fingerprint unchanged; changing a mapping changes
it even if the version label stays the same. With `dictionary = NULL`, the
package uses `lexicon::hash_lemmas` and records the installed lexicon version
and the actual table's fingerprint. This records context-free lookup, not
POS disambiguation or evidence that the labels are linguistically correct.

The example has two occurrences of `third`. Suppose an analysis specifies
word-form labels for ordinal expressions, while leaving the fractional
expression for separate review. Editing the dictionary would affect both.
Instead, record the reviewed token position and supply an aligned lemma vector.
This is an illustration of an explicit counting scheme, not a recommendation
to retain numeric fraction labels in every study.

```{r}
if (requireNamespace("textstem", quietly = TRUE)) {
  revisions <- data.frame(token_index = 4L, before_lemma = "3",
    after_lemma = "third", reason = "ordinal word-form counting scheme")
  stopifnot(baseline$tokens$surface[revisions$token_index] == "third",
    baseline$tokens$lemma[revisions$token_index] == revisions$before_lemma)
  reviewed_lemmas <- baseline$tokens$lemma
  reviewed_lemmas[revisions$token_index] <- revisions$after_lemma
  reviewed <- lexdiv_lemmatize(input, lemmas = reviewed_lemmas,
    backend_id = "authored-context-review", backend_version = "1")
  changes <- lexdiv_compare_annotations(baseline, reviewed)
  changes$changes[, c("token_index", "surface", "before_lemma", "after_lemma")]
  saved <- tempfile(fileext = ".rds")
  saveRDS(list(dictionary = dictionary, before = baseline, after = reviewed,
    revisions = revisions, audit = changes), saved)
  restored <- readRDS(saved)
  replay <- lexdiv_lemmatize(input, method = "textstem",
    dictionary = restored$dictionary, dictionary_id = "authored-demo",
    dictionary_version = "1")
  stopifnot(identical(replay, restored$before))
  unlink(saved)
}
```

Use a persistent path for real analyses. The prepared object retains dictionary
metadata, not the dictionary table; saving only its hash cannot reproduce
lookup. Retain the actual dictionary, backend versions, locale, both annotation
sets and revision reasons. For a POS-tagged analysis also pass the aligned UPOS
vector and its backend identity when reannotating. Missing tags are not inferred.
Older saved annotations remain readable; missing dictionary records cannot
establish strict lemma comparability in `lexdiv_content_overlap()`.

## Inspect surprising transformations

```{r}
example <- lexdiv_tokenize("second compound missing", case = "lower")
example <- lexdiv_lemmatize(example, lemmas = c("2", "two words", NA),
  backend_id = "authored-transformations", backend_version = "1")
example_profile <- nj8_profile(example, unit = "lemma")
diagnostic <- nj8_diagnostics(example_profile)
diagnostic$unit_mappings
diagnostic$exclusion_reasons
```

The numeric and whitespace flags describe changes; they do not establish an
error. The phrase `two words` remains one supplied lexical unit. Missing lemmas
are excluded, not counted as unmatched. `lookup_changed` can reflect query
normalization or a flemma headword-conflict decision. Keep the original
`profile$lookup` for occurrence positions and the prepared objects for annotation
details. No automatic spelling correction, compound splitting or numeric-label
removal is performed by the diagnostics.

## Keep window sensitivity separate from vocabulary coverage

The existing plan API compares common windows on ordered tokens. These repeated
tokens are an illustration of parameter effects, not learner observations.

```{r}
documents <- list(clustered = rep(c("river", "stone", "cloud"), each = 20),
                  interleaved = rep(c("river", "stone", "cloud"), 20))
method <- lexdiv_methods()$method_id[lexdiv_methods()$metric_id == "mattr"]
plan <- lexdiv_plan(presets = character(), grids = lexdiv_grid(
  method, "window_length", c(10, 20), request_id_prefix = "window"))
results <- lexdiv_profile_batch(documents, plan)
results[, c("document_id", "request_id", "value", "status")]
```

Use the same requested windows for every text; retain missing results when a
text is too short. Do not choose a different window for each writer to improve
its score or average unlike conditions together. Lexical diversity and NJ8
coverage answer different questions, so neither provides a combined proficiency
score. Generalizing to proficiency or writing quality needs an appropriate
external criterion and a design that respects writers, tasks and raters.

## Distinguish window size from position effects

Text length and the reduction parameter are separate sensitivity questions
([Bestgen, 2024](https://doi.org/10.1111/lang.12630)). MATTR also weights
positions unequally through its overlapping windows
([Bestgen, 2025](https://doi.org/10.1016/j.rmal.2024.100168)). This authored
counterexample holds the number and frequency of tokens fixed and moves one
unique token from an endpoint to the middle.

```{r}
positions <- list(endpoint = c("x", rep("a", 6)),
  middle = c(rep("a", 3), "x", rep("a", 3)))
position_scores <- lexdiv_metrics_batch(positions,
  metrics = c("ttr", "mattr", "hdd"), window_length = 3, sample_size = 3)
position_scores[, c("document_id", "metric_id", "N", "V", "value")]

position_plan <- lexdiv_plan(presets = character(), specs = lexdiv_spec(
  method, list(window_length = 3), request_id = "mattr_3"))
position_profile <- lexdiv_mattr_profile(positions$endpoint, position_plan)
position_profile$exposure[, c("position", "exposure_count")]
```

There are five complete windows. At the endpoint, `x` appears in one window;
in the middle it appears in three. MATTR is therefore 0.4 and 8/15 (about 0.533),
respectively. TTR and HD-D are identical across the two orders because their
calculations here depend on the same frequency distribution. This checks a
mathematical property, not which index best captures learners' proficiency.
Exposure counts describe window membership, not each token's causal contribution.
Keep the canonical MATTR calculation; dividing its score by exposure counts
would define a different, unvalidated index.

The local windows overlap and are not independent observations. A future
length/position study should retain the parent document ID, distinguish
contiguous excerpts from randomly sampled tokens, and use uncertainty estimates
that respect repeated observations within writers. A single MATTR window
comparison does not establish text-length invariance.

## Save a reproducible analysis

Save the complete profiles and prepared objects, not just a top-word list:
`saveRDS(list(prepared = annotated, profile = lemma, audit = audit_lemma), path)`.
Use `write.csv(audit_lemma$unmatched_terms, path, row.names = FALSE)` for inspection
in a spreadsheet. These objects can contain text-derived data; choose a
destination appropriate to the corpus. Record the package version, selected
unit, annotation backend, normalization, exclusions and NJ8 citation alongside
any reported coverage.

For the condition comparison, also save
`saveRDS(list(prepared = annotated, conditions = conditions, runs = condition_runs,
results = condition_results), path)`. Save the complete objects before exporting
display tables to CSV; a CSV alone does not preserve their annotation records.
