---
title: "Connect contextual model outputs to research occurrences"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Connect contextual model outputs to research occurrences}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ldfreq)
has_quanteda <- requireNamespace("quanteda", quietly = TRUE) &&
  packageVersion("quanteda") >= "4.5.0" && isTRUE(l10n_info()[["UTF-8"]])
```

## The research problem

For a word such as *bank*, a model can provide a representation of each
occurrence in context. The research question might be whether those
representations distinguish financial from river-edge uses. Computing vectors
does not establish that distinction: the target, context, candidate inventory
and an independent evaluation must also be specified.

`lexdiv_import_contextual()` connects external model output to source occurrences
used for KWIC review. It checks joins, preserves missing output and keeps model
suggestions separate from human decisions. It does not run BERT, install Python,
download weights, select a meaning, or validate a semantic measure. The importer
needs no model software. Creating the review uses optional quanteda; the
English/Japanese example below requires a UTF-8 R session.

## An offline example with authored values

These sentences, candidates, vectors and scores are **authored illustrations**,
not model predictions, corpus observations or psycholinguistic norms. Japanese
`人気` illustrates different readings of the same spelling, not a demonstrated
related-sense manipulation. Use real annotations for a real study.

```{r prepare, eval=has_quanteda}
words <- list(c("The", "bank", "lent", "money", "."),
              c("The", "river", "bank", "flooded", "."),
              c("その", "店", "は", "人気", "が", "ある", "。"),
              c("ここ", "は", "人気", "が", "ない", "。"))
tokens <- data.frame(document_id = "example",
  segment_id = rep(paste0("s", 1:4), lengths(words)),
  token_index = sequence(lengths(words)), surface = unlist(words, use.names = FALSE))
segments <- data.frame(document_id = "example", segment_id = paste0("s", 1:4),
  text = c("The bank lent money.", "The river bank flooded.",
           "その店は人気がある。", "ここは人気がない。"))
annotations <- lexdiv_import_annotations(tokens, segments, list(
  language = "en/ja", analyzer = "authored", analyzer_version = "1",
  dictionary = "none", dictionary_version = "none", unit = "authored",
  normalization = "none"))
candidates <- data.frame(term = c("bank", "bank", "人気", "人気"),
  candidate_id = c("financial", "river", "popularity", "human-presence"),
  label = c("financial institution", "river edge", "にんき", "ひとけ"))
review <- lexdiv_ambiguity_review(annotations, c("bank", "人気"), candidates,
  list(resource_id = "authored", resource_version = "1",
    source_reference = "Authored candidates, not a dictionary", data_license = "MIT"))

# Export anchors BEFORE computation. Never reconstruct them after a model run.
input <- review$occurrences[c("review_id", "occurrence_id", "segment_text",
                              "surface", "start", "end")]
data <- input[1:3, ]
data$status <- c("processed", "processed", "skipped")
data$reason <- c("Authored transport example", "Authored transport example",
                 "Illustrated failure, not an actual model limitation")
vectors <- matrix(c(1, 0, 0, 1), nrow = 2,
                  dimnames = list(data$occurrence_id[1:2], NULL))
scores <- data.frame(occurrence_id = rep(data$occurrence_id[1:2], each = 2),
  candidate_id = rep(c("financial", "river"), 2), score = c(2, -1, 0.5, 0.5))
model <- list(model_id = "authored-example", model_revision = "1",
  tokenizer_id = "authored", tokenizer_revision = "1",
  software = "illustration", software_version = "1",
  context_policy = "full segment; no truncation",
  representation = "authored two-dimensional vectors, not model output",
  score_definition = "authored scores; larger is preferred; not probabilities")
imported <- lexdiv_import_contextual(review, data, model, vectors, scores)
imported$summary
imported$occurrences[c("surface", "status", "model_status", "has_embedding")]
imported$suggestions[c("pre", "keyword", "post", "label", "score", "human_status")]
stopifnot(identical(imported$review, review),
          imported$summary$processed_proportion == 0.5,
          all(imported$occurrences$status == "unreviewed"))
```

Two of four occurrences have output. Skipped and omitted (`not_returned`)
occurrences remain in the denominator. Tied and negative scores are retained;
importing them does not create probabilities or human decisions. The suggestions
table contains KWIC and candidate labels; `segment_text` retains the full original
segment. Scores may cover only part of an inventory: an absent score is missing,
not zero. Use the ordinary review decision table to record later human selections
and keep the earlier review as well. Model-assisted judgments are not independent
human judgments; conduct a blinded human pass before showing suggestions when
independent accuracy or agreement is the research question.

## Compare scores with a declared reference

`lexdiv_evaluate_contextual()` predicts a candidate only when **every candidate
in the supplied inventory is scored** and one is best. Partial lists and tied
best scores abstain. This conservative policy is not a WSD benchmark convention
or proof that the inventory contains every meaning. Specify whether larger or
smaller scores are preferred. The optional absolute `tie_tolerance` treats nearby
scores as ties; choose it before final evaluation, not by tuning on test labels.

These reference labels are also authored illustrations, not independent human
gold labels. Agreement here is not evidence about BERT accuracy. Keep the
reference separate from the imported model output and its original human review.

```{r evaluate-scores, eval=has_quanteda}
decisions <- input[c("review_id", "occurrence_id")]
decisions$status <- c("selected", "selected", "selected", "unresolved")
decisions$candidate_id <- c("financial", "river", "popularity", NA_character_)
decisions$reviewer <- "authored-example"
decisions$reason <- c(rep("Authored criterion for this example", 3),
                      "Illustration of an unresolved reference")
reference_review <- lexdiv_ambiguity_review(annotations, c("bank", "人気"),
  candidates, review$provenance$resource, decisions)
evaluation <- lexdiv_evaluate_contextual(imported, reference_review,
  direction = "higher", reference_info = list(
    reference_id = "authored-demonstration", annotation_protocol = "Authored examples, not human ratings",
    model_exposure = "unknown", evaluation_role = "development"))
evaluation$summary[c("agreement", "paired", "agreement_among_paired",
  "predictions", "occurrences", "prediction_coverage",
  "reference_selected", "matches_among_reference_selected")]
evaluation$review_queue[c("surface", "prediction_status", "model_status",
  "reference_status", "reference_reason", "outcome")]
stopifnot(evaluation$summary$agreement_among_paired == 1,
          evaluation$summary$prediction_coverage == 1/4,
          evaluation$summary$matches_among_reference_selected == 1/3,
          identical(evaluation$reference, reference_review),
          identical(evaluation$model_output, imported))
saved_evaluation <- tempfile(fileext = ".rds")
saveRDS(evaluation, saved_evaluation)
stopifnot(identical(readRDS(saved_evaluation), evaluation))
unlink(saved_evaluation)
```

The one comparable case agrees (1/1), but only one of four occurrences receives
a prediction. Among the three occurrences with selected reference labels, one
receives a matching prediction (1/3). "100% agreement" alone would hide most
cases. The KWIC `review_queue` retains the tied English example, the skipped
Japanese example and the unresolved reference. `terms` repeats counts by surface,
including absent targets; `confusion` keeps candidate IDs scoped to each surface.
Overall proportions weight occurrences rather than word types. A singleton
inventory produces a trivial prediction, counted in `single_candidate_predictions`;
inspect these separately when evaluating ambiguity.

All inputs remain unchanged. An unresolved or unreviewed reference is not a wrong
label and is not counted as agreement. Embedding-only output yields no predictions:
a separately specified classifier or scoring procedure is still needed.
`reference_scored` identifies whether a selected reference candidate appeared in
the scores, distinguishing a missing reference candidate from missing rivals.

Use `model_exposure = "not_shown"` only when annotators did not see model output,
and `evaluation_role = "held_out"` only when examples were not used to fit or
choose the scoring procedure or evaluation settings. These declarations are not
verified by the package. Model-assisted judgments (`shown`) and development
examples are valid descriptive inputs, but are not independent held-out accuracy.

WSD evaluation depends on consistent annotation guidelines and sense inventories
([Raganato et al., 2017](https://aclanthology.org/E17-1010/)). Performance on answered
cases and coverage are also distinguished in selective classification
([Geifman & El-Yaniv, 2017](https://papers.neurips.cc/paper_files/paper/2017/hash/4a8423d5e91fda00bb7e46540e2b0cf1-Abstract.html)).
This evaluator implements neither their training algorithms nor their statistical
guarantees; it reports counts and descriptive proportions.

For a candidate-frequency baseline, estimate its scores on separate training
data, import them against the same unchanged review, and evaluate with the same
reference and predeclared policy. Keep each method's coverage and compare their
overlapping evaluable occurrence IDs as well. A higher conditional agreement
based on fewer, easier cases is not evidence of improvement. Do not derive
baseline frequencies from held-out labels or equate different candidate inventories.

## Score separate training and query examples in R

`lexdiv_score_contextual()` bridges imported vectors and the evaluator. Its
centroid baseline averages the raw target vectors for each selected training
candidate, then computes cosine similarity to a query vector. Averaging labeled
contextual embeddings is a starting point in
[Loureiro & Jorge (2019), section 4.1](https://aclanthology.org/P19-1569/).
This small baseline does not implement their full LMMS method, WordNet
propagation, gloss representations or reported benchmark results.

The comparison baseline uses selected training-label counts for the same surface.
Unlike centroid fitting, it can use selected examples without embeddings.
Always inspect both `n_selected` and `n_vectors`. Neither score is a probability.
The two methods remain separate: there is no automatic frequency fallback.

Create training and query sources separately **before** model computation.
Stable document IDs must not overlap; renamed copies of identical target-bearing
segments are also rejected. Candidate tables/resources and embedding declarations
(including model/tokenizer revisions, layer/aggregation and software) must match.
Matching declarations are not proof that external software followed them.

The following sentences, labels and vectors are authored for illustration.
They are not an independent annotation study or a trained BERT benchmark.

```{r score-baselines, eval=has_quanteda}
train_words <- list(c("Her", "bank", "approved", "a", "loan", "."),
                    c("A", "bank", "holds", "deposits", "."),
                    c("We", "rested", "by", "the", "bank", "."),
                    c("人気", "の", "店", "だ", "。"),
                    c("周囲", "に", "人気", "が", "ない", "。"))
train_tokens <- data.frame(document_id = "training-example",
  segment_id = rep(paste0("t", 1:5), lengths(train_words)),
  token_index = sequence(lengths(train_words)), surface = unlist(train_words, use.names = FALSE))
train_segments <- data.frame(document_id = "training-example", segment_id = paste0("t", 1:5),
  text = c("Her bank approved a loan.", "A bank holds deposits.", "We rested by the bank.",
           "人気の店だ。", "周囲に人気がない。"))
train_annotations <- lexdiv_import_annotations(train_tokens, train_segments,
  annotations$provenance$annotation)
train_review <- lexdiv_ambiguity_review(train_annotations, c("bank", "人気"), candidates,
  review$provenance$resource)
train_input <- train_review$occurrences[c("review_id", "occurrence_id", "segment_text", "surface", "start", "end")]
train_data <- train_input
train_data$status <- "processed"; train_data$reason <- "Authored training vectors"
train_vectors <- rbind(c(2,0), c(2,1), c(0,2), c(2,0), c(0,2))
rownames(train_vectors) <- train_input$occurrence_id
train_imported <- lexdiv_import_contextual(train_review, train_data, model, train_vectors)
train_decisions <- train_input[c("review_id", "occurrence_id")]
train_decisions$status <- "selected"
train_decisions$candidate_id <- c("financial", "financial", "river", "popularity", "human-presence")
train_decisions$reviewer <- "authored-example"
train_decisions$reason <- "Intended labels of authored sentences; not independent human judgments"
train_reference <- lexdiv_ambiguity_review(train_annotations, c("bank", "人気"), candidates,
  review$provenance$resource, train_decisions)
scored <- lexdiv_score_contextual(imported, train_imported, train_reference,
  training_info = list(training_id = "authored-training",
    annotation_protocol = "Author-specified intended labels",
    partition_protocol = "Separate authored documents and target contexts",
    model_exposure = "unknown"))
scored$candidates[c("term", "candidate_id", "n_selected", "n_vectors", "prototype_status")]
scored$query_occurrences[c("surface", "model_status", "centroid_status", "frequency_status")]
baseline_evaluations <- lapply(scored[c("centroid", "frequency")], function(output)
  lexdiv_evaluate_contextual(output, reference_review, "higher",
    evaluation$provenance$reference_info))
comparison <- do.call(rbind, lapply(baseline_evaluations, function(result)
  result$summary[c("agreement", "paired", "prediction_coverage", "agreement_among_paired")]))
comparison
baseline_evaluations$centroid$review_queue[c("keyword", "pre", "post", "prediction_status", "model_reason")]
stopifnot(comparison$agreement_among_paired == c(1, 0.5),
          all(comparison$prediction_coverage == 0.5),
          identical(scored$query, imported), identical(scored$reference, train_reference))
saved_scores <- tempfile(fileext = ".rds")
saveRDS(scored, saved_scores)
stopifnot(identical(readRDS(saved_scores), scored))
unlink(saved_scores)
```

The centroid method matches both English reference choices in this constructed
example; the frequency baseline picks the financial candidate for both. This
is a demonstration of different rules, not empirical evidence of improvement.
The Japanese queries have no imported vectors, so no centroid scores are made.
Frequency scores exist for those queries, but tied training counts mean that
the evaluator abstains. Scoring-stage `processed` therefore does not mean a
unique prediction, an available embedding or a correct interpretation.

A missing candidate prototype stays missing and causes evaluator abstention
when the score list is incomplete. Frequency zero means no selected training
example of that candidate; an entirely unobserved surface receives no scores.
Zero vectors and canceling means are recorded separately. Training counts can
be small or uneven; inspect them before interpreting a candidate comparison.

For a real run, use the optional embedding script below separately on the two
source rosters with identical settings, import both outputs, and supply the
training review and independent query reference. Do not relabel query outputs
as training data or manufacture new IDs to bypass the split check. Participant
overlap, near duplicates, pretraining contamination and earlier model-selection
leakage are not detected automatically. This baseline cannot generalize to a
new surface with no training labels; a word-held-out study needs another method.
Save the entire `scored` object: each score output contains training fingerprints
and required declarations, while the full object retains all training examples,
prototype vectors and exclusion reasons.

## A study folder that survives a new R session

The installed `examples/contextual-study/` template connects preparation,
development, a frozen analysis, final scoring and replay. It follows the useful
file handoffs in [Masaki Eguchi's LDA II project template](https://github.com/egumasa/lda2-proj-template),
using existing ldfreq functions and local R files. No code or data from that
project is included. No cloud account, Python, model or new dependency is needed
to run this illustration; optional quanteda and a UTF-8 R session are required.

**Every sentence, label, group and vector is authored.** The 14 documents are
six training, four development and four test illustrations. Group IDs demonstrate
split checks; they do not identify independent participants. The word *bank*
and the readings of Japanese `人気` illustrate different candidate inventories,
not validated equivalent ambiguity manipulations. The example's test evaluation
uses `evaluation_role = "unknown"`, not a claim of independent held-out accuracy.

| Script | What it does | Saved handoff |
|---|---|---|
| `01-prepare.R` | Creates source-checked occurrences, a document/group partition table, two authored judgments and a separate reference, plus vectors and a missing-vector case | `inputs/`: design, metadata, request CSVs, model and reference RDS files, readable decision CSV snapshots |
| `02-evaluate.R` | Checks document/group separation and exact target-context copies across partitions, evaluates development examples, freezes the settings and training inputs, then writes test scores before opening test reference labels | `analysis/`: development results, frozen inputs/settings/session, test scores and evaluations |
| `03-report.R` | Recomputes evaluation from saved scores and references, checks agreement with the saved result, and creates coverage and common-occurrence comparison tables | An R object `report`; this stage reads files without changing them |

Copy the scripts to a **new study directory**. Preparation and evaluation refuse
existing output directories, so a later run cannot silently replace an earlier
one. The report can be replayed repeatedly. If a stage fails after writing a
partial result, retain it for diagnosis; fix the input and use a new run directory.
Directories and RDS files are not access controls, tamper-proof registration or
proof that a researcher did not inspect test examples.

```{r study-template, eval=has_quanteda}
template <- system.file("examples", "contextual-study", package = "ldfreq", mustWork = TRUE)
study_dir <- tempfile("ldfreq-study-") # Choose a permanent new folder for your study.
dir.create(study_dir)
stopifnot(all(file.copy(list.files(template, pattern = "[.]R$", full.names = TRUE), study_dir)))
run_stage <- function(script) {
  previous <- setwd(study_dir)
  on.exit(setwd(previous))
  env <- new.env(parent = globalenv())
  sys.source(script, envir = env)
  env
}
invisible(run_stage("01-prepare.R"))
invisible(run_stage("02-evaluate.R"))
study_report <- run_stage("03-report.R")$report
study_report$comparison[c("method", "occurrences", "predictions", "paired",
  "agreement_among_paired", "matches_among_reference_selected")]
study_report$common_comparison
study_report$judgments$summary
```

The centroid method predicts 3/4 examples and matches 2/2 selected references
among those predictions. The frequency baseline predicts 4/4 and matches 2/3
selected references. Both match **2/3 of all selected references**. On the two
examples where both methods and the reference are available, the match counts
are 2/2 and 1/2. These are different denominators and populations, not competing
estimates of the same quantity. Common-case coverage is only 2/4. The unresolved
fourth reference is not counted as correct or incorrect for either method.

The two authored judges jointly select a candidate in 3/4 cases and agree in
2/3 of those cases. Their individual judgments survive adjudication. An unresolved
case remains unresolved in the reference. `report$pairs` retains group/language
metadata, predictions, reference status and KWIC; `report$terms` and
`report$review_queue` keep word-level results and cases needing inspection.
The counts are too small, and the values too constructed, to support a claim
about English/Japanese accuracy, proficiency or psychological validity.

After changing to the copied study directory, these commands also run in three
separate R sessions. Replaying the report does not need the live `inputs/` files
or another model inference:

```sh
Rscript --vanilla 01-prepare.R
Rscript --vanilla 02-evaluate.R
Rscript --vanilla 03-report.R
```

### Replace the authored inputs for a real study

Keep the three file handoffs, and adapt the preparation script to your source
annotations, candidate inventory and independently collected decisions. In
particular:

1. Define the target population, sense/reading criteria, sampling and grouping
   before final evaluation. Use real author/speaker IDs for a study of new
   participants. For a study of new documents from known participants, define
   the grouping accordingly and describe that narrower claim. A supplied group
   ID is checked for overlap, not for its truth. Near-duplicate contexts and
   pretraining contamination still require separate assessment.
2. Create unreviewed occurrence rosters and collect each annotator's decisions
   without model suggestions. Retain their initial tables, disagreement review
   and reasons for adjudication separately. Rebuild each reference with
   `lexdiv_ambiguity_review(..., decisions = decisions)`, using its unchanged
   source and candidate resource. The CSV files here are readable snapshots;
   **editing them does not update the saved RDS inputs**. Read your finalized
   table explicitly (for these CSVs, use `na.strings = ""` and UTF-8), pass it to
   the review API, then save that rebuilt object in the reference bundle.
3. Replace the authored vectors with actual source-aligned outputs imported by
   `lexdiv_import_contextual()`. Keep the same declared model/layer/aggregation
   across partitions. Use development data to choose those settings, leaving
   final evaluation untouched. The scorer uses training labels only; this
   example keeps the original training data after development rather than
   refitting on development labels. Candidate inventories must agree, but a
   candidate need not occur in every partition. Report rare/missing candidates.
   If a partition yields no vectors, preserve a zero-row matrix with the known
   model dimension and column names; `NULL` does not declare those dimensions.
4. Replace the authored protocol and `unknown` declarations only with accurate
   descriptions of your study. Freeze the selected settings and save test
   scores before examining test labels. Do not choose a model or tolerance by
   repeatedly trying the final evaluation set. Keep skipped examples, abstentions
   and unresolved references in their respective denominators.

For statistical comparisons, define the estimand and account for repeated
documents/participants, word inventories and unequal missingness. The template
supplies descriptive tables, not confidence intervals or a test of superiority.
Candidate-frequency counts describe the labeled training sample, which may be
balanced or purposefully sampled; they are not population sense frequencies or
TUBELEX sense-specific frequencies. Source text in saved reviews retains the
original corpus's sharing restrictions. No test-set secrecy is enforced by RDS.

```{r study-cleanup, eval=has_quanteda, include=FALSE}
unlink(study_dir, recursive = TRUE)
```

## Compute actual target embeddings with a cached model

The installed `examples/contextual-embeddings.py` is an optional, explicitly
invoked Python example using Hugging Face Transformers and PyTorch. Obtain those
dependencies and the chosen model separately under their applicable terms.
The script uses only the local Hugging Face cache, a specified full commit,
CPU inference and safetensors; it never downloads files or loads custom remote
code. A missing model or incompatible tokenizer stops with an error.

It extracts the last hidden layer and averages only subwords that exactly cover
the target span. This is **not** the complete sentence-transformers embedding
pipeline, even when its base model is used. It is not a trained WSD model and
does not produce candidate scores.

```{r model-call, eval=FALSE}
# Explicit optional step: jsonlite is a suggested package.
# 'review' and 'input' are the unchanged objects created above.
request_path <- tempfile(fileext = ".json")
output_path <- tempfile(fileext = ".json")
jsonlite::write_json(list(schema_version = "0.1.0", occurrences = input),
  request_path, dataframe = "rows", auto_unbox = TRUE, pretty = TRUE)
script <- system.file("examples", "contextual-embeddings.py", package = "ldfreq")
python <- Sys.which("python3") # Must contain torch and transformers.
stopifnot(nzchar(python))
# Illustration, not a validated Japanese WSD model. Cache this revision first.
model_id <- "sentence-transformers/paraphrase-multilingual-MiniLM-L12-v2"
revision <- "e8f8c211226b894fcb81acc59f3b34ba3efd5f42"
status <- system2(python, c(shQuote(script), "--input", shQuote(request_path),
  "--output", shQuote(output_path), "--model", shQuote(model_id),
  "--revision", shQuote(revision)))
stopifnot(status == 0L)
payload <- jsonlite::fromJSON(output_path)
stopifnot(identical(payload$schema_version, "0.1.0"))
embeddings <- NULL
if (length(payload$embeddings$occurrence_ids)) {
  embeddings <- as.matrix(payload$embeddings$values)
  rownames(embeddings) <- payload$embeddings$occurrence_ids
}
# A zero-row JSON array needs the typed, empty input roster in R.
returned <- payload$data
if (!length(returned)) {
  returned <- input[FALSE, ]
  returned$status <- returned$reason <- character()
}
actual <- lexdiv_import_contextual(review, returned, payload$model, embeddings)
actual$summary
actual$occurrences[c("surface", "model_status", "model_reason")]
saveRDS(actual, "contextual-model-result.rds")
```

The script processes one occurrence at a time; it is not a high-throughput corpus
engine. A fast tokenizer must expose original character offsets. Python uses
zero-based/end-exclusive offsets; R's request uses one-based/inclusive Unicode
codepoints, so the script subtracts one from `start` only. It rejects subwords
crossing the target boundary, gaps, overlaps and unknown target tokens. Overlong
segments are skipped without truncation. These strict checks can exclude valid
linguistic items that cannot be represented with exact spans: report that coverage.
See the [Hugging Face tokenizer documentation](https://huggingface.co/docs/transformers/main_classes/tokenizer).

The full original segment is used regardless of the visible KWIC window.
Sentence/turn boundaries are supplied by the researcher. This initial importer
does not accept cropped, truncated or multi-segment contexts. Record tokenizer
normalization, model revision, layer and aggregation. Do not mix models or
dimensions in a matrix. The importer checks declared anchors, not whether external
software actually followed them. Source text in JSON/RDS retains its original
sharing restrictions; no input corpus or model weights are bundled in ldfreq.

## Review categorical proposals, without inventing scores

Sometimes a model returns a candidate choice or an abstention, rather than a
score for every candidate. Do not turn unchosen candidates into zero scores
to make them fit `lexdiv_import_contextual()`. The explicitly sourced recipe
below keeps these categorical proposals in separate columns beside the KWIC.
Its helpers are installed examples, not exported package functions.

This first example is entirely offline. All texts and proposals are authored;
they demonstrate transport and review, not model accuracy. It needs quanteda
and UTF-8 for the same reason as the earlier KWIC example.

```{r categorical-proposals, eval=has_quanteda}
proposal_tools <- new.env(parent = baseenv())
for (script in c("candidate-proposals.R", "candidate-proposals-demo.R")) {
  sys.source(system.file("examples", script, package = "ldfreq", mustWork = TRUE),
             envir = proposal_tools)
}
illustration <- proposal_tools$candidate_proposals_example
worksheet <- illustration$imported
worksheet$summary
worksheet$occurrences[c("document_id", "pre", "keyword", "post",
  "proposal_status", "proposal_candidate_id", "human_status")]
```

Three of four occurrences were requested. The authored proposals choose
*popularity* and *human presence* for two occurrences of `人気`, and leave a
third unresolved. `駅` was not requested. Every original occurrence remains
in the worksheet, and the original human-decision columns are unchanged.

`prepare_candidate_proposals(review, occurrence_ids)` selects exact occurrence
IDs from an unchanged ambiguity review and retains complete segment text,
1-based inclusive character positions, and the supplied candidate IDs/labels.
Inspect its `$anchors` and `$candidates` before sending anything. Human judgments
are not included in this plan. `import_candidate_proposals(review, plan, data,
model)` accepts these four character columns:

| Column | Meaning |
|---|---|
| `occurrence_id` | An ID from the selected plan; rows can be reordered or partially returned. |
| `proposal_status` | `selected`, `unresolved`, `candidate_missing`, or `boundary_review`. |
| `candidate_id` | A supplied candidate for this occurrence's surface when selected; otherwise `NA_character_`. |
| `reason` | A nonblank explanation of the proposal, not evidence that it is correct. |

`candidate_missing` means that the supplied inventory is insufficient;
`boundary_review` asks a person to inspect tokenization. Neither changes the
text or token boundaries. Omitted requested rows become `not_returned`;
unrequested rows become `not_requested`. The result retains `$review`, `$plan`,
`$model`, `$proposals`, `$occurrences` and counts in `$summary`.

Read the original context before entering a human decision. This demonstration
explicitly enters two authored decisions; it does not automatically accept the
model's proposals or constitute independent human validation.

```{r categorical-human-review, eval=has_quanteda}
human_decisions <- illustration$review$occurrences[1:2,
  c("review_id", "occurrence_id")]
human_decisions$status <- "selected"
human_decisions$candidate_id <- c("popularity", "human-presence")
human_decisions$reviewer <- "authored-review-demonstration"
human_decisions$reason <- c("Being liked by young people", "Nobody present in the park")
confirmed <- lexdiv_ambiguity_review(illustration$review$source,
  illustration$review$summary$term, illustration$review$candidates,
  illustration$review$provenance$resource, decisions = human_decisions)
confirmed_worksheet <- proposal_tools$import_candidate_proposals(confirmed,
  illustration$plan, illustration$proposals, illustration$model)
confirmed_worksheet$occurrences[c("surface", "proposal_status", "human_status")]
proposal_file <- tempfile(fileext = ".rds")
saveRDS(confirmed_worksheet, proposal_file)
stopifnot(identical(readRDS(proposal_file), confirmed_worksheet))
unlink(proposal_file)
```

## Optionally obtain proposals through your OpenAI account

**This step sends the chosen texts and candidate labels to OpenAI and incurs
charges on your API account.** It is never run by installing/loading ldfreq,
the offline examples, ordinary analysis or the package tests. Python is not
needed. Install the optional `httr2` and `jsonlite` packages yourself, and set
`OPENAI_API_KEY` in your local R environment (for example, in your private
`.Renviron`). Do not put the key in shared scripts, saved results or a repository.

Inspect the plan and choose text you may send to that service. The recipe uses
the [Responses API](https://developers.openai.com/api/reference/overview) with
[Structured Outputs](https://developers.openai.com/api/docs/guides/structured-outputs).
The following chunk is not evaluated when the guide is built:

```{r paid-candidate-proposals, eval=FALSE}
# Continue from the offline example above. Set OPENAI_API_KEY privately first.
# If needed, run install.packages(c("httr2", "jsonlite")) once.
api_plan <- illustration$plan
api_plan$anchors
api_plan$candidates
dir.create("proposal-runs", showWarnings = FALSE)
model_id <- "gpt-4.1-mini-2025-04-14" # Explicit snapshot, not a best-model claim.
run_directory <- file.path("proposal-runs", "japanese-example-01")
api_run <- proposal_tools$openai_candidate_proposals(
  api_plan, model = model_id, run_dir = run_directory,
  allow_paid = TRUE, max_output_tokens = 1024L)
api_run$status
api_run$usage
api_worksheet <- proposal_tools$import_candidate_proposals(
  illustration$review, api_plan, api_run$proposals, api_run$model)
api_worksheet$occurrences[c("pre", "keyword", "post", "proposal_status",
  "proposal_candidate_id", "proposal_reason", "human_status")]
saveRDS(api_worksheet, file.path(run_directory, "worksheet.rds"))

# Reuse the same request and saved result without sending or paying again.
replayed <- proposal_tools$openai_candidate_proposals(
  api_plan, model = model_id, run_dir = run_directory,
  max_output_tokens = 1024L)
stopifnot(identical(replayed, api_run))
```

Each new run sends one request for 1–20 occurrences, with at most 40 KB of JSON
input. The output-token limit bounds requested generation, **not the total
currency cost**; input tokens are also billed. Check current
[model pricing](https://developers.openai.com/api/docs/pricing) before a paid run.
The model must support Responses and Structured Outputs. No model fallback,
automatic retry or automatic batch loop is performed.

The run directory contains `request.rds` (plan, instructions, schema and settings)
and `result.rds` (response, parsed proposals, returned model ID, usage and status).
Neither stores the authentication key. Keep both files: identical requests reuse
the result, while changed requests are rejected for that directory. If execution
stops after reserving a directory but before saving a result, the recipe blocks
automatic resubmission because the first request may already have incurred charges.
Inspect the run before deliberately choosing a new directory. This is a sequential
recipe, not a general concurrent job manager.

`partial`, `refused`, `incomplete`, `invalid_output`, `http_error`, and
`transport_error` are distinct run statuses. Invalid or unsuccessful output
produces no candidate choices; all requested rows remain `not_returned` in the
worksheet, and the run status explains why. Inspect both. Complete responses can
still be linguistically wrong. Responses are requested with `store = FALSE`;
this is not a claim of zero provider retention (see
[OpenAI data controls](https://developers.openai.com/api/docs/guides/your-data)).
Saved input and response files contain the selected text and retain its sharing
restrictions. Human-confirmed judgments remain a separate operation.

## Use existing R tools where they fit

[text](https://www.r-text.org/) already calls Transformer models from R through
Python. `text::textEmbed()` supports token/layer aggregation. An adapter must
additionally demonstrate exact per-occurrence target alignment: do not label
sentence or word-type vectors as target-token vectors. For Japanese, review
preprocessing explicitly: the documented `remove_non_ascii` default is `TRUE`.
See [textEmbed](https://www.r-text.org/reference/textEmbed.html).

ldfreq adds checked source joins, missing-output coverage and connections to
human KWIC reviews and lexical resources. It does not claim a new embedding
algorithm or better accuracy. Either R or Python can supply values satisfying
the contract. The core R analyses require neither model environment.

## Evaluate a research claim separately

A useful study asks whether contextual representations distinguish meanings of
the **target word** beyond a simple candidate-frequency baseline. Define the
inventory and obtain independent human judgments first; WLSP row counts are
not a gold-standard sense inventory. GlossBERT illustrates an additional
context/definition model for WSD, not a capability obtained by averaging vectors
([Huang et al., 2019](https://aclanthology.org/D19-1355/)).

Separate examples used to choose a model, layer, threshold or sense prototype
from final evaluation examples. For generalization to unseen words or documents,
split by those units rather than mixing their occurrences across sets. Report
candidate coverage, processed/all-occurrence coverage, conditional accuracy on
independently resolved cases and unresolved cases separately. Inspect errors by
language, reading, frequency, genre and segmentation. Abstentions must remain
visible. For usability, compare existing tools and this joined workflow on the
same task: erroneous joins, missed cases, correction time and reproducibility.
Human review time, semantic accuracy, norm validity and learner knowledge require
evidence beyond a software round trip; the authored examples supply none of it.

```{r save, eval=has_quanteda}
saved <- tempfile(fileext = ".rds")
saveRDS(imported, saved)
stopifnot(identical(readRDS(saved), imported))
unlink(saved)
```
