---
title: "Connect contextual model outputs to research occurrences"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Connect contextual model outputs to research occurrences}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ldfreq)
has_quanteda <- requireNamespace("quanteda", quietly = TRUE) &&
  packageVersion("quanteda") >= "4.5.0" && isTRUE(l10n_info()[["UTF-8"]])
```

## The research problem

For a word such as *bank*, a model can provide a representation of each
occurrence in context. The research question might be whether those
representations distinguish financial from river-edge uses. Computing vectors
does not establish that distinction: the target, context, candidate inventory
and an independent evaluation must also be specified.

`lexdiv_import_contextual()` connects external model output to source occurrences
used for KWIC review. It checks joins, preserves missing output and keeps model
suggestions separate from human decisions. It does not run BERT, install Python,
download weights, select a meaning, or validate a semantic measure. The importer
needs no model software. Creating the review uses optional quanteda; the
English/Japanese example below requires a UTF-8 R session.

## An offline example with authored values

These sentences, candidates, vectors and scores are **authored illustrations**,
not model predictions, corpus observations or psycholinguistic norms. Japanese
`人気` illustrates different readings of the same spelling, not a demonstrated
related-sense manipulation. Use real annotations for a real study.

```{r prepare, eval=has_quanteda}
words <- list(c("The", "bank", "lent", "money", "."),
              c("The", "river", "bank", "flooded", "."),
              c("その", "店", "は", "人気", "が", "ある", "。"),
              c("ここ", "は", "人気", "が", "ない", "。"))
tokens <- data.frame(document_id = "example",
  segment_id = rep(paste0("s", 1:4), lengths(words)),
  token_index = sequence(lengths(words)), surface = unlist(words, use.names = FALSE))
segments <- data.frame(document_id = "example", segment_id = paste0("s", 1:4),
  text = c("The bank lent money.", "The river bank flooded.",
           "その店は人気がある。", "ここは人気がない。"))
annotations <- lexdiv_import_annotations(tokens, segments, list(
  language = "en/ja", analyzer = "authored", analyzer_version = "1",
  dictionary = "none", dictionary_version = "none", unit = "authored",
  normalization = "none"))
candidates <- data.frame(term = c("bank", "bank", "人気", "人気"),
  candidate_id = c("financial", "river", "popularity", "human-presence"),
  label = c("financial institution", "river edge", "にんき", "ひとけ"))
review <- lexdiv_ambiguity_review(annotations, c("bank", "人気"), candidates,
  list(resource_id = "authored", resource_version = "1",
    source_reference = "Authored candidates, not a dictionary", data_license = "MIT"))

# Export anchors BEFORE computation. Never reconstruct them after a model run.
input <- review$occurrences[c("review_id", "occurrence_id", "segment_text",
                              "surface", "start", "end")]
data <- input[1:3, ]
data$status <- c("processed", "processed", "skipped")
data$reason <- c("Authored transport example", "Authored transport example",
                 "Illustrated failure, not an actual model limitation")
vectors <- matrix(c(1, 0, 0, 1), nrow = 2,
                  dimnames = list(data$occurrence_id[1:2], NULL))
scores <- data.frame(occurrence_id = rep(data$occurrence_id[1:2], each = 2),
  candidate_id = rep(c("financial", "river"), 2), score = c(2, -1, 0.5, 0.5))
model <- list(model_id = "authored-example", model_revision = "1",
  tokenizer_id = "authored", tokenizer_revision = "1",
  software = "illustration", software_version = "1",
  context_policy = "full segment; no truncation",
  representation = "authored two-dimensional vectors, not model output",
  score_definition = "authored scores; larger is preferred; not probabilities")
imported <- lexdiv_import_contextual(review, data, model, vectors, scores)
imported$summary
imported$occurrences[c("surface", "status", "model_status", "has_embedding")]
imported$suggestions[c("pre", "keyword", "post", "label", "score", "human_status")]
stopifnot(identical(imported$review, review),
          imported$summary$processed_proportion == 0.5,
          all(imported$occurrences$status == "unreviewed"))
```

Two of four occurrences have output. Skipped and omitted (`not_returned`)
occurrences remain in the denominator. Tied and negative scores are retained;
importing them does not create probabilities or human decisions. The suggestions
table contains KWIC and candidate labels; `segment_text` retains the full original
segment. Scores may cover only part of an inventory: an absent score is missing,
not zero. Use the ordinary review decision table to record later human selections
and keep the earlier review as well. Model-assisted judgments are not independent
human judgments; conduct a blinded human pass before showing suggestions when
independent accuracy or agreement is the research question.

## Compare scores with a declared reference

`lexdiv_evaluate_contextual()` predicts a candidate only when **every candidate
in the supplied inventory is scored** and one is best. Partial lists and tied
best scores abstain. This conservative policy is not a WSD benchmark convention
or proof that the inventory contains every meaning. Specify whether larger or
smaller scores are preferred. The optional absolute `tie_tolerance` treats nearby
scores as ties; choose it before final evaluation, not by tuning on test labels.

These reference labels are also authored illustrations, not independent human
gold labels. Agreement here is not evidence about BERT accuracy. Keep the
reference separate from the imported model output and its original human review.

```{r evaluate-scores, eval=has_quanteda}
decisions <- input[c("review_id", "occurrence_id")]
decisions$status <- c("selected", "selected", "selected", "unresolved")
decisions$candidate_id <- c("financial", "river", "popularity", NA_character_)
decisions$reviewer <- "authored-example"
decisions$reason <- c(rep("Authored criterion for this example", 3),
                      "Illustration of an unresolved reference")
reference_review <- lexdiv_ambiguity_review(annotations, c("bank", "人気"),
  candidates, review$provenance$resource, decisions)
evaluation <- lexdiv_evaluate_contextual(imported, reference_review,
  direction = "higher", reference_info = list(
    reference_id = "authored-demonstration", annotation_protocol = "Authored examples, not human ratings",
    model_exposure = "unknown", evaluation_role = "development"))
evaluation$summary[c("agreement", "paired", "agreement_among_paired",
  "predictions", "occurrences", "prediction_coverage",
  "reference_selected", "matches_among_reference_selected")]
evaluation$review_queue[c("surface", "prediction_status", "model_status",
  "reference_status", "reference_reason", "outcome")]
stopifnot(evaluation$summary$agreement_among_paired == 1,
          evaluation$summary$prediction_coverage == 1/4,
          evaluation$summary$matches_among_reference_selected == 1/3,
          identical(evaluation$reference, reference_review),
          identical(evaluation$model_output, imported))
saved_evaluation <- tempfile(fileext = ".rds")
saveRDS(evaluation, saved_evaluation)
stopifnot(identical(readRDS(saved_evaluation), evaluation))
unlink(saved_evaluation)
```

The one comparable case agrees (1/1), but only one of four occurrences receives
a prediction. Among the three occurrences with selected reference labels, one
receives a matching prediction (1/3). "100% agreement" alone would hide most
cases. The KWIC `review_queue` retains the tied English example, the skipped
Japanese example and the unresolved reference. `terms` repeats counts by surface,
including absent targets; `confusion` keeps candidate IDs scoped to each surface.
Overall proportions weight occurrences rather than word types. A singleton
inventory produces a trivial prediction, counted in `single_candidate_predictions`;
inspect these separately when evaluating ambiguity.

All inputs remain unchanged. An unresolved or unreviewed reference is not a wrong
label and is not counted as agreement. Embedding-only output yields no predictions:
a separately specified classifier or scoring procedure is still needed.
`reference_scored` identifies whether a selected reference candidate appeared in
the scores, distinguishing a missing reference candidate from missing rivals.

Use `model_exposure = "not_shown"` only when annotators did not see model output,
and `evaluation_role = "held_out"` only when examples were not used to fit or
choose the scoring procedure or evaluation settings. These declarations are not
verified by the package. Model-assisted judgments (`shown`) and development
examples are valid descriptive inputs, but are not independent held-out accuracy.

WSD evaluation depends on consistent annotation guidelines and sense inventories
([Raganato et al., 2017](https://aclanthology.org/E17-1010/)). Performance on answered
cases and coverage are also distinguished in selective classification
([Geifman & El-Yaniv, 2017](https://papers.neurips.cc/paper_files/paper/2017/hash/4a8423d5e91fda00bb7e46540e2b0cf1-Abstract.html)).
This evaluator implements neither their training algorithms nor their statistical
guarantees; it reports counts and descriptive proportions.

For a candidate-frequency baseline, estimate its scores on separate training
data, import them against the same unchanged review, and evaluate with the same
reference and predeclared policy. Keep each method's coverage and compare their
overlapping evaluable occurrence IDs as well. A higher conditional agreement
based on fewer, easier cases is not evidence of improvement. Do not derive
baseline frequencies from held-out labels or equate different candidate inventories.

## Score separate training and query examples in R

`lexdiv_score_contextual()` bridges imported vectors and the evaluator. Its
centroid baseline averages the raw target vectors for each selected training
candidate, then computes cosine similarity to a query vector. Averaging labeled
contextual embeddings is a starting point in
[Loureiro & Jorge (2019), section 4.1](https://aclanthology.org/P19-1569/).
This small baseline does not implement their full LMMS method, WordNet
propagation, gloss representations or reported benchmark results.

The comparison baseline uses selected training-label counts for the same surface.
Unlike centroid fitting, it can use selected examples without embeddings.
Always inspect both `n_selected` and `n_vectors`. Neither score is a probability.
The two methods remain separate: there is no automatic frequency fallback.

Create training and query sources separately **before** model computation.
Stable document IDs must not overlap; renamed copies of identical target-bearing
segments are also rejected. Candidate tables/resources and embedding declarations
(including model/tokenizer revisions, layer/aggregation and software) must match.
Matching declarations are not proof that external software followed them.

The following sentences, labels and vectors are authored for illustration.
They are not an independent annotation study or a trained BERT benchmark.

```{r score-baselines, eval=has_quanteda}
train_words <- list(c("Her", "bank", "approved", "a", "loan", "."),
                    c("A", "bank", "holds", "deposits", "."),
                    c("We", "rested", "by", "the", "bank", "."),
                    c("人気", "の", "店", "だ", "。"),
                    c("周囲", "に", "人気", "が", "ない", "。"))
train_tokens <- data.frame(document_id = "training-example",
  segment_id = rep(paste0("t", 1:5), lengths(train_words)),
  token_index = sequence(lengths(train_words)), surface = unlist(train_words, use.names = FALSE))
train_segments <- data.frame(document_id = "training-example", segment_id = paste0("t", 1:5),
  text = c("Her bank approved a loan.", "A bank holds deposits.", "We rested by the bank.",
           "人気の店だ。", "周囲に人気がない。"))
train_annotations <- lexdiv_import_annotations(train_tokens, train_segments,
  annotations$provenance$annotation)
train_review <- lexdiv_ambiguity_review(train_annotations, c("bank", "人気"), candidates,
  review$provenance$resource)
train_input <- train_review$occurrences[c("review_id", "occurrence_id", "segment_text", "surface", "start", "end")]
train_data <- train_input
train_data$status <- "processed"; train_data$reason <- "Authored training vectors"
train_vectors <- rbind(c(2,0), c(2,1), c(0,2), c(2,0), c(0,2))
rownames(train_vectors) <- train_input$occurrence_id
train_imported <- lexdiv_import_contextual(train_review, train_data, model, train_vectors)
train_decisions <- train_input[c("review_id", "occurrence_id")]
train_decisions$status <- "selected"
train_decisions$candidate_id <- c("financial", "financial", "river", "popularity", "human-presence")
train_decisions$reviewer <- "authored-example"
train_decisions$reason <- "Intended labels of authored sentences; not independent human judgments"
train_reference <- lexdiv_ambiguity_review(train_annotations, c("bank", "人気"), candidates,
  review$provenance$resource, train_decisions)
scored <- lexdiv_score_contextual(imported, train_imported, train_reference,
  training_info = list(training_id = "authored-training",
    annotation_protocol = "Author-specified intended labels",
    partition_protocol = "Separate authored documents and target contexts",
    model_exposure = "unknown"))
scored$candidates[c("term", "candidate_id", "n_selected", "n_vectors", "prototype_status")]
scored$query_occurrences[c("surface", "model_status", "centroid_status", "frequency_status")]
baseline_evaluations <- lapply(scored[c("centroid", "frequency")], function(output)
  lexdiv_evaluate_contextual(output, reference_review, "higher",
    evaluation$provenance$reference_info))
comparison <- do.call(rbind, lapply(baseline_evaluations, function(result)
  result$summary[c("agreement", "paired", "prediction_coverage", "agreement_among_paired")]))
comparison
baseline_evaluations$centroid$review_queue[c("keyword", "pre", "post", "prediction_status", "model_reason")]
stopifnot(comparison$agreement_among_paired == c(1, 0.5),
          all(comparison$prediction_coverage == 0.5),
          identical(scored$query, imported), identical(scored$reference, train_reference))
saved_scores <- tempfile(fileext = ".rds")
saveRDS(scored, saved_scores)
stopifnot(identical(readRDS(saved_scores), scored))
unlink(saved_scores)
```

The centroid method matches both English reference choices in this constructed
example; the frequency baseline picks the financial candidate for both. This
is a demonstration of different rules, not empirical evidence of improvement.
The Japanese queries have no imported vectors, so no centroid scores are made.
Frequency scores exist for those queries, but tied training counts mean that
the evaluator abstains. Scoring-stage `processed` therefore does not mean a
unique prediction, an available embedding or a correct interpretation.

A missing candidate prototype stays missing and causes evaluator abstention
when the score list is incomplete. Frequency zero means no selected training
example of that candidate; an entirely unobserved surface receives no scores.
Zero vectors and canceling means are recorded separately. Training counts can
be small or uneven; inspect them before interpreting a candidate comparison.

For a real run, use the optional embedding script below separately on the two
source rosters with identical settings, import both outputs, and supply the
training review and independent query reference. Do not relabel query outputs
as training data or manufacture new IDs to bypass the split check. Participant
overlap, near duplicates, pretraining contamination and earlier model-selection
leakage are not detected automatically. This baseline cannot generalize to a
new surface with no training labels; a word-held-out study needs another method.
Save the entire `scored` object: each score output contains training fingerprints
and required declarations, while the full object retains all training examples,
prototype vectors and exclusion reasons.

## Compute actual target embeddings with a cached model

The installed `examples/contextual-embeddings.py` is an optional, explicitly
invoked Python example using Hugging Face Transformers and PyTorch. Obtain those
dependencies and the chosen model separately under their applicable terms.
The script uses only the local Hugging Face cache, a specified full commit,
CPU inference and safetensors; it never downloads files or loads custom remote
code. A missing model or incompatible tokenizer stops with an error.

It extracts the last hidden layer and averages only subwords that exactly cover
the target span. This is **not** the complete sentence-transformers embedding
pipeline, even when its base model is used. It is not a trained WSD model and
does not produce candidate scores.

```{r model-call, eval=FALSE}
# Explicit optional step: jsonlite is a suggested package.
# 'review' and 'input' are the unchanged objects created above.
request_path <- tempfile(fileext = ".json")
output_path <- tempfile(fileext = ".json")
jsonlite::write_json(list(schema_version = "0.1.0", occurrences = input),
  request_path, dataframe = "rows", auto_unbox = TRUE, pretty = TRUE)
script <- system.file("examples", "contextual-embeddings.py", package = "ldfreq")
python <- Sys.which("python3") # Must contain torch and transformers.
stopifnot(nzchar(python))
# Illustration, not a validated Japanese WSD model. Cache this revision first.
model_id <- "sentence-transformers/paraphrase-multilingual-MiniLM-L12-v2"
revision <- "e8f8c211226b894fcb81acc59f3b34ba3efd5f42"
status <- system2(python, c(shQuote(script), "--input", shQuote(request_path),
  "--output", shQuote(output_path), "--model", shQuote(model_id),
  "--revision", shQuote(revision)))
stopifnot(status == 0L)
payload <- jsonlite::fromJSON(output_path)
stopifnot(identical(payload$schema_version, "0.1.0"))
embeddings <- NULL
if (length(payload$embeddings$occurrence_ids)) {
  embeddings <- as.matrix(payload$embeddings$values)
  rownames(embeddings) <- payload$embeddings$occurrence_ids
}
# A zero-row JSON array needs the typed, empty input roster in R.
returned <- payload$data
if (!length(returned)) {
  returned <- input[FALSE, ]
  returned$status <- returned$reason <- character()
}
actual <- lexdiv_import_contextual(review, returned, payload$model, embeddings)
actual$summary
actual$occurrences[c("surface", "model_status", "model_reason")]
saveRDS(actual, "contextual-model-result.rds")
```

The script processes one occurrence at a time; it is not a high-throughput corpus
engine. A fast tokenizer must expose original character offsets. Python uses
zero-based/end-exclusive offsets; R's request uses one-based/inclusive Unicode
codepoints, so the script subtracts one from `start` only. It rejects subwords
crossing the target boundary, gaps, overlaps and unknown target tokens. Overlong
segments are skipped without truncation. These strict checks can exclude valid
linguistic items that cannot be represented with exact spans: report that coverage.
See the [Hugging Face tokenizer documentation](https://huggingface.co/docs/transformers/main_classes/tokenizer).

The full original segment is used regardless of the visible KWIC window.
Sentence/turn boundaries are supplied by the researcher. This initial importer
does not accept cropped, truncated or multi-segment contexts. Record tokenizer
normalization, model revision, layer and aggregation. Do not mix models or
dimensions in a matrix. The importer checks declared anchors, not whether external
software actually followed them. Source text in JSON/RDS retains its original
sharing restrictions; no input corpus or model weights are bundled in ldfreq.

## Use existing R tools where they fit

[text](https://www.r-text.org/) already calls Transformer models from R through
Python. `text::textEmbed()` supports token/layer aggregation. An adapter must
additionally demonstrate exact per-occurrence target alignment: do not label
sentence or word-type vectors as target-token vectors. For Japanese, review
preprocessing explicitly: the documented `remove_non_ascii` default is `TRUE`.
See [textEmbed](https://www.r-text.org/reference/textEmbed.html).

ldfreq adds checked source joins, missing-output coverage and connections to
human KWIC reviews and lexical resources. It does not claim a new embedding
algorithm or better accuracy. Either R or Python can supply values satisfying
the contract. The core R analyses require neither model environment.

## Evaluate a research claim separately

A useful study asks whether contextual representations distinguish meanings of
the **target word** beyond a simple candidate-frequency baseline. Define the
inventory and obtain independent human judgments first; WLSP row counts are
not a gold-standard sense inventory. GlossBERT illustrates an additional
context/definition model for WSD, not a capability obtained by averaging vectors
([Huang et al., 2019](https://aclanthology.org/D19-1355/)).

Separate examples used to choose a model, layer, threshold or sense prototype
from final evaluation examples. For generalization to unseen words or documents,
split by those units rather than mixing their occurrences across sets. Report
candidate coverage, processed/all-occurrence coverage, conditional accuracy on
independently resolved cases and unresolved cases separately. Inspect errors by
language, reading, frequency, genre and segmentation. Abstentions must remain
visible. For usability, compare existing tools and this joined workflow on the
same task: erroneous joins, missed cases, correction time and reproducibility.
Human review time, semantic accuracy, norm validity and learner knowledge require
evidence beyond a software round trip; the authored examples supply none of it.

```{r save, eval=has_quanteda}
saved <- tempfile(fileext = ".rds")
saveRDS(imported, saved)
stopifnot(identical(readRDS(saved), imported))
unlink(saved)
```
