---
title: "Connect contextual model outputs to research occurrences"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Connect contextual model outputs to research occurrences}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ldfreq)
has_quanteda <- requireNamespace("quanteda", quietly = TRUE) &&
  packageVersion("quanteda") >= "4.5.0" && isTRUE(l10n_info()[["UTF-8"]])
```

## The research problem

For a word such as *bank*, a model can provide a representation of each
occurrence in context. The research question might be whether those
representations distinguish financial from river-edge uses. Computing vectors
does not establish that distinction: the target, context, candidate inventory
and an independent evaluation must also be specified.

`lexdiv_import_contextual()` connects external model output to source occurrences
used for KWIC review. It checks joins, preserves missing output and keeps model
suggestions separate from human decisions. It does not run BERT, install Python,
download weights, select a meaning, or validate a semantic measure. The importer
needs no model software. Creating the review uses optional quanteda; the
English/Japanese example below requires a UTF-8 R session.

## An offline example with authored values

These sentences, candidates, vectors and scores are **authored illustrations**,
not model predictions, corpus observations or psycholinguistic norms. Japanese
`人気` illustrates different readings of the same spelling, not a demonstrated
related-sense manipulation. Use real annotations for a real study.

```{r prepare, eval=has_quanteda}
words <- list(c("The", "bank", "lent", "money", "."),
              c("The", "river", "bank", "flooded", "."),
              c("その", "店", "は", "人気", "が", "ある", "。"),
              c("ここ", "は", "人気", "が", "ない", "。"))
tokens <- data.frame(document_id = "example",
  segment_id = rep(paste0("s", 1:4), lengths(words)),
  token_index = sequence(lengths(words)), surface = unlist(words, use.names = FALSE))
segments <- data.frame(document_id = "example", segment_id = paste0("s", 1:4),
  text = c("The bank lent money.", "The river bank flooded.",
           "その店は人気がある。", "ここは人気がない。"))
annotations <- lexdiv_import_annotations(tokens, segments, list(
  language = "en/ja", analyzer = "authored", analyzer_version = "1",
  dictionary = "none", dictionary_version = "none", unit = "authored",
  normalization = "none"))
candidates <- data.frame(term = c("bank", "bank", "人気", "人気"),
  candidate_id = c("financial", "river", "popularity", "human-presence"),
  label = c("financial institution", "river edge", "にんき", "ひとけ"))
review <- lexdiv_ambiguity_review(annotations, c("bank", "人気"), candidates,
  list(resource_id = "authored", resource_version = "1",
    source_reference = "Authored candidates, not a dictionary", data_license = "MIT"))

# Export anchors BEFORE computation. Never reconstruct them after a model run.
input <- review$occurrences[c("review_id", "occurrence_id", "segment_text",
                              "surface", "start", "end")]
data <- input[1:3, ]
data$status <- c("processed", "processed", "skipped")
data$reason <- c("Authored transport example", "Authored transport example",
                 "Illustrated failure, not an actual model limitation")
vectors <- matrix(c(1, 0, 0, 1), nrow = 2,
                  dimnames = list(data$occurrence_id[1:2], NULL))
scores <- data.frame(occurrence_id = rep(data$occurrence_id[1:2], each = 2),
  candidate_id = rep(c("financial", "river"), 2), score = c(2, -1, 0.5, 0.5))
model <- list(model_id = "authored-example", model_revision = "1",
  tokenizer_id = "authored", tokenizer_revision = "1",
  software = "illustration", software_version = "1",
  context_policy = "full segment; no truncation",
  representation = "authored two-dimensional vectors, not model output",
  score_definition = "authored scores; larger is preferred; not probabilities")
imported <- lexdiv_import_contextual(review, data, model, vectors, scores)
imported$summary
imported$occurrences[c("surface", "status", "model_status", "has_embedding")]
imported$suggestions[c("pre", "keyword", "post", "label", "score", "human_status")]
stopifnot(identical(imported$review, review),
          imported$summary$processed_proportion == 0.5,
          all(imported$occurrences$status == "unreviewed"))
```

Two of four occurrences have output. Skipped and omitted (`not_returned`)
occurrences remain in the denominator. Tied and negative scores are retained;
importing them does not create probabilities or human decisions. The suggestions
table contains KWIC and candidate labels; `segment_text` retains the full original
segment. Scores may cover only part of an inventory: an absent score is missing,
not zero. Use the ordinary review decision table to record later human selections
and keep the earlier review as well. Model-assisted judgments are not independent
human judgments; conduct a blinded human pass before showing suggestions when
independent accuracy or agreement is the research question.

## Compute actual target embeddings with a cached model

The installed `examples/contextual-embeddings.py` is an optional, explicitly
invoked Python example using Hugging Face Transformers and PyTorch. Obtain those
dependencies and the chosen model separately under their applicable terms.
The script uses only the local Hugging Face cache, a specified full commit,
CPU inference and safetensors; it never downloads files or loads custom remote
code. A missing model or incompatible tokenizer stops with an error.

It extracts the last hidden layer and averages only subwords that exactly cover
the target span. This is **not** the complete sentence-transformers embedding
pipeline, even when its base model is used. It is not a trained WSD model and
does not produce candidate scores.

```{r model-call, eval=FALSE}
# Explicit optional step: jsonlite is a suggested package.
# 'review' and 'input' are the unchanged objects created above.
request_path <- tempfile(fileext = ".json")
output_path <- tempfile(fileext = ".json")
jsonlite::write_json(list(schema_version = "0.1.0", occurrences = input),
  request_path, dataframe = "rows", auto_unbox = TRUE, pretty = TRUE)
script <- system.file("examples", "contextual-embeddings.py", package = "ldfreq")
python <- Sys.which("python3") # Must contain torch and transformers.
stopifnot(nzchar(python))
# Illustration, not a validated Japanese WSD model. Cache this revision first.
model_id <- "sentence-transformers/paraphrase-multilingual-MiniLM-L12-v2"
revision <- "e8f8c211226b894fcb81acc59f3b34ba3efd5f42"
status <- system2(python, c(shQuote(script), "--input", shQuote(request_path),
  "--output", shQuote(output_path), "--model", shQuote(model_id),
  "--revision", shQuote(revision)))
stopifnot(status == 0L)
payload <- jsonlite::fromJSON(output_path)
stopifnot(identical(payload$schema_version, "0.1.0"))
embeddings <- NULL
if (length(payload$embeddings$occurrence_ids)) {
  embeddings <- as.matrix(payload$embeddings$values)
  rownames(embeddings) <- payload$embeddings$occurrence_ids
}
# A zero-row JSON array needs the typed, empty input roster in R.
returned <- payload$data
if (!length(returned)) {
  returned <- input[FALSE, ]
  returned$status <- returned$reason <- character()
}
actual <- lexdiv_import_contextual(review, returned, payload$model, embeddings)
actual$summary
actual$occurrences[c("surface", "model_status", "model_reason")]
saveRDS(actual, "contextual-model-result.rds")
```

The script processes one occurrence at a time; it is not a high-throughput corpus
engine. A fast tokenizer must expose original character offsets. Python uses
zero-based/end-exclusive offsets; R's request uses one-based/inclusive Unicode
codepoints, so the script subtracts one from `start` only. It rejects subwords
crossing the target boundary, gaps, overlaps and unknown target tokens. Overlong
segments are skipped without truncation. These strict checks can exclude valid
linguistic items that cannot be represented with exact spans: report that coverage.
See the [Hugging Face tokenizer documentation](https://huggingface.co/docs/transformers/main_classes/tokenizer).

The full original segment is used regardless of the visible KWIC window.
Sentence/turn boundaries are supplied by the researcher. This initial importer
does not accept cropped, truncated or multi-segment contexts. Record tokenizer
normalization, model revision, layer and aggregation. Do not mix models or
dimensions in a matrix. The importer checks declared anchors, not whether external
software actually followed them. Source text in JSON/RDS retains its original
sharing restrictions; no input corpus or model weights are bundled in ldfreq.

## Use existing R tools where they fit

[text](https://www.r-text.org/) already calls Transformer models from R through
Python. `text::textEmbed()` supports token/layer aggregation. An adapter must
additionally demonstrate exact per-occurrence target alignment: do not label
sentence or word-type vectors as target-token vectors. For Japanese, review
preprocessing explicitly: the documented `remove_non_ascii` default is `TRUE`.
See [textEmbed](https://www.r-text.org/reference/textEmbed.html).

ldfreq adds checked source joins, missing-output coverage and connections to
human KWIC reviews and lexical resources. It does not claim a new embedding
algorithm or better accuracy. Either R or Python can supply values satisfying
the contract. The core R analyses require neither model environment.

## Evaluate a research claim separately

A useful study asks whether contextual representations distinguish meanings of
the **target word** beyond a simple candidate-frequency baseline. Define the
inventory and obtain independent human judgments first; WLSP row counts are
not a gold-standard sense inventory. GlossBERT illustrates an additional
context/definition model for WSD, not a capability obtained by averaging vectors
([Huang et al., 2019](https://aclanthology.org/D19-1355/)).

Separate examples used to choose a model, layer, threshold or sense prototype
from final evaluation examples. For generalization to unseen words or documents,
split by those units rather than mixing their occurrences across sets. Report
candidate coverage, processed/all-occurrence coverage, conditional accuracy on
independently resolved cases and unresolved cases separately. Inspect errors by
language, reading, frequency, genre and segmentation. Abstentions must remain
visible. For usability, compare existing tools and this joined workflow on the
same task: erroneous joins, missed cases, correction time and reproducibility.
Human review time, semantic accuracy, norm validity and learner knowledge require
evidence beyond a software round trip; the authored examples supply none of it.

```{r save, eval=has_quanteda}
saved <- tempfile(fileext = ".rds")
saveRDS(imported, saved)
stopifnot(identical(readRDS(saved), imported))
unlink(saved)
```
