---
title: "Using your own lexical reference data across a corpus"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using your own lexical reference data across a corpus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ldfreq)
```

## Choose a question and a reference unit

Suppose you want to describe familiarity and age-of-acquisition values for
words in a collection of texts. The authors may be native speakers or language
learners; the same computation applies. Its interpretation depends on the
reference population, text sampling and word unit, not the author's L1 label.

`lexdiv_norm_profile_batch()` applies one caller-supplied reference table to
multiple documents. It retains each document and each variable, rather than
pooling all texts or combining different scales. This guide uses authored
texts and numbers, not observed norms or a finding about register differences.
No external dataset or network connection is needed to run it.

## Preserve documents and their metadata

Use a unique document ID independently of writer/speaker ID, since the same
person may contribute several texts. Real metadata should record corpus/version,
register, task/topic, author/speaker, documented language background and sampling
conditions. Keep unknown information missing; do not infer L1 from a genre label.

```{r}
texts <- data.frame(
  document_id = c("demo_essay", "demo_turn", "demo_blank"),
  corpus_id = "authored_demo_v1",
  register = c("essay_example", "conversation_example", "essay_example"),
  participant_id = NA_character_,
  text = c("We read the book and discuss the policy.",
           "I read the book. I read it again.", "")
)
stopifnot(!anyNA(texts$document_id), !anyDuplicated(texts$document_id))
prepared <- lexdiv_tokenize_batch(texts, tokenizer = "english",
  normalization = "NFC", case = "lower")
terms <- lapply(prepared, function(x) x$tokens$surface)
```

These are plain text examples, not raw XML or unprocessed conversation
transcripts. Select speaker/turn boundaries and handle transcription markup
before tokenization. Do not join unrelated documents to make them long enough
for a metric. For other languages, supply an appropriate external segmentation
and prepared lookup terms; this English tokenizer and the bundled English
resources are not evidence of multilingual validation.

## Describe the reference table, variables and population

The toy table uses lowercase surface forms. A real lemma-based resource instead
needs compatible lemmas, including its POS convention if relevant. Exact
matching does not silently correct case, accents, spelling or inflection.

```{r}
norms <- data.frame(term = c("read", "book", "discuss", "policy"),
  familiarity = c(6, 7, 5, NA_real_),
  aoa_years = c(6, 4, 9, 12))
specs <- data.frame(
  measure_id = c("familiarity", "aoa"),
  value_column = c("familiarity", "aoa_years"),
  construct_id = c("subjective_familiarity", "age_of_acquisition"),
  value_unit = c("seven_point_rating", "years"),
  direction = c("higher", "descriptive"),
  language = "English", variety = "unspecified",
  population_id = "authored_not_participant_data",
  collection_year = "not_observed",
  valid_min = c(1, 0), valid_max = c(7, NA_real_))
resource <- list(resource_id = "authored_norms", resource_version = "1",
  creator = "Example author", source_reference = "Authored demonstration",
  data_license = "Example numbers; no external dataset",
  transformation_id = "none", lookup_unit = "lowercase_surface",
  resource_key_normalization_id = "caller-prepared-v1")

profiles <- lexdiv_norm_profile_batch(terms, norms, "term", specs, resource)
profiles$summary[, c("document_id", "measure_id", "weighting", "estimate",
  "value_unit", "status", "missing_reason")]
```

`direction` describes the variable; it does not turn a high score into good
writing. Years and rating points remain separate scales. The same interface can
use documented frequency measures, but record their units, reference corpus,
denominator and transformations. A per-million frequency, a log frequency and
a Zipf value are not interchangeable. For another source, make a separate call
with its own resource record. The bundled TUBELEX adapter additionally verifies
its fixed resource; this generic function records caller assertions and does
not fingerprint or authenticate the supplied table.

## Keep coverage alongside each estimate

Three rates answer different questions:

| Rate | Numerator / denominator | Question |
|---|---|---|
| Resource coverage | Matched units / all input units | How much of the document occurs in the reference? |
| Value coverage | Units with observed values / all input units | How much contributes to this variable's mean? |
| Annotation coverage | Units with observed values / matched units | How complete is this variable among matched entries? |

In the essay, `policy` matches the reference but lacks familiarity. Other words
are absent from the reference entirely. Both stay visible, with different
reasons. Neither receives zero. The blank document retains rows with
`empty_input`; it does not disappear from the study.

```{r}
profiles$coverage
profiles$lookup[profiles$lookup$term == "policy",
  c("document_id", "term", "measure_id", "lookup_status", "value_status")]
profiles$diagnostics$documents
```

Token weighting gives repeated words repeated weight within a document. Type
weighting counts distinct exact lookup terms once. Neither weights writers or
documents for a group comparison. Group averages would require another explicit
decision about equal-document versus token weighting, missing results and
sampling dependencies. The batch function deliberately returns document rows.

## Join metadata by ID and save the complete analysis

Metric/profile tables are long: each document contributes multiple rows.
Join metadata by unique document ID, preserving every result row and its order.
Additional columns in an input data frame are not automatically copied into
the batch output.

```{r}
metadata <- texts[, c("document_id", "corpus_id", "register", "participant_id")]
row_index <- match(profiles$summary$document_id, metadata$document_id)
stopifnot(!anyNA(row_index), !anyDuplicated(metadata$document_id),
  !any(setdiff(names(metadata), "document_id") %in% names(profiles$summary)))
analysis <- cbind(profiles$summary,
  metadata[row_index, setdiff(names(metadata), "document_id"), drop = FALSE])
rownames(analysis) <- NULL
stopifnot(nrow(analysis) == nrow(profiles$summary))

# One variable and weighting, with all documents and missing rows retained.
report <- analysis[analysis$measure_id == "familiarity" &
  analysis$weighting == "token", c("document_id", "register", "estimate",
    "input_units", "observed_value_units", "resource_coverage",
    "value_coverage", "annotation_coverage", "status", "missing_reason")]
report

saved <- tempfile(fileext = ".rds")
saveRDS(list(texts = texts, prepared = prepared, terms = terms,
  norms = norms, specs = specs, resource = resource, profiles = profiles,
  analysis = analysis, session = sessionInfo()), saved)
restored <- readRDS(saved)
replayed <- lexdiv_norm_profile_batch(restored$terms, restored$norms, "term",
  restored$specs, restored$resource)
stopifnot(identical(replayed, restored$profiles),
  identical(restored$analysis, analysis))
unlink(saved)
```

Use a persistent path for an actual study. Save the resource table as well as
its citation/version and preprocessing; a version label alone does not preserve
its contents. Obtain the source table under its applicable terms before use.
Research data saved in your project do not become package data automatically.

For this authored example, the useful answer is which words and denominators
contribute to each mean. It does not establish a register difference or a
participant's knowledge. In a real study, examine coverage differences before
comparing estimates, use appropriate independent documents or speaker/author
clusters, and justify the reference population. See
[designing comparisons](designing-comparisons.html) for the broader study design.

## Know when an input stops the calculation

Duplicate IDs, malformed term vectors and invalid reference specifications stop
the whole call. This follows the single-document norm API. Valid empty,
unmatched and unannotated documents remain in the output. The global `max_rows`
limit covers lookup, summary and coverage rows; reduce the requested output or
choose an explicit larger limit after considering its memory needs. Splitting
documents does not evade this bound within one call.
