---
title: "Trace adjective–noun dependencies from source context to document counts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trace adjective–noun dependencies from source context to document counts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ldfreq)
```

## Which noun does an adjective modify?

In *Big red balloons float*, both *big* and *red* modify *balloons*.
Adjacent bigrams cannot represent the first relation as a two-word sequence.
`lexdiv_amod_pairs()` instead reads supplied basic Universal Dependencies (UD)
and connects each selected pair to its two source positions, context and
document counts. It does not run a parser or decide whether a phrase is idiomatic.

The initial feature is deliberately specific: a dependent tagged `ADJ`, a head
tagged `NOUN`, and an `amod` relation (including subtypes). `PROPN` and `PRON`
heads are excluded. [UD's amod relation](https://universaldependencies.org/u/dep/amod.html)
is broader than this selected feature. The function also does not reproduce
the Penn-tag selection in Kyle and Eguchi (2024) exactly.

The research question is whether annotation differences change the pairs and
counts used in an analysis. A correct total does not establish that the same
occurrences were found. A structurally valid tree can still have the wrong head.

## Supply complete sentences and basic trees

Use `lexdiv_import_annotations()` to import the complete, unnormalized output
of an existing parser or human annotation process. Keep punctuation and original
sentence text. The input contract is:

| Table | Required information |
|---|---|
| `segments` | `document_id`, `segment_id`, `text`; one sentence per segment, including empty inputs |
| token table | Same IDs; `token_index` starting at 1 in each sentence; exact `surface`; `upos`, `head`, `deprel` |
| optional lexical units | `lemma`, if lemma type counts are requested |
| provenance | Language, analyzer and version, dictionary and version, token unit and normalization; add model identity and any explicit projection choices |

`head` is numeric and refers to `token_index` in the **same sentence**; root has
head 0. `deprel` and `upos` use UD labels. Convert missing annotation fields to
typed `NA`, not the literal underscore used by CoNLL-U. Do not convert legitimate
surface strings such as `NA` into missing values. When adapting a parser table,
map its columns explicitly and preserve the relation between IDs and heads.

This is not a general CoNLL-U reader. Multiword-token range records, empty nodes,
and syntactic words without individually matching original surfaces are outside
the current input contract. Enhanced edges in `deps`/`DEPS` are rejected;
explicitly select the basic tree and record the projection when that is the
research target. See the [CoNLL-U specification](https://universaldependencies.org/format.html).
The importer never reconstructs original text from normalized token strings.

Known head errors, self-links, cycles, inconsistent root labels and multiple
roots raise errors. A valid complete tree has one root. These checks cannot
verify that a declared segment is linguistically one sentence or that its
syntactic analysis is correct.

## Run an offline English/Japanese example

The installed example supplies authored annotations. It requires no model,
corpus download or additional package. The Japanese sentence includes *赤い鳥*
to exercise original Japanese source positions and the same declared feature;
this does not establish comparable feature coverage across languages.

```{r}
env <- new.env(parent = baseenv())
sys.source(system.file("examples", "amod-pairs.R", package = "ldfreq",
  mustWork = TRUE), env)
example <- env$amod_pairs_example
reference <- lexdiv_amod_pairs(example$inputs$reference, unit = "lemma")
reference$occurrences[c("document_id", "dependent_term", "head_term",
  "token_distance", "pre", "keyword", "post")]
```

`keyword` spans both endpoints and all intervening text. The separate
`dependent_start`/`dependent_end` and `head_start`/`head_end` columns identify
the selected words. Positions are one-based, inclusive Unicode codepoints
within the original sentence. Keep document and segment IDs with these columns
when returning to KWIC or comparing occurrences. A row number is not an
occurrence identity.

## Report counts beside coverage

```{r}
predicted <- example$predicted
predicted$documents[c("document_id", "tokens", "analyzed_tokens", "token_coverage",
  "observed_pairs", "pairs", "pairs_with_unit", "types", "status")]
predicted$segments[c("document_id", "segment_id", "missing_heads",
  "missing_relations", "missing_upos", "status")]
```

Any missing head, relation or UPOS excludes that whole sentence. This conservative
rule prevents an unknown label from being silently classified as a negative.
The `partial` document retains one observed pair from a complete sentence, but
its total pair count is unknown because its other sentence is incomplete.
`zero` has a complete analysis with no selected relation; `unavailable` has no
analyzed sentence. Their totals are respectively 0 and `NA`. Empty inputs have
zero counts, an `empty` status and undefined token coverage.

Token coverage uses all supplied syntactic tokens, **including punctuation**,
as its denominator. `counts` and `observed_types` summarize the observed subset,
not an estimate for excluded sentences. Surface matching is case-sensitive;
selecting `unit = "lemma"` uses exactly the supplied lemmas. If a detected pair
is missing either lemma, it remains in the occurrence table but contributes no
known type key. `pairs_with_unit` reports this second source of incompleteness,
and a complete type total is `NA`.

## Compare occurrences as well as totals

```{r}
example$differences
subset(example$comparison, document_id == "wrong_head")
```

The `wrong_head` document has two pairs under both annotations, yet one intended
pair was lost and another introduced. The example joins **both endpoints** by
source IDs and positions, yielding one true positive, one false positive and
one false negative relative to the authored reference. Comparing word strings
alone would also lose the identity of repeated occurrences.

Incomplete documents remain in the differences table but are not scored as
complete-reference comparisons. Their unavailable predictions are not reported
as perfect zero-error results. The example is a transparent comparison recipe,
not a new general dependency-evaluation API. It assumes common original text and
segmentation; use `lexdiv_align_annotations()` to examine segmentation changes
before defining a cross-segmentation dependency evaluation.

```{r document-differences, fig.width=6.5, fig.height=3.2, dev.args=list(pointsize=11)}
draw_pair_changes <- function(example, monochrome = FALSE) {
  d <- example$differences
  keep <- d$evaluated & example$reference$documents$tokens > 0
  d <- d[keep, ]
  changed <- d$delta_pairs != 0
  old <- par(family = "sans", bty = "l", las = 1, mar = c(4, 7, 1, 1))
  on.exit(par(old))
  plot(d$delta_pairs, seq_len(nrow(d)), type = "n", yaxt = "n",
    xlim = c(-1.2, .2), xlab = "Pair Count Difference (Prediction - Reference)", ylab = "")
  abline(v = 0, lty = 2, col = "grey60")
  axis(2, at = seq_len(nrow(d)), labels = d$document_id, tick = FALSE)
  points(d$delta_pairs, seq_len(nrow(d)), pch = ifelse(changed, 17, 16),
    col = if (monochrome) "black" else ifelse(changed, "#D55E00", "#0072B2"))
  invisible(d)
}
draw_pair_changes(example)
```

This is an illustration of four nonempty documents, not an estimate of population
effects or parser accuracy. The zero difference for `wrong_head` is why the
occurrence comparison must accompany the plot. Use `monochrome = TRUE` for black
symbols. Place manuscript figure numbers, titles and explanatory notes outside
the image; see [the report guide](from-text-to-report.html) for final-size export.

## Save inputs and reproduce the result

```{r}
path <- tempfile(fileext = ".rds")
saveRDS(example$inputs, path)
saved <- readRDS(path)
replayed <- lexdiv_amod_pairs(saved$predicted, unit = saved$unit)
stopifnot(identical(replayed, example$predicted))
unlink(path)
```

For an empirical study, preserve annotation guidelines, sampling and reviewer
decisions separately. Candidate KWIC review alone cannot discover all missed
positive cases; independent complete-sentence annotation is needed to assess
recall. Authored fixtures verify implementation behavior, not the accuracy of
an English or Japanese parser, and a common API does not establish measurement
equivalence between languages.

No reference corpus is bundled for this feature. MI requires dependency-specific
joint counts, compatible marginals and opportunity totals; unigram frequencies
or adjacent-bigram frequencies do not supply those quantities automatically.
The present output ends at source-linked occurrences, counts and coverage.
