---
title: "Trace adjective–noun dependencies from source context to document counts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trace adjective–noun dependencies from source context to document counts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ldfreq)
# Optional local model; never acquired or run merely by installing ldfreq.
model_file <- Sys.getenv("LDFREQ_UDPIPE_MODEL")
run_udpipe <- nzchar(model_file) && requireNamespace("udpipe", quietly = TRUE)
```

## Which noun does an adjective modify?

In *Big red balloons float*, both *big* and *red* modify *balloons*.
Adjacent bigrams cannot represent the first relation as a two-word sequence.
`lexdiv_amod_pairs()` instead reads supplied basic Universal Dependencies (UD)
and connects each selected pair to its two source positions, context and
document counts. It does not run a parser or decide whether a phrase is idiomatic.

The initial feature is deliberately specific: a dependent tagged `ADJ`, a head
tagged `NOUN`, and an `amod` relation (including subtypes). `PROPN` and `PRON`
heads are excluded. [UD's amod relation](https://universaldependencies.org/u/dep/amod.html)
is broader than this selected feature. The function also does not reproduce
the Penn-tag selection in Kyle and Eguchi (2024) exactly.

The research question is whether annotation differences change the pairs and
counts used in an analysis. A correct total does not establish that the same
occurrences were found. A structurally valid tree can still have the wrong head.

## Supply complete sentences and basic trees

Use `lexdiv_import_annotations()` to import the complete, unnormalized output
of an existing parser or human annotation process. Keep punctuation and original
sentence text. The input contract is:

| Table | Required information |
|---|---|
| `segments` | `document_id`, `segment_id`, `text`; one sentence per segment, including empty inputs |
| token table | Same IDs; `token_index` starting at 1 in each sentence; exact `surface`; `upos`, `head`, `deprel` |
| optional lexical units | `lemma`, if lemma type counts are requested |
| provenance | Language, analyzer and version, dictionary and version, token unit and normalization; add model identity and any explicit projection choices |

`head` is numeric and refers to `token_index` in the **same sentence**; root has
head 0. `deprel` and `upos` use UD labels. Convert missing annotation fields to
typed `NA`, not the literal underscore used by CoNLL-U. Do not convert legitimate
surface strings such as `NA` into missing values. When adapting a parser table,
map its columns explicitly and preserve the relation between IDs and heads.

This is not a general CoNLL-U reader. Multiword-token range records, empty nodes,
and syntactic words without individually matching original surfaces are outside
the current input contract. Enhanced edges in `deps`/`DEPS` are rejected;
explicitly select the basic tree and record the projection when that is the
research target. See the [CoNLL-U specification](https://universaldependencies.org/format.html).
The importer never reconstructs original text from normalized token strings.

Known head errors, self-links, cycles, inconsistent root labels and multiple
roots raise errors. A valid complete tree has one root. These checks cannot
verify that a declared segment is linguistically one sentence or that its
syntactic analysis is correct.

## Run an offline English/Japanese example

The installed example supplies authored annotations. It requires no model,
corpus download or additional package. The Japanese sentence includes *赤い鳥*
to exercise original Japanese source positions and the same declared feature;
this does not establish comparable feature coverage across languages.

```{r}
env <- new.env(parent = baseenv())
sys.source(system.file("examples", "amod-pairs.R", package = "ldfreq",
  mustWork = TRUE), env)
example <- env$amod_pairs_example
reference <- lexdiv_amod_pairs(example$inputs$reference, unit = "lemma")
reference$occurrences[c("document_id", "dependent_term", "head_term",
  "token_distance", "pre", "keyword", "post")]
```

`keyword` spans both endpoints and all intervening text. The separate
`dependent_start`/`dependent_end` and `head_start`/`head_end` columns identify
the selected words. Positions are one-based, inclusive Unicode codepoints
within the original sentence. Keep document and segment IDs with these columns
when returning to KWIC or comparing occurrences. A row number is not an
occurrence identity.

## Report counts beside coverage

```{r}
predicted <- example$predicted
predicted$documents[c("document_id", "tokens", "analyzed_tokens", "token_coverage",
  "observed_pairs", "pairs", "pairs_with_unit", "types", "status")]
predicted$segments[c("document_id", "segment_id", "missing_heads",
  "missing_relations", "missing_upos", "status")]
```

Any missing head, relation or UPOS excludes that whole sentence. This conservative
rule prevents an unknown label from being silently classified as a negative.
The `partial` document retains one observed pair from a complete sentence, but
its total pair count is unknown because its other sentence is incomplete.
`zero` has a complete analysis with no selected relation; `unavailable` has no
analyzed sentence. Their totals are respectively 0 and `NA`. Empty inputs have
zero counts, an `empty` status and undefined token coverage.

Token coverage uses all supplied syntactic tokens, **including punctuation**,
as its denominator. `counts` and `observed_types` summarize the observed subset,
not an estimate for excluded sentences. Surface matching is case-sensitive;
selecting `unit = "lemma"` uses exactly the supplied lemmas. If a detected pair
is missing either lemma, it remains in the occurrence table but contributes no
known type key. `pairs_with_unit` reports this second source of incompleteness,
and a complete type total is `NA`.

## Compare occurrences as well as totals

```{r}
example$differences
subset(example$comparison, document_id == "wrong_head")
```

The `wrong_head` document has two pairs under both annotations, yet one intended
pair was lost and another introduced. The example joins **both endpoints** by
source IDs and positions, yielding one true positive, one false positive and
one false negative relative to the authored reference. Comparing word strings
alone would also lose the identity of repeated occurrences.

Incomplete documents remain in the differences table but are not scored as
complete-reference comparisons. Their unavailable predictions are not reported
as perfect zero-error results. The example is a transparent comparison recipe,
not a new general dependency-evaluation API. It assumes common original text and
segmentation; use `lexdiv_align_annotations()` to examine segmentation changes
before defining a cross-segmentation dependency evaluation.

```{r document-distributions, fig.width=6.5, fig.height=3.2, dev.args=list(pointsize=11)}
draw_pair_distributions <- function(example, monochrome = FALSE) {
  d <- example$differences
  keep <- d$evaluated & example$reference$documents$tokens > 0
  d <- d[keep, ]
  counts <- seq.int(0, max(d$reference_pairs, d$predicted_pairs))
  reference_n <- tabulate(match(d$reference_pairs, counts), nbins = length(counts))
  prediction_n <- tabulate(match(d$predicted_pairs, counts), nbins = length(counts))
  colors <- if (monochrome) c("black", "black") else c("#0072B2", "#D55E00")
  old <- par(family = "sans", bty = "l", las = 1, mar = c(4, 4.5, 1, 1))
  on.exit(par(old))
  ymax <- max(reference_n, prediction_n)
  plot(counts, reference_n, type = "n", xaxt = "n", yaxt = "n",
    ylim = c(0, ymax + 1), xlab = "Adjective–Noun Pairs per Document", ylab = "Documents")
  axis(1, at = counts)
  axis(2, at = seq.int(0, ymax))
  points(counts, reference_n, pch = 21, bg = "white", col = colors[1], cex = 1.5, lwd = 1.5)
  points(counts, prediction_n, pch = 17, col = colors[2], cex = .8)
  legend("topright", c("Reference", "Prediction"), col = colors, pch = c(21, 17),
    pt.bg = "white", pt.cex = c(1.5, .8), bty = "n")
  invisible(d)
}
draw_pair_distributions(example)
```

This overlays the observed count distributions for four nonempty documents;
it does not estimate population effects or parser accuracy. An open circle and
a smaller triangle remain visible when frequencies coincide. These few discrete
counts do not justify a smooth density curve. Identical count distributions can
still conceal different source occurrences, so retain the endpoint comparison.
Use `monochrome = TRUE` for black symbols. Place figure numbers, titles and notes outside
the image; see [the report guide](from-text-to-report.html) for final-size export.

## Save inputs and reproduce the result

```{r}
path <- tempfile(fileext = ".rds")
saveRDS(example$inputs, path)
saved <- readRDS(path)
replayed <- lexdiv_amod_pairs(saved$predicted, unit = saved$unit)
stopifnot(identical(replayed, example$predicted))
unlink(path)
```

For an empirical study, preserve annotation guidelines, sampling and reviewer
decisions separately. Candidate KWIC review alone cannot discover all missed
positive cases; independent complete-sentence annotation is needed to assess
recall. Authored fixtures verify implementation behavior, not the accuracy of
an English or Japanese parser, and a common API does not establish measurement
equivalence between languages.

No reference corpus is bundled for this feature. MI requires dependency-specific
joint counts, compatible marginals and opportunity totals; unigram frequencies
or adjacent-bigram frequencies do not supply those quantities automatically.
The present output ends at source-linked occurrences, counts and coverage.

## Connect a real parser in R

The optional [UDPipe R package](https://bnosac.github.io/udpipe/docs/doc2.html)
can supply tokenization, tags, lemmas and dependency parses without Python.
This recipe uses its complete `udpipe_annotate()` output and the original
sentence roster. The `sentence` column and reconstructed offsets in a parser
table are not a substitute for your original source text.

Install `udpipe` separately and select a local model appropriate to your study.
For example, the English EWT UD 2.5 model is available through
`udpipe::udpipe_download_model("english-ewt", model_dir = "models")`.
This is an explicit download, outside package installation and routine analysis.
The [model distributor](https://github.com/jwijffels/udpipe.models.ud.2.5) states
CC BY-NC-SA 4.0 for these models; model terms are separate from the package's
code license. Check the selected model's terms and training data. No model is
bundled here, and a general English model is not validated for your register
or learner population merely because this example runs.

To execute the following optional chunks, set `LDFREQ_UDPIPE_MODEL` to the local
model file before rendering this guide, or run the code with `model_file` set
to that path. These chunks are not run in ordinary package builds. The helper
is an explicitly sourced recipe, not an exported API.

```{r udpipe-source, eval=run_udpipe}
source(system.file("examples", "udpipe-amod.R", package = "ldfreq", mustWork = TRUE), local = TRUE)
stopifnot(file.exists(model_file), file.info(model_file)$size > 0)
model_sha256 <- digest::digest(file = model_file, algo = "sha256")
model <- udpipe::udpipe_load_model(model_file)

# Four authored inputs, including the empty document. This is not corpus data.
selected_docs <- c("english", "wrong_head", "zero", "empty")
roster <- subset(example$inputs$reference$segments, document_id %in% selected_docs,
                 select = c(document_id, segment_id, text))
roster$parser_id <- paste0("segment_", seq_len(nrow(roster)))
# Each caller-supplied sentence occupies one line; never silently remove line breaks.
stopifnot(!any(grepl("[\r\n]", roster$text)))
raw_annotation <- udpipe::udpipe_annotate(model, x = roster$text,
  doc_id = roster$parser_id, tokenizer = "tokenizer=presegmented", tagger = "default", parser = "default")
metadata <- list(language = "en", analyzer = "udpipe",
  analyzer_version = as.character(packageVersion("udpipe")),
  dictionary = "embedded-in-model", dictionary_version = model_sha256,
  unit = "UDPipe-syntactic-word", normalization = "none",
  model = basename(model_file), model_sha256 = model_sha256,
  model_source = "https://github.com/jwijffels/udpipe.models.ud.2.5",
  model_license = "CC BY-NC-SA 4.0",
  tokenizer = "tokenizer=presegmented", tagger = "default", parser = "default")
parsed <- import_udpipe_sentences(raw_annotation, roster, metadata)
actual <- lexdiv_amod_pairs(parsed, unit = "lemma")
actual$occurrences[c("document_id", "dependent_term", "head_term", "pre", "keyword", "post")]
actual$documents[c("document_id", "tokens", "pairs", "types", "token_coverage", "status")]
```

Change the model metadata when using a different model. Save the complete raw
output, including `x`, `conllu` and `errors`. The recipe verifies that `x`
matches the ordered original roster and rejects parser errors, unknown/reordered
IDs, multiple sentences in one input segment, multiword-token ranges, empty
nodes and enhanced edges. It maps integer token IDs without renumbering and
imports all surfaces, including punctuation, before selecting pairs. Annotation
underscores become missing fields through UDPipe's reader; literal surface
strings such as `NA` or `_` are preserved. Empty inputs stay in the roster.
Missing annotation values remain missing and receive the existing sentence-level
coverage treatment; a parser-reported failure stops this recipe for inspection.

For full documents, first supply and retain a sentence roster with its mapping
to the original document. This recipe does not automatically split raw documents
or invent document-global offsets. The
[presegmented tokenizer](https://ufal.mff.cuni.cz/udpipe/1/api-reference)
respects the supplied sentence line while still predicting token boundaries.
Use it only after deciding the sentence boundaries; passing a whole paragraph
on one line would force that paragraph into one parse. Newlines within a segment
require an explicit, recorded boundary decision, not silent whitespace replacement.
Saved outputs with multiple parser sentences per segment are still rejected.
Never silently discard unsupported rows to make an
import succeed. A saved annotated table alone cannot attest that filtering or
normalization did not occur upstream.

## Compare parser occurrences to the authored example

```{r udpipe-compare, eval=run_udpipe}
ref_occ <- subset(example$reference$occurrences, document_id %in% roster$document_id)
key <- c("document_id", "segment_id", "dependent_start", "dependent_end", "head_start", "head_end")
ref <- ref_occ[key]; ref$in_reference <- rep(TRUE, nrow(ref))
pred <- actual$occurrences[key]; pred$in_prediction <- rep(TRUE, nrow(pred))
occurrence_comparison <- merge(ref, pred, by = key, all = TRUE, sort = FALSE)
occurrence_comparison$in_reference[is.na(occurrence_comparison$in_reference)] <- FALSE
occurrence_comparison$in_prediction[is.na(occurrence_comparison$in_prediction)] <- FALSE
occurrence_comparison$outcome <- with(occurrence_comparison,
  ifelse(in_reference & in_prediction, "tp", ifelse(in_reference, "fn", "fp")))

document_comparison <- actual$documents
ref_rows <- match(document_comparison$document_id, example$reference$documents$document_id)
document_comparison$reference_pairs <- example$reference$documents$pairs[ref_rows]
document_comparison$delta_pairs <- document_comparison$pairs - document_comparison$reference_pairs
stopifnot(!anyNA(ref_rows), all(document_comparison$status %in% c("complete", "empty")))
document_comparison[c("document_id", "reference_pairs", "pairs", "delta_pairs", "status")]
occurrence_comparison
```

Endpoint comparison assumes the same original text and sentence boundaries.
This short illustration also requires complete annotations; it stops before
reporting comparisons if any document is incomplete. For a research sample,
report those failures rather than excluding them from the study roster.

With UDPipe 0.8.16 and English EWT UD 2.5 model SHA-256
`784bd0fa85e3d831fd02a55290d0acfd05c953159dc38cc33d52e1b28add9957`,
the *Big red balloons float.* input illustrates the need for this comparison:
two pairs are returned, but *big* modifies the predicted noun *float* instead
of *balloons* in the authored reference. An unchanged count conceals a false
positive and a false negative relative to that reference. These small, visible
authored cases demonstrate a workflow, not independent accuracy estimates.

```{r udpipe-plot, eval=run_udpipe, fig.width=6.5, fig.height=3.2, dev.args=list(pointsize=11)}
# Overlay the original count distributions; color is the default.
parser_plot <- list(differences = transform(document_comparison, evaluated = TRUE,
                      predicted_pairs = pairs),
                    reference = actual)
draw_pair_distributions(parser_plot)
# draw_pair_distributions(parser_plot, monochrome = TRUE)
```

```{r udpipe-replay, eval=run_udpipe}
record <- list(raw_annotation = raw_annotation, segments = roster, provenance = metadata,
               imported = parsed, pairs = actual, comparison = occurrence_comparison,
               document_comparison = document_comparison, session = sessionInfo())
record_path <- tempfile(fileext = ".rds") # Use a study-specific path to retain it.
saveRDS(record, record_path)
saved <- readRDS(record_path)
reimported <- import_udpipe_sentences(saved$raw_annotation, saved$segments, saved$provenance)
stopifnot(identical(reimported, saved$imported),
          identical(lexdiv_amod_pairs(reimported, unit = "lemma"), saved$pairs))
unlink(record_path)
```

Replay needs the optional UDPipe package to read its saved output, but does not
load a model or rerun inference. Keep the reader version as well as the original
analyzer/model metadata; a reader-version change is a provenance change, not
proof that the linguistic results changed. Use `lexdiv_amod_pairs(saved$imported)`
with the original `unit` argument to replay extraction without UDPipe at all.
