---
title: "Designing comparable lexical-diversity analyses"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Designing comparable lexical-diversity analyses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ldfreq)
```

## Start with a question and a comparison unit

Suppose the question is whether writers use a greater variety of word forms
locally in responses to the same writing task. The observations are documents;
repeated responses by the same writer are not independent participants. Fix
the task, population, lexical unit, preprocessing, and common window size before
comparing groups. A result about local surface-form variety is not automatically
a result about vocabulary knowledge, proficiency, or writing quality.

The examples below are synthetic. They isolate effects of length, order, and
window size so that we can explain what the calculations answer. They cannot
establish an effect in learners or the validity of a chosen window for a task.

## Why use a common MATTR window?

MATTR averages the proportion of distinct types over all complete windows of a
specified size. A common window makes its local denominator explicit. Global
TTR supplies a useful contrast: it depends on the full document length.

```{r}
vocabulary <- c("river", "stone", "tree", "cloud")
documents <- list(
  blocked = rep(vocabulary, each = 30),
  interleaved = rep(vocabulary, 30),
  interleaved_long = rep(vocabulary, 60)
)
primary <- lexdiv_metrics_batch(documents,
  metrics = c("ttr", "mattr"), window_length = 20)
primary[, c("document_id", "metric_id", "N", "V", "value", "status")]
```

The first two sequences have identical length and type frequencies. Their TTRs
are equal, but MATTR differs because one clusters repetition and the other
spreads the same four words across every window. Doubling the interleaved
sequence halves TTR while leaving its MATTR unchanged at 4/20. This controlled
example answers why local and whole-document variety need not agree; it does
not show that MATTR is length-independent for arbitrary texts.

In a real raw-text analysis, tokenize every document with the same explicit
choices, such as NFC and lowercase, and retain the resulting provenance.
Lowercasing merges case distinctions; selecting lemmas changes type identity;
restricting to content words changes both order spacing and the denominator.
Use annotation coverage to check whether exclusions differ between groups.
Do not choose a different lexical unit only for documents with poor coverage.

## Separate the main comparison from parameter sensitivity

Window 20 is an illustrative choice, not a recommended universal value. A
second common window asks whether the conclusion is sensitive to the local
scale. Do not pool the scores as interchangeable measurements.

```{r}
mattr_method <- lexdiv_methods()$method_id[
  lexdiv_methods()$metric_id == "mattr"]
plan <- lexdiv_plan(presets = character(), grids = lexdiv_grid(
  mattr_method, "window_length", c(20, 40), request_id_prefix = "window"))
sensitivity <- lexdiv_profile_batch(documents, plan)
sensitivity[, c("document_id", "request_id", "value", "requested_parameters")]
wide <- lexdiv_widen(sensitivity)
```

Interleaved MATTR changes from 4/20 to 4/40 solely because its window changes.
A plot must therefore select one request, for example
`plot(sensitivity, request_id = "window_1")`. The default wide table retains
the contract, method, parameters, and counts. Request labels are local to a
plan: when combining independently constructed plans, separate differing
conditions or use `names_from = "specification_id"`.

Documents shorter than the requested window receive a missing result; the
package does not silently shrink their window. Report these exclusions by
group. A computable score or a passed token-floor screen does not establish
acceptable precision or eliminate length effects. Estimate uncertainty with
a design that respects writers, tasks, and other sampling dependencies; these
descriptive metric functions do not perform that inferential analysis.

## A stable definition can still have an awkward boundary

The canonical MTLD definition requires at least ten tokens for a complete
factor and uses an unclamped linear residual factor. A short residual can
therefore create a discontinuity even when only one token is added.

```{r}
mtld_boundary <- data.frame(N = c(50L, 51L, 52L, 59L, 60L))
mtld_boundary$MTLD <- vapply(mtld_boundary$N, function(n) {
  lexdiv_metrics(rep("a", n), metrics = "mtld")$value
}, numeric(1))
mtld_boundary
```

For this intentionally extreme sequence, the 59-token result is about 7.22
and the 60-token result is 10. This follows the stated factor and tail rules;
changing or clamping those rules would define a different method. Inspect
directional/factor diagnostics and length sensitivity when MTLD is central to
the question. The variant APIs compare their explicitly stated definitions;
agreement on a label is not proof of agreement with another implementation.

Expected-TTR D has a different limitation: near-all-unique texts yield very
large estimates because the target curve approaches one. Package 0.2.0 fixes
numerical cancellation here, but numerical accuracy does not make that
boundary substantively informative. Report `near_saturation`, fit diagnostics,
sample sizes, and package version; all-unique input has an unbounded fit.

## Reference frequency answers a separate question

If the question also concerns how frequent the forms are in general YouTube
subtitles, use TUBELEX as a separate corpus-relative measure. Its Treebank
segmentation is not supplied by the general Unicode tokenizer in this package.
Prepare and record a compatible external pipeline; these hand-prepared terms
illustrate only one sentence.

```{r}
terms <- c("I", "do", "n't", "think", "it", "'s", "a", "book")
frequency <- tubelex_profile(terms)
frequency$summary[, c("weighting", "coverage", "mean_zipf")]
frequency$provenance$tokenization_alignment
```

The mean describes matched terms, not all terms with unknown words assigned
zero frequency. A change in mean can accompany a change in coverage, so inspect
both and the unmatched terms. Character-vector alignment remains unverified:
even 100% coverage cannot confirm the intended segmentation. The explicit
apostrophe normalization option harmonizes typography, not token boundaries.

This measure does not answer whether an essay is better or whether its writer
knows more vocabulary. Those claims need an appropriate external criterion,
reference-corpus justification, and study design.

## What to preserve and report

Preserve document IDs, task and writer identifiers, the prepared tokens and
preprocessing provenance, exclusions and coverage, the complete long result,
method and contract IDs, parameters, and the R/package session information.
Keep exact text and tokens under the study's data-access rules rather than
placing them automatically in a public report.

The synthetic results support a precise conclusion: equal whole-document
vocabularies can have different local variety, and a changed window changes
the question. They provide implementation and interpretation checks. An
empirical group conclusion still requires its actual effect size, uncertainty,
sampling design, exclusions, and sensitivity results.
