---
title: "English tokenization and document input"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{English tokenization and document input}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ldfreq)
```

## Choose what counts as a word

An essay can include contractions, web addresses and dates. Counting pieces of
a URL as vocabulary changes both lexical diversity and vocabulary-list coverage.
Before comparing texts, choose the same segmentation and inspect representative
examples from your corpus. ldfreq offers an English lexical tokenizer that runs
in R with its existing `stringi` dependency, without Python, model downloads,
or an external service.

Select it explicitly with `tokenizer = "english"`. The default `"unicode"`
retains the original segmentation for compatibility with existing analyses.

```{r}
text <- "John’s well-known book costs 3.14. See https://example.org on 2026-10-04."
english <- lexdiv_tokenize(text, tokenizer = "english", case = "lower")
english$tokens
english$provenance$excluded_spans
```

`John's` and `well-known` each count as one surface word. The price, URL, and
date do not enter the lexical denominator. Apostrophe and hyphen typography are
canonicalized after the requested Unicode normalization and case conversion:
left/right curly apostrophes become `'`, and Unicode hyphen/non-breaking hyphen
become `-`. This does not expand contractions or provide lemmas.

| Input | English lexical rule |
|---|---|
| `can't`, `won't`, `John's` | Keep whole; no grammatical interpretation |
| `well-known`, `COVID-19`, `2nd` | Keep as one word |
| `U.S.A.`, `e.g.` | Keep dotted initialisms with periods |
| `Dr.` | Retain `Dr`; discard the final punctuation |
| `3.14`, `-12`, `1,000`, `2e-3`, `75%` | One number-like span; exclude by default |
| `2026-10-04`, `3/4`, `10:30` | One number-like span; exclude by default |
| HTTP(S)/www URL, dotted-domain email address | Exclude recognized full span |
| Standalone punctuation, emoji, combining mark | Ignore |

With `keep_numbers = TRUE`, number-like expressions are retained and flagged
with `is_number = TRUE`. Numeric recognition is not date or number validation;
for example, an invalid calendar date can still be a number-like expression.
URL recognition can include trailing punctuation. Bare domains without www or
HTTP(S), malformed addresses, abbreviations requiring a dictionary, and
language-specific segmentation are outside these rules. Check token rows when
those cases are important to your corpus.

Offsets count Unicode code points in the **processed text**, after normalization,
case conversion and typography changes. They are not byte positions or offsets
into the original text. The full source is not retained, but tokens and
`excluded_spans` contain observed text; this is an analysis record, not redaction.
The exclusion table lists recognized URLs, emails and removed numbers, not every
punctuation character. SHA-256 fingerprints detect accidental edits when the
object is reused. They do not establish authenticity or reconstruct source text.

## Check the effect on your results

The two rules answer different operational questions. This small sensitivity
example illustrates the consequences; it does not establish that either rule
is universally superior.

```{r}
comparison_text <- "Read read at https://example.org on 2026-10-04."
prepared <- lapply(c("unicode", "english"), function(rule) {
  lexdiv_tokenize(comparison_text, tokenizer = rule, case = "lower")
})
names(prepared) <- c("unicode", "english")
comparison <- lexdiv_metrics_text_batch(prepared, metrics = "ttr")
knitr::kable(comparison$results[, c("document_id", "N", "V", "value")])
lapply(prepared, function(x) x$tokens$surface)
```

Here, removing non-lexical spans changes the number of tokens and types, and
therefore TTR. A higher TTR under one rule is not evidence of better writing or
better tokenization. Report the tokenizer ID/version and settings along with
the metric definition and parameters. For NJ8, a retained surface form such as
`John's`, `can't` or `U.S.A.` can still be off-list. Lemma analyses require an
explicit annotation step, and the coverage report must remain with the scores.
Both R tokenizers differ from the bundled TUBELEX Treebank segmentation; see
the separate [TUBELEX input guide](tubelex-input.html) when source-aligned
segmentation is required.

## Analyze an ID/text table

Use one row per document, with explicit unique IDs and character text. Other
columns, such as writer or task, stay in the study table for a later join by ID.
Empty text is a valid zero-token document. `NA` text is an error naming the
document; decide whether it denotes missing data before analysis rather than
silently replacing it with an empty string.

```{r}
essays <- data.frame(
  essay_id = c("essay_02", "essay_01", "empty"),
  response = c("The student reads a book and discusses the book.",
               "A reader explores a story with a friend.", "")
)
tokens <- lexdiv_tokenize_batch(
  essays, id_col = "essay_id", text_col = "response",
  tokenizer = "english", case = "lower"
)
analysis <- lexdiv_metrics_text_batch(
  tokens, metrics = c("ttr", "mattr"), window_length = 5
)
analysis$results
levels <- nj8_profile_batch(tokens, unit = "surface")
levels$coverage
```

The output keeps the input document order. Empty documents retain metric rows
with missing values and reasons, plus a preprocessing record. They have no token
audit rows. Short documents never reduce another document's requested window.
For raw text alone, `lexdiv_metrics_text_batch(essays, id_col = "essay_id",
text_col = "response", tokenizer = "english", case = "lower", ...)` combines
the two steps. For existing tokenizations, omit tokenizer arguments: their
recorded settings are authoritative. Annotated tokenizations can be analyzed
with `unit = "lemma"` or `"flemma"` and the same explicit word-inclusion policy
used by `lexdiv_metrics_text()`.

## Import CSV or UTF-8 text files

Base R is sufficient. Read ID/text columns as character to retain identifiers
such as `001` and literal responses such as `TRUE`. Choose missing-value markers
deliberately for your data.

```{r, eval=FALSE}
essays <- read.csv(
  "essays.csv", fileEncoding = "UTF-8", colClasses = "character",
  na.strings = "<MISSING>", check.names = FALSE
)
tokens <- lexdiv_tokenize_batch(essays, tokenizer = "english", case = "lower")

# Explicit file-to-ID mapping, with each file representing one document.
files <- c(essay_01 = "texts/essay_01.txt", essay_02 = "texts/essay_02.txt")
texts <- vapply(files, function(path) {
  paste(readLines(path, encoding = "UTF-8", warn = FALSE), collapse = "\n")
}, character(1))
tokens <- lexdiv_tokenize_batch(texts, tokenizer = "english", case = "lower")
```

This file recipe joins lines with newline separators and does not record source
file bytes. The tokenization hash identifies the string passed to the tokenizer.
Retain the original corpus and its import procedure separately when byte-level
source reconstruction is needed.

## Save the complete analysis

```{r}
record <- list(tokens = tokens, diversity = analysis, nj8 = levels,
               session = sessionInfo())
path <- tempfile(fileext = ".rds")
saveRDS(record, path)
stopifnot(identical(readRDS(path), record))
unlink(path)
wide <- lexdiv_widen(analysis$results)
```

Use `wide` for a join or a CSV export, while keeping the RDS record for the
tokenization and exclusion decisions. In methods, state that the English
tokenizer preserves contractions and hyphenated words, identify the number
policy and case/normalization choices, and distinguish surface coverage from
lemma coverage. Cite ldfreq and the New JACET 8000 source recorded in
`levels$provenance` when reporting NJ8 results. The
[text-to-report example](from-text-to-report.html) develops the interpretation
and reporting of the resulting metrics.
