---
title: "English tokenization and document input"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{English tokenization and document input}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ldfreq)
```

## Choose what counts as a word

An essay can include contractions, web addresses and dates. Counting pieces of
a URL as vocabulary changes both lexical diversity and vocabulary-list coverage.
Before comparing texts, choose the same segmentation and inspect representative
examples from your corpus. ldfreq offers an English lexical tokenizer that runs
in R with its existing `stringi` dependency, without Python, model downloads,
or an external service.

Select it explicitly with `tokenizer = "english"`. The default `"unicode"`
retains the original segmentation for compatibility with existing analyses.

Tokenization does not correct learner spelling or grammar. Preserve the source
and declare any correction policy separately. The
[reviewed-text example](auditing-vocabulary-profiles.html#handle-learner-errors-without-losing-the-observed-text)
keeps approved, rejected and unresolved edits and compares complete original
and revised texts without treating off-list words as errors.

```{r}
text <- "John’s well-known book costs 3.14. See https://example.org on 2026-10-04."
english <- lexdiv_tokenize(text, tokenizer = "english", case = "lower")
english$tokens
english$provenance$excluded_spans
```

`John's` and `well-known` each count as one surface word. The price, URL, and
date do not enter the lexical denominator. Apostrophe and hyphen typography are
canonicalized after the requested Unicode normalization and case conversion:
left/right curly apostrophes become `'`, and Unicode hyphen/non-breaking hyphen
become `-`. This does not expand contractions or provide lemmas.

| Input | English lexical rule |
|---|---|
| `can't`, `won't`, `John's` | Keep whole; no grammatical interpretation |
| `well-known`, `COVID-19`, `2nd` | Keep as one word |
| `U.S.A.`, `e.g.` | Keep dotted initialisms with periods |
| `Dr.` | Retain `Dr`; discard the final punctuation |
| `3.14`, `-12`, `1,000`, `2e-3`, `75%` | One number-like span; exclude by default |
| `2026-10-04`, `3/4`, `10:30` | One number-like span; exclude by default |
| HTTP(S)/www URL, dotted-domain email address | Exclude recognized full span |
| Standalone punctuation, emoji, combining mark | Ignore |

With `keep_numbers = TRUE`, number-like expressions are retained and flagged
with `is_number = TRUE`. Numeric recognition is not date or number validation;
for example, an invalid calendar date can still be a number-like expression.
URL recognition can include trailing punctuation. Bare domains without www or
HTTP(S), malformed addresses, abbreviations requiring a dictionary, and
language-specific segmentation are outside these rules. Check token rows when
those cases are important to your corpus.

Offsets count Unicode code points in the **processed text**, after normalization,
case conversion and typography changes. They are not byte positions or offsets
into the original text. The full source is not retained, but tokens and
`excluded_spans` contain observed text; this is an analysis record, not redaction.
The exclusion table lists recognized URLs, emails and removed numbers, not every
punctuation character. SHA-256 fingerprints detect accidental edits when the
object is reused. They do not establish authenticity or reconstruct source text.

## Check the effect on your results

The two rules answer different operational questions. This small sensitivity
example illustrates the consequences; it does not establish that either rule
is universally superior.

```{r}
comparison_text <- "Read read at https://example.org on 2026-10-04."
prepared <- lapply(c("unicode", "english"), function(rule) {
  lexdiv_tokenize(comparison_text, tokenizer = rule, case = "lower")
})
names(prepared) <- c("unicode", "english")
comparison <- lexdiv_metrics_text_batch(prepared, metrics = "ttr")
knitr::kable(comparison$results[, c("document_id", "N", "V", "value")])
lapply(prepared, function(x) x$tokens$surface)
```

Here, removing non-lexical spans changes the number of tokens and types, and
therefore TTR. A higher TTR under one rule is not evidence of better writing or
better tokenization. Report the tokenizer ID/version and settings along with
the metric definition and parameters. For NJ8, a retained surface form such as
`John's`, `can't` or `U.S.A.` can still be off-list. Lemma analyses require an
explicit annotation step, and the coverage report must remain with the scores.
Both R tokenizers differ from the bundled TUBELEX Treebank segmentation; see
the separate [TUBELEX input guide](tubelex-input.html) when source-aligned
segmentation is required.

## Declare spelling, name, numeral and symbol policies

These are measurement choices, not automatically errors in the writer's text.
Keep source forms, analysis keys and reference-lookup keys separate. The raw
tokenizer does not infer national variety, proper names or the meaning of a
number. Current behavior and possible alternatives are:

| Item | Current English lexical tokenizer | Study-specific alternative |
|---|---|---|
| `colour` / `color` | Separate surface forms; no variety conversion | Reviewed equivalence for counting, lookup, or both, declared separately |
| `Rose` / `rose` | Retains both; `case = "lower"` merges their spellings | Annotate the original case; select proper-noun occurrences using supplied POS |
| `2026`, `3.14`, `75%` | Removed with `keep_numbers = FALSE`; retained with `TRUE` | Retain for numeracy/discourse questions; record the numeric-expression rule |
| `two`, `2nd`, `COVID-19` | Retained even with `keep_numbers = FALSE` | Use explicit linguistic annotation or reviewed categories if broader exclusion is needed |
| Punctuation, `$`, `+`, emoji | Standalone marks are not lexical tokens | Import complete annotations when these marks or their boundaries matter |
| `can't`, `well-known` | Internal apostrophes/hyphens retained | A different segmentation needs its own declared tokenizer/reference compatibility |

### English spelling varieties

`colour` and `color` are accepted variety spellings
([Cambridge Dictionary](https://dictionary.cambridge.org/dictionary/english/color)).
The bundled NJ8 snapshot matches `color`, `center` and `theater`, but does not
match the surface queries `colour`, `centre` and `theatre`. This is a lookup
boundary, not evidence that the British spellings are errors or more advanced
words. NJ8's `expand_parenthetical` only expands alternatives actually present
in the resource; it is not a comprehensive British/American conversion table.

```{r}
spelling_lookup <- nj8_profile(c("colour", "color", "centre", "center",
  "theatre", "theater"), unit = "surface")
spelling_lookup$lookup[, c("surface_term", "lookup_term", "matched", "level")]
```

For surface-form research, keep the distinction. For a declared equivalence
analysis, retain a versioned, reviewed mapping and show the original result
alongside it. Mapping only the lookup key can improve reference coverage without
changing the diversity calculation. Merging types is a separate decision.
There is no universal need to prefer US over UK forms, and English varieties
are not exhausted by those two labels. Avoid global suffix substitutions or
rewriting names/titles. Some mappings depend on POS and sense; for example,
[practise is a UK verb form](https://dictionary.cambridge.org/dictionary/english/practise),
while [practice also serves as a noun](https://dictionary.cambridge.org/dictionary/english/practice).
Spelling equivalence is distinct from inflectional lemmatization and synonymy.

For frequency resources such as TUBELEX, querying the US form is **not** a
pooled frequency for both spellings. Pooling requires compatible underlying
counts and denominators and a declared grouping; do not add log/Zipf values.
Document ranges cannot simply be added because a document may contain both
forms. Surface-form psycholinguistic properties such as orthographic length or
form-specific familiarity must remain attached to the displayed form unless
there is evidence for combining them. No frequency/norm pooling is implemented
by the example below.

### Proper nouns and numerals

`word_inclusion = "content"` means `ADJ`, `ADV`, `NOUN`, **`PROPN`** and `VERB`.
It therefore includes proper nouns; it is not a "remove names" switch. For a
different selection, import complete annotations and create an explicit view.
Capitalization alone cannot decide whether a token is a name: `Rose` can name a
person, while an ordinary noun may start a sentence. Keep case for annotation,
then separately declare case handling for counting. A POS filter is not full
named-entity recognition: some words within names/titles retain other POS tags,
as explained in the [UD PROPN specification](https://universaldependencies.org/u/pos/PROPN.html).
Excluding complete entities requires reviewed spans or an identified NER layer.
Replacing all names with `<NAME>` creates an artificial repeated type; it is
neither simple exclusion nor an unchanged lexical-diversity analysis. Selection
also does not anonymize the saved source or audit tables.

`is_number` is a tokenizer pattern flag, not a semantic numeral annotation.
With English tokenization it recognizes digit-based numeric expressions,
including dates and percentages. `two` and `IV` are not caught by that flag;
`2nd` and alphanumeric product names remain word-like tokens. In contrast,
[UD NUM](https://universaldependencies.org/u/pos/NUM.html) includes cardinal
number words, but ordinarily places English ordinals such as `second` under
ADJ. Thus even removing `NUM` does not mean removing every expression with a
numerical meaning. The legacy `tokenizer = "unicode"` has a narrower pure-number
rule; state which tokenizer was used.

Retaining `1,000` and `1000` does not automatically merge their forms or parse
their values. The tokenizer is not a locale-aware numeric parser and does not
validate dates. For quantity analysis, define decimal/grouping conventions,
units, dates, identifiers and ambiguous forms separately. Replacing all numeric
expressions with `<NUM>` would again change the measured type distribution.

### Symbols and boundaries

[UD distinguishes symbols from punctuation](https://universaldependencies.org/u/pos/SYM.html):
currency signs, mathematical operators and emoji can carry meaning. Ignoring
them can be reasonable for a declared word-based measure, but cannot support a
claim about their absence or their communicative functions. The raw English
tokenizer turns `C++` into `C`, `R&D` into `R` and `D`, and `#topic` into `topic`.
It does not interpret `&` as `and`, expand currency signs, preserve social-media
entities or provide a punctuation-error measure. Curly apostrophes and lexical
hyphens are canonicalized; an em dash separates words. Inspect such cases before
using a lexical tokenizer for specialist texts.

For punctuation/symbol research, use complete externally supplied segmentation
with `lexdiv_import_annotations()`. It retains every supplied surface and aligns
it to the original text; a reviewed token can therefore preserve `C++`. This
does not make its POS automatically correct. Use the same tokenization policy
across comparison groups and retain the source record.

For adjacent n-grams, do not remove marks and renumber the remaining words.
Dropping `+` from `cats + dogs` must not silently create an adjacent `cats dogs`
pair under a policy that treats excluded positions as boundaries. The example
retains full original token positions, so `lexdiv_ngrams()` respects the gap.
`lexdiv_as_quanteda()` can also retain such gaps as padding. A raw lexical
tokenizer has already removed punctuation before numbering tokens; it cannot
reconstruct these boundaries later. Sentence/segment boundaries must be supplied
explicitly. Windowed diversity on a selected word sequence has a different
contract: its windows count selected tokens, not original character distance.

### Run the policy comparison

The installed example supplies eight authored documents and their complete
token rows, including a technical name, missing POS and an empty document. It
compares name/numeral selections and spelling equivalence through existing
functions. Three illustrative noun aliases are explicitly reviewed; this is
not an automatically validated or comprehensive spelling dictionary.

```{r}
policy_example <- new.env(parent = environment())
sys.source(system.file("examples", "lexical-counting-policy.R", package = "ldfreq",
  mustWork = TRUE), envir = policy_example)
counting <- policy_example$lexical_counting_example
subset(counting$metrics, metric_id == "ttr" &
  document_id %in% c("variants", "names", "numerals") &
  policy %in% c("words_and_numerals", "without_names_or_numerals", "spelling_equivalence"),
  select = c(policy, document_id, selected_known, types_known, reportable_value))
subset(counting$coverage, document_id == "variants" & policy == "words_and_numerals")
```

For `The colour is color.`, surface TTR is 1 and TTR with declared spelling
equivalence is 0.75. Exact NJ8 token coverage is 2/4; reviewed alias lookup is
3/4. The lookup comparison is also available without merging diversity types.
In `Rose saw a rose.`, lowercased surface TTR increases from 0.75 to 1 after
removing the supplied PROPN occurrence. Excluding names does not have a fixed
direction of effect; inspect N and V together.

`$annotation` retains full source/positions, `$selections` keeps every token and
its exclusion reason, and `$runs` retains complete metric/reference objects and
query-to-source links. `$aliases` and `$alias_info` retain the mapping, review
metadata and fingerprint. `conditional_value` summarizes only selected tokens
with known POS. If a selection cannot be resolved because POS is missing,
`reportable_value` stays missing; neither the unknown token nor its document is
silently treated as a complete case. Reference coverage is explicitly conditional
on selected known tokens. The three-token MATTR window is illustrative only.

```{r}
policy_path <- tempfile(fileext = ".rds")
saveRDS(list(result = counting, session = sessionInfo()), policy_path)
stopifnot(identical(readRDS(policy_path)$result, counting))
unlink(policy_path)
```

For a study, save to a persistent path and retain the annotation model/version
and review procedure. The authored tags here demonstrate accounting and source
alignment, not POS/NER accuracy, psycholinguistic validity or a preferred
universal exclusion policy.

## Analyze an ID/text table

Use one row per document, with explicit unique IDs and character text. Other
columns, such as writer or task, stay in the study table for a later join by ID.
Empty text is a valid zero-token document. `NA` text is an error naming the
document; decide whether it denotes missing data before analysis rather than
silently replacing it with an empty string.

```{r}
essays <- data.frame(
  essay_id = c("essay_02", "essay_01", "empty"),
  response = c("The student reads a book and discusses the book.",
               "A reader explores a story with a friend.", "")
)
tokens <- lexdiv_tokenize_batch(
  essays, id_col = "essay_id", text_col = "response",
  tokenizer = "english", case = "lower"
)
analysis <- lexdiv_metrics_text_batch(
  tokens, metrics = c("ttr", "mattr"), window_length = 5
)
analysis$results
levels <- nj8_profile_batch(tokens, unit = "surface")
levels$coverage
```

The output keeps the input document order. Empty documents retain metric rows
with missing values and reasons, plus a preprocessing record. They have no token
audit rows. Short documents never reduce another document's requested window.
For raw text alone, `lexdiv_metrics_text_batch(essays, id_col = "essay_id",
text_col = "response", tokenizer = "english", case = "lower", ...)` combines
the two steps. For existing tokenizations, omit tokenizer arguments: their
recorded settings are authoritative. Annotated tokenizations can be analyzed
with `unit = "lemma"` or `"flemma"` and the same explicit word-inclusion policy
used by `lexdiv_metrics_text()`.

## Import CSV or UTF-8 text files

Base R is sufficient. Read ID/text columns as character to retain identifiers
such as `001` and literal responses such as `TRUE`. Choose missing-value markers
deliberately for your data.

```{r, eval=FALSE}
essays <- read.csv(
  "essays.csv", fileEncoding = "UTF-8", colClasses = "character",
  na.strings = "<MISSING>", check.names = FALSE
)
tokens <- lexdiv_tokenize_batch(essays, tokenizer = "english", case = "lower")

# Explicit file-to-ID mapping, with each file representing one document.
files <- c(essay_01 = "texts/essay_01.txt", essay_02 = "texts/essay_02.txt")
texts <- vapply(files, function(path) {
  paste(readLines(path, encoding = "UTF-8", warn = FALSE), collapse = "\n")
}, character(1))
tokens <- lexdiv_tokenize_batch(texts, tokenizer = "english", case = "lower")
```

This file recipe joins lines with newline separators and does not record source
file bytes. The tokenization hash identifies the string passed to the tokenizer.
Retain the original corpus and its import procedure separately when byte-level
source reconstruction is needed.

## Save the complete analysis

```{r}
record <- list(tokens = tokens, diversity = analysis, nj8 = levels,
               session = sessionInfo())
path <- tempfile(fileext = ".rds")
saveRDS(record, path)
stopifnot(identical(readRDS(path), record))
unlink(path)
wide <- lexdiv_widen(analysis$results)
```

Use `wide` for a join or a CSV export, while keeping the RDS record for the
tokenization and exclusion decisions. In methods, state that the English
tokenizer preserves contractions and hyphenated words, identify the number
policy and case/normalization choices, and distinguish surface coverage from
lemma coverage. Cite ldfreq and the New JACET 8000 source recorded in
`levels$provenance` when reporting NJ8 results. The
[text-to-report example](from-text-to-report.html) develops the interpretation
and reporting of the resulting metrics.
