---
title: "From English text to a report"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From English text to a report}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ldfreq)
```

## Question and unit of analysis

Suppose we want to describe how two texts about reading differ in local
surface-word variety and coverage by New JACET 8000. These are separate
questions: repeating familiar words can change diversity without changing
list membership. The observations here are two project-authored sentences;
they do not represent learner groups or establish proficiency differences.

We use the same English tokenizer, NFC normalization, lowercase conversion, word selection, and
10-token MATTR window for both texts. The small window makes the example
executable; it is not a recommended universal window. In a study, justify the
common window from the task and document-length distribution before comparing
results. Keep writer and task IDs with document IDs in the study's own table.

```{r}
texts <- c(
  first = "The student reads a book and discusses the book with a friend.",
  second = "The student explores a story and shares several ideas with a friend."
)
prepared <- lexdiv_tokenize_batch(
  texts, tokenizer = "english", normalization = "NFC", case = "lower"
)
diversity <- lexdiv_metrics_text_batch(
  prepared, metrics = c("ttr", "mattr"), window_length = 10
)
metrics <- diversity$results
knitr::kable(metrics[, c("document_id", "metric_id", "N", "V", "value", "status")])
```

The first text repeats `the`, `a`, and `book`; the second repeats `a`.
Both contain 12 tokens, but the first has 9 distinct surface forms and the second
has 11. MATTR rises from 0.800 to 0.933: on average, the second text contains
about 1.33 more distinct forms per 10-token window. This describes greater
local word variety under the common settings; it cannot establish that one text
is better. For other documents, inspect `status`
and `missing_reason`: a text below 10 tokens gets no value for this requested
MATTR, and its presence does not shrink the other documents' windows.

## Add level coverage without changing the word unit

The bundled NJ8 resource is a representative-lemma list. We deliberately request
surface-form coverage, so `reads` is looked up as `reads`, not silently converted
to `read`. This answers a surface-matching question. For lemma-based research,
explicitly annotate every document with the same lemma procedure, report its
coverage, and select `unit = "lemma"`. Do not switch units only for documents
with poor matching.

```{r}
levels <- nj8_profile_batch(prepared, unit = "surface")
knitr::kable(levels$coverage[, c(
  "document_id", "eligible_tokens", "matched_tokens", "token_coverage", "type_coverage"
)])
levels$lookup[!levels$lookup$matched, c("document_id", "term")]
```

The unmatched rows show which surface forms contribute to off-list coverage.
NJ8 token coverage is 83.3% (10/12) for the first text and 75.0% (9/12) for the
second. Thus, the text with greater local variety has lower surface-list
coverage in this example. Inflected forms such as `reads` can be off-list even
when their lemma is familiar.
They are not automatically rare, difficult, or erroneous words. All eligible
terms remain in the denominator, including those not in NJ8.

For each document, the cumulative Level 8 token proportion equals its total
token coverage. Lower cumulative levels describe the fraction found at or below
that level, not a learner's vocabulary knowledge.

```{r}
level8 <- subset(levels$summary, weighting == "token" & level == 8)
stopifnot(isTRUE(all.equal(
  level8$cumulative_proportion, levels$coverage$token_coverage
)))
plot(nj8_profile(prepared[[1]], unit = "surface"), weighting = "token")
```

## Produce an analysis table and inspect it

`lexdiv_widen()` retains each metric's definition, parameters, and counts.
Its checks prevent different MATTR windows from entering one metric column
as if they were the same specification. Coverage is then joined by explicit
IDs, preserving document order.

```{r}
analysis <- lexdiv_widen(metrics)
coverage <- levels$coverage[match(analysis$document_id, levels$coverage$document_id), ]
stopifnot(identical(analysis$document_id, coverage$document_id))
analysis$nj8_surface_token_coverage <- coverage$token_coverage
analysis$nj8_surface_type_coverage <- coverage$type_coverage
knitr::kable(analysis[, c(
  "document_id", "ttr__value", "mattr__value",
  "nj8_surface_token_coverage", "nj8_surface_type_coverage"
)], col.names = c("Document", "TTR", "MATTR", "NJ8 token coverage", "NJ8 type coverage"),
digits = 3)
plot(metrics, metric_id = "mattr")
```

Report differences in the selected metric and differences in coverage
separately. These observations support a descriptive answer for the supplied
texts under the stated settings. They do not answer whether a difference
persists across writers or tasks; that requires a sampling design and an
appropriate analysis of independent and repeated observations.

If the research question also concerns frequency in YouTube subtitles, follow
`vignette("tubelex-input", package = "ldfreq")`. Keep that segmentation and its
frequency results distinct from the surface units selected above.

## Save enough to reproduce and report the analysis

Save the full results and preprocessing record, not just a CSV of scores.
The following round trip keeps list-columns and resource identity intact.
The saved object contains the input terms; handle it under the study's data
access rules. It is not intended for automatic public sharing.

```{r}
record <- list(
  prepared = prepared, diversity = diversity, metrics = metrics, levels = levels,
  analysis = analysis, session = sessionInfo()
)
path <- tempfile(fileext = ".rds")
saveRDS(record, path)
stopifnot(identical(readRDS(path), record))
unlink(path)
levels$provenance[c("resource_version", "resource_bundled", "resource_citation")]
```

A methods statement for this example could read:

> We tokenized both texts with the ldfreq English word tokenizer (0.1.0),
> using NFC normalization and lowercase
> surface forms. We calculated TTR over each full text and MATTR with a
> common 10-token window.
> New JACET 8000 coverage used the bundled jacet2016-8000-v1 table with the
> default NFKC/lowercase lookup normalization and parenthetical aliases.
> Off-list forms remained in the coverage denominator. We retained the full
> results, preprocessing records, and R/package session information.

In a real report, include the actual package version, selected `method_id`,
parameters, document counts, exclusions, task/writer structure, and the reason
for the window choice. Report effect sizes and uncertainty using the study's
sampling units; these descriptive functions do not choose an inferential model.

## What to cite

Use `citation("ldfreq")` for the software. Cite the resources and methods
actually used, in addition to the software:

- **New JACET 8000:** JACET Basic Word Revision Committee (Ed.). (2016).
  *The New JACET List of 8000 Basic Words*. Tokyo: Kirihara Shoten.
  The original Japanese citation and permission notice are installed in
  `licenses/nj8/NOTICE.md`.
- **MATTR:** Covington, M. A., & McFall, J. D. (2010). Cutting the Gordian knot:
  The moving-average type-token ratio (MATTR). *Journal of Quantitative
  Linguistics, 17*(2), 94–100. <https://doi.org/10.1080/09296171003643098>.
- **MTLD and HD-D, if used:** McCarthy, P. M., & Jarvis, S. (2010). MTLD,
  vocd-D, and HD-D: A validation study of sophisticated approaches to lexical
  diversity assessment. *Behavior Research Methods, 42*, 381–392.
  <https://doi.org/10.3758/BRM.42.2.381>.
- **TUBELEX, if used:** Nohejl et al. (2025). Beyond film subtitles: Is YouTube
  the best approximation of spoken vocabulary? *COLING 2025*.
  <https://aclanthology.org/2025.coling-main.641/>.

A method citation does not establish identity with every implementation carrying
the same metric name. Retain the exact ldfreq definition and selected settings.
