---
title: "Getting started with ldfreq"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ldfreq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`ldfreq` begins at an assumption-transparent boundary: an ordered character
vector of tokens. The package does not guess how raw prose should be tokenized,
normalized, or lemmatized.

```{r}
library(ldfreq)

tokens <- c("the", "cat", "saw", "the", "other", "cat")
result <- lexdiv_metrics(tokens, metrics = c("ttr", "rttr", "yule_k"))
result
```

## Reading results

Each requested metric receives one row. The `status` and `missing_reason`
columns state whether a value is available; `requested_parameters` and
`effective_parameters` are list-columns. Contract and schema identifiers are
ordinary columns so that they survive row binding and tabular serialization.

Short documents do not cause the requested method to change. For example, an
MATTR window larger than the document yields a structured missing result rather
than a smaller implicit window.

```{r}
lexdiv_metrics(tokens, metrics = "mattr", window_length = 50L)
```

## Interpreting values

Maas a-squared and Yule's K use lower values for more lexical diversity. TTR,
RTTR, CTTR, Herdan's C, MSTTR, MATTR, MTLD, HD-D, and Yule's I use higher values.
This direction is meaningful only within the same method, parameters,
tokenization, normalization, sampling design, and comparable texts. Scores from
different metric IDs are not interchangeable because their scales and length
sensitivities differ.

The metrics describe lexical-distribution properties of the supplied tokens.
They are not direct measures of language proficiency, writing quality, reader
response, or communicative effectiveness. A separate validated study design is
needed to make claims about those constructs.

## Multiple documents

Batch input must carry explicit, unique document identifiers.

```{r}
documents <- list(
  document_a = c("one", "two", "one"),
  document_b = c("alpha", "beta", "gamma", "delta")
)

lexdiv_metrics_batch(documents, metrics = c("ttr", "hdd"))
```

## Explicit method profiles

A profile plan records parameter variants before computation. The
`length_50_100` preset adds MSTTR and MATTR settings without duplicating MTLD or
turning descriptive token floors into alternative metric values.

```{r}
plan <- lexdiv_plan(presets = "length_50_100")
profile <- lexdiv_profile(tokens, plan)
head(profile)

screen <- lexdiv_screen(profile)
head(screen)
```

Quality floors and requested screens are advisory token-count checks. They do
not change values or parameters, censor documents, or establish validity or
reliability. Read them together with `status`, `missing_reason`, `N`, `V`, and
the method and parameter identities.

Use `lexdiv_methods()` and `lexdiv_presets()` to inspect the frozen registries.
The `expected_ttr_d` candidate is deferred from v0.1 and intentionally absent
from the public method set.

## Offline installed-package check

An installed smoke script exercises the public workflows without network access
or an external runtime:

```{r}
smoke_path <- system.file("examples", "offline-smoke.R", package = "ldfreq")
stopifnot(nzchar(smoke_path))
source(smoke_path, local = TRUE)
```
