---
title: "Getting started with ldfreq"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ldfreq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

There are two common starting points. If you already have one character value
per token, use `lexdiv_metrics()`. If you have ordinary prose, use
`lexdiv_metrics_text()`. The second route records how the text was tokenized,
normalized, and optionally lemmatized before it calls the same metric core.

```{r}
library(ldfreq)

tokens <- c("the", "cat", "saw", "the", "other", "cat")
result <- lexdiv_metrics(tokens, metrics = c("ttr", "rttr", "yule_k"))
result
```

## Raw text without hidden preprocessing

The shortest raw-text workflow is a direct call to `lexdiv_metrics_text()`.
Normalization, case, and number handling are explicit arguments rather than
global options.

```{r}
raw_result <- lexdiv_metrics_text(
  "Cats and cat ran run.",
  normalization = "NFC",
  case = "preserve",
  metrics = "ttr"
)
raw_result
```

Call `lexdiv_tokenize()` separately when you want to inspect or annotate the
tokens first. Its offsets refer to the processed text, and its provenance keeps
source and processed text hashes without storing the source text itself.

```{r}
tokenization <- lexdiv_tokenize(
  "Cats and cat ran run.",
  normalization = "NFC",
  case = "preserve"
)
tokenization
```

Lemma and content-word analyses require an explicit annotation source. The
example annotations are project-authored fixtures, not a recommended universal
lemmatizer. Missing annotations would be excluded and counted in
`unit_coverage` rather than imputed.

```{r}
annotated <- lexdiv_lemmatize(
  tokenization,
  lemmas = c("cat", "and", "cat", "run", "run"),
  upos = c("NOUN", "CCONJ", "NOUN", "VERB", "VERB"),
  backend_id = "vignette-fixture",
  backend_version = "1"
)

lemma_content <- lexdiv_metrics_text(
  annotated,
  unit = "lemma",
  word_inclusion = "content",
  metrics = "ttr"
)
lemma_content$results
lemma_content$token_audit
lemma_content$preprocessing
```

For a quick English lemma baseline, the optional `textstem` backend can be
selected explicitly. It supplies lemmas only; it does not infer UPOS tags, so a
content-word analysis still needs tags from a separately identified source.

```{r}
if (requireNamespace("textstem", quietly = TRUE)) {
  automatic_lemmas <- lexdiv_lemmatize(
    lexdiv_tokenize("The cats were running and studies.", case = "lower"),
    method = "textstem"
  )
  automatic_lemmas$tokens[, c("surface", "lemma")]
  automatic_lemmas$provenance$annotation
}
```

This separation is deliberate: changing forms to lemmas or restricting the
denominator to content words changes the measurement. It is not merely a file
cleaning step.

Flemma is a separate lexical unit because it groups forms without preserving
POS distinctions. `lexdiv_flemmatize()` accepts a caller-supplied local AntBNC
file; it neither bundles nor downloads that resource. Unknown forms retain a
normalized identity value, while token rows record AntBNC, override, or
identity matching.

```{r, eval=FALSE}
flemmas <- lexdiv_flemmatize(
  tokenization,
  "/path/to/antbnc_lemmas_ver_004.txt",
  resource_version = "004"
)
lexdiv_metrics_text(flemmas, unit = "flemma", metrics = "ttr")
```

The raw AntBNC mapping approximates, but does not reproduce, New Word Level
Checker. NWLC documents manual word-list alignment. New JACET profiles
therefore expose AntBNC-versus-headword conflicts and accept explicit upstream
overrides.

## Corpus-relative frequency with coverage

`tubelex_frequency_profile()` applies either an explicit source-aligned query
normalization or identity matching. It returns original terms, lookup terms,
unmatched rows, token/type coverage, and matched-only conditional summaries.

```{r}
frequency <- tubelex_frequency_profile(tokenization)
frequency$summary
frequency$coverage
frequency$lookup
```

TUBELEX is a general YouTube subtitle frequency space. Its values describe
frequency and prevalence relative to that corpus; they are not direct measures
of proficiency, writing quality, academic register, or context-independent
lexical sophistication.

## Caller-supplied lexical frequency levels

`new_jacet8000_profile()` maps ranks in a caller-supplied New JACET 8000 table
to eight 1,000-rank levels. The package contains no copy of the list and never
downloads one. Exact and cumulative token/type rates keep off-list items in the
denominator. The official `j8_2016.xlsx` workbook can be passed directly; its
`新J8` sheet and Japanese rank/lemma column names are detected automatically.

```{r}
synthetic_nj8 <- data.frame(
  NJ8 = c(1L, 1001L, 6001L, 8000L),
  Word = c("the", "develop", "rare", "extreme")
)
levels <- new_jacet8000_profile(annotated, synthetic_nj8, unit = "lemma")
levels$summary
levels$coverage
```

Use `plot(levels)` for token proportions and their cumulative curve, or
`plot(levels, weighting = "type")` for normalized types.

For an object created by `lexdiv_flemmatize()`, set `unit = "flemma"` and use
`flemma_conflict = "antbnc"`, `"wordlist"`, or `"error"`. The lossless lookup
retains the chosen resolution and the alternative surface-headword rank/level.

## Reading results

Each requested metric receives one row. The `status` and `missing_reason`
columns state whether a value is available; `requested_parameters` and
`effective_parameters` are list-columns. Contract and schema identifiers are
ordinary columns so that they survive row binding and tabular serialization.

| `status` | Meaning |
|---|---|
| `ok` | The requested method produced a value. |
| `missing` | The document was valid, but the method's frozen domain or requested parameter did not permit a value. |
| `invalid_input` | At least one invalid token made the document invalid; it was not silently deleted. |

Common `missing_reason` values distinguish empty input, insufficient tokens for
a formula, a document shorter than the requested window/segment/sample, a zero
denominator, and a missing MTLD factor. Structural request errors such as an
unknown metric ID stop the call instead of creating a result row.

Short documents do not cause the requested method to change. For example, an
MATTR window larger than the document yields a structured missing result rather
than a smaller implicit window.

```{r}
lexdiv_metrics(tokens, metrics = "mattr", window_length = 50L)
```

## Interpreting values

Within the same exact method and design, Maas a-squared and Yule's K
conventionally decrease as repetition decreases. TTR, RTTR, CTTR, Herdan's C,
MSTTR, MATTR, MTLD, HD-D, and Yule's I conventionally increase with observed
lexical variety or lower repetition. These methods primarily operationalize
lexical variety and repetition rather than the full multidimensional
lexical-diversity construct. Direction is meaningful only within the same
method, parameters, tokenization, normalization, sampling design, and comparable
texts. Scores from different metric IDs are not interchangeable because their
scales and length sensitivities differ.

The metrics describe lexical-distribution properties of the supplied tokens.
They are not direct measures of language proficiency, writing quality, reader
response, or communicative effectiveness. A separate validated study design is
needed to make claims about those constructs.

## Multiple documents

Batch input must carry explicit, unique document identifiers.

```{r}
documents <- list(
  document_a = c("one", "two", "one"),
  document_b = c("alpha", "beta", "gamma", "delta")
)

lexdiv_metrics_batch(documents, metrics = c("ttr", "hdd"))
```

## Explicit method profiles

A profile plan records parameter variants before computation. The
`length_50_100` preset adds MSTTR and MATTR settings without duplicating MTLD or
turning descriptive token floors into alternative metric values.

```{r}
plan <- lexdiv_plan(presets = "length_50_100")
profile <- lexdiv_profile(tokens, plan)
head(profile)

screen <- lexdiv_screen(profile)
head(screen)
```

The frozen `below_quality_floor` field and requested screens are advisory
token-count evidence checks. Here, `quality` does not mean writing quality,
measurement validity, or reliability. The screens do not change values or
parameters or censor documents. Read them together with `status`,
`missing_reason`, `N`, `V`, and the method and parameter identities.

Use `lexdiv_methods()` and `lexdiv_presets()` to inspect the versioned
registries. The separately defined `expected_ttr_d` method is not included in
the version 0.1.0 public method set.

## Definition sensitivity outside the frozen core

Maas and sequential MTLD have same-name formula and aggregation variants. The
separate variant surface makes those choices comparable without changing the
canonical registry or silently replacing a method.

```{r}
variant_catalog <- lexdiv_variant_ids()
variant_catalog

variant_results <- lexdiv_variant_metrics(
  rep(tokens, 10),
  mtld_thresholds = c(0.72, 0.92)
)
variant_results[, c(
  "family", "method_id", "reference_label", "value", "status"
)]
```

The comparison-scope column is part of interpretation. A formula-level
TAALED-relevant row is not an official preprocessing or full-pipeline
compatibility claim.

## Auditing the formal contracts

The help pages explain ordinary use. Exact formulas, domains, normalization
rules, result schemas, and resource boundaries are installed as JSON contracts:

```{r}
contract_files <- c(
  "lexical-diversity-contract.json",
  "ldfreq-preprocessing-contract.json",
  "lexical-diversity-variant-contract.json",
  "lexical-level-profile-contract.json",
  "tubelex-frequency-profile-contract.json"
)
contract_paths <- system.file("spec", contract_files, package = "ldfreq")
stopifnot(all(nzchar(contract_paths)))
basename(contract_paths)
```

Use `jsonlite::read_json(contract_paths[[1]], simplifyVector = FALSE)` when an
R list representation is convenient. Reading the files does not run a metric,
download a resource, or expose caller file paths.

## Offline installed-package check

An installed smoke script exercises the public workflows without network access
or an external runtime:

```{r}
smoke_path <- system.file("examples", "offline-smoke.R", package = "ldfreq")
stopifnot(nzchar(smoke_path))
source(smoke_path, local = TRUE)
```
