---
title: "Getting started with ldfreq"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ldfreq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`ldfreq` begins at an assumption-transparent boundary: an ordered character
vector of tokens. The package does not guess how raw prose should be tokenized,
normalized, or lemmatized.

```{r}
library(ldfreq)

tokens <- c("the", "cat", "saw", "the", "other", "cat")
result <- lexdiv_metrics(tokens, metrics = c("ttr", "rttr", "yule_k"))
result
```

## Reading results

Each requested metric receives one row. The `status` and `missing_reason`
columns state whether a value is available; `requested_parameters` and
`effective_parameters` are list-columns. Contract and schema identifiers are
ordinary columns so that they survive row binding and tabular serialization.

Short documents do not cause the requested method to change. For example, an
MATTR window larger than the document yields a structured missing result rather
than a smaller implicit window.

```{r}
lexdiv_metrics(tokens, metrics = "mattr", window_length = 50L)
```

## Multiple documents

Batch input must carry explicit, unique document identifiers.

```{r}
documents <- list(
  document_a = c("one", "two", "one"),
  document_b = c("alpha", "beta", "gamma", "delta")
)

lexdiv_metrics_batch(documents, metrics = c("ttr", "hdd"))
```

## Explicit method profiles

A profile plan records parameter variants before computation. The
`length_50_100` preset adds MSTTR and MATTR settings without duplicating MTLD or
turning descriptive token floors into alternative metric values.

```{r}
plan <- lexdiv_plan(presets = "length_50_100")
profile <- lexdiv_profile(tokens, plan)
head(profile)

screen <- lexdiv_screen(profile)
head(screen)
```

Use `lexdiv_methods()` and `lexdiv_presets()` to inspect the frozen registries.
The `expected_ttr_d` candidate is deferred from v0.1 and intentionally absent
from the public method set.
