---
title: "Japanese lexical norms: from stimulus items to documented matches"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Japanese lexical norms: from stimulus items to documented matches}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ldfreq)
```

Suppose you are selecting Japanese words for a lexical-decision or reading
experiment. You need to inspect each item's normative ratings, how many people
rated it, and which items have no match. Keep those rows and experimental
conditions together before comparing groups. A document average alone cannot
show that a particular stimulus was unmatched.

This guide uses existing `lexdiv_norm_profile()` calculations and an optional
local-file example. No external norm values, participant data or dictionaries
are bundled. The executable illustration uses **invented values**; sections
using real data run only when you explicitly supply the separately obtained
files. The example script can be loaded in R without a shell or Python.

## Choose a construct and check the data terms

| Resource | Information | Data terms and scope |
|---|---|---|
| [WLSP-familiarity v4.0](https://github.com/masayu-a/WLSP-familiarity) | Familiarity/use estimates for knowing, reading, writing, speaking and listening | CC BY-NC-SA 3.0. Local-file candidate review below; record choices remain explicit |
| [WLSP-norms v1.0](https://github.com/masayu-a/WLSP-norms) | Subjective polysemousness estimates | CC BY-NC-SA 3.0. [Reviewed record lookup and KWIC connection](japanese-polysemy.html); values are not sense counts |
| [AWD-J core](https://sociocom.naist.jp/awd-j/) | Human abstractness ratings for 15,220 words | CC BY 4.0. Scale-direction descriptions conflict; no automatic transformation here. EX contains model predictions |
| [Japanese AoA](https://osf.io/fawmq/) | Retrospective acquisition ratings for 5,736 words | Data: CC BY-NC 4.0, independently of the article's CC BY license. Local-file example below |
| [JALEX](https://osf.io/qr2sg/) | Lexical-decision response times and accuracy | The data's `license.txt` specifies CC BY-NC 4.0. Current data version/exclusion rules need separate review |
| [Japanese BOI](https://osf.io/gkjmc/) | Body-object interaction ratings | OSF project: CC BY 4.0. Local-file example below; publication/file scope caveat applies |

These conditions were checked on 2026-10-05. Free access does not by itself
permit redistribution. Follow attribution and other applicable conditions for
the actual data, including joined tables or extracts; this package's MIT code
license does not relicense imported data. An OSF project's license does not
automatically clear third-party columns in a merged table.

Familiarity, frequency, acquisition age, abstractness, BOI and reaction time
measure different things. Do not average them into an undocumented difficulty
score. Adult native-speaker ratings are not an individual Japanese learner's
acquisition history, known vocabulary, or ability to use a word appropriately.

## Keep a row for every stimulus

Use `item_id` as the study's unique identifier. A word can occur in several
items or conditions: the lookup form need not be unique. `term` is the exact
spelling to look up. Keep the displayed stimulus, reading, sense and any mapping
decisions in additional columns if they differ from the lookup form.

```{r items}
items <- data.frame(
  item_id = c("a1", "a2", "b1", "b2", "b3"),
  condition = c("A", "A", "B", "B", "B"),
  term = c("学校", "研究", "学校", "辞書", "未登録の例語")
)
```

For a CSV, use `read.csv("stimuli.csv", fileEncoding = "UTF-8",
colClasses = "character")` when IDs must retain leading zeros. Convert numeric
design variables separately. The real-data helper below rejects duplicate or
missing IDs, missing lookup forms, and output-column collisions. It preserves
input order and extra columns; it does not tokenize, normalize or resolve senses.

## Inspect matches before interpreting averages

The following table is **authored demonstration data, not Japanese norms**.
It includes a matched word with a missing annotation, as well as a query word
absent from the table. Both produce `NA`, with different reasons.

```{r offline-profile}
toy <- data.frame(term = c("学校", "研究", "辞書"),
                  rating = c(2, 5, NA_real_))
specs <- data.frame(measure_id = "illustration", value_column = "rating",
  construct_id = "authored_illustration", value_unit = "invented_rating_points",
  direction = "descriptive", language = "Japanese", variety = "unspecified",
  population_id = "authored_not_participant_data", collection_year = "not_observed",
  valid_min = 1, valid_max = 7)
resource <- list(resource_id = "authored_ja_norms", resource_version = "1",
  creator = "Example author", source_reference = "Invented values in this guide",
  data_license = "MIT: authored illustration", transformation_id = "none",
  lookup_unit = "authored_spelling", resource_key_normalization_id = "identity-v1")
example <- lexdiv_norm_profile(items$term, toy, "term", specs, resource)
item_table <- cbind(items, example$lookup[, c("value", "lookup_status", "value_status")])
item_table
example$summary[, c("weighting", "estimate", "resource_coverage",
                     "value_coverage", "annotation_coverage")]
stopifnot(identical(item_table$item_id, items$item_id),
          isTRUE(all.equal(example$summary$estimate, c(3, 3.5))))
```

For item-row (`token`) weighting, 4/5 rows match the resource, 3/5 have a value,
and 3/4 matched rows have an annotation. The mean is `(2 + 5 + 2) / 3 = 3`.
For distinct-spelling (`type`) weighting, the mean is `(2 + 5) / 2 = 3.5`.
Here “token” means each supplied item row, not an independently observed corpus
token. Neither mean should replace item-level information in the experiment.

Inspect missingness by condition rather than silently dropping unmatched rows:

```{r condition-coverage}
condition_coverage <- do.call(rbind, lapply(split(item_table, item_table$condition),
  function(x) data.frame(condition = x$condition[1], items = nrow(x),
    with_value = sum(x$value_status == "observed"),
    without_value = sum(x$value_status != "observed"),
    value_coverage = mean(x$value_status == "observed"))))
rownames(condition_coverage) <- NULL
condition_coverage
```

In this illustration, condition A has values for both items, whereas B has
values for only one of three. A comparison of available-value means would use
different subsets. Review the unmatched forms and selection rules before
interpreting such a comparison. Do not equate missing values with zero, rarity,
low familiarity, or lack of participant knowledge.

## Use a separately obtained AoA or BOI file

Download `data_AoA.csv` from the [AoA project](https://osf.io/fawmq/) or
`data_boi.csv` from the [BOI project](https://osf.io/gkjmc/) after checking the
terms for your intended use. Each can be used independently. Use the aggregate
file, not raw participant responses or a merged collection of third-party norms.

The helper checks the inspected file's SHA-256; a changed file fails rather
than silently acquiring the old metadata. Both inspected files contain 5,736
unique `Word` keys. A future version requires reviewing the source and updating
the example, not disabling the hash check without examining the change.

```{r real-data, eval=FALSE}
source(system.file("examples", "japanese-norms.R", package = "ldfreq"))
aoa <- profile_japanese_norm_items(items, "data_AoA.csv", resource = "aoa")
aoa$items                  # Original IDs/conditions plus source IDs, n, SD and values
aoa$profile$coverage        # Resource, value and annotation coverage
aoa$source                 # Citation, data license, input hash and limitation

# Independent alternative: obtaining AoA is not required to use BOI.
boi <- profile_japanese_norm_items(items, "data_boi.csv", resource = "boi")
```

AoA `norm_mean` is the mean of **seven age-category ratings**, not years.
Categories run from 1 (0–1 years) to 7 (12+ years); the instructions also assign
unknown acquisition age to 7. The aggregate cannot separate those responses.
See [Mochizuki & Ota (2025)](https://doi.org/10.3389/flang.2025.1605224).

BOI `norm_mean` is the original mean rating. The
[2026 publication](https://doi.org/10.3758/s13428-025-02939-1) and OSF project
title say 5,637 words, whereas the inspected CSV and analysis script contain
5,736. The content of the [correction](https://doi.org/10.3758/s13428-026-03000-5)
has not been verified here. Treat this as a reproducible lookup into the pinned
file, not replication of the corrected publication's analysis sample.

`norm_n` and `norm_sd` describe the source ratings, not the sample size or
uncertainty of your own experiment. A normative SD is not a standard error for
a condition difference. Preserve source sampling, repeated participants/items,
and your experiment's task when choosing a statistical model.

## Join resources and save the record

Join by the **study item ID**, after separately matching each resource's
spelling. Do not assume that two data providers' numerical source IDs are equal.
Do not use an inner join that discards unmatched stimuli.

```{r save-real-data, eval=FALSE}
index <- match(aoa$items$item_id, boi$items$item_id)
stopifnot(!anyNA(index), !anyDuplicated(boi$items$item_id))
combined <- aoa$items
combined$boi_mean <- boi$items$norm_mean[index]
combined$boi_value_status <- boi$items$norm_value_status[index]
# Source details for both resources stay in the RDS, alongside the flat table.
saveRDS(list(aoa = aoa, boi = boi, combined = combined,
             session = sessionInfo()), "japanese-norm-analysis.rds")
write.csv(combined, "japanese-norm-items.csv", row.names = FALSE,
          fileEncoding = "UTF-8")
restored <- readRDS("japanese-norm-analysis.rds")
stopifnot(identical(restored$combined, combined))
```

## Why spelling alone sometimes fails

WLSP includes multiple readings for spellings such as `人気` (`にんき` and
`ひとけ`). Even spelling plus reading is not necessarily unique in the inspected
file: `上手` includes repeated `じょうず` records with different estimates.
Inspect the original record IDs, headings, reading and classification before
defining a correspondence. The generic profile deliberately rejects duplicate
keys; it does not pick the first row or average distinct entries automatically.

Keep a separate, explicit correspondence table when a study item needs a
particular source record. Review each mapping in context. A morphological
analyzer's lemma is not evidence of a uniquely correct psycholinguistic norm
match. For running text, start with the [Japanese annotation workflow](japanese-annotations.html)
and retain source positions, forms, dictionary and segmentation decisions.

## Review WLSP candidates and record your choices

The bundled `wlsp-items.R` example reads a separately obtained
`bunruidb-fam.csv` from **WLSP-familiarity v4.0**, with its verified file hash.
It returns every exact-spelling candidate, retaining record ID, heading,
reading, classification, record type and the five published estimates for
knowing, writing, reading, speaking and listening. Numeric estimates are not
converted into probabilities, raw rating points or individual knowledge scores.
The data remain subject to CC BY-NC-SA 3.0; no WLSP ratings are bundled.

For occurrences in running text, the [KWIC review guide](ambiguity-review.html)
uses the same local-file reader through `wlsp_ambiguity_candidates()` to retain
record IDs, readings and classifications as contextual review candidates.
It also shows how to compare two reviewers while keeping unresolved cases.

The following are authored study items. The reading and context specify the
intended stimulus; the helper keeps them as metadata and does not use them to
automatically discard candidates or resolve word senses.

```{r wlsp-candidates, eval=FALSE}
source(system.file("examples", "wlsp-items.R", package = "ldfreq"))
stimuli <- data.frame(
  item_id = c("p1", "p2", "p3", "p4", "p5"),
  condition = c("A", "B", "A", "B", "B"),
  term = c("人気", "人気", "学校", "上手", "未登録の例語"),
  intended_reading = c("にんき", "ひとけ", "がっこう", "じょうず", NA),
  context = c("この商品は人気がある。", "夜の公園には人気がない。",
              "学校に通う。", "彼は料理が上手だ。", "作成した未照合例。")
)
path <- "bunruidb-fam.csv"
review <- review_wlsp_items(stimuli, path)
review$items[, c("item_id", "term", "wlsp_candidate_count", "wlsp_status")]
review$candidates[, c("item_id", "record_id", "heading", "reading",
                     "classification", "classification_label", "know")]
```

For this pinned file the candidate counts are 2, 2, 1, 7 and 0. None is selected
automatically, including the single candidate for `学校`. The five study rows
remain separate from the twelve candidate rows. Candidate multiplicity must not
increase an item's weight in subsequent analyses.

After inspecting the candidates, record one source record per chosen item and
a reason. This example selects the two intended readings of `人気` and the
reviewed `学校` entry. It leaves `上手` unresolved: reading alone does not make
its correspondence unique.

```{r wlsp-decisions, eval=FALSE}
chosen <- subset(review$candidates,
  (item_id == "p1" & reading == "にんき") |
  (item_id == "p2" & reading == "ひとけ") | item_id == "p3")
stopifnot(nrow(chosen) == 3L,
          identical(chosen$item_id, c("p1", "p2", "p3")))
decisions <- data.frame(
  item_id = chosen$item_id, record_id = chosen$record_id,
  reason = c("Reviewed heading/reading for the intended popularity stimulus",
             "Reviewed heading/reading for the intended human-presence stimulus",
             "Reviewed the single school entry against the stimulus")
)
resolved <- review_wlsp_items(stimuli, path, decisions)
resolved$items[, c("item_id", "wlsp_status", "wlsp_record_id",
                   "wlsp_reason", "wlsp_know")]
```

`selected` records an explicit correspondence, `unreviewed` means candidates
exist but none was chosen, and `unmatched` means no exact-spelling candidate
exists. Unselected values stay `NA`; so do missing values in a selected source
record. Inspect selection status **and** the value you intend to analyze.
The helper rejects unknown item/record IDs, multiple choices for one item,
blank reasons, choices whose source spelling differs from the item, and files
that differ from the inspected snapshot. It does not validate a reviewer's
linguistic judgment. Repeated spellings across different item IDs are allowed.

Do not read these decisions as validated sense-specific familiarity norms.
The upstream description says that estimation changed to surface forms from
version 3, yet the inspected table still contains different estimates for some
repeated heading/reading combinations. A recorded mapping makes the choice
auditable; it does not settle that methodological issue. Uncertain mappings can
remain unresolved or be compared in explicitly labeled sensitivity analyses.

Keep the denominator and reasons visible, including items without values:

```{r wlsp-save, eval=FALSE}
selection_coverage <- do.call(rbind, lapply(
  split(resolved$items, resolved$items$condition), function(x) data.frame(
    condition = x$condition[1], items = nrow(x),
    with_candidates = sum(x$wlsp_candidate_count > 0),
    selected = sum(x$wlsp_status == "selected"),
    with_know_value = sum(!is.na(x$wlsp_know)))))
selection_coverage
write.csv(decisions, "wlsp-decisions.csv", row.names = FALSE, fileEncoding = "UTF-8")
# Source record IDs such as 002375 have leading zeros: preserve them as text.
saved_decisions <- read.csv("wlsp-decisions.csv", fileEncoding = "UTF-8",
                           colClasses = "character")
restored <- review_wlsp_items(stimuli, path, saved_decisions)
stopifnot(identical(restored$items, resolved$items))
saveRDS(list(review = resolved, coverage = selection_coverage,
             session = sessionInfo()), "wlsp-item-review.rds")
```

Report which of the five estimates you used and why. The helper returns
item-level values without pooling across candidate records or experimental
conditions. In this example, condition A has two chosen items with values and
B has one of three: report the remaining unreviewed and unmatched items before
interpreting condition means. The example caps expansion at one million
candidate pairs; review smaller item sets if needed.

## Report what the analysis can answer

Record the resource/file version, population, scale, lookup form and any mapping
or exclusion rule. Report coverage for each condition and the subset used for
each comparison. Examine distributions as well as averages when selecting
stimuli; similar means alone do not establish balanced conditions.

This workflow documents normative features of the chosen items. It does not
validate a proficiency score, diagnose knowledge from an unmatched word, or
establish causal effects of familiarity or acquisition age on response time.
For corpus-level profiles, see [multiple-document norms](corpus-reference-profiles.html).
To combine these sources with Japanese TUBELEX, review item-level missingness
and inspect condition distributions, follow the [stimulus-selection workflow](japanese-stimuli.html).
