---
title: "Japanese annotations: from original text to lexical measures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Japanese annotations: from original text to lexical measures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ldfreq)
dic_dir <- Sys.getenv("LDFREQ_UNIDIC_LITE_1_0_8", unset = "")
run_gibasa <- nzchar(dic_dir) && requireNamespace("gibasa", quietly = TRUE)
```

For stimulus-item ratings and separately obtained AoA/BOI tables, see
[Japanese lexical norms](japanese-norms.html). That workflow retains item IDs
and norm-matching decisions independently of morphological segmentation.

Japanese text needs an explicit word-unit decision. For example, counting
`国際連合` as one item or `国際 / 連合` as two changes both vocabulary diversity
and the available bigrams. The default Unicode tokenizer in ldfreq does not
perform Japanese morphological analysis. Instead, use a Japanese analyzer and
bring its complete annotations into `lexdiv_import_annotations()`.

The importer checks the supplied surfaces against original text, derives source
positions, retains dictionary metadata and leaves missing annotations missing.
Use existing R packages for morphology and concordances, and ldfreq to keep
those input decisions visible in lexical analyses. This is an integration and
audit contribution, not a new Japanese segmentation algorithm or a validated
Japanese proficiency scale.

## An offline example

The package's token-table importer supports R >= 4.1. The optional
gibasa 1.1.3 recipe requires **R >= 4.2**, as well as the separately supplied
dictionary. On R 4.1, import annotations prepared elsewhere or use the offline
example below; gibasa is not required to install or use the core package.

These annotations are authored for illustration, not analyzer output or a
Japanese gold standard. The original examples in this guide are covered by the
package's MIT license; no corpus or dictionary is redistributed.

```{r authored}
segments <- data.frame(
  document_id = c("essay", "empty"), segment_id = c("s1", "s1"),
  text = c("国際連合で国際協力を学ぶ。", "")
)
annotations <- data.frame(
  document_id = "essay", segment_id = "s1", token_index = 1:8,
  surface = c("国際", "連合", "で", "国際", "協力", "を", "学ぶ", "。"),
  lemma = c("国際", "連合", "で", "国際", "協力", "を", "学ぶ", NA_character_)
)
imported <- lexdiv_import_annotations(annotations, segments, list(
  language = "ja", analyzer = "authored", analyzer_version = "1",
  dictionary = "none", dictionary_version = "not-applicable",
  unit = "authored-fine", normalization = "none"
))
imported$tokens
```

`start` and `end` are **1-based inclusive Unicode codepoint positions within
the segment's original text**. They are not whole-document positions, byte
offsets, UTF-16 offsets or positions after normalization. The empty document
remains in `imported$documents`, with a zero-token segment. A sentence or turn
is a caller-supplied segment; the importer does not infer these boundaries.

Import **before filtering**. Apart from whitespace, every original character
must be covered, including punctuation. A skipped repeated word, dropped
punctuation, or normalized replacement causes an error. A supplied `start` or
`end` column must match the derived coordinates. This verifies the alignment,
not the correctness of POS, lexical units, or the source's sentence boundaries.

```{r measures}
words <- imported$tokens[imported$tokens$surface != "。", ]
lexdiv_metrics(words$surface, metrics = "ttr")
phrases <- lexdiv_ngrams(words, "authored-ja-surface-no-punctuation-v1",
  term_col = "surface", documents = imported$documents$document_id)
phrases$documents
```

The seven retained tokens contain six distinct surface strings: TTR is 6/7.
There are six adjacent bigram and five trigram opportunities. These counts
verify the example's arithmetic; a seven-token text cannot establish the
reliability of a diversity measure. Choose length-appropriate measures and
report non-computable requests rather than changing their parameters silently.

## Use gibasa and a local UniDic dictionary in R

[gibasa](https://paithiov909.github.io/gibasa/) supplies the MeCab interface;
ldfreq does not reimplement it. Install gibasa separately with
`install.packages("gibasa")`. Its optional installation requirements are
separate from those of ldfreq. A UTF-8 compiled dictionary is also required.

This recipe is for **unidic-lite 1.0.8, based on UniDic 2.1.2**, with the
26-field binary schema. It is a concrete integration example, not a claim that
this older dictionary is best for every contemporary Japanese dataset.
[unidic-lite](https://github.com/polm/unidic-lite) modifies the original UniDic
release; record both identities. Obtain its versioned archive from the
[release files](https://pypi.org/project/unidic-lite/1.0.8/#files), verify the
published SHA-256, and extract it. R's `download.file()` and `untar()` can do
this; no Python runtime is needed to use the extracted dictionary in gibasa.
Keep the included dictionary license and source information with it. Set
`LDFREQ_UNIDIC_LITE_1_0_8` to the extracted `unidic_lite/dicdir` directory before
rendering this guide, or assign that path to `dic_dir` and run the following
chunks interactively. The package build runs the authored examples offline;
it does not download a dictionary.

Supply one sentence or turn per segment, with a separate roster for original
document IDs. The following strings are new examples, not sampled corpus text.
The full-width letters and an invented string expose normalization and
missing-feature decisions.

```{r japanese-source}
ja_segments <- data.frame(document_id = c("essay", "essay", "empty"),
  segment_id = c("s1", "s2", "s1"),
  text = c("国際連合で国際協力を学ぶ。", "ＡＩで学びました。 qzxv😀", ""))
```

gibasa needs a readable MeCab configuration file even when `sys_dic` is
explicit. This helper retains an existing `~/.mecabrc` (which gibasa's bundled
MeCab checks before `MECABRC`), or uses a temporary empty configuration when
that file is absent. It records the configuration hash, verifies that only the
requested system dictionary is active, and restores the prior environment on
exit. It does not edit the home configuration or request normalization or
automatic sentence splitting. Use a path without spaces for this gibasa recipe.

```{r gibasa-analysis, eval=run_gibasa}
analyze_unidic <- function(segments, dic_dir) {
  rc <- path.expand("~/.mecabrc")
  temporary_rc <- !file.exists(rc)
  if (temporary_rc) {
    rc <- tempfile("mecabrc-")
    writeLines("", rc)
  }
  previous <- Sys.getenv("MECABRC", unset = NA_character_)
  on.exit({
    if (is.na(previous)) Sys.unsetenv("MECABRC") else Sys.setenv(MECABRC = previous)
    if (temporary_rc) unlink(rc)
  })
  Sys.setenv(MECABRC = rc)
  info <- gibasa::dictionary_info(sys_dic = dic_dir, user_dic = "")
  stopifnot(nrow(info) == 1L, info$type == 0L,
    tolower(gsub("-", "", info$charset)) == "utf8",
    normalizePath(info$file_path) == normalizePath(file.path(dic_dir, "sys.dic")))
  texts <- setNames(segments$text, paste0("input_", seq_len(nrow(segments))))
  raw <- gibasa::tokenize(texts, sys_dic = dic_dir, user_dic = "",
    split = FALSE, partial = FALSE)
  expanded <- as.data.frame(gibasa::prettify(raw,
    into = gibasa::get_dict_features("unidic26")))
  source_row <- match(as.character(raw$doc_id), names(texts))
  tokens <- data.frame(document_id = segments$document_id[source_row],
    segment_id = segments$segment_id[source_row], token_index = raw$token_id,
    surface = raw$token, raw_feature = raw$feature)
  features <- setdiff(names(expanded), c("doc_id", "sentence_id", "token_id", "token"))
  list(tokens = cbind(tokens, expanded[features]),
    configuration_sha256 = digest::digest(file = rc, algo = "sha256"))
}
analysis <- analyze_unidic(ja_segments, dic_dir)
ja_annotations <- analysis$tokens
ja_annotations[c("surface", "orthBase", "lemma", "lForm", "POS1")]
```

Use the feature schema for the **actual dictionary**, not the default IPA
schema of `prettify()`. The raw feature string is retained to make the expansion
auditable, including quoted commas and missing fields. gibasa does not expose
MeCab's unknown-node flag in these rows: a missing lemma is observable, but
is not an independently verified unknown-word classification. Do not fabricate
a definitive unknown flag from it.

```{r gibasa-import, eval=run_gibasa}
dictionary_files <- c("sys.dic", "unk.dic", "char.bin", "matrix.bin", "dicrc")
dictionary_hashes <- setNames(lapply(file.path(dic_dir, dictionary_files),
  function(path) digest::digest(file = path, algo = "sha256")),
  paste0(dictionary_files, "_sha256"))
ja <- lexdiv_import_annotations(ja_annotations, ja_segments, c(list(
  language = "ja", analyzer = "gibasa/MeCab",
  analyzer_version = as.character(utils::packageVersion("gibasa")),
  dictionary = "unidic-lite", dictionary_version = "1.0.8 (UniDic 2.1.2)",
  unit = "UniDic-short-unit-analysis", normalization = "none",
  settings = "unidic26; split=FALSE; partial=FALSE; no user dictionary",
  configuration_sha256 = analysis$configuration_sha256,
  dictionary_source = "https://pypi.org/project/unidic-lite/1.0.8/"
), dictionary_hashes))
ja$tokens[c("document_id", "segment_id", "token_index", "surface",
  "orthBase", "lemma", "start", "end")]
```

Metadata and hashes record the declared dictionary snapshot. They do not
independently identify or license an arbitrary dictionary path. If you change
dictionary release, schema, user dictionary or settings, update the metadata
and verify the feature mapping again. Retain user-dictionary hashes as well
when used. RDS saves the imported tables without requiring gibasa or the
dictionary when reloading them for analysis.

## Choose lexical forms and preserve exclusions

[UniDic's terminology](https://clrd.ninjal.ac.jp/unidic/glossary.html)
distinguishes the observed surface, `orthBase` (orthographic base) and `lemma`
(lexeme label). In this example, surface `学び` has base and lemma `学ぶ`;
`まし` maps to `ます`. A lemma label alone need not identify a unique lexeme:
homographs may require reading, POS or dictionary IDs. Do not use a first-match
join to resolve collisions with a reference table.

This explicit example excludes auxiliary symbols and whitespace tokens,
retains particles and auxiliary verbs, and requires an observed `orthBase`.
Inspect and save the exclusions rather than substituting surface forms for
missing bases. Change this selection if the research question calls for a
different lexical population, and change the preprocessing label accordingly.

```{r gibasa-selection, eval=run_gibasa}
t <- ja$tokens
punctuation <- !is.na(t$POS1) & t$POS1 %in% c("補助記号", "空白")
has_whitespace <- stringi::stri_detect_regex(t$surface, "\\p{White_Space}")
missing_base <- is.na(t$orthBase) | !nzchar(t$orthBase)
keep <- !punctuation & !has_whitespace & !missing_base
selection <- data.frame(document_id = t$document_id, segment_id = t$segment_id,
  token_index = t$token_index, punctuation, has_whitespace, missing_base, retained = keep)
selected <- t[keep, ]
selected$term <- selected$orthBase
by_document <- setNames(lapply(ja$documents$document_id,
  function(id) selected$term[selected$document_id == id]), ja$documents$document_id)
results <- lexdiv_metrics_batch(by_document, metrics = "ttr")
grams <- lexdiv_ngrams(selected, "ja-unidic-lite-1.0.8-orthBase-selection-v1",
  documents = ja$documents$document_id)
results
grams$documents
```

The diversity calculation describes the retained sequence. N-grams also
respect segment boundaries and the **original** `token_index`: excluding a
particle, punctuation or an unavailable base leaves a gap, not a new adjacent
pair. Do not renumber `selected`. With quanteda installed,
`lexdiv_as_quanteda(selected, ja$segments, term_col = "term")` preserves those
gaps for concordances and other quanteda operations in a UTF-8 R session.

```{r save-analysis, eval=FALSE}
saveRDS(list(imported = ja, selection = selection, metrics = results, ngrams = grams),
  "japanese-analysis.rds")
```

## Frequency resources and interpretation

The bundled `tubelex_profile()` and `nj8_profile()` resources are **English**;
do not treat unmatched Japanese tokens as rare English words. Japanese
[TUBELEX](https://github.com/naist-nlp/tubelex) has different surface/base/lemma
and dictionary-version variants. Sharing MeCab does not by itself establish
compatible keys: normalization, dictionary version, POS definition and
selection also matter. The [Japanese stimulus guide](japanese-stimuli.html)
connects reviewed `orthBase` keys to one separately acquired, pinned Japanese
TUBELEX table. It records denominators and key transformations without claiming
to reproduce upstream tokenization. Its unigram tables cannot supply n-gram
co-occurrences.

For an external Japanese norm table, prepare and check unique keys and use
`lexdiv_norm_profile()` with the table's actual metadata and missingness. For
n-gram frequencies, supply permitted local reference counts through the
[n-gram workflow](ngram-profiles.html). Neither approach requires redistributing
the underlying corpus. Access to a source does not itself establish a right
to redistribute it; check the terms of the actual artifact used.

Analyzing Japanese learner texts, native-speaker texts, or instructional
materials is possible with this input path. Software alignment tests do not
validate learner-error tokenization, lexical sophistication, or equivalence
between English and Japanese scores. Evaluate those questions against suitable
annotated data and external outcomes. Comparing segmentations with splits and
merges remains a separate development task: `lexdiv_compare_annotations()`
still requires identical token rows and cannot perform that comparison.
