---
title: "Japanese annotations: from original text to lexical measures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Japanese annotations: from original text to lexical measures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ldfreq)
dic_dir <- Sys.getenv("LDFREQ_UNIDIC_LITE_1_0_8", unset = "")
run_gibasa <- nzchar(dic_dir) && requireNamespace("gibasa", quietly = TRUE)
```

For stimulus-item ratings and separately obtained AoA/BOI tables, see
[Japanese lexical norms](japanese-norms.html). That workflow retains item IDs
and norm-matching decisions independently of morphological segmentation.

Japanese text needs an explicit word-unit decision. For example, counting
`国際連合` as one item or `国際 / 連合` as two changes both vocabulary diversity
and the available bigrams. The default Unicode tokenizer in ldfreq does not
perform Japanese morphological analysis. Instead, use a Japanese analyzer and
bring its complete annotations into `lexdiv_import_annotations()`.

The importer checks the supplied surfaces against original text, derives source
positions, retains dictionary metadata and leaves missing annotations missing.
Use existing R packages for morphology and concordances, and ldfreq to keep
those input decisions visible in lexical analyses. This is an integration and
audit contribution, not a new Japanese segmentation algorithm or a validated
Japanese proficiency scale.

## An offline example

The package's token-table importer supports R >= 4.1. The optional
gibasa 1.1.3 recipe requires **R >= 4.2**, as well as the separately supplied
dictionary. On R 4.1, import annotations prepared elsewhere or use the offline
example below; gibasa is not required to install or use the core package.

These annotations are authored for illustration, not analyzer output or a
Japanese gold standard. The original examples in this guide are covered by the
package's MIT license; no corpus or dictionary is redistributed.

```{r authored}
segments <- data.frame(
  document_id = c("essay", "empty"), segment_id = c("s1", "s1"),
  text = c("国際連合で国際協力を学ぶ。", "")
)
annotations <- data.frame(
  document_id = "essay", segment_id = "s1", token_index = 1:8,
  surface = c("国際", "連合", "で", "国際", "協力", "を", "学ぶ", "。"),
  lemma = c("国際", "連合", "で", "国際", "協力", "を", "学ぶ", NA_character_)
)
imported <- lexdiv_import_annotations(annotations, segments, list(
  language = "ja", analyzer = "authored", analyzer_version = "1",
  dictionary = "none", dictionary_version = "not-applicable",
  unit = "authored-fine", normalization = "none"
))
imported$tokens
```

`start` and `end` are **1-based inclusive Unicode codepoint positions within
the segment's original text**. They are not whole-document positions, byte
offsets, UTF-16 offsets or positions after normalization. The empty document
remains in `imported$documents`, with a zero-token segment. A sentence or turn
is a caller-supplied segment; the importer does not infer these boundaries.

Import **before filtering**. Apart from whitespace, every original character
must be covered, including punctuation. A skipped repeated word, dropped
punctuation, or normalized replacement causes an error. A supplied `start` or
`end` column must match the derived coordinates. This verifies the alignment,
not the correctness of POS, lexical units, or the source's sentence boundaries.

Whitespace annotations may be partial. For example, an analyzer can emit CR
tokens while omitting LF from CRLF line endings. A whitespace-prefixed token
uses the earliest exact match across whitespace-only gaps; non-whitespace
characters cannot be skipped. Original line endings and positions are retained.

```{r measures}
words <- imported$tokens[imported$tokens$surface != "。", ]
lexdiv_metrics(words$surface, metrics = "ttr")
phrases <- lexdiv_ngrams(words, "authored-ja-surface-no-punctuation-v1",
  term_col = "surface", documents = imported$documents$document_id)
phrases$documents
```

The seven retained tokens contain six distinct surface strings: TTR is 6/7.
There are six adjacent bigram and five trigram opportunities. These counts
verify the example's arithmetic; a seven-token text cannot establish the
reliability of a diversity measure. Choose length-appropriate measures and
report non-computable requests rather than changing their parameters silently.

## Use gibasa and a local UniDic dictionary in R

[gibasa](https://paithiov909.github.io/gibasa/) supplies the MeCab interface;
ldfreq does not reimplement it. Install gibasa separately with
`install.packages("gibasa")`. Its optional installation requirements are
separate from those of ldfreq. A UTF-8 compiled dictionary is also required.

This recipe is for **unidic-lite 1.0.8, based on UniDic 2.1.2**, with the
26-field binary schema. It is a concrete integration example, not a claim that
this older dictionary is best for every contemporary Japanese dataset.
[unidic-lite](https://github.com/polm/unidic-lite) modifies the original UniDic
release; record both identities. Obtain its versioned archive from the
[release files](https://pypi.org/project/unidic-lite/1.0.8/#files), verify the
published SHA-256, and extract it. R's `download.file()` and `untar()` can do
this; no Python runtime is needed to use the extracted dictionary in gibasa.
Keep the included dictionary license and source information with it. Set
`LDFREQ_UNIDIC_LITE_1_0_8` to the extracted `unidic_lite/dicdir` directory before
rendering this guide, or assign that path to `dic_dir` and run the following
chunks interactively. The package build runs the authored examples offline;
it does not download a dictionary.

Supply one sentence or turn per segment, with a separate roster for original
document IDs. The following strings are new examples, not sampled corpus text.
The full-width letters and an invented string expose normalization and
missing-feature decisions.

```{r japanese-source}
ja_segments <- data.frame(document_id = c("essay", "essay", "empty"),
  segment_id = c("s1", "s2", "s1"),
  text = c("国際連合で国際協力を学ぶ。", "ＡＩで学びました。 qzxv😀", ""))
```

gibasa needs a readable MeCab configuration file even when `sys_dic` is
explicit. This helper retains an existing `~/.mecabrc` (which gibasa's bundled
MeCab checks before `MECABRC`), or uses a temporary empty configuration when
that file is absent. It records the configuration hash, verifies that only the
requested system dictionary is active, and restores the prior environment on
exit. It does not edit the home configuration or request normalization or
automatic sentence splitting. Use a path without spaces for this gibasa recipe.

```{r gibasa-analysis, eval=run_gibasa}
analyze_unidic <- function(segments, dic_dir) {
  rc <- path.expand("~/.mecabrc")
  temporary_rc <- !file.exists(rc)
  if (temporary_rc) {
    rc <- tempfile("mecabrc-")
    writeLines("", rc)
  }
  previous <- Sys.getenv("MECABRC", unset = NA_character_)
  on.exit({
    if (is.na(previous)) Sys.unsetenv("MECABRC") else Sys.setenv(MECABRC = previous)
    if (temporary_rc) unlink(rc)
  })
  Sys.setenv(MECABRC = rc)
  info <- gibasa::dictionary_info(sys_dic = dic_dir, user_dic = "")
  stopifnot(nrow(info) == 1L, info$type == 0L,
    tolower(gsub("-", "", info$charset)) == "utf8",
    normalizePath(info$file_path) == normalizePath(file.path(dic_dir, "sys.dic")))
  texts <- setNames(segments$text, paste0("input_", seq_len(nrow(segments))))
  raw <- gibasa::tokenize(texts, sys_dic = dic_dir, user_dic = "",
    split = FALSE, partial = FALSE)
  expanded <- as.data.frame(gibasa::prettify(raw,
    into = gibasa::get_dict_features("unidic26")))
  source_row <- match(as.character(raw$doc_id), names(texts))
  tokens <- data.frame(document_id = segments$document_id[source_row],
    segment_id = segments$segment_id[source_row], token_index = raw$token_id,
    surface = raw$token, raw_feature = raw$feature)
  features <- setdiff(names(expanded), c("doc_id", "sentence_id", "token_id", "token"))
  list(tokens = cbind(tokens, expanded[features]),
    configuration_sha256 = digest::digest(file = rc, algo = "sha256"))
}
analysis <- analyze_unidic(ja_segments, dic_dir)
ja_annotations <- analysis$tokens
ja_annotations[c("surface", "orthBase", "lemma", "lForm", "POS1")]
```

Use the feature schema for the **actual dictionary**, not the default IPA
schema of `prettify()`. The raw feature string is retained to make the expansion
auditable, including quoted commas and missing fields. gibasa does not expose
MeCab's unknown-node flag in these rows: a missing lemma is observable, but
is not an independently verified unknown-word classification. Do not fabricate
a definitive unknown flag from it.

```{r gibasa-import, eval=run_gibasa}
dictionary_files <- c("sys.dic", "unk.dic", "char.bin", "matrix.bin", "dicrc")
dictionary_hashes <- setNames(lapply(file.path(dic_dir, dictionary_files),
  function(path) digest::digest(file = path, algo = "sha256")),
  paste0(dictionary_files, "_sha256"))
ja <- lexdiv_import_annotations(ja_annotations, ja_segments, c(list(
  language = "ja", analyzer = "gibasa/MeCab",
  analyzer_version = as.character(utils::packageVersion("gibasa")),
  dictionary = "unidic-lite", dictionary_version = "1.0.8 (UniDic 2.1.2)",
  unit = "UniDic-short-unit-analysis", normalization = "none",
  settings = "unidic26; split=FALSE; partial=FALSE; no user dictionary",
  configuration_sha256 = analysis$configuration_sha256,
  dictionary_source = "https://pypi.org/project/unidic-lite/1.0.8/"
), dictionary_hashes))
ja$tokens[c("document_id", "segment_id", "token_index", "surface",
  "orthBase", "lemma", "start", "end")]
```

Metadata and hashes record the declared dictionary snapshot. They do not
independently identify or license an arbitrary dictionary path. If you change
dictionary release, schema, user dictionary or settings, update the metadata
and verify the feature mapping again. Retain user-dictionary hashes as well
when used. RDS saves the imported tables without requiring gibasa or the
dictionary when reloading them for analysis.

## Choose lexical forms and preserve exclusions

[UniDic's terminology](https://clrd.ninjal.ac.jp/unidic/glossary.html)
distinguishes the observed surface, `orthBase` (orthographic base) and `lemma`
(lexeme label). In this example, surface `学び` has base and lemma `学ぶ`;
`まし` maps to `ます`. A lemma label alone need not identify a unique lexeme:
homographs may require reading, POS or dictionary IDs. Do not use a first-match
join to resolve collisions with a reference table.

This explicit example excludes auxiliary symbols and whitespace tokens,
retains particles and auxiliary verbs, and requires an observed `orthBase`.
Inspect and save the exclusions rather than substituting surface forms for
missing bases. Change this selection if the research question calls for a
different lexical population, and change the preprocessing label accordingly.

```{r gibasa-selection, eval=run_gibasa}
t <- ja$tokens
punctuation <- !is.na(t$POS1) & t$POS1 %in% c("補助記号", "空白")
has_whitespace <- stringi::stri_detect_regex(t$surface, "\\p{White_Space}")
missing_base <- is.na(t$orthBase) | !nzchar(t$orthBase)
keep <- !punctuation & !has_whitespace & !missing_base
selection <- data.frame(document_id = t$document_id, segment_id = t$segment_id,
  token_index = t$token_index, punctuation, has_whitespace, missing_base, retained = keep)
selected <- t[keep, ]
selected$term <- selected$orthBase
by_document <- setNames(lapply(ja$documents$document_id,
  function(id) selected$term[selected$document_id == id]), ja$documents$document_id)
results <- lexdiv_metrics_batch(by_document, metrics = "ttr")
grams <- lexdiv_ngrams(selected, "ja-unidic-lite-1.0.8-orthBase-selection-v1",
  documents = ja$documents$document_id)
results
grams$documents
```

The diversity calculation describes the retained sequence. N-grams also
respect segment boundaries and the **original** `token_index`: excluding a
particle, punctuation or an unavailable base leaves a gap, not a new adjacent
pair. Do not renumber `selected`. With quanteda installed,
`lexdiv_as_quanteda(selected, ja$segments, term_col = "term")` preserves those
gaps for concordances and other quanteda operations in a UTF-8 R session.

```{r save-analysis, eval=FALSE}
saveRDS(list(imported = ja, selection = selection, metrics = results, ngrams = grams),
  "japanese-analysis.rds")
```

## Kana spellings and contextual lexical identity

Kana spelling is an observation, not evidence that a writer cannot produce a
kanji. Task instructions, input method and stylistic choices can also matter.
Keep two outcomes separate: what was written, and which lexical item a reviewed
occurrence represents. Inferring kanji knowledge requires additional evidence,
such as a production task with known opportunities and permitted writing aids.

[UniDic's hierarchy](https://clrd.ninjal.ac.jp/unidic/glossary.html) retains
orthographic variants below lexical entries. `orthBase` alone does **not** merge
all kana/kanji spellings; choosing `lemma` also pools inflectional variation.
For an orthography-only comparison, hold lexical identity and grammatical form
constant. A reading alone cannot distinguish homophones. An analyzer's selected
entry is a candidate annotation, not independent evidence that the writer meant
that entry. Segmentation errors require boundary review before lexical review.

The following example reuses the existing KWIC decision API. All text,
boundaries and candidates are authored; it requires optional **quanteda** in a
UTF-8 R session, but no downloaded corpus, dictionary, Python or model. The IDs identify illustrative
lexical items, not word families or a validated inventory of word senses.

```{r kana-review, eval=requireNamespace("quanteda", quietly=TRUE) && isTRUE(l10n_info()[["UTF-8"]])}
source(system.file("examples", "japanese-orthography.R", package = "ldfreq"))
spelling <- japanese_orthography_example
spelling$occurrences[, c("document_id", "segment_id", "pre", "keyword", "post",
  "status", "lexical_id", "lexical_label", "reason")]
spelling$counts
```

Only the declared **target occurrences** enter these counts; particles, verbs
and punctuation still remain in the source annotations and KWIC context.

| Authored target set | Observed spellings | Reviewed lexical identities |
|---|---|---|
| `りんご`, `リンゴ`, `林檎` | 3 types in 3 occurrences | 1 type in the same 3 occurrences |
| `はし` in crossing and eating contexts | 1 type in 2 selected occurrences | 2 types in the same 2 occurrences |
| A third `はし` in a seeing context | Its surface is known | Identity remains unresolved |
| Invented `ぷにょ語` | Its surface is known | No candidate supplied; no identity fallback |

The three HASHI occurrences have selection coverage 2/3. Their full-target
`lexical_V` remains `NA`; the two resolved occurrences support a conditional
comparison, with `common_surface_V = 1` and `common_lexical_V = 2`. Do not compare
a complete surface count with an incomplete lexical count as though their
populations matched. Even a single supplied candidate stays unreviewed until
explicitly selected. An absent candidate is not proof of a misspelling.

Original `surface`, `start` and `end` never change. Selected IDs and labels are
separate columns; the example does not rewrite kana into kanji. Saving the whole
object preserves the source, candidate inventory, decisions and reviewer reasons:

```{r save-kana-review, eval=FALSE}
saveRDS(spelling, "japanese-orthography-review.rds")
restored <- readRDS("japanese-orthography-review.rds")
```

### Review boundaries before lexical identity

The same example file also supplies `japanese_boundary_example`. In the authored
sentence `はしではしをつかいます。`, the intended meaning is using chopsticks on
a bridge. Compare two complete annotations of **the same original text**:

- Split alternative: `はし / で / は / し / を / つかい / ます / 。`
- Reviewed alternative: `はし / で / はし / を / つかい / ます / 。`

A local probe with gibasa 1.1.3 and unidic-lite 1.0.8 (UniDic 2.1.2) produced
the split alternative. This is a concrete integration case, not an estimate of
analyzer accuracy on learner writing. The offline example supplies both
annotations explicitly and requires neither that analyzer nor its dictionary.

```{r kana-boundaries, eval=requireNamespace("quanteda", quietly=TRUE) && isTRUE(l10n_info()[["UTF-8"]])}
boundary <- japanese_boundary_example
subset(boundary$alignment$groups, relation != "exact",
  select = c(start, end, pre, keyword, post, predicted_n, reference_n))
boundary$boundary_decisions
boundary$scores[, c("condition", "document_id", "N", "V", "value", "status")]
boundary$review$occurrences[, c("start", "end", "keyword", "candidate_id", "reason")]
```

The [alignment workflow](annotation-alignment.html) links original characters
4--5 to the two split tokens and the one reviewed token. The decision and its
reason are saved alongside both complete annotation objects. Import the reviewed
annotations afresh; editing an imported object's token rows in place invalidates
its identity. No lemma or POS is inherited automatically from a fragment.

The surface counts exclude only the full stop and retain all other tokens:
the split alternative has N = 7 and V = 7; the reviewed alternative has N = 6
and V = 5. These illustrative TTR values describe the supplied sequences, not
proficiency. Restricting the comparison to exact-span matches would discard the
changed tokens and fail to measure the effect under study.

Boundary selection does not decide whether either occurrence means bridge,
chopsticks or edge. Regenerate KWIC review on the new complete annotations and
record lexical decisions separately. The old split exposes one whole-token
HASHI, while the reviewed annotation exposes two. Decisions from a previous
segmentation have a different review ID and cannot be silently reused.

This example accepts an analyst's complete alternative segmentation; it does
not automatically repair boundaries or select the intended meaning. Keep the
whole `boundary` object when saving. Do not close gaps left by excluded or
unresolved tokens when forming n-grams or moving-window measures.

[JASWRIC](https://language.sakura.ne.jp/jaswric/) illustrates the practical issue:
it supplies original transcriptions and edited texts, including script changes,
for Japanese L1 student writing. Its supplied morphological analysis is based
on the edited version. Such paired materials can support a later sensitivity
study, but the authored example above is not a corpus-based accuracy evaluation.
Script proportions are descriptive; a kanji-production score additionally needs
a justified set of eligible words and acceptable spellings.

## Review inflected forms spanning several tokens

A morphological candidate can cover several analyzer tokens. For example,
`食べられる` can be represented by `食べ / られる`, while its potential, passive
and honorific analyses remain different candidates. A span identifies a source
occurrence; it does **not** automatically redefine that occurrence as one word
in a diversity calculation. Inflectional analyses also differ from lexical
senses and pedagogical word-family membership.

The following explicitly sourced recipe works offline with authored text,
annotations and six authored format-demonstration rows. It uses the existing
annotation importer and stringi's exact search; no new analyzer or exported
API is introduced. The earlier `lexdiv_ambiguity_review()` workflow continues
to handle single-token lexical candidates.

```{r morphology-spans-authored}
source(system.file("examples", "junimorph-spans.R", package = "ldfreq"))
source(system.file("examples", "junimorph-spans-demo.R", package = "ldfreq"))
span_demo <- junimorph_spans_example
span_demo$reviewed$occurrences[, c("pre", "keyword", "post", "start", "end",
  "token_from", "token_to", "candidate_count", "status")]
span_demo$counts
```

This example selects two occurrences, leaves a repeated `食べられる` unresolved,
keeps `開ける` unreviewed, and identifies an invented form with no supplied
candidate. A requested but absent form and an empty document are retained in
the term/document summaries. On the **same two selected spans**, two forms map
to one lemma label. The complete five-target lemma total is unavailable. All
26 original annotation tokens remain unchanged before and after the review.
These are illustrative decisions, not an accuracy assessment.

Unlike the earlier quanteda token-window example, this span recipe's `window`
is a number of **original Unicode codepoints** on either side. KWIC includes
original spaces and line endings. Matches cannot cross a supplied segment.
Exact substring matches whose endpoints fall inside analyzer tokens remain
visible as `boundary_mismatch`; lexical decisions on them are rejected until
the analyst supplies and imports a complete reviewed segmentation.

All overlapping matches remain present. `overlap_n` and
`selected_overlap_n` expose overlapping source ranges; their counts cannot be
summed as a count of independent words. The recipe never applies longest-match
selection, merges token rows, removes exclusions or automatically selects a
single candidate. Its document summaries count diagnostic spans, not word tokens.

### Connect a locally acquired J-UniMorph table

[J-UniMorph](https://github.com/cl-tohoku/J-UniMorph) supplies a three-column
`jpn` table (lemma, form, feature bundle) under CC BY 4.0. Its filtered table
and the pre-filtering hit-count file are different inputs. Obtain `jpn`
separately; the package includes only the reader and authored examples.

```{r junimorph-local-input, eval=FALSE}
reference <- read_junimorph("local-resources/jpn", list(
  resource_id = "J-UniMorph",
  resource_version = "jpn git blob 6b001fa9d33139bd9b85ffe70f74c1143ae9df24",
  source_reference = "https://github.com/cl-tohoku/J-UniMorph",
  data_license = "CC BY 4.0"
))
# Use the actual revision of your file; its SHA-256 is recorded automatically.
# ja is the complete annotation import above, before any filtering.
span_review <- review_morphology_spans(ja, reference)
span_review$documents
span_review$candidates[span_review$candidates$form == "食べられる", ]

# After inspecting an occurrence and the corresponding candidate table:
decision <- span_review$occurrences[chosen_row, c("review_id", "occurrence_id")]
decision$status <- "selected"             # or "unresolved"
decision$candidate_id <- chosen_candidate_id # NA_character_ for unresolved
decision$reviewer <- "analyst-id"
decision$reason <- "Record the contextual evidence or reason for withholding judgment"
reviewed <- review_morphology_spans(ja, reference, decisions = decision)
saveRDS(reviewed, "morphology-span-review.rds")
restored <- readRDS("morphology-span-review.rds")
replayed <- review_morphology_spans(restored$source, restored$reference,
  targets = restored$terms$form, decisions = restored$decisions,
  window = restored$provenance$window_codepoints)
stopifnot(identical(restored, replayed))
```

Source row IDs, original feature strings, duplicate records and file identity
are preserved. Changing the source annotations, reference table or target set
invalidates old decisions. Unreviewed and unresolved occurrences are distinct;
even one listed candidate requires an explicit selection. The workflow accepts
recorded decisions without claiming to verify their linguistic correctness.

The inspected snapshot has 12,687 rows, 107 lemma labels and 10,848 forms. It is
a limited inflection reference, not a comprehensive Japanese vocabulary list.
Matching is exact: kana/kanji variation, spelling errors and missing forms are
not silently normalized. A missing match is not an unknown-word diagnosis.
The example scans one form at a time and is intended for bounded local analyses.

In the 28-essay diagnostic subset described below, the full table produced
1,309 exact spans: 587 agreed with both supplied token boundaries, including
409 multi-token spans; 722 had a boundary mismatch. Of the aligned spans, 98
had multiple candidate rows. All real-corpus decisions remain unreviewed;
these counts measure matching behavior, not error rates or contextual accuracy.
Nested forms can describe overlapping text and shorter portions of an
inflectional sequence. Inspect the whole context before interpreting a hit.

## Read locally acquired learner essays

The [NINJAL Essay Database](https://mmsrv.ninjal.ac.jp/essay/essay_05.html)
distributes a text ZIP and separate essay/writer metadata workbooks. Obtain
these files directly from the provider under its **CC BY-NC-ND 4.0** terms.
They are not included in ldfreq. The following explicitly sourced reader needs
only the local ZIP and metadata tables; optional `readxl` reads the workbooks.

```{r ninjal-local-input, eval=FALSE}
source(system.file("examples", "ninjal-essays.R", package = "ldfreq"))
data_dir <- "ninjal-essay-data"
essays <- as.data.frame(readxl::read_excel(
  file.path(data_dir, "sakubun.xlsx"), col_types = "text"))
writers <- as.data.frame(readxl::read_excel(
  file.path(data_dir, "shipitsusha.xlsx"), col_types = "text"))

# Start with a small, explicitly selected subset of local files.
files <- head(essays[["日本語作文txt"]], 5)
corpus <- read_ninjal_essays(file.path(data_dir, "sakubun_txt.zip"),
  essays, writers, files = files, encoding = "UTF-8")
corpus$documents[, c("document_id", "writer_id", "task", "l1_reported",
  "writer_metadata_status", "source_encoding", "utf8_bom_removed")]

# With the optional analyzer configured above:
analysis <- analyze_unidic(corpus$segments, dic_dir)
# Import these complete annotations with the actual analyzer/dictionary
# metadata as above, before selecting lexical forms or excluding punctuation.
saveRDS(corpus, "ninjal-essays-local.rds")
```

The UTF-8 default is a declaration, not an encoding guess. For a mixed archive,
supply a named vector with each selected member's inspected encoding:

```{r ninjal-local-encoding, eval=FALSE}
files <- essays[["日本語作文txt"]]
encodings <- setNames(rep("UTF-8", length(files)), files)
# A local list of member names verified as CP932, not a list of error words.
cp932_files <- readLines(file.path(data_dir, "verified-cp932-members.txt"), encoding = "UTF-8")
stopifnot(all(cp932_files %in% files))
encodings[cp932_files] <- "CP932"
corpus <- read_ninjal_essays(file.path(data_dir, "sakubun_txt.zip"),
  essays, writers, files = files, encoding = encodings)
```

Do not silence a decoding failure by replacing unreadable bytes or guessing a
different encoding for every failure. The reader rejects failed conversions
and failed byte round-trips. It records a removed leading UTF-8 BOM separately;
all subsequent characters and line endings remain unchanged. Imported character
positions refer to that decoded text, not ZIP bytes. Raw member hashes and the
archive hash retain the connection to the downloaded files.

IDs come from the official metadata mapping, not filename slicing. A missing
writer row stays `missing` with an unavailable L1; duplicate join keys stop the
reader instead of duplicating essays. Source L1 labels, including nonresponse
labels, are retained verbatim. Collection country is not inferred to be L1.

In a local input check of the distribution downloaded on 2026-10-06, the ZIP
and essay workbook contained **1,777** matching files/records: 1,764 UTF-8 and
13 inspected CP932 files; 179 had a leading UTF-8 BOM. One essay lacked a writer
metadata match. These are observations of those downloaded files, not the
overview page's historical total or guarantees about future editions.

A diagnostic subset of 28 essays from one task (four per seven reported L1
labels, including Japanese) produced 10,932 supplied annotation rows. All source
spans and non-whitespace text were checked. The selected lexical population had
9,554 tokens, of which 139 had no lemma annotation. Missing lemma is not an error
rate or a verified unknown-word classification. The input contains complete
distributed TXT, including titles and transcription notation. This verifies
input, alignment and counting, not segmentation accuracy, kanji knowledge or
group differences. Keep corpus text and individual review tables locally;
the earlier authored examples provide the distributable workflow.

## Frequency resources and interpretation

The bundled `tubelex_profile()` and `nj8_profile()` resources are **English**;
do not treat unmatched Japanese tokens as rare English words. Japanese
[TUBELEX](https://github.com/naist-nlp/tubelex) has different surface/base/lemma
and dictionary-version variants. Sharing MeCab does not by itself establish
compatible keys: normalization, dictionary version, POS definition and
selection also matter. The [Japanese stimulus guide](japanese-stimuli.html)
connects reviewed `orthBase` keys to one separately acquired, pinned Japanese
TUBELEX table. It records denominators and key transformations without claiming
to reproduce upstream tokenization. Its unigram tables cannot supply n-gram
co-occurrences.

Surface frequency for a kana spelling and frequency for a reviewed lexical
entry answer different questions. Do not silently substitute the frequency of
its kanji spelling or sum homophones merely because they share a reading.
Use a compatible lexical-unit table or report that the requested frequency is
unavailable; a lexical selection does not create sense-specific frequencies.

For an external Japanese norm table, prepare and check unique keys and use
`lexdiv_norm_profile()` with the table's actual metadata and missingness. For
n-gram frequencies, supply permitted local reference counts through the
[n-gram workflow](ngram-profiles.html). Neither approach requires redistributing
the underlying corpus. Access to a source does not itself establish a right
to redistribute it; check the terms of the actual artifact used.

Analyzing Japanese learner texts, native-speaker texts, or instructional
materials is possible with this input path. Software alignment tests do not
validate learner-error tokenization, lexical sophistication, or equivalence
between English and Japanese scores. Evaluate those questions against suitable
annotated data and external outcomes. Use `lexdiv_align_annotations()` for
source-span comparisons involving splits and merges; the label comparison in
`lexdiv_compare_annotations()` requires identical token rows. Span alignment
alone does not establish which segmentation is linguistically correct.
