---
title: "Japanese stimulus selection: frequency, familiarity, AoA and BOI"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Japanese stimulus selection: frequency, familiarity, AoA and BOI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ldfreq)
```

For a Japanese word-recognition experiment, you may want to vary BOI while
inspecting frequency, familiarity and acquisition ratings as other item
characteristics. This guide joins those sources by **study item ID**, retains
unresolved and unmatched items, and shows distributions and missingness within
each condition. It provides a review table, not automatic stimulus matching or
evidence that conditions are balanced. The small item set below illustrates
data handling; it is not a recommended experimental design.

The code uses existing `lexdiv_norm_profile()` calculations and three installed
example scripts. These scripts are explicitly sourced helpers, not exported
package functions. All operations after file acquisition run in R, without
Python. Source tables and derived norm values are obtained separately; no
external values are embedded in this guide. Each real-data chunk below requires
the specified local files and is not executed during package installation.

## Fix the frequency table and lexical unit

Use [TUBELEX (Nohejl et al., 2025)](https://aclanthology.org/2025.coling-main.641/),
Japanese `tubelex-ja-base-pos.tsv.xz`, at commit
`7cb5fb36add76b83a266d1967536e1a1d3faa513`. This is the **orthographic base**
(`orthBase`) table using unidic-lite 1.0.8 (UniDic 2.1.2), not the surface,
lemma or UniDic 3.1.0 table. The helper checks the exact compressed-file SHA-256:
`be1a66ac600d5e7efe6a301f63fba5321352742ae5dbaaa231529cdf2ab11710`.

The [upstream format](https://github.com/naist-nlp/tubelex/blob/7cb5fb36add76b83a266d1967536e1a1d3faa513/README.md)
provides counts and the number of videos/channels containing each key.
The final `[TOTAL]` row supplies denominators: 165,932,178 tokens, 100,660
videos and 30,550 channels. Counts include special tokens such as `<num>`;
the denominator is not recomputed after selecting study items or POS.
The helper returns `count / total_count * 1e6`, `videos / total_videos` and
`channels / total_channels`, without smoothing. These last two are proportions,
unlike the log scores of the bundled **English** `tubelex_profile()`.

Acquire the frequency file explicitly, once:

```{r acquire-frequency, eval=FALSE}
data_dir <- "japanese-resources"
dir.create(data_dir, showWarnings = FALSE)
revision <- "7cb5fb36add76b83a266d1967536e1a1d3faa513"
frequency_path <- file.path(data_dir, "tubelex-ja-base-pos.tsv.xz")
if (!file.exists(frequency_path)) {
  download.file(paste0("https://raw.githubusercontent.com/naist-nlp/tubelex/",
    revision, "/frequencies/tubelex-ja-base-pos.tsv.xz"),
    frequency_path, mode = "wb")
}
```

Retain the [upstream BSD-3-Clause notice](https://github.com/naist-nlp/tubelex/blob/7cb5fb36add76b83a266d1967536e1a1d3faa513/LICENSE)
and cite the paper when using TUBELEX. Aggregate-frequency availability does not
make the original subtitles a redistributable corpus. For the other sources,
follow the [Japanese norms guide](japanese-norms.html): WLSP-familiarity v4.0
is CC BY-NC-SA 3.0; the inspected AoA data are CC BY-NC 4.0; the BOI OSF
project specifies CC BY 4.0. The package code license does not replace those
terms for your joined table.

## Review the keys before joining

Keep `term` as the norm lookup spelling and `orth_base` as the reviewed frequency
key. For example, `食べた` can use frequency base `食べる`, while AoA/BOI lookup
still uses `食べた`; substituting a base-form rating would be a separate decision.
`人気` in two contexts receives the same base-key frequency. The TUBELEX
`pos` column is the majority POS for a key, not a disambiguated frequency.
Neither a reading nor a WLSP record selection splits that aggregate count.

```{r prepare-items, eval=FALSE}
items <- data.frame(
  item_id = c("a1", "a2", "a3", "a4", "b1", "b2", "b3", "b4", "b5"),
  condition = c(rep("A", 4), rep("B", 5)),
  term = c("学校", "犬", "人気", "食べた", "猫", "椅子", "人気", "未登録の例語", "青い空"),
  context = c("school", "animal", "popularity", "ate", "animal", "chair",
              "presence of people", "unmatched illustration", "multiword phrase"),
  orth_base = c("学校", "犬", "人気", "食べる", "猫", "椅子", "人気", "未登録の例語", NA_character_),
  base_reason = c(rep("Manually reviewed base for this illustration", 7),
    "Deliberately absent test key, not an observed lexical base",
    "Multiple words; no single orthBase assigned")
)
source(system.file("examples", "japanese-tubelex.R", package = "ldfreq"))
frequency <- profile_japanese_tubelex_items(items, frequency_path)
frequency$items[, c("item_id", "term", "orth_base", "tubelex_lookup_term",
                    "tubelex_status", "tubelex_per_million")]
```

The helper performs NFKC then lowercase **after** you supply the base. It keeps
the original forms and a normalization-change flag. It does not trim strings,
guess a base, tokenize phrases, or replace an unmatched key with a lemma. A
missing base is `unresolved_base`, an absent supplied key is `unmatched`, and
both retain `NA` counts, not zero. Its generic profile contains only queried
items; `profile_item_ids` maps those rows back to the full item table. Use the
full table as the denominator for stimulus selection.

For corpus tokens, use the optional gibasa/UniDic recipe in
[Japanese annotations](japanese-annotations.html). After inspecting `selected`
from that guide, its original positions and `orthBase` can be connected as follows:

```{r annotation-connection, eval=FALSE}
token_items <- data.frame(item_id = paste0("token-", seq_len(nrow(selected))),
  document_id = selected$document_id, segment_id = selected$segment_id,
  token_index = selected$token_index, term = selected$surface,
  orth_base = selected$orthBase,
  base_reason = "gibasa 1.1.3; UniDic 2.1.2 / unidic-lite 1.0.8; reviewed selection")
token_frequency <- profile_japanese_tubelex_items(token_items, frequency_path)
saveRDS(list(imported = ja, selection = selection, frequency = token_frequency),
        "japanese-token-frequency.rds")
```

This is a reviewed base-key lookup, **not a reproduction of the full TUBELEX
tokenizer**. Upstream converts fullwidth tilde before tokenization, handles
numbers/placeholders and uses surface fallback for unknown nodes; the gibasa
recipe does not reproduce all those steps or expose a verified unknown-node
flag. Do not normalize raw text before annotation merely to increase coverage:
that can alter segmentation and source positions. Record any separate text
transformation, dictionary hashes and excluded tokens. The saved imported
object retains those declarations and positions. This lookup does not establish
accuracy on learner errors or equivalence between English and Japanese measures.

## Combine ratings while keeping decisions and missingness

Place the separately acquired, pinned files documented in the norms guide in
`data_dir`. Review all WLSP candidates before recording your choices; even a
unique candidate requires an explicit decision. The selections below illustrate
those decisions for the stated contexts, not validated sense-specific norms.

```{r combine-sources, eval=FALSE}
source(system.file("examples", "japanese-norms.R", package = "ldfreq"))
source(system.file("examples", "wlsp-items.R", package = "ldfreq"))
aoa <- profile_japanese_norm_items(items, file.path(data_dir, "data_AoA.csv"), "aoa")
boi <- profile_japanese_norm_items(items, file.path(data_dir, "data_boi.csv"), "boi")
wlsp_path <- file.path(data_dir, "bunruidb-fam.csv")
review <- review_wlsp_items(items, wlsp_path)
review$candidates
decisions <- data.frame(item_id = c("a1", "a2", "a3", "b1", "b2", "b3"),
  record_id = c("022471", "062994", "033333", "063080", "053482", "002375"),
  reason = c("School context", "Animal, not occupational metaphor", "Popularity reading ninki",
             "Animal context", "Furniture context", "Presence reading hitoke"))
wlsp <- review_wlsp_items(items, wlsp_path, decisions)

# Each item appears once. Never merge by spelling: 人気 has two study IDs.
joined <- frequency$items
for (name in c("aoa", "boi", "wlsp")) {
  result <- list(aoa = aoa, boi = boi, wlsp = wlsp)[[name]]$items
  index <- match(joined$item_id, result$item_id)
  stopifnot(!anyDuplicated(result$item_id), !anyNA(index), nrow(result) == nrow(items))
  fields <- setdiff(names(result), names(items))
  added <- result[index, fields, drop = FALSE]
  if (name != "wlsp") names(added) <- sub("^norm_", paste0(name, "_"), fields)
  stopifnot(!any(names(added) %in% names(joined)))
  joined[names(added)] <- added
}
stopifnot(identical(joined$item_id, items$item_id))
joined[, c("item_id", "condition", "tubelex_status", "wlsp_status",
           "aoa_lookup_status", "boi_lookup_status")]
```

AoA is a mean of seven **rating categories**, not years; category 7 includes
both 12+ years and unknown age. BOI is a separate seven-point rating. WLSP
`know` is a published Bayesian estimate, not an NTT seven-point rating. Adult
native-speaker norms do not measure an individual learner's knowledge or ability
to employ a word appropriately. The BOI publication/correction scope and WLSP
estimation caveats remain as documented in the norms guide and helper metadata.

## Inspect every condition and each available subset

```{r inspect-conditions, eval=FALSE}
measure_columns <- c(frequency = "tubelex_per_million", familiarity = "wlsp_know",
                     aoa = "aoa_mean", boi = "boi_mean")
coverage <- do.call(rbind, lapply(split(joined, joined$condition), function(x) {
  do.call(rbind, lapply(names(measure_columns), function(measure) {
    value <- x[[measure_columns[[measure]]]]
    observed <- value[!is.na(value)]
    q <- if (length(observed)) unname(quantile(observed, c(0, .25, .5, .75, 1))) else rep(NA_real_, 5)
    data.frame(condition = x$condition[1], measure, total = nrow(x),
      observed = length(observed), missing = sum(is.na(value)),
      mean = if (length(observed)) mean(observed) else NA_real_,
      sd = if (length(observed) > 1) sd(observed) else NA_real_,
      min = q[1], q25 = q[2], median = q[3], q75 = q[4], max = q[5])
  }))
}))
rownames(coverage) <- NULL
coverage
complete <- complete.cases(joined[unname(measure_columns)])
table(condition = joined$condition, all_four_observed = complete)
joined[!complete, c("item_id", "condition", "tubelex_status", "wlsp_status",
                    "aoa_lookup_status", "boi_lookup_status")]

# Show the individual observations, not just condition means.
pdf("japanese-stimulus-distributions.pdf", width = 8, height = 7)
par(mfrow = c(2, 2))
for (measure in names(measure_columns)) {
  value <- joined[[measure_columns[[measure]]]]
  if (measure == "frequency") value <- log10(value) # matched positive counts only
  if (any(is.finite(value))) {
    stripchart(value ~ joined$condition, vertical = TRUE, method = "stack", pch = 16,
      xlab = "Condition", ylab = if (measure == "frequency") "log10(per million)" else measure)
  } else { plot.new(); message(measure, " has no observed values; see the coverage table.") }
}
dev.off()
```

Select potential confounds based on the research question, not on which columns
make condition means look similar. Decide in advance how to handle missing
items, acceptable imbalance and correlations among predictors. Equal means or
non-significant tests do not establish balance. Check the complete subset
separately: excluding unmatched items can change the lexical population. Record
any replacements and repeat the inspection on the final item IDs. Do not sum
these measures into a composite difficulty score.

```{r save-and-reload, eval=FALSE}
bundle <- list(items = joined, coverage = coverage, complete = complete,
  frequency = frequency, aoa = aoa, boi = boi, wlsp = wlsp, session = sessionInfo())
saveRDS(bundle, "japanese-stimulus-review.rds")
write.csv(joined, "japanese-stimulus-review.csv", row.names = FALSE, fileEncoding = "UTF-8")
restored <- readRDS("japanese-stimulus-review.rds")
stopifnot(identical(restored$items, joined), identical(restored$wlsp$decisions, decisions))
```

RDS preserves source hashes, denominators, mapping decisions and the complete
profile objects; CSV is a convenient review table but lacks that provenance.
Keep the acquired files in a permitted local archive. Sharing joined values
requires respecting each source's terms. To report the analysis, specify the
study population and task, lexical unit, resource versions, selected WLSP
estimate, base-form choices, per-condition missingness and retained item IDs.
