---
title: "Japanese annotations: from original text to lexical measures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Japanese annotations: from original text to lexical measures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ldfreq)
dic_dir <- Sys.getenv("LDFREQ_UNIDIC_LITE_1_0_8", unset = "")
run_gibasa <- nzchar(dic_dir) && requireNamespace("gibasa", quietly = TRUE)
run_file_review <- requireNamespace("quanteda", quietly = TRUE) &&
  isTRUE(l10n_info()[["UTF-8"]])
knitr::read_chunk(system.file("examples", "japanese-file-workflow.R",
  package = "ldfreq", mustWork = TRUE))
knitr::read_chunk(system.file("examples", "japanese-document-profile-demo.R",
  package = "ldfreq", mustWork = TRUE))
knitr::read_chunk(system.file("examples", "japanese-frequency-demo.R",
  package = "ldfreq", mustWork = TRUE))
```

For stimulus-item ratings and separately obtained AoA/BOI tables, see
[Japanese lexical norms](japanese-norms.html). That workflow retains item IDs
and norm-matching decisions independently of morphological segmentation.

Japanese text needs an explicit word-unit decision. For example, counting
`国際連合` as one item or `国際 / 連合` as two changes both vocabulary diversity
and the available bigrams. The default Unicode tokenizer in ldfreq does not
perform Japanese morphological analysis. Instead, use a Japanese analyzer and
bring its complete annotations into `lexdiv_import_annotations()`.

The importer checks the supplied surfaces against original text, derives source
positions, retains dictionary metadata and leaves missing annotations missing.
Use existing R packages for morphology and concordances, and ldfreq to keep
those input decisions visible in lexical analyses. This is an integration and
audit contribution, not a new Japanese segmentation algorithm or a validated
Japanese proficiency scale.

For a complete first workflow, start with
[Read files, review words and save the analysis](#japanese-file-workflow).
It uses small authored TXT/CSV files and prepared annotations, so you can learn
the decisions without installing an analyzer. Then use the
[gibasa recipe](#use-gibasa-and-a-local-unidic-dictionary-in-r) to supply your own
annotations. File reading and morphological analysis are separate steps.

## An offline example

The package's token-table importer supports R >= 4.1. The optional
gibasa 1.1.3 recipe requires **R >= 4.2**, as well as the separately supplied
dictionary. On R 4.1, import annotations prepared elsewhere or use the offline
example below; gibasa is not required to install or use the core package.

These annotations are authored for illustration, not analyzer output or a
Japanese gold standard. The original examples in this guide are covered by the
package's MIT license; no corpus or dictionary is redistributed.

```{r authored}
segments <- data.frame(
  document_id = c("essay", "empty"), segment_id = c("s1", "s1"),
  text = c("国際連合で国際協力を学ぶ。", "")
)
annotations <- data.frame(
  document_id = "essay", segment_id = "s1", token_index = 1:8,
  surface = c("国際", "連合", "で", "国際", "協力", "を", "学ぶ", "。"),
  lemma = c("国際", "連合", "で", "国際", "協力", "を", "学ぶ", NA_character_)
)
imported <- lexdiv_import_annotations(annotations, segments, list(
  language = "ja", analyzer = "authored", analyzer_version = "1",
  dictionary = "none", dictionary_version = "not-applicable",
  unit = "authored-fine", normalization = "none"
))
imported$tokens
```

`start` and `end` are **1-based inclusive Unicode codepoint positions within
the segment's original text**. They are not whole-document positions, byte
offsets, UTF-16 offsets or positions after normalization. The empty document
remains in `imported$documents`, with a zero-token segment. A sentence or turn
is a caller-supplied segment; the importer does not infer these boundaries.

Import **before filtering**. Apart from whitespace, every original character
must be covered, including punctuation. A skipped repeated word, dropped
punctuation, or normalized replacement causes an error. A supplied `start` or
`end` column must match the derived coordinates. This verifies the alignment,
not the correctness of POS, lexical units, or the source's sentence boundaries.

Whitespace annotations may be partial. For example, an analyzer can emit CR
tokens while omitting LF from CRLF line endings. A whitespace-prefixed token
uses the earliest exact match across whitespace-only gaps; non-whitespace
characters cannot be skipped. Original line endings and positions are retained.

```{r measures}
words <- imported$tokens[imported$tokens$surface != "。", ]
lexdiv_metrics(words$surface, metrics = "ttr")
phrases <- lexdiv_ngrams(words, "authored-ja-surface-no-punctuation-v1",
  term_col = "surface", documents = imported$documents$document_id)
phrases$documents
```

The seven retained tokens contain six distinct surface strings: TTR is 6/7.
There are six adjacent bigram and five trigram opportunities. These counts
verify the example's arithmetic; a seven-token text cannot establish the
reliability of a diversity measure. Choose length-appropriate measures and
report non-computable requests rather than changing their parameters silently.

## Compare word units on the same original text {#compare-japanese-word-units}

A different word unit changes the denominator as well as the vocabulary being
counted. Keep the original documents fixed, import both complete annotations,
and compare their source spans before calculating scores. Here is a second,
coarser **authored** annotation of the same teaching text; neither annotation
is presented as an official short-unit or long-unit analysis.

```{r ja-unit-comparison}
coarse_rows <- data.frame(document_id = "essay", segment_id = "s1",
  token_index = 1:6,
  surface = c("国際連合", "で", "国際協力", "を", "学ぶ", "。"))
coarse <- lexdiv_import_annotations(coarse_rows, segments, list(
  language = "ja", analyzer = "authored", analyzer_version = "1",
  dictionary = "none", dictionary_version = "not-applicable",
  unit = "authored-coarse", normalization = "none"
))
unit_alignment <- lexdiv_align_annotations(imported, coarse)
unit_alignment$groups[c("keyword", "relation", "predicted_n", "reference_n")]

count_units <- function(x) {
  t <- x$tokens
  retained <- t$surface != "。"  # This teaching example's explicit selection.
  tokens <- setNames(lapply(x$documents$document_id, function(id) {
    t$surface[t$document_id == id & retained]
  }), x$documents$document_id)
  lexdiv_metrics_batch(tokens, metrics = "ttr")
}
count_units(imported)
count_units(coarse)
```

The same sentence now has five retained tokens and five types, versus seven
tokens and six types before. TTR rises from 6/7 to 1 without a change in the
writer's text. The empty document stays missing in both conditions. Span
alignment identifies the merges; it does not decide which analysis is best.

In real data, check the **covered original characters after selection**, too.
A punctuation token in a fine analysis may be inside a retained compound in
a coarser analysis. Applying `!upos %in% c("PUNCT", "SYM")` to both tables can
therefore select different material. Do not remove that internal punctuation
by rewriting a token silently. Retain the discrepancy or justify a shared
source-span selection for the research question.

### What the executed comparisons show

The [reproducible unit comparison](https://github.com/Ryuya-dot-com/ldfreq/tree/main/experiments/japanese-units)
uses all 543 test sentences from UD Japanese GSD/GSDLUW r2.18. Their published
SUW/LUW annotations have identical sentence IDs and original text. The
[upstream description](https://github.com/UniversalDependencies/UD_Japanese-GSDLUW/blob/71fd68633ab8c3439a48db3f3718c9ed80be9c6c/README.md)
specifies manually annotated layers following BCCWJ conventions. These are
news/blog sentences, not learner essays. The corpus is obtained separately
under CC BY-SA 4.0; the R package bundles neither its text nor its annotations.

| All-token sentence summary | SUW | LUW |
| --- | ---: | ---: |
| Median N | 21 | 17 |
| Median TTR | 0.9130 | 0.9259 |
| MATTR50 computable / all sentences | 29 / 543 | 10 / 543 |
| MTLD tail-only / computable | 352 / 384 | 331 / 346 |

All-token counting includes punctuation to isolate boundaries over the same
covered characters; it is not a recommended lexical policy. Excluding UPOS
PUNCT/SYM produces different coverage in 54/543 sentence pairs. The comparison
records all cases and the 489-pair same-coverage subset separately. MATTR50
holds the window's token count fixed, not its original-text span. Its paired
all-token comparison has only ten eligible sentences, not 543. Sentences are
never concatenated to manufacture eligible documents.

The many tail-only MTLD values illustrate why a finite score is insufficient
evidence of precision. Consult [factor support and length checks](designing-comparisons.html#inspect-mtld-factor-support).
The advisory 50-token floor is not a validated Japanese reliability threshold.

A separate check reuses saved Sudachi A/B/C analyses of 28 actual Japanese
learner essays. Surface-form scores change in 24/28 essays between A and C:

| Metric | A median | C median | Paired Spearman rho |
| --- | ---: | ---: | ---: |
| TTR | 0.4316 | 0.4328 | 0.9907 |
| MATTR50 | 0.7082 | 0.7081 | 0.9880 |
| MTLD | 43.9420 | 43.6272 | 0.9496 |

All 28 pairs are finite and retain the same original character coverage; none
has a tail-only MTLD flag. These small-sample observations do not validate
precision, proficiency discrimination or interchangeable score scales.
**Sudachi C is not relabeled NINJAL LUW.** The public sentence comparison and
the saved essay mode comparison answer different questions.

For a methods section, report the actual source/version, unit, form, selection,
window and MTLD definition. For example, the essay check above can be described
as follows (replace these conditions with those actually used in your study):

> We compared surface-form lexical diversity in the same 28 essays using saved
> SudachiPy 0.6.11 / SudachiDict-core 20260428 annotations in modes A, B and C.
> Original text, including titles, was retained. Whitespace and auxiliary-symbol
> tokens were excluded; particles, auxiliaries and numbers were retained.
> MATTR used 50-token windows. MTLD used threshold .72, strict `<`, no minimum
> factor length, linear tail credit and the mean of directional scores. We
> checked original-character coverage, paired missingness and MTLD factor
> support before comparing scores. These comparisons assessed sensitivity to
> the analysis mode, not proficiency or measurement precision.

## Read files, review words and save the analysis {#japanese-file-workflow}

Suppose the first line is a title and your question concerns vocabulary in the
body. Deleting the title changes later character positions; counting every
target hit includes words outside your intended sample. This example retains
the whole file, selects the body by its original positions, and reports body
word counts separately from the lexical identities you review.

The five authored files contain spelling variants, homophones, an invented
unlisted form, punctuation only, and an empty file. Their annotations are
**prepared by the example's author**, not produced by a hidden tokenizer. All
files are MIT teaching material; no learner corpus or dictionary is bundled.
The CSV also contains authored `POS1` and `goshu` labels for the descriptive
profile below; `<MISSING>` marks unassigned features. These are not UniDic results.
Optional quanteda >= 4.5.0 and a UTF-8 R session are needed for KWIC review.
No Python or external model is required for this walkthrough.

To run the complete example at once:

```{r ja-files-quick-start, eval=FALSE}
source(system.file("examples", "japanese-file-workflow.R", package = "ldfreq"))
as.data.frame(ja_file_record$metrics)  # All retained body words
ja_file_record$counts          # Reviewed target occurrences only
ja_file_output                 # Directory containing the complete RDS and count CSV
```

The following steps show the same recipe. The files and annotations are separate
inputs; replacing a TXT file also requires complete annotations of its new text.

### 1. Read the original files and import prepared annotations

```{r ja-files-read, eval=run_file_review, results='hide'}
```

Each original file is one segment here. Whitespace gaps, including its line
breaks, are preserved. These are not inferred sentence boundaries. The importer
checks every non-whitespace character against `annotations.csv` before any
title or punctuation is excluded. `ja_file_manifest` records encoding, byte
hashes and a removed leading BOM, if any. The [file input guide](english-tokenization.html#import-text-files)
explains explicit UTF-8/CP932 decoding in more detail.

| Object | One row represents | Keep it for |
|---|---|---|
| `ja_file_segments` | One original file/segment, including empty files | Original text and document IDs |
| `ja_file_import$tokens` | One prepared token with original character positions | Complete annotation before selection |
| `ja_file_selection` | One original token and its inclusion reason | Body and punctuation exclusions |
| `ja_file_after$occurrences` | One target occurrence, not one candidate | Context, status and recorded decisions |
| `ja_file_record` | The complete saved analysis | Reopening without the analyzer or source files |

### 2. Select the body without rewriting the text

```{r ja-files-ranges, eval=run_file_review}
```

```{r ja-files-scope, eval=run_file_review, echo=FALSE}
```

The first body starts at character 5, after the title `リンゴ` and its newline.
The second starts at character 4, after `はし` and its newline. Those are
explicit decisions about these files, **not a rule that every first line is a
title**. Empty files retain unavailable bounds. The helper is limited to one
segment and one contiguous body range per document; it rejects a range that
cuts through a token. A boundary change requires a new complete annotation,
as shown [below](#review-boundaries-before-lexical-identity).

This recipe excludes only the literal full stop `。` after selecting the body.
It retains particles and verbs. Other datasets need their own declared policy
for punctuation, numbers and proper nouns. Neither the complete import nor its
original token indices are edited.

<details>
<summary>Show the body-range validation and selection code</summary>

```{r ja-files-scope-code, ref.label='ja-files-scope', eval=FALSE, echo=TRUE}
```

</details>

### 3. Inspect KWIC and record contextual decisions

```{r ja-files-review, eval=run_file_review}
```

Title hits remain visible with `in_body = FALSE`; they are unsubmitted and do
not enter the body counts. The body contains three spellings intended as the
fruit, two contextually identified uses of `はし`, a third unresolved `はし`,
and an invented word with no candidate. Even a single candidate needs an
explicit decision. The assignments in the code express the author's intended
meanings; they are not automatic correction or evidence of annotation accuracy.

For your own research, use the
[CSV decision worksheet](ambiguity-review.html#edit-decision-worksheet) to inspect,
edit and reapply decisions by their fixed IDs. Keep title/body membership
beside the worksheet; do not delete the original annotations to hide title hits.
Changing the source, segmentation or candidate inventory requires a new review.
The body range is a separate selection: save it with the review and rerun the
selection/count steps when it changes. The review ID does not certify a body
range that was never part of its input.

### 4. Report whole-body measures and target counts separately

```{r ja-files-counts, eval=run_file_review, include=FALSE}
```

```{r ja-files-body-table, eval=run_file_review, echo=FALSE}
knitr::kable(as.data.frame(ja_file_metrics)[c("document_id", "N", "V", "value", "status")],
  digits = 3, col.names = c("Document", "Body tokens", "Surface types", "TTR", "Status"))
```

The first two bodies contain 5 and 9 tokens, with 4 and 6 surface types. These
TTR values illustrate accounting on tiny texts, not a recommended proficiency
measure. Punctuation-only and empty documents both have no retained words;
their source counts distinguish them, while TTR remains missing for both.

```{r ja-files-target-table, eval=run_file_review, echo=FALSE}
knitr::kable(ja_file_counts[c("document_id", "target_N", "selected_N",
  "unresolved_N", "no_candidates_N", "selection_coverage")], digits = 2,
  col.names = c("Document", "Targets", "Selected", "Unresolved", "No candidate", "Coverage"))
```

```{r ja-files-type-table, eval=run_file_review, echo=FALSE}
knitr::kable(ja_file_counts[c("document_id", "common_surface_V", "common_lexical_V",
  "target_lexical_V")], col.names = c("Document", "Selected surface types",
  "Selected lexical types", "All-target lexical types"))
```

The variant spellings have 3 surface types and 1 selected lexical identity on
the same 3 targets. The two selected homophone occurrences have 1 surface type
and 2 identities. The third homophone is unresolved, so the complete target
lexical-type count stays `NA`, with coverage 2/3. The invented form also stays
unavailable, rather than receiving its own identity as a fallback. For a
document with no targets, zero target types is an empty-set count, not a claim
that its writer knows zero words; selection coverage is unavailable.

`ja_file_metrics` uses every retained body word. `ja_file_counts` describes only
the declared targets. Do not replace the former's V with the latter's selected
identity count. An all-word lexical-identity measure would require an explicit
identity policy and adequate coverage for the whole word population.

<details>
<summary>Show how the body measures and target tables are calculated</summary>

```{r ja-files-counts-code, ref.label='ja-files-counts', eval=FALSE, include=TRUE, echo=TRUE}
```

</details>

### 5. Save enough to resume

```{r ja-files-save, eval=run_file_review}
```

The recipe uses a temporary directory to avoid overwriting study files. In your
project, save the same complete `ja_file_record` to a chosen persistent path:

```{r ja-files-persistent-save, eval=FALSE}
saveRDS(ja_file_record, "japanese-file-analysis.rds", version = 2)
saved <- readRDS("japanese-file-analysis.rds")
saved$counts
saved$body_ranges
saved$selection[saved$selection$reason != "retained", ]
```

The RDS retains the original text, input hashes, annotations, body ranges,
inclusion reasons, candidates, before/after decisions, measures and session.
The CSV is an inspection table and cannot replace that complete record. You can
read the RDS without Python or a dictionary; reapplying KWIC decisions requires
quanteda. For other analyzers, import their complete original surfaces and
actual metadata into the same workflow; never replace source text with a
normalized form to make the import succeed.

## Describe script, word origin and POS by document

Use a document profile to describe **what was counted** before interpreting a
diversity score. The installed `japanese-document-profile.R` recipe takes a
complete import, its complete selection table, an explicit POS grouping and a
condition label. It returns ordinary data frames and keeps its inputs. It is
an explicitly sourced example, not an exported function or a new analyzer.

Continue from the file workflow above:

```{r ja-profile-load, eval=run_file_review}
```

```{r ja-profile-build, eval=run_file_review}
```

The example map groups some POS labels as `content` or `function`. Pronouns,
affixes, conjunctions and other unlisted labels remain **unmapped**; this is
a deliberately partial classification, not a universal content-word definition.
For your study, supply the labels and grouping that match your dictionary and
research question. The recipe does not change which tokens you selected.

```{r ja-profile-documents, eval=run_file_review, echo=FALSE}
knitr::kable(ja_profile$documents[c("document_id", "source_N", "outside_body_N",
  "excluded_inside_N", "retained_N", "retained_codepoints")],
  col.names = c("Document", "Source tokens", "Outside body", "Excluded in body",
    "Retained tokens", "Retained codepoints"))
```

For each document, source tokens equal outside-body tokens plus exclusions
inside the body plus retained tokens. The original file, including its title,
line endings and spaces, remains available. Character counts have two distinct
populations: `original_text` covers every original segment, whereas
`retained_tokens` covers only the surfaces of retained tokens, excluding gaps.
The latter is not the length of the original body including whitespace.

```{r ja-profile-scripts, eval=run_file_review, echo=FALSE}
shown_scripts <- subset(ja_profile$characters,
  document_id == "variants" & population == "retained_tokens" & n > 0)
knitr::kable(shown_scripts[c("category", "n", "denominator", "proportion")],
  digits = 2, col.names = c("Character category", "Codepoints", "All retained codepoints", "Proportion"))
```

The three fruit spellings appear in 5 retained tokens totaling 10 codepoints:
2 Han-script, 5 hiragana and 3 katakana. Their supplied origin labels tell a
different story:

```{r ja-profile-origins, eval=run_file_review, echo=FALSE}
shown_origins <- subset(ja_profile$features, feature == "origin" &
  (n > 0 | (status == "missing" & denominator == 0)))
knitr::kable(shown_origins[c("document_id", "category", "status", "n", "denominator", "proportion")],
  digits = 2, col.names = c("Document", "Origin label", "Status", "Tokens", "All retained tokens", "Proportion"))
```

All three fruit spellings have the authored `漢` label; the two particles have
`和`. Writing a word in katakana does not make this recipe assign it a loanword
label. Actual [UniDic output](https://clrd.ninjal.ac.jp/unidic/faq.html) supplies
`goshu` independently of surface spelling. Its labels, including proper-name
and symbol categories, are preserved as supplied. No origin is inferred for a
missing label, and character proportions are not scores of kanji knowledge.

The character unit is the **original Unicode codepoint**, with no normalization.
Thus precomposed `が` has one codepoint; `か` followed by a combining dakuten has
two. These are not grapheme-cluster or visual-character counts. The recipe uses
[Unicode Script properties](https://www.unicode.org/reports/tr24/) through
stringi and saves its ICU/Unicode versions. Its disjoint categories first
separate six shared kana signs (`ー`, `ｰ`, `゛`, `゜`, `ﾞ`, `ﾟ`), then combining
marks, decimal digits, whitespace, punctuation, symbols, Han, hiragana, katakana,
Latin and other codepoints. Han-script includes signs such as `々`; it is not an
educational kanji inventory. Half-width forms remain half-width, and emoji
sequences can contain several codepoints in different categories.

Inspect the three tables and the source-linked rows directly:

```{r ja-profile-inspect, eval=run_file_review}
ja_profile$documents[c("document_id", "pos_missing_N", "origin_missing_N", "pos_unmapped_N")]
subset(ja_profile$tokens, retained &
  (pos_status == "missing" | pos_group_status == "unmapped"),
  select = c(document_id, segment_id, start, end, surface, pos_value, pos_group_status))
```

| Output | Unit and denominator |
|---|---|
| `$documents` | One document under the declared condition, including empty documents; token and codepoint totals are separate. |
| `$characters` | One document/population/category; `denominator` includes every codepoint in that population, including whitespace and other characters where present. |
| `$features` | One document/feature/category/status; all retained tokens are the denominator, including missing and unmapped entries. Features are `pos`, `origin` and `pos_group`. |
| `$tokens` | Every original token with source anchors, selection reasons, raw labels, classification statuses and category-specific `char_` counts. |

`missing` means `NA`, blank-only text or literal `*`; the raw value survives in
`$tokens` and `$imported`. `unmapped` means a POS label exists but has no entry
in your grouping table. `field_present = FALSE` distinguishes an absent input
column from a present column with missing entries. Feature columns must be
character vectors. Supply `pos_col` and
`origin_col` if your columns have different names. A zero count with a positive
denominator is zero; an empty denominator produces `NA` proportions.

Lexical-review decisions remain in `ja_file_record$after`: a decision about a
target's lexical identity does not validate every POS/origin label or change
these descriptions automatically. The profile keeps the full selection and
input; the file-workflow record additionally keeps body ranges, decisions and
the existing N/V/TTR results. Save them together:

```{r ja-profile-save, eval=run_file_review}
```

For a persistent study directory, change the output path explicitly. The CSVs
are analysis/inspection tables; the full RDS retains the original sources and
conditions. Recompute from the saved inputs and policy without rerunning a model:

```{r ja-profile-replay, eval=run_file_review}
```

Exact replay here uses the same software versions; retain the saved counts and
ICU/Unicode metadata when comparing results across software upgrades.
Changed source or partial/stale selection rows are rejected. Reordering the
selection table is allowed because its original token anchors are checked.
No API call, fee or additional dependency is needed for these profiles.

To run both the file workflow and profiling example together in a fresh session:

```{r ja-profile-complete, eval=FALSE}
source(system.file("examples", "japanese-document-profile-demo.R", package = "ldfreq"))
ja_profile$documents
ja_file_output  # Temporary folder; choose a persistent output for your study.
```

## Connect document selections to reference frequency

A frequency mean can change because the selected words changed, because lookup
keys changed, or because some words could not be scored. Keep the original
occurrences and all three denominators visible. This recipe connects the document
profile above to the existing `lexdiv_norm_profile_batch()` calculation.

The following reference has **fictional teaching numbers, not TUBELEX data or
Japanese norms**. Its zero and missing value are deliberate examples. No download,
analyzer or API call is needed. Start with the complete file/profile example above.

```{r ja-frequency-load, eval=run_file_review}
```

```{r ja-frequency-keys, eval=run_file_review}
```

This table-specific key mapping reuses the authored lexical decisions. It does
not generate UniDic `orthBase` values. One ambiguous `はし` remains unresolved;
the invented `ぷにょ語` has a supplied key but no entry in the teaching table.
Every original token, including exclusions and title words, has an anchored row.
In your study, record the source and reason for each key. A lexical decision
alone does not establish compatibility with a particular frequency table.

```{r ja-frequency-reference, eval=run_file_review}
```

```{r ja-frequency-accounting, echo=FALSE, eval=run_file_review}
knitr::kable(ja_frequency$documents[c("document_id", "retained_N", "keyed_N",
  "matched_N", "unresolved_key_N", "unmatched_N", "matched_coverage")],
  row.names = FALSE, digits = 3,
  col.names = c("Document", "Retained", "Key supplied", "Matched", "Key unresolved", "Unmatched", "Coverage"))
```

Here `homophones` has 9 retained tokens, 8 supplied keys and 8 matches:
**full-token coverage is 8/9**, even though all 8 supplied keys match. `unlisted`
has coverage zero. Documents with no retained tokens have `NA` coverage and
means. The accounting identity is retained = matched + unmatched + unresolved.

```{r ja-frequency-summary, eval=run_file_review}
subset(ja_frequency$norms$summary, measure_id == "per_million",
  select = c(document_id, weighting, estimate, input_units, observed_value_units, missing_reason))
```

The existing norm summaries use **resolved keys** as input units. Token means
weight occurrences; type means use unique exact lookup keys, not independently
identified lexemes. Both means use observed matched values only. For `variants`,
the fictional token mean is 406 and the type mean is 505. For `homophones`, the
frequency mean includes the observed zero for `見る`; its missing channel value
is omitted only from the channel mean, whose observed denominator is shown.
Always report these conditional summaries beside `$documents`, whose coverage
includes unresolved keys. Type coverage for unresolved lexical identities is
not estimated. `$occurrences` retains source anchors, keys, reasons, statuses
and each `value_` column, including excluded occurrences for inspection.

```{r ja-frequency-coverage-plot, eval=run_file_review && requireNamespace("ggplot2", quietly=TRUE), fig.width=7, fig.height=3.3, out.width="100%"}
plot_japanese_frequency_coverage(ja_frequency)
# Use monochrome = TRUE for black and white.
```

The points use the complete retained-token denominator. Rows without a point
have no retained tokens; they are not plotted at zero. The figure has no title
or subtitle. Add the figure number, title and note in your manuscript.

### Compare selection conditions on common original spans

```{r ja-frequency-compare, eval=run_file_review}
```

`$all` retains each full selection. `$common` retains only the same original
document/segment/start/end/surface with a match and **all supplied measures
observed in every condition**. This is a conservative complete-case comparison:
it does not align a long token to two shorter ones or impute missing values.
Token indices may change after segmentation, so commonality uses original spans,
not row numbers or token indices. Different original texts or reference tables
are rejected; comparison across reference resources is a separate operation.

```{r ja-frequency-common-summary, echo=FALSE, eval=run_file_review}
shown_frequency <- do.call(rbind, lapply(c("all", "common"), function(scope)
  do.call(rbind, lapply(names(ja_frequency_comparison[[scope]]), function(condition) {
    x <- ja_frequency_comparison[[scope]][[condition]]
    z <- subset(x$norms$summary, weighting == "token" & measure_id == "per_million")
    data.frame(scope, condition, document = z$document_id,
      retained = x$documents$retained_N, observed = z$observed_value_units, mean = z$estimate)
  }))))
knitr::kable(shown_frequency[shown_frequency$document %in% c("variants", "homophones"), ],
  row.names = FALSE, digits = 3)
```

The common rows have identical frequency means here because both conditions use
the same keys and reference. Differences in their full-selection means describe
which tokens entered the analysis. They do not establish a change in proficiency.
The missing channel annotation also excludes `見る` from this all-measure common
set, although its frequency itself is observed as zero. Choose your measure set
before forming the intersection and report the retained counts.

```{r ja-frequency-save, eval=run_file_review}
```

The full RDS saves original text, complete annotations, earlier lexical decisions,
selection/key policies, the actual reference table and both comparison scopes.
CSV tables support inspection and analysis; they do not replace that record.
Choose a persistent output folder for a study. Replay needs the saved reference,
not a model, API key or a new download:

```{r ja-frequency-replay, eval=run_file_review}
```

Run the entire authored workflow in a fresh session with:

```{r ja-frequency-complete, eval=FALSE}
source(system.file("examples", "japanese-frequency-demo.R", package = "ldfreq"))
ja_frequency$documents
ja_file_output
```

## Use gibasa and a local UniDic dictionary in R

[gibasa](https://paithiov909.github.io/gibasa/) supplies the MeCab interface;
ldfreq does not reimplement it. Install gibasa separately with
`install.packages("gibasa")`. Its optional installation requirements are
separate from those of ldfreq. A UTF-8 compiled dictionary is also required.

This recipe is for **unidic-lite 1.0.8, based on UniDic 2.1.2**, with the
26-field binary schema. It is a concrete integration example, not a claim that
this older dictionary is best for every contemporary Japanese dataset.
[unidic-lite](https://github.com/polm/unidic-lite) modifies the original UniDic
release; record both identities. Obtain its versioned archive from the
[release files](https://pypi.org/project/unidic-lite/1.0.8/#files), verify the
published SHA-256, and extract it. R's `download.file()` and `untar()` can do
this; no Python runtime is needed to use the extracted dictionary in gibasa.
Keep the included dictionary license and source information with it. Set
`LDFREQ_UNIDIC_LITE_1_0_8` to the extracted `unidic_lite/dicdir` directory before
rendering this guide, or assign that path to `dic_dir` and run the following
chunks interactively. The package build runs the authored examples offline;
it does not download a dictionary.

Supply one sentence or turn per segment, with a separate roster for original
document IDs. The following strings are new examples, not sampled corpus text.
The full-width letters and an invented string expose normalization and
missing-feature decisions.

```{r japanese-source}
ja_segments <- data.frame(document_id = c("essay", "essay", "empty"),
  segment_id = c("s1", "s2", "s1"),
  text = c("国際連合で国際協力を学ぶ。", "ＡＩで学びました。 qzxv😀", ""))
```

gibasa needs a readable MeCab configuration file even when `sys_dic` is
explicit. This helper retains an existing `~/.mecabrc` (which gibasa's bundled
MeCab checks before `MECABRC`), or uses a temporary empty configuration when
that file is absent. It records the configuration hash, verifies that only the
requested system dictionary is active, and restores the prior environment on
exit. It does not edit the home configuration or request normalization or
automatic sentence splitting. Use a path without spaces for this gibasa recipe.

```{r gibasa-analysis, eval=run_gibasa}
analyze_unidic <- function(segments, dic_dir) {
  rc <- path.expand("~/.mecabrc")
  temporary_rc <- !file.exists(rc)
  if (temporary_rc) {
    rc <- tempfile("mecabrc-")
    writeLines("", rc)
  }
  previous <- Sys.getenv("MECABRC", unset = NA_character_)
  on.exit({
    if (is.na(previous)) Sys.unsetenv("MECABRC") else Sys.setenv(MECABRC = previous)
    if (temporary_rc) unlink(rc)
  })
  Sys.setenv(MECABRC = rc)
  info <- gibasa::dictionary_info(sys_dic = dic_dir, user_dic = "")
  stopifnot(nrow(info) == 1L, info$type == 0L,
    tolower(gsub("-", "", info$charset)) == "utf8",
    normalizePath(info$file_path) == normalizePath(file.path(dic_dir, "sys.dic")))
  texts <- setNames(segments$text, paste0("input_", seq_len(nrow(segments))))
  raw <- gibasa::tokenize(texts, sys_dic = dic_dir, user_dic = "",
    split = FALSE, partial = FALSE)
  expanded <- as.data.frame(gibasa::prettify(raw,
    into = gibasa::get_dict_features("unidic26")))
  source_row <- match(as.character(raw$doc_id), names(texts))
  tokens <- data.frame(document_id = segments$document_id[source_row],
    segment_id = segments$segment_id[source_row], token_index = raw$token_id,
    surface = raw$token, raw_feature = raw$feature)
  features <- setdiff(names(expanded), c("doc_id", "sentence_id", "token_id", "token"))
  list(tokens = cbind(tokens, expanded[features]),
    configuration_sha256 = digest::digest(file = rc, algo = "sha256"))
}
analysis <- analyze_unidic(ja_segments, dic_dir)
ja_annotations <- analysis$tokens
ja_annotations[c("surface", "orthBase", "lemma", "lForm", "POS1")]
```

Use the feature schema for the **actual dictionary**, not the default IPA
schema of `prettify()`. The raw feature string is retained to make the expansion
auditable, including quoted commas and missing fields. gibasa does not expose
MeCab's unknown-node flag in these rows: a missing lemma is observable, but
is not an independently verified unknown-word classification. Do not fabricate
a definitive unknown flag from it.

```{r gibasa-import, eval=run_gibasa}
dictionary_files <- c("sys.dic", "unk.dic", "char.bin", "matrix.bin", "dicrc")
dictionary_hashes <- setNames(lapply(file.path(dic_dir, dictionary_files),
  function(path) digest::digest(file = path, algo = "sha256")),
  paste0(dictionary_files, "_sha256"))
ja <- lexdiv_import_annotations(ja_annotations, ja_segments, c(list(
  language = "ja", analyzer = "gibasa/MeCab",
  analyzer_version = as.character(utils::packageVersion("gibasa")),
  dictionary = "unidic-lite", dictionary_version = "1.0.8 (UniDic 2.1.2)",
  unit = "UniDic-short-unit-analysis", normalization = "none",
  settings = "unidic26; split=FALSE; partial=FALSE; no user dictionary",
  configuration_sha256 = analysis$configuration_sha256,
  dictionary_source = "https://pypi.org/project/unidic-lite/1.0.8/"
), dictionary_hashes))
ja$tokens[c("document_id", "segment_id", "token_index", "surface",
  "orthBase", "lemma", "start", "end")]
```

Metadata and hashes record the declared dictionary snapshot. They do not
independently identify or license an arbitrary dictionary path. If you change
dictionary release, schema, user dictionary or settings, update the metadata
and verify the feature mapping again. Retain user-dictionary hashes as well
when used. RDS saves the imported tables without requiring gibasa or the
dictionary when reloading them for analysis.

## Choose lexical forms and preserve exclusions

[UniDic's terminology](https://clrd.ninjal.ac.jp/unidic/glossary.html)
distinguishes the observed surface, `orthBase` (orthographic base) and `lemma`
(lexeme label). In this example, surface `学び` has base and lemma `学ぶ`;
`まし` maps to `ます`. A lemma label alone need not identify a unique lexeme:
homographs may require reading, POS or dictionary IDs. Do not use a first-match
join to resolve collisions with a reference table.

This explicit example excludes auxiliary symbols and whitespace tokens,
retains particles and auxiliary verbs, and requires an observed `orthBase`.
Inspect and save the exclusions rather than substituting surface forms for
missing bases. Change this selection if the research question calls for a
different lexical population, and change the preprocessing label accordingly.

```{r gibasa-selection, eval=run_gibasa}
t <- ja$tokens
punctuation <- !is.na(t$POS1) & t$POS1 %in% c("補助記号", "空白")
has_whitespace <- stringi::stri_detect_regex(t$surface, "\\p{White_Space}")
missing_base <- is.na(t$orthBase) | !nzchar(t$orthBase)
keep <- !punctuation & !has_whitespace & !missing_base
selection <- data.frame(document_id = t$document_id, segment_id = t$segment_id,
  token_index = t$token_index, punctuation, has_whitespace, missing_base, retained = keep)
selected <- t[keep, ]
selected$term <- selected$orthBase
by_document <- setNames(lapply(ja$documents$document_id,
  function(id) selected$term[selected$document_id == id]), ja$documents$document_id)
results <- lexdiv_metrics_batch(by_document, metrics = "ttr")
grams <- lexdiv_ngrams(selected, "ja-unidic-lite-1.0.8-orthBase-selection-v1",
  documents = ja$documents$document_id)
results
grams$documents
```

The diversity calculation describes the retained sequence. N-grams also
respect segment boundaries and the **original** `token_index`: excluding a
particle, punctuation or an unavailable base leaves a gap, not a new adjacent
pair. Do not renumber `selected`. With quanteda installed,
`lexdiv_as_quanteda(selected, ja$segments, term_col = "term")` preserves those
gaps for concordances and other quanteda operations in a UTF-8 R session.

```{r save-analysis, eval=FALSE}
saveRDS(list(imported = ja, selection = selection, metrics = results, ngrams = grams),
  "japanese-analysis.rds")
```

## Kana spellings and contextual lexical identity

Kana spelling is an observation, not evidence that a writer cannot produce a
kanji. Task instructions, input method and stylistic choices can also matter.
Keep two outcomes separate: what was written, and which lexical item a reviewed
occurrence represents. Inferring kanji knowledge requires additional evidence,
such as a production task with known opportunities and permitted writing aids.

[UniDic's hierarchy](https://clrd.ninjal.ac.jp/unidic/glossary.html) retains
orthographic variants below lexical entries. `orthBase` alone does **not** merge
all kana/kanji spellings; choosing `lemma` also pools inflectional variation.
For an orthography-only comparison, hold lexical identity and grammatical form
constant. A reading alone cannot distinguish homophones. An analyzer's selected
entry is a candidate annotation, not independent evidence that the writer meant
that entry. Segmentation errors require boundary review before lexical review.

The following example reuses the existing KWIC decision API. All text,
boundaries and candidates are authored; it requires optional **quanteda** in a
UTF-8 R session, but no downloaded corpus, dictionary, Python or model. The IDs identify illustrative
lexical items, not word families or a validated inventory of word senses.

```{r kana-review, eval=requireNamespace("quanteda", quietly=TRUE) && isTRUE(l10n_info()[["UTF-8"]])}
source(system.file("examples", "japanese-orthography.R", package = "ldfreq"))
spelling <- japanese_orthography_example
spelling$occurrences[, c("document_id", "segment_id", "pre", "keyword", "post",
  "status", "lexical_id", "lexical_label", "reason")]
spelling$counts
```

Only the declared **target occurrences** enter these counts; particles, verbs
and punctuation still remain in the source annotations and KWIC context.

| Authored target set | Observed spellings | Reviewed lexical identities |
|---|---|---|
| `りんご`, `リンゴ`, `林檎` | 3 types in 3 occurrences | 1 type in the same 3 occurrences |
| `はし` in crossing and eating contexts | 1 type in 2 selected occurrences | 2 types in the same 2 occurrences |
| A third `はし` in a seeing context | Its surface is known | Identity remains unresolved |
| Invented `ぷにょ語` | Its surface is known | No candidate supplied; no identity fallback |

The three HASHI occurrences have selection coverage 2/3. Their full-target
`lexical_V` remains `NA`; the two resolved occurrences support a conditional
comparison, with `common_surface_V = 1` and `common_lexical_V = 2`. Do not compare
a complete surface count with an incomplete lexical count as though their
populations matched. Even a single supplied candidate stays unreviewed until
explicitly selected. An absent candidate is not proof of a misspelling.

Original `surface`, `start` and `end` never change. Selected IDs and labels are
separate columns; the example does not rewrite kana into kanji. Saving the whole
object preserves the source, candidate inventory, decisions and reviewer reasons:

```{r save-kana-review, eval=FALSE}
saveRDS(spelling, "japanese-orthography-review.rds")
restored <- readRDS("japanese-orthography-review.rds")
```

### Review boundaries before lexical identity

The same example file also supplies `japanese_boundary_example`. In the authored
sentence `はしではしをつかいます。`, the intended meaning is using chopsticks on
a bridge. Compare two complete annotations of **the same original text**:

- Split alternative: `はし / で / は / し / を / つかい / ます / 。`
- Reviewed alternative: `はし / で / はし / を / つかい / ます / 。`

A local probe with gibasa 1.1.3 and unidic-lite 1.0.8 (UniDic 2.1.2) produced
the split alternative. This is a concrete integration case, not an estimate of
analyzer accuracy on learner writing. The offline example supplies both
annotations explicitly and requires neither that analyzer nor its dictionary.

```{r kana-boundaries, eval=requireNamespace("quanteda", quietly=TRUE) && isTRUE(l10n_info()[["UTF-8"]])}
boundary <- japanese_boundary_example
subset(boundary$alignment$groups, relation != "exact",
  select = c(start, end, pre, keyword, post, predicted_n, reference_n))
boundary$boundary_decisions
boundary$scores[, c("condition", "document_id", "N", "V", "value", "status")]
boundary$review$occurrences[, c("start", "end", "keyword", "candidate_id", "reason")]
```

The [alignment workflow](annotation-alignment.html) links original characters
4--5 to the two split tokens and the one reviewed token. The decision and its
reason are saved alongside both complete annotation objects. Import the reviewed
annotations afresh; editing an imported object's token rows in place invalidates
its identity. No lemma or POS is inherited automatically from a fragment.

The surface counts exclude only the full stop and retain all other tokens:
the split alternative has N = 7 and V = 7; the reviewed alternative has N = 6
and V = 5. These illustrative TTR values describe the supplied sequences, not
proficiency. Restricting the comparison to exact-span matches would discard the
changed tokens and fail to measure the effect under study.

Boundary selection does not decide whether either occurrence means bridge,
chopsticks or edge. Regenerate KWIC review on the new complete annotations and
record lexical decisions separately. The old split exposes one whole-token
HASHI, while the reviewed annotation exposes two. Decisions from a previous
segmentation have a different review ID and cannot be silently reused.

This example accepts an analyst's complete alternative segmentation; it does
not automatically repair boundaries or select the intended meaning. Keep the
whole `boundary` object when saving. Do not close gaps left by excluded or
unresolved tokens when forming n-grams or moving-window measures.

[JASWRIC](https://language.sakura.ne.jp/jaswric/) illustrates the practical issue:
it supplies original transcriptions and edited texts, including script changes,
for Japanese L1 student writing. Its supplied morphological analysis is based
on the edited version. Such paired materials can support a later sensitivity
study, but the authored example above is not a corpus-based accuracy evaluation.
Script proportions are descriptive; a kanji-production score additionally needs
a justified set of eligible words and acceptable spellings.

## Review inflected forms spanning several tokens

A morphological candidate can cover several analyzer tokens. For example,
`食べられる` can be represented by `食べ / られる`, while its potential, passive
and honorific analyses remain different candidates. A span identifies a source
occurrence; it does **not** automatically redefine that occurrence as one word
in a diversity calculation. Inflectional analyses also differ from lexical
senses and pedagogical word-family membership.

The following explicitly sourced recipe works offline with authored text,
annotations and six authored format-demonstration rows. It uses the existing
annotation importer and stringi's exact search; no new analyzer or exported
API is introduced. The earlier `lexdiv_ambiguity_review()` workflow continues
to handle single-token lexical candidates.

```{r morphology-spans-authored}
source(system.file("examples", "junimorph-spans.R", package = "ldfreq"))
source(system.file("examples", "junimorph-spans-demo.R", package = "ldfreq"))
span_demo <- junimorph_spans_example
span_demo$reviewed$occurrences[, c("pre", "keyword", "post", "start", "end",
  "token_from", "token_to", "candidate_count", "status")]
span_demo$counts
```

This example selects two occurrences, leaves a repeated `食べられる` unresolved,
keeps `開ける` unreviewed, and identifies an invented form with no supplied
candidate. A requested but absent form and an empty document are retained in
the term/document summaries. On the **same two selected spans**, two forms map
to one lemma label. The complete five-target lemma total is unavailable. All
26 original annotation tokens remain unchanged before and after the review.
These are illustrative decisions, not an accuracy assessment.

Unlike the earlier quanteda token-window example, this span recipe's `window`
is a number of **original Unicode codepoints** on either side. KWIC includes
original spaces and line endings. Matches cannot cross a supplied segment.
Exact substring matches whose endpoints fall inside analyzer tokens remain
visible as `boundary_mismatch`; lexical decisions on them are rejected until
the analyst supplies and imports a complete reviewed segmentation.

All overlapping matches remain present. `overlap_n` and
`selected_overlap_n` expose overlapping source ranges; their counts cannot be
summed as a count of independent words. The recipe never applies longest-match
selection, merges token rows, removes exclusions or automatically selects a
single candidate. Its document summaries count diagnostic spans, not word tokens.

### Connect a locally acquired J-UniMorph table

[J-UniMorph](https://github.com/cl-tohoku/J-UniMorph) supplies a three-column
`jpn` table (lemma, form, feature bundle) under CC BY 4.0. Its filtered table
and the pre-filtering hit-count file are different inputs. Obtain `jpn`
separately; the package includes only the reader and authored examples.

```{r junimorph-local-input, eval=FALSE}
reference <- read_junimorph("local-resources/jpn", list(
  resource_id = "J-UniMorph",
  resource_version = "jpn git blob 6b001fa9d33139bd9b85ffe70f74c1143ae9df24",
  source_reference = "https://github.com/cl-tohoku/J-UniMorph",
  data_license = "CC BY 4.0"
))
# Use the actual revision of your file; its SHA-256 is recorded automatically.
# ja is the complete annotation import above, before any filtering.
span_review <- review_morphology_spans(ja, reference)
span_review$documents
span_review$candidates[span_review$candidates$form == "食べられる", ]

# After inspecting an occurrence and the corresponding candidate table:
decision <- span_review$occurrences[chosen_row, c("review_id", "occurrence_id")]
decision$status <- "selected"             # or "unresolved"
decision$candidate_id <- chosen_candidate_id # NA_character_ for unresolved
decision$reviewer <- "analyst-id"
decision$reason <- "Record the contextual evidence or reason for withholding judgment"
reviewed <- review_morphology_spans(ja, reference, decisions = decision)
saveRDS(reviewed, "morphology-span-review.rds")
restored <- readRDS("morphology-span-review.rds")
replayed <- review_morphology_spans(restored$source, restored$reference,
  targets = restored$terms$form, decisions = restored$decisions,
  window = restored$provenance$window_codepoints)
stopifnot(identical(restored, replayed))
```

Source row IDs, original feature strings, duplicate records and file identity
are preserved. Changing the source annotations, reference table or target set
invalidates old decisions. Unreviewed and unresolved occurrences are distinct;
even one listed candidate requires an explicit selection. The workflow accepts
recorded decisions without claiming to verify their linguistic correctness.

The inspected snapshot has 12,687 rows, 107 lemma labels and 10,848 forms. It is
a limited inflection reference, not a comprehensive Japanese vocabulary list.
Matching is exact: kana/kanji variation, spelling errors and missing forms are
not silently normalized. A missing match is not an unknown-word diagnosis.
The example scans one form at a time and is intended for bounded local analyses.

In the 28-essay diagnostic subset described below, the full table produced
1,309 exact spans: 587 agreed with both supplied token boundaries, including
409 multi-token spans; 722 had a boundary mismatch. Of the aligned spans, 98
had multiple candidate rows. All real-corpus decisions remain unreviewed;
these counts measure matching behavior, not error rates or contextual accuracy.
Nested forms can describe overlapping text and shorter portions of an
inflectional sequence. Inspect the whole context before interpreting a hit.

## Read locally acquired learner essays

The [NINJAL Essay Database](https://mmsrv.ninjal.ac.jp/essay/essay_05.html)
distributes a text ZIP and separate essay/writer metadata workbooks. Obtain
these files directly from the provider under its **CC BY-NC-ND 4.0** terms.
They are not included in ldfreq. The following explicitly sourced reader needs
only the local ZIP and metadata tables; optional `readxl` reads the workbooks.

```{r ninjal-local-input, eval=FALSE}
source(system.file("examples", "ninjal-essays.R", package = "ldfreq"))
data_dir <- "ninjal-essay-data"
essays <- as.data.frame(readxl::read_excel(
  file.path(data_dir, "sakubun.xlsx"), col_types = "text"))
writers <- as.data.frame(readxl::read_excel(
  file.path(data_dir, "shipitsusha.xlsx"), col_types = "text"))

# Start with a small, explicitly selected subset of local files.
files <- head(essays[["日本語作文txt"]], 5)
corpus <- read_ninjal_essays(file.path(data_dir, "sakubun_txt.zip"),
  essays, writers, files = files, encoding = "UTF-8")
corpus$documents[, c("document_id", "writer_id", "task", "l1_reported",
  "writer_metadata_status", "source_encoding", "utf8_bom_removed")]

# With the optional analyzer configured above:
analysis <- analyze_unidic(corpus$segments, dic_dir)
# Import these complete annotations with the actual analyzer/dictionary
# metadata as above, before selecting lexical forms or excluding punctuation.
saveRDS(corpus, "ninjal-essays-local.rds")
```

The UTF-8 default is a declaration, not an encoding guess. For a mixed archive,
supply a named vector with each selected member's inspected encoding:

```{r ninjal-local-encoding, eval=FALSE}
files <- essays[["日本語作文txt"]]
encodings <- setNames(rep("UTF-8", length(files)), files)
# A local list of member names verified as CP932, not a list of error words.
cp932_files <- readLines(file.path(data_dir, "verified-cp932-members.txt"), encoding = "UTF-8")
stopifnot(all(cp932_files %in% files))
encodings[cp932_files] <- "CP932"
corpus <- read_ninjal_essays(file.path(data_dir, "sakubun_txt.zip"),
  essays, writers, files = files, encoding = encodings)
```

Do not silence a decoding failure by replacing unreadable bytes or guessing a
different encoding for every failure. The reader rejects failed conversions
and failed byte round-trips. It records a removed leading UTF-8 BOM separately;
all subsequent characters and line endings remain unchanged. Imported character
positions refer to that decoded text, not ZIP bytes. Raw member hashes and the
archive hash retain the connection to the downloaded files.

IDs come from the official metadata mapping, not filename slicing. A missing
writer row stays `missing` with an unavailable L1; duplicate join keys stop the
reader instead of duplicating essays. Source L1 labels, including nonresponse
labels, are retained verbatim. Collection country is not inferred to be L1.

In a local input check of the distribution downloaded on 2026-10-06, the ZIP
and essay workbook contained **1,777** matching files/records: 1,764 UTF-8 and
13 inspected CP932 files; 179 had a leading UTF-8 BOM. One essay lacked a writer
metadata match. These are observations of those downloaded files, not the
overview page's historical total or guarantees about future editions.

A diagnostic subset of 28 essays from one task (four per seven reported L1
labels, including Japanese) produced 10,932 supplied annotation rows. All source
spans and non-whitespace text were checked. The selected lexical population had
9,554 tokens, of which 139 had no lemma annotation. Missing lemma is not an error
rate or a verified unknown-word classification. The input contains complete
distributed TXT, including titles and transcription notation. This verifies
input, alignment and counting, not segmentation accuracy, kanji knowledge or
group differences. Keep corpus text and individual review tables locally;
the earlier authored examples provide the distributable workflow.

## Frequency resources and interpretation

The bundled `tubelex_profile()` and `nj8_profile()` resources are **English**;
do not treat unmatched Japanese tokens as rare English words. Japanese
[TUBELEX](https://github.com/naist-nlp/tubelex) has different surface/base/lemma
and dictionary-version variants. Sharing MeCab does not by itself establish
compatible keys: normalization, dictionary version, POS definition and
selection also matter. The [Japanese stimulus guide](japanese-stimuli.html)
connects reviewed `orthBase` keys to one separately acquired, pinned Japanese
TUBELEX table. It records denominators and key transformations without claiming
to reproduce upstream tokenization. Its unigram tables cannot supply n-gram
co-occurrences.

Surface frequency for a kana spelling and frequency for a reviewed lexical
entry answer different questions. Do not silently substitute the frequency of
its kanji spelling or sum homophones merely because they share a reading.
Use a compatible lexical-unit table or report that the requested frequency is
unavailable; a lexical selection does not create sense-specific frequencies.

To apply the same document workflow to the **separately acquired, pinned Japanese
TUBELEX base table**, first follow the acquisition instructions in
[Japanese stimulus selection](japanese-stimuli.html). Continue from the actual
`ja` import in the gibasa section; do not pass the teaching reference above as
TUBELEX. This example selects the whole source except auxiliary symbols and
whitespace, while retaining missing-base tokens in the full denominator.

```{r ja-frequency-tubelex, eval=FALSE}
source(system.file("examples", "japanese-frequency-documents.R", package = "ldfreq"))
source(system.file("examples", "japanese-tubelex.R", package = "ldfreq"))
t <- ja$tokens
anchors <- c("document_id", "segment_id", "token_index", "start", "end", "surface")
mask <- t[anchors]
mask$in_body <- TRUE
excluded <- (!is.na(t$POS1) & t$POS1 %in% c("補助記号", "空白")) |
  stringi::stri_detect_regex(t$surface, "\\p{White_Space}")
mask$retained <- !excluded
mask$reason <- ifelse(excluded, "auxiliary_symbol_or_whitespace", "retained")
profile <- japanese_document_profile(ja, mask,
  c("名詞" = "content", "動詞" = "content", "助詞" = "function", "助動詞" = "function"),
  condition = "whole-source-with-missing-bases")
items <- t[anchors]
items$item_id <- paste0("source-", seq_len(nrow(t)))
items$term <- t$surface
items$orth_base <- t$orthBase
absent <- is.na(items$orth_base) | !nzchar(stringi::stri_trim_both(items$orth_base)) | items$orth_base == "*"
items$orth_base[absent] <- NA_character_
items$base_reason <- ifelse(absent, "No dictionary orthBase assigned",
  "Recorded unidic-lite 1.0.8 orthBase; not an independent contextual judgment")
frequency_path <- file.path("japanese-resources", "tubelex-ja-base-pos.tsv.xz")
eligible <- nzchar(stringi::stri_trim_both(items$term))
items$base_reason[!eligible] <- "Whitespace token not sent to word-key lookup"
lookup <- profile_japanese_tubelex_items(items[eligible, ], frequency_path)
keys <- items[anchors]
keys$lookup_term <- lookup$items$tubelex_lookup_term[match(items$item_id, lookup$items$item_id)]
keys$key_reason <- items$base_reason
reference <- lookup$reference
reference$source <- lookup$source  # File hash, totals, normalization and versions.
frequency_documents <- japanese_frequency_documents(profile, keys, reference)
frequency_documents$documents
frequency_documents$norms$summary
saveRDS(list(frequency = frequency_documents, lookup = lookup), "japanese-frequency.rds")
```

The local reader now returns its complete inspected aggregate in `$reference`,
so document batches and later replay use the same table rather than a queried-only
excerpt. Keys receive the reader's declared NFKC/lowercase transformation;
original surfaces and positions stay unchanged. The majority POS remains
reference information, not a POS- or sense-specific frequency. Saving the full
reference consumes more space than saving a summary alone. No Japanese aggregate
or source corpus is added to the package, and normal examples never download it.

For an external Japanese norm table, prepare and check unique keys and use
`lexdiv_norm_profile()` with the table's actual metadata and missingness. For
n-gram frequencies, supply permitted local reference counts through the
[n-gram workflow](ngram-profiles.html). Neither approach requires redistributing
the underlying corpus. Access to a source does not itself establish a right
to redistribute it; check the terms of the actual artifact used.

Analyzing Japanese learner texts, native-speaker texts, or instructional
materials is possible with this input path. Software alignment tests do not
validate learner-error tokenization, lexical sophistication, or equivalence
between English and Japanese scores. Evaluate those questions against suitable
annotated data and external outcomes. Use `lexdiv_align_annotations()` for
source-span comparisons involving splits and merges; the label comparison in
`lexdiv_compare_annotations()` requires identical token rows. Span alignment
alone does not establish which segmentation is linguistically correct.
