---
title: "Preprocessing, parameter sensitivity, and frequency coverage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preprocessing, parameter sensitivity, and frequency coverage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
library(ldfreq)
```

## What `ldfreq` adds to the R ecosystem

R already has mature lexical-diversity implementations. In particular,
`quanteda.textstats::textstat_lexdiv()` integrates with `quanteda` tokens and
document-feature matrices, while `koRpus::lex.div()` includes raw-character,
lemma, HD-D, MTLD, and MTLD-MA workflows. `ldfreq` does not claim that these
metrics were absent from R.

The narrower contribution of `ldfreq` is an auditable measurement boundary:

| Question | `ldfreq` representation |
|---|---|
| Which formula and boundary variant ran? | `method_id` and metric contract |
| Which window, segment, or sample was requested? | requested and effective parameter list-columns |
| Was a short-text request computable? | structured `status` and `missing_reason` |
| Did a screen change or censor a value? | no; screens are separate rows |
| Which raw-text decisions created the units? | preprocessing contract and token audit |
| Which resource produced a frequency value? | version, manifest hash, formula, and lookup identity |
| How much of the text matched that resource? | adjacent token/type coverage |

This makes `ldfreq` complementary to broad text-processing ecosystems rather
than a replacement for them.

## Why preprocessing is a measurement decision

Caltabellotta, Van Steendam, Noreillie, and Peters (2026,
<https://doi.org/10.1016/j.asw.2026.101039>) found that lemmatization and the
inclusion of all words versus content words materially changed MATTR and
frequency-based sophistication results. Their empirical best-performing choice
in one English/French study should not be converted into a universal software
default. `ldfreq` therefore exposes the alternatives and their coverage.

```{r}
tokenization <- lexdiv_tokenize("Cats and cat ran run.")
annotated <- lexdiv_lemmatize(
  tokenization,
  lemmas = c("cat", "and", "cat", "run", "run"),
  upos = c("NOUN", "CCONJ", "NOUN", "VERB", "VERB"),
  backend_id = "vignette-fixture",
  backend_version = "1",
  upos_backend_id = "vignette-upos-fixture",
  upos_backend_version = "1"
)

surface <- lexdiv_metrics_text(tokenization, metrics = "ttr")
lemma_all <- lexdiv_metrics_text(
  annotated,
  unit = "lemma",
  metrics = "ttr"
)
lemma_content <- lexdiv_metrics_text(
  annotated,
  unit = "lemma",
  word_inclusion = "content",
  metrics = "ttr"
)

data.frame(
  operationalization = c("surface/all", "lemma/all", "lemma/content"),
  N = c(surface$results$N, lemma_all$results$N, lemma_content$results$N),
  V = c(surface$results$V, lemma_all$results$V, lemma_content$results$V),
  TTR = c(
    surface$results$value,
    lemma_all$results$value,
    lemma_content$results$value
  )
)
```

The point is not that TTR is preferred. The example demonstrates that the
lexical unit and denominator change the estimand before any diversity formula
is selected.

### Exact content-word overlap between two texts

The same annotation boundary supports type-set overlap. A second text must be
tokenized and annotated explicitly; `ldfreq` does not guess its UPOS tags or
lemmas.

```{r}
comparison <- lexdiv_lemmatize(
  lexdiv_tokenize("Cats and birds ran."),
  lemmas = c("cat", "and", "bird", "run"),
  upos = c("NOUN", "CCONJ", "NOUN", "VERB"),
  backend_id = "vignette-fixture",
  backend_version = "1",
  upos_backend_id = "vignette-upos-fixture",
  upos_backend_version = "1"
)

content_overlap <- lexdiv_content_overlap(
  annotated,
  comparison,
  unit = "lemma",
  details = "terms",
  document_ids = c("essay", "reference")
)
content_overlap$summary[, c(
  "measure_id", "value", "numerator", "denominator", "status"
)]
content_overlap$coverage
content_overlap$shared_terms
```

Because `details = "terms"` retains exact lexical strings, the result provenance
and print method disclose their presence. The default comparability policy is
fail-closed unless an analysis explicitly selects `mismatch = "warn"` or
`"allow"`. The comparison is unit-relevant: surface uses tokenizer and UPOS
settings; lemma also uses its annotation method/backend; flemma instead uses
the fixed flemma adapter/parser and declared resource/override settings while
ignoring lemma backend identity because the flemma adapter consumes surface
forms. UPOS labels are always supplied separately from lemma labels. Strict
flemma comparison requires present, equal caller-declared resource versions.
Two inputs with no overrides are comparable on the override setting; otherwise
both need the same declared override version. The four-column comparability
table discloses label values and match status but does not compare file bytes or
content hashes.

All five rows use distinct exact types, but not the same denominator. Jaccard
uses the union, Dice uses the sum of set sizes, the two coverage measures use A
or B, and the overlap coefficient uses the smaller set. A strict subset can
therefore have overlap coefficient 1 while Jaccard remains below 1. Missing
UPOS, non-content UPOS, and missing selected units are counted separately, so
the coverage table is part of the interpretation rather than an optional
quality label.

Changing `unit = "lemma"` to `"surface"` changes the estimand and may change
the value. None of these exact lexical-sharing measures detects plagiarism or
directly measures semantic similarity, coherence, proficiency, or writing
quality.

### Directional reference coverage across documents

When the research question is how much of several documents belongs to one
reference vocabulary, pairwise Jaccard or Dice values introduce the wrong
denominator. `lexdiv_reference_coverage()` instead retains both the number of
matched document items and the full document denominator.

```{r}
coverage_documents <- list(
  essay_a = c("cat", "cat", "run", "rare"),
  essay_b = c("bird", "run", "unknown")
)
reference_coverage <- lexdiv_reference_coverage(
  coverage_documents,
  reference = c("cat", "bird", "run"),
  reference_id = "analysis_list",
  details = "terms"
)
reference_coverage$summary[, c(
  "document_id", "weighting", "value", "numerator", "denominator", "status"
)]
reference_coverage$terms
```

The token row counts every repeated document occurrence; the type row counts
each distinct document term once. Repeating an entry in the reference changes
neither value. The function does not normalize or annotate terms, so comparable
surface, lemma, or flemma vectors must be constructed explicitly upstream.
Term detail is opt-in because it copies exact lexical strings into the result.

### Flemma as a caller-supplied resource transformation

For NWLC-oriented sensitivity analysis, `lexdiv_flemmatize()` reads a local
AntBNC lemma-list file supplied by the analyst. The payload is never bundled or
downloaded. Unknown forms use normalized-surface identity fallback so they
remain visible as off-list items, and explicit overrides take precedence over
the raw AntBNC mapping.

```{r}
antbnc_fixture <- tempfile(fileext = ".txt")
writeLines(
  c(
    "cat\t->\tcat\tcats",
    "run\t->\tran\trun",
    "the\t->\tthe"
  ),
  antbnc_fixture,
  useBytes = TRUE
)

flemma_annotations <- lexdiv_flemmatize(
  tokenization,
  antbnc_fixture,
  resource_version = "synthetic-vignette-fixture"
)
unlink(antbnc_fixture)

flemma_annotations$tokens[, c(
  "surface", "flemma", "flemma_matched", "flemma_match_rule"
)]
lexdiv_metrics_text(flemma_annotations, unit = "flemma", metrics = "ttr")
```

For a real analysis, replace the synthetic fixture with a legitimately
obtained local AntBNC file and declare its version. The adapter and parser
identities are fixed by `ldfreq`; `resource_version` and, when overrides are
non-empty, `override_version` are caller labels rather than inferred content
identities. They are disclosed in provenance and overlap comparability output,
so use path-free, non-sensitive labels and never put secrets or private hashes
in them. The local file name and content hash are not retained in public
provenance; a source-byte digest is used only for caching and is never copied
to public provenance or results.

This is an explicit approximation boundary. NWLC reports using a manually
modified AntBNC mapping aligned to its selected word lists. Raw AntBNC can map
a surface form that is an independent New JACET headword to a different family
lemma. `nj8_profile()` records these conflicts and lets an analysis
retain AntBNC, prefer the word-list headword, or stop for explicit review.

### Resource-defined word families

**Choose a family definition before counting.** A lemma usually links inflected
forms; a derivational family may also include words such as `reusability`.
Flemma and word family are not interchangeable. Bauer and Nation
(1993, <https://doi.org/10.1093/ijl/6.4.253>) describe graded inclusion criteria;
an inventory's inclusion level is not a frequency band or evidence that a
learner knows every member.

Experimental `lexdiv_family_profile()` accepts a plain form–family table and
complete `lexdiv_import_annotations()` output. It needs no corpus download,
Python, model or additional dependency. This offline example contains **authored
decisions**, not verified entries from Nation's lists:

```{r}
family_env <- new.env(parent = baseenv())
sys.source(system.file("examples", "word-families.R", package = "ldfreq",
  mustWork = TRUE), family_env)
family_example <- family_env$word_families_example
families <- family_example$profile
families$dictionary
families$documents[c("document_id", "selected_tokens", "matched_tokens",
  "unresolved_tokens", "family_types", "family_ttr", "status")]
family_example$comparison
```

In `use uses reusability use.`, excluding supplied `PUNCT` leaves four tokens.
They contain three surface types, two authored lemmas/flemmas and one authored
family: TTR is .75, .50 and .25 respectively. Every comparison uses the same
four ordered occurrences. The window of three for MATTR is only a short
arithmetic illustration, not a recommended setting for research.

The `bank quux.` document is incomplete: `bank` has two candidate families and
`quux` is unlisted. Neither is silently assigned a family, and complete family
types/TTR are `NA`. Inspect original context and both candidates:

```{r}
families$occurrences[families$occurrences$status %in% c("ambiguous", "unlisted"),
  c("document_id", "segment_id", "token_index", "pre", "keyword", "post", "status")]
families$candidates[families$candidates$document_id == "unresolved", ]
families$members
```

#### Use Nation's bundled BNC/COCA inventory

`bnccoca_data()` returns Nation's **Level 6, Version 1.0.0** family lists
under **CC BY-SA 4.0**, including attribution and ShareAlike conditions.
Its `dictionary` contains 75,679 headword/member rows in 25,000 families;
each of the 25 frequency/range bands contains 1,000 families.
The source is [Nation's official resource page](https://www.wgtn.ac.nz/lals/resources/paul-nations-resources/vocabulary-analysis-programs).
The source spellings and memberships are preserved, including uppercase keys.

```{r}
sys.source(system.file("examples", "bnccoca-families.R", package = "ldfreq",
  mustWork = TRUE), family_env)
nation <- family_env$bnccoca_example
nation$profile$documents[c("document_id", "selected_tokens", "matched_tokens",
  "unresolved_tokens", "family_types", "family_ttr", "status")]
nation$candidates[c("document_id", "headword", "frequency_band")]
```

The example explicitly uses `normalization = "nfkc_lower"`. Its first text,
`use uses colour color`, contains four tokens and two source families.
The second, `use reusability`, has an unlisted form: **REUSABILITY is absent
from this snapshot**, even though the earlier authored example assigns it to
USE. Complete family types and TTR therefore remain `NA`. Inspect the original
KWIC, rather than inventing a family or interpreting noncoverage as an error:

```{r}
nation$profile$occurrences[nation$profile$occurrences$status == "unlisted",
  c("document_id", "pre", "keyword", "post", "status")]
```

`reference <- bnccoca_data()` exposes `reference$dictionary` and
`reference$resource` directly for `lexdiv_family_profile(annotations,
reference$dictionary, reference$resource, normalization = "nfkc_lower")`.
The result retains the table, source IDs, citation and license. Save the whole
`reference` as well to retain conversion metadata and supplementary lists.
The installed example joins frequency bands by **record ID**; matching a
candidate is not a corpus frequency estimate or a learner knowledge score.

Inclusion **Level 6** and frequency **band 6** are different variables.
The Level 3 partial lists are a separate resource, not obtainable by filtering
these frequency bands. The first bands also reflect pedagogical decisions.
Each member does not have a separate Bauer–Nation affix-level annotation.

`reference$supplementary` keeps proper names, marginal words, transparent
compounds and acronyms separate, with `NA` frequency bands. Membership in a
proper-name list does not label the use of that string in context. Placeholder
slots 26–30 and Range software are excluded. Any explicit combination with
the core dictionary needs its own declared scope and candidate checks.
See `?bnccoca_data` and the installed `licenses/bnccoca/NOTICE.md`.

#### Compare counting units without hiding unlisted words

Family membership and dictionary coverage answer different questions. A smaller
type count can reflect grouping words into families, or dropping words that the
inventory does not list. Compare units on **the same occurrence IDs**, while
reporting coverage of the original selection separately.

The following explicitly sourced recipe uses the authored lemmas in the example
above. It runs no analyzer and adds no exported API. Surface and lemma types use
the supplied strings verbatim; case/spelling decisions should be made explicitly
upstream. Family lookup retains its declared `nfkc_lower` normalization.

```{r}
sys.source(system.file("examples", "family-count-comparison.R", package = "ldfreq",
  mustWork = TRUE), family_env)
counts <- family_env$compare_family_counts(nation$profile)
counts$documents[c("document_id", "selected_tokens", "matched_tokens",
  "unlisted_tokens", "ambiguous_tokens", "missing_lemma_tokens",
  "common_tokens", "token_coverage")]
counts$comparison[c("document_id", "scope", "unit", "N", "V", "ttr", "status")]
```

For `use uses colour color`, all four occurrences are covered: surface, lemma,
and family types are 4, 3, and 2. `colour` and `color` remain distinct supplied
lemmas but share a source family. For `use reusability`, the full selection has
two tokens and an unavailable family total/TTR. The `common_resolved` comparison
contains only `use`: N = 1 and TTR = 1 for each unit. **That conditional value is
not the family TTR of the original two-token text.**

`all_selected` keeps each unit's unavailable totals as `NA` when its assignments
are incomplete. `common_resolved` is the intersection of selected occurrences
with a known lemma and an assigned family. `observed_types` counts known values
without claiming a complete total. Empty selections have zero types and undefined
TTR. The recipe also accepts an explicit `selected` logical vector in profile row
order, limited to the profile's known-selected rows; its returned `selection`
keeps document/segment/token IDs. Do not reuse a vector after reordering rows.

Keep unresolved occurrences available for inspection:

```{r}
o <- counts$profile$occurrences
o[counts$selection$selected & o$status != "matched",
  c("document_id", "token_index", "pre", "keyword", "post", "status")]
```

Use this subset for type counts and TTR only. Removing unresolved occurrences
does not justify closing gaps for moving-window measures or n-grams. A string
match is not an accuracy score, and supplementary-list membership does not
establish a proper noun in context. Save the full result and the reference:

```{r, eval=FALSE}
saveRDS(list(counts = counts, reference = nation$reference,
  session = sessionInfo()), "family-comparison.rds")
saved <- readRDS("family-comparison.rds")
recounted <- family_env$compare_family_counts(saved$counts$profile,
  selected = saved$counts$selection$selected)
stopifnot(identical(recounted, saved$counts))
```

This workflow was also checked locally using 140 original essays from
**ICNALE GRA V2.1**, retaining 31,902 saved lexical tokens and their lemmas.
Document counts and assignments agreed with a separate direct dictionary lookup;
30,892 tokens matched the core inventory. This is an implementation check on a
fixed input, not independent validation of contextual family assignments.
ICNALE texts and individual records are not included in the package; the example
above uses authored text. Dataset source: Ishikawa, S. (2024), *The ICNALE Global
Rating Archives (V2.1)*, Kobe University.

#### Supply a different inventory

For a different inventory, prepare character columns `record_id`, `form`, `family_id`
and optionally `upos`. Repeated forms are allowed; assign only when all matching
records agree on one family. Optional POS matching requires exact UD tags.
If your inventory contains inclusion levels or alternative definitions, select
the intended rows explicitly **before** lookup and record that rule in
`family_definition`. Preserve extra level/band columns for inspection.

```{r, eval=FALSE}
dictionary <- read.csv("my-family-table.csv", colClasses = "character",
  fileEncoding = "UTF-8")
resource <- list(resource_id = "my-family-table", resource_version = "1",
  language = "en", source_reference = "Full inventory citation",
  data_license = "Terms applying to this locally obtained table",
  family_definition = "Exact inclusion criteria and filtering used",
  lookup_unit = "surface")
families <- lexdiv_family_profile(annotations, dictionary, resource,
  unit = "surface", normalization = "identity", exclude_pos = c("PUNCT", "SYM"))
saveRDS(families, "family-profile.rds")
restored <- readRDS("family-profile.rds")
replayed <- lexdiv_family_profile(restored$annotations, restored$dictionary,
  restored$provenance$resource, unit = restored$provenance$unit,
  normalization = restored$provenance$normalization,
  exclude_pos = restored$provenance$exclude_pos,
  context_chars = restored$provenance$context_chars, review = restored$review)
stopifnot(identical(restored, replayed))
```

Use `normalization = "nfkc_lower"` only when NFKC, trimming and English
lowercasing fit the inventory. `unit = "lemma"` or `"flemma"` uses an already
supplied annotation column; it does not run an analyzer. Family IDs remain
unchanged. Missing POS makes selection unknown when exclusions are requested;
`conditional_token_coverage` then describes known-selected tokens only, while
whole-selection coverage is unavailable. Read coverage beside the unresolved
counts, not as an annotation-accuracy estimate.

For other metrics, the installed example passes ordered family IDs from a
**complete** document to `lexdiv_metrics()`. Never remove unresolved tokens and
join the survivors for MATTR. The pooled summary is also not an average of
document TTRs. Do not resolve a context-dependent form by changing the global
table for all its occurrences. Affix decomposition, inventory licensing and
empirical validation remain separate tasks.

#### Apply contextual judgments to individual occurrences

When a declared inventory distinguishes two families for the same spelling,
`review_input` prepares the existing `lexdiv_ambiguity_review()` workflow.
Use `review =` to apply the resulting judgments to the family profile. This
optional review preparation uses quanteda; basic family lookup and replay of
saved judgments do not require loading quanteda or running a language model.

```{r}
if (requireNamespace("quanteda", quietly = TRUE)) {
  sys.source(system.file("examples", "word-family-review.R", package = "ldfreq",
    mustWork = TRUE), family_env)
  contextual <- family_env$word_family_review_example
  contextual$reviewed$occurrences[
    contextual$reviewed$occurrences$surface == "bank",
    c("segment_id", "pre", "keyword", "post", "lookup_status",
      "status", "family_id", "review_record_id", "reviewer", "reason")]
  contextual$reviewed$documents[c("document_id", "selected_tokens",
    "review_selected_tokens", "unresolved_tokens", "family_types", "family_ttr")]
}
```

The authored sentences are `The bank lends money.` and `The river bank floods.`
The two occurrences receive different families under this **illustrative**
definition. Excluding punctuation still leaves eight tokens. Before review,
two occurrences are ambiguous and complete family TTR is unavailable. After
the two supplied judgments, seven families give TTR = 7/8. The original
dictionary and automatic lookup remain unchanged in the saved result.
These choices do not assert that every inventory separates the two meanings,
or that grouping by meaning and grouping by derivation are the same operation.

For your own profile, use its prepared inputs without editing their candidate
set or resource snapshot. Decisions identify **record IDs**, not family IDs:

```{r, eval=FALSE}
stopifnot(length(families$review_input$targets) > 0)
review <- do.call(lexdiv_ambiguity_review,
  c(list(x = families$annotations), families$review_input))
write.csv(review$occurrences, "family-decisions.csv", row.names = FALSE, na = "")
# Review the source/context and fill status, candidate_id, reviewer and reason.
# Use selected + a record ID, or unresolved + an empty candidate_id.
# Omit unreviewed rows from this decision file; source tokens remain retained.
decisions <- read.csv("family-decisions.csv", colClasses = "character",
  na.strings = "", check.names = FALSE)
review <- do.call(lexdiv_ambiguity_review,
  c(list(x = families$annotations), families$review_input, list(decisions = decisions)))
p <- families$provenance
reviewed <- lexdiv_family_profile(families$annotations, families$dictionary,
  p$resource, unit = p$unit, normalization = p$normalization,
  exclude_pos = p$exclude_pos, context_chars = p$context_chars, review = review)
saveRDS(reviewed, "reviewed-families.rds")
```

Character CSV input preserves IDs such as `01` and literal `NA`; empty fields
are missing. A `selected` judgment applies only to that source occurrence.
An explicit `unresolved` judgment also withholds a previously automatic match;
unreviewed occurrences keep their lookup result. Report `review_selected_tokens`
and `review_unresolved_tokens` beside coverage, and inspect `lookup_status`
and `lookup_family_id` when explaining what changed. Original contexts and the
complete review, including judgment reasons, remain in `$review`.

The generic review display lists candidate records by original surface. If
the same surface has different POS/lemma annotations across occurrences,
consult `$candidates` using document/segment/token IDs to check eligibility.
A selected record for the wrong POS or lemma, or a decision on an excluded or
unknown-selection token, is rejected. Review cannot invent missing inventory
entries, fix annotations or silently reinclude excluded tokens.

Changes to the source annotations, dictionary contents/order, resource
definition, selected unit, normalization or exclusions invalidate the review
snapshot. Recreate the review and reassess judgments; do not copy old IDs into
a new snapshot. Context-window changes alone do not invalidate judgments.
Independent reviewers can use the same prepared input and compare their
results with `lexdiv_compare_ambiguity()` before applying an adjudicated review.
This preserves disagreement; it does not establish independent judgment or
linguistic correctness by itself.

### Source-linked roots and affixes

The installed `word-parts.R` recipe counts **supplied analyses** at original
occurrences. It is an explicitly sourced example helper, not an exported API
or an automatic morphology analyzer. It needs no Python, model, network or
quanteda. Family membership remains a separate input and is not inferred.

```{r}
parts_env <- new.env(parent = baseenv())
sys.source(system.file("examples", "word-parts.R", package = "ldfreq",
  mustWork = TRUE), parts_env)
sys.source(system.file("examples", "word-parts-demo.R", package = "ldfreq",
  mustWork = TRUE), parts_env)
word_parts <- parts_env$word_parts_example$profile
word_parts$occurrences[c("surface", "status", "analysis_id",
  "observed_root_occurrences", "observed_prefix_occurrences",
  "observed_suffix_occurrences", "observed_inflection_occurrences",
  "observed_derivation_occurrences")]
word_parts$documents[c("document_id", "selected_tokens", "complete_coverage",
  "observed_affix_occurrences", "affix_occurrences", "affixes_per_token", "status")]
```

These are **authored teaching examples**, not imported MorphoLex or Nation
entries. The first document contains five selected tokens. `teachers` contributes
one root, derivational `-er` and inflectional `-s`; comparative `-er` in `smaller`
has a different ID. The five tokens contain seven affix occurrences, so affixes
per token is 7/5. The proportion of tokens containing affixes is 1. These are
different denominators. The alternate authored analysis leaves `transmit`
undivided: inspect `parts_env$word_parts_example$alternative` to see six affix
occurrences and five root types, compared with seven and four in the first
analysis. Neither choice establishes what a learner knows or mentally decomposes.

The second document retains ambiguous `bank`, explicitly unanalysed `went`,
and unlisted `quux`; its complete totals are unavailable, not zero. Empty
documents retain zero counts and undefined proportions. A `partial` analysis
contributes its known parts to `observed_*` counts but leaves full totals
unavailable. `complete` is always relative to the declared **analysis scope**:
a complete derivational analysis can omit inflection by design.

Prepare two plain data frames with character identifiers:

| Table | Required columns | Meaning |
|---|---|---|
| `analyses` | `analysis_id`, `form`, `completeness` | Unique analysis IDs; exact surface forms; `complete`, `partial` or `unanalysed`. Optional `upos` requires exact UD POS matching. Multiple eligible analyses remain ambiguous. |
| `parts` | `analysis_id`, `part_index`, `part_id`, `canonical`, `role`, `process`, `boundness` | One row per part, including repeated parts. Roles: `root`, `prefix`, `suffix`. Processes: `none` for roots; `inflection`, `derivation`, `unknown` for affixes. Boundness: `free`, `bound`, `unknown`; affixes must be bound. |

`part_index` is consecutive display order **within each analysis**, not an
attachment sequence or a derivation tree. A complete analysis must contain a
root and classify its affix processes. Rootless or other unsupported analyses
remain partial/unanalysed. Multiple roots are permitted. Parts with a shared ID
must have consistent canonical form and classification. Homographic affixes
with different functions need different IDs when the source distinguishes them.
Additional metadata columns are retained; they are not interpreted as trees,
validated substring coordinates, or evidence of semantic transparency.

The authored `mit/miss` example uses one root ID with a retained `realization`
column. This expresses an explicit analytical choice, not a rule that unifies
similar strings. A **base** can already contain several parts before another
affix is attached; this recipe counts listed parts and does not reconstruct
those formation stages. Nonaffixal processes such as conversion or irregular
inflection are not diagnosed by a zero affix count.

The required `resource` declarations are `resource_id`, `resource_version`,
`language`, `source_reference`, `data_license`, `analysis_scope`, and
`analysis_basis`. `word_parts_profile(annotations, analyses, parts, resource,
exclude_pos = character(), context_chars = 30L, max_rows = 1e6)` accepts complete
`lexdiv_import_annotations()` output. Matching is exact and case-sensitive;
there is no lemma fallback or silent normalization. Missing POS makes exclusion
unknown, and whole-selection coverage is then unavailable. Row ceilings apply
to each reference table, source tokens and candidate/part expansion, not total
memory. Keep inputs scoped to the analysis at hand.

```{r}
# Full tables remain available beside aggregate results.
word_parts$part_occurrences
word_parts$frequency
word_parts$occurrences[word_parts$occurrences$status == "ambiguous",
  c("document_id", "segment_id", "token_index", "pre", "keyword", "post")]
word_parts$candidates
```

`frequency` reports observed part occurrences by document and part ID; absent
rows are not assertions of complete zero counts. Read it with `documents`.
Alternative candidates do not contribute to frequency until an analysis is
uniquely eligible. This recipe supports inspection, but does not yet apply
occurrence-specific reviewer decisions or automatically choose a candidate.

```{r, eval=FALSE}
saveRDS(word_parts, "word-parts.rds")
restored <- readRDS("word-parts.rds")
stopifnot(identical(do.call(parts_env$word_parts_profile, restored$inputs), restored))
# For caller-prepared CSV input, preserve IDs and distinguish literal NA from missing.
analyses <- read.csv("analyses.csv", colClasses = "character", na.strings = "",
  fileEncoding = "UTF-8", check.names = FALSE)
parts <- read.csv("parts.csv", colClasses = "character", na.strings = "",
  fileEncoding = "UTF-8", check.names = FALSE)
parts$part_index <- as.numeric(parts$part_index) # validated by the recipe
```

#### Different resources answer different morphological questions

Nation's lists group forms into educational families. MorphoLex provides
segmentation and morphological variables. [MorphyNet](https://github.com/kbatsuren/MorphyNet)
separates derivational source–target relations from inflectional lemma–form
relations. Its English derivational table records `transmit` → `retransmit`
with prefix `re`, and `transmit` → `transmitter` with suffix `er`. A relation
is not a complete word segmentation; graph connectivity does not establish a
Nation family. MorphyNet's 15 languages do not include Japanese. It uses
CC BY-SA 3.0. A nine-row teaching excerpt is bundled; the full file is read
from a separately obtained local copy as shown below.

For Japanese, [J-UniMorph](https://github.com/cl-tohoku/J-UniMorph) is a
CC BY 4.0 candidate for **inflectional features**, not an English-style
educational family inventory. Its filtered `jpn` snapshot has 12,687 records
for 107 lemmas. For example, `食べられる` has potential, passive and honorific
analyses; `開ける` also permits distinct lemma analyses. Keep the original
features and multiple candidates. These forms can span several UniDic short
units, so matching them requires an explicit source-span or whole-form unit.
Joining by a single short-unit token would change the question.

Neither resource is a contextual analyzer supplied by ldfreq. Japanese
inflection, productive derivation, compound structure and homophony require
different relations and evidence. UniDic lemmas or shared kanji alone do not
define a derivational family; verb forms in J-UniMorph do not constitute a
complete Japanese vocabulary or a frequency norm. See Matsuzaki et al. (2024),
[J-UniMorph](https://aclanthology.org/2024.sigmorphon-1.2/), for its scope.

#### Inspect MorphyNet formation relations

`morphynet_read_derivations()` reads the official six-column derivational TSV
in R, preserving original words, POS, affixes and line IDs. It needs no Python,
model or extra package. For the complete English v1 reference, obtain
[`eng.derivational.v1.tsv`](https://github.com/kbatsuren/MorphyNet/blob/main/eng/eng.derivational.v1.tsv)
from the official repository, retain its CC BY-SA 3.0 terms, and read it:

```{r, eval=FALSE}
reference <- morphynet_read_derivations("eng.derivational.v1.tsv",
  language = "en", resource_version = "English v1")
reference$resource$source_sha256
saveRDS(reference, "morphynet-reference.rds")
```

The English file checked for this guide has 225,131 relations and SHA-256
`5920edacc1888b14464fc5cd96beea0a721221d56d1dc0e49de22f4c7c537c50`.
Language/version are your declarations; the reader does not authenticate every
possible supplied file. It rejects invalid formats and retains a hash of the
actual input. POS symbols remain unchanged, including `U`; the source row
`teach -> teacher` has N/N tags, not an automatically corrected V/N pair.
Supplied relations can include multiword fields, which are retained literally.

The installed example uses **nine unchanged rows** selected for six words,
with a mapping to their original source lines. This small excerpt is not a
reference for estimating vocabulary coverage. The contexts and reviewer
decisions are authored illustrations:

```{r}
morphynet_env <- new.env(parent = baseenv())
sys.source(system.file("examples", "morphynet-relations.R", package = "ldfreq",
  mustWork = TRUE), morphynet_env)
morphynet <- morphynet_env$morphynet_example
morphynet$reference$relations[
  morphynet$reference$relations$target_word == "reusability",
  c("source_word", "target_word", "source_pos", "target_pos", "morpheme", "affix_position")]
```

Reusability has incoming relations from `reuse` with `ability`, `reusable`
with `ity`, and `usability` with `re`. **Do not count these as three affixes in
one occurrence.** They are separate formation relations that can coexist.
Nor does their presence supply missing membership in Nation's inventory.

The example builds a candidate table by matching each exact `target_word` to
the requested surface, then uses the existing review API when quanteda is
available. Source/target words and POS remain visible beside the relation ID.
A study must declare what selecting one relation means; context alone does
not guarantee a unique morphological derivation.

```{r}
if (!is.null(morphynet$reviewed)) {
  morphynet$reviewed$occurrences[c("segment_id", "pre", "keyword", "post",
    "candidate_count", "status", "reason")]
  morphynet$selected[c("segment_id", "surface", "source_word", "target_word",
    "morpheme", "affix_position")]
}
```

There are five targeted occurrences. The two instances of `reusability`
retain all three relations; the illustration selects `reusable + ity` for
one and withholds selection for the other. The single incoming relations of
`retransmit` and `transmitter` also remain unreviewed until explicitly chosen.
`teachers` has no candidate in this excerpt: that is not an affix count of
zero, a lack of inflection or evidence that a learner made an error. The
complete source token/segment roster, including the empty document, is retained.
Review summaries describe the declared targets, not all corpus tokens.

```{r, eval=FALSE}
saveRDS(morphynet, "morphynet-review.rds")
restored <- readRDS("morphynet-review.rds")
replayed <- lexdiv_ambiguity_review(restored$annotations, restored$targets,
  restored$candidates, restored$resource,
  decisions = restored$reviewed$decisions)
stopifnot(identical(replayed, restored$reviewed))
```

Reading saved tables requires no quanteda; regenerating KWIC review does.
The review API currently matches whitespace-free single-token surfaces.
Do not delete spaces from multiword targets or silently substitute lemmas to
force a match. Case normalization, lemma-based alignment, complete derivation
trees and validated morphological-productivity estimates are separate tasks.
For adaptation or redistribution, retain source attribution and applicable
ShareAlike terms; see `licenses/morphynet/NOTICE.md`.

#### Use the bundled MorphoLex reference

MorphoLex-en supplies canonical segmentation and root/affix variables
(Sánchez-Gutiérrez et al., 2018, <https://doi.org/10.3758/s13428-017-0981-8>).
The package includes all 34 worksheets (68,624 word records and the aggregate
prefix/suffix/root tables), plus the original data dictionary, from
<https://github.com/hugomailhot/MorphoLex-en>. **These data use CC BY-NC-SA 4.0,
not the MIT license of the R code.** Noncommercial sharing and adaptation
require attribution and the applicable ShareAlike conditions; commercial use
is not granted by that license. See the installed `licenses/morpholex/NOTICE.md`.

```{r}
morpholex <- morpholex_data(c("0-1-1", "All roots"))
word_rows <- morpholex$sheets[["0-1-1"]]
word_rows[word_rows$Word %in% c("teacher", "teachers"),
  c("Word", "MorphoLexSegm", "ROOT1_FamSize", "ROOT1_Freq_HAL")]
morpholex$provenance$data_license
```

Tables retain character values, including identifiers and missing cells.
Convert a numeric measure explicitly when needed, for example
`as.numeric(word_rows$ROOT1_FamSize)`. `morpholex_data()` without arguments
returns all sheets; `$provenance$sheet_catalog` lists their names and dimensions.
Two source `Word` cells (ELP IDs 42162 and 65908) are logical values and appear
as `TRUE` and `FALSE`; an exact lowercase query will not match them. The source
values are preserved, so any case normalization should be an explicit decision.
No `readxl`, Python, model, or internet connection is needed for bundled data.
The word-parts recipe can use the same snapshot:

```{r}
sys.source(system.file("examples", "morpholex-word-parts.R", package = "ldfreq",
  mustWork = TRUE), parts_env)
forms <- c("transmit", "transport", "transmission", "teacher", "teachers")
reference <- parts_env$read_morpholex_parts(
  sheets = c("0-1-0", "0-1-1", "1-1-0"),
  words = forms)
reference$analyses
reference$unlisted_words # absent from the selected sheets, not necessarily the workbook
source <- data.frame(document_id = "example", segment_id = "s",
  text = paste(forms, collapse = " "))
tokens <- data.frame(document_id = "example", segment_id = "s",
  token_index = seq_along(forms), surface = forms)
imported <- lexdiv_import_annotations(tokens, source,
  list(language = "en", analyzer = "authored", analyzer_version = "1",
    dictionary = "none", dictionary_version = "1", unit = "word", normalization = "none"))
profile <- parts_env$word_parts_profile(imported, reference$analyses,
  reference$parts, reference$resource)
profile$documents[c("root_occurrences", "affix_occurrences", "complete_coverage")]
```

These five source entries give five root occurrences and four affix occurrences.
Save the complete reference/profile pair to keep both inputs and results:

```{r, eval=FALSE}
saveRDS(list(reference = reference, profile = profile), "morpholex-profile.rds")
# Optional: read an explicitly obtained workbook instead; requires readxl.
local_reference <- parts_env$read_morpholex_parts(path = "MorphoLEX_en.xlsx",
  sheets = c("0-1-0", "0-1-1", "1-1-0"), words = forms)
```

The reader retains all columns of selected source rows, selected sheet names,
requested/missing words and a workbook SHA-256. It extracts recorded canonical
parts and checks them against the reported PRS signature and morpheme count.
Unsupported syntax or inconsistent counts remain `unanalysed`; the recipe
never repairs an entry by guessing. It does not validate linguistic accuracy
or the derivation tree, and preserves root boundness as unknown. Original POS
and numeric variables remain in `source_rows`; POS is not silently mapped to UD.

In the inspected workbook, `transmit` is undivided, `transport` is
`trans- + port`, and `transmission` uses canonical `transmit + -ion`.
Both `teacher` and `teachers` use `teach + -er`. Thus, its complete counts describe
**reported derivational segmentation, excluding inflection**. The whole-word
TUBELEX frequency, MorphoLex's cumulative root frequency/family size, and
observed root counts in a target corpus remain separate variables with their
own reference populations. Do not infer missing inflections, root meanings,
learner knowledge, or productivity from these counts.

#### Connect families and morphology at the same occurrences

To explain a vocabulary profile, keep family membership, recorded segmentation
and incoming formation relations alongside the **same source occurrence**.
The explicitly sourced recipe below reuses `lexdiv_family_profile()`, the
word-parts recipe and `lexdiv_ambiguity_review()`. It requires optional quanteda
in a UTF-8 R session. Text and decisions are authored; references are the
bundled Nation/MorphoLex snapshots and the nine-row MorphyNet teaching excerpt.

```{r morphology-link-example, eval=requireNamespace("quanteda", quietly=TRUE) && isTRUE(l10n_info()[["UTF-8"]])}
source(system.file("examples", "morphology-link.R", package = "ldfreq"))
source(system.file("examples", "morphology-link-demo.R", package = "ldfreq"))
linked <- morphology_link_example$reviewed
linked$occurrences[, c("surface", "family_status", "morpholex_lookup_status",
  "morpholex_review_status", "morphynet_candidate_count", "morphynet_review_status")]
linked$documents[, c("document_id", "eligible_N", "morpholex_selected_complete_N",
  "reviewed_suffix_instances", "full_suffix_instances", "selected_suffix_relations")]
```

The five occurrences include two instances of `reusability`. Each has three
MorphyNet candidates; the example selects `reusable + ity` for one occurrence
and withholds selection for the other. Only one relation enters the selected
relation table. The MorphoLex analysis chosen for `teacher` contributes one
recorded root and one derivational suffix. The `teachers` choice is withheld.
Full-population suffix totals remain `NA`; an unlisted or unreviewed occurrence
is not assigned zero affixes. The empty document remains in the output.

Use the outputs for different purposes:

| Output | Unit and use |
|---|---|
| `occurrences` | One row per eligible original token, with unchanged source positions/family assignments and separate lookup/review states |
| `documents` | One row per document; eligible denominator, resource match coverage, selected/withheld counts and explicitly reviewed part/relation totals |
| `reviews` | Complete KWIC reviews and candidate tables, including reviewer identities and reasons |
| `reviewed_parts` | Parts of explicitly selected **complete** segmentations, retaining root/prefix/suffix and process labels |
| `reviewed_relations` | Explicitly selected one-step relations, retaining original source/target words and POS; not a full affix decomposition |

The selection vector must identify eligible rows of the complete family input.
Context-only punctuation, numbers or other excluded chunks stay outside that
denominator. Repeated equal surfaces remain separate occurrences. A decision
for an excluded occurrence is rejected; no row is silently duplicated by a
one-to-many reference join. Candidate alternatives are kept in the review
tables, rather than summed as observed morphemes.

Morphology matching uses the supplied surface **exactly**, without a lemma
fallback or POS inference. The family profile retains its separately declared
lookup policy. The source annotations record any earlier case/spelling
transformation; attaching morphology does not change that processing. An
explicitly selected reference analysis is still not independently verified
linguistic truth, and MorphoLex's derivational scope still excludes inflection.

```{r morphology-link-save, eval=FALSE}
saveRDS(linked, "linked-morphology.rds")
restored <- readRDS("linked-morphology.rds")
stopifnot(identical(do.call(link_morphology_candidates, restored$inputs), restored))
# For your corpus, pass its existing family profile, an explicit token selection,
# the MorphoLex adapter output and your locally read complete MorphyNet reference:
linked_corpus <- link_morphology_candidates(family, morpholex_reference,
  morphynet_reference, selected = eligible_tokens)
# Inspect candidates and submit decisions using each review's occurrence/review IDs.
```

A local input check with **ICNALE GRA V2.1**, 140 original essays, reused 31,902
saved lexical tokens and preserved every source ID, position, family assignment
and document denominator. MorphoLex matched 30,690 occurrences, with 30,680
passing the adapter's complete-segmentation checks and 10 retained as
unanalysed. The full English MorphyNet v1 table matched 4,592 occurrences,
including 401 with multiple incoming relations; 4,348 occurrences matched all
three inventories. These are exact-surface lookup counts for resources with
different purposes, not comparative accuracy or learner morphology scores.
Real-corpus choices remain unreviewed. The restricted texts and individual
outputs are kept locally; the executable example above supplies public text.

## Length and parameter sensitivity

Bestgen (2024, <https://doi.org/10.1111/lang.12630>) distinguishes ordinary
text-length dependence from sensitivity to the reduction or window parameter.
Bestgen (2025, <https://doi.org/10.1016/j.rmal.2024.100168>) further shows why
MATTR is useful for within-text fluctuation but should not be treated as a
parameter-free universal score.

`lexdiv_plan()` and `lexdiv_profile()` preserve parameter variants as different
requests:

```{r}
long_tokens <- rep(
  c("one", "two", "three", "one", "four", "five", "one", "six"),
  20
)
plan <- lexdiv_plan(presets = "length_50_100")
profile <- lexdiv_profile(long_tokens, plan)
profile[profile$metric_id %in% c("mattr", "msttr"), c(
  "request_id", "metric_id", "value", "status", "N", "V"
)]
```

The whole-text mean alone hides the local trajectory and the fact that endpoint
positions occur in fewer complete windows. `lexdiv_mattr_profile()` reuses the
same MATTR specifications and exposes both quantities without retaining token
strings:

```{r, fig.width=5, fig.height=3}
methods <- lexdiv_methods()
mattr_method <- methods$method_id[methods$metric_id == "mattr"]
mattr_plan <- lexdiv_plan(
  presets = character(),
  grids = lexdiv_grid(
    mattr_method,
    "window_length",
    c(50, 100),
    request_id_prefix = "mattr"
  )
)
local_mattr <- lexdiv_mattr_profile(long_tokens, mattr_plan)
local_mattr$diagnostics
head(local_mattr$windows)
head(local_mattr$exposure)
plot(local_mattr, request_id = "mattr_1")
```

For each request, `nominal_observation_weight` sums to one. It describes the
moving-window accounting scheme, not the lexical importance of a token. The
local rows support custom graphics and downstream models, but do not constitute
an inferential stability test.

A downstream study may combine several indices, fit a PCA, or optimize a
window for a target population. Such sample-dependent modeling is not embedded
as a universal package score.

## Maas and MTLD are variant families, not single formulas

The version-pinned [TAALED 0.32 documentation](https://pypi.org/project/taaled/0.32/)
exposes distinctions that a shared label can hide. Its Maas calculation uses
the base-10, a-squared form, whereas the versioned `ldfreq` core uses natural logs
and a-squared. The base-10 value equals the natural-log value multiplied by
`log(10)`: rankings may be unchanged while coefficients, cutoffs, and raw-scale
comparisons are not.

For sequential MTLD, the versioned core reports the arithmetic mean of forward
and reverse tokens-per-factor scores. TAALED 0.32 exposes a pooled adjusted
mean-factor-length value, a mean of the two directional mean-factor-length
values, and an Original-style directional score mean. The final-token rule and
zero-denominator policy also matter. A label such as `MTLD` therefore does not
identify a method on its own.

`lexdiv_variant_metrics()` expands these choices in a separate contract:

```{r}
variant_tokens <- rep(
  c("one", "two", "three", "one", "four", "five"),
  12
)
variant_results <- lexdiv_variant_metrics(
  variant_tokens,
  mtld_thresholds = c(0.72, 0.92)
)
variant_results[, c(
  "family", "method_id", "reference_label", "value", "status",
  "comparison_scope"
)]
```

This is a sensitivity crosswalk, not a translation of TAALED code or an
official compatibility layer. The TAALED-relevant rows fix only the stated
formula, sequential factorization, and aggregation. Exact end-to-end comparison
still requires the same token sequence and a separately version-pinned TAALED
run; preprocessing, short-input, and missing-value policies must be compared
independently.

## Frequency is relative to a corpus

Bottini and Le Foll (2025, <https://doi.org/10.1075/ijlcr.23029.bot>) show that
mean-frequency measures can be influenced more by the reference corpus's mode
and register than by learner proficiency. The TUBELEX output is therefore named
as frequency and prevalence, not as an unqualified sophistication score.

```{r}
# Resource-compatible segmentation must be prepared separately.
# This manually prepared sequence illustrates only the sentence shown.
frequency <- tubelex_profile(c("I", "do", "n't", "think", "it", "'s", "a", "book"))
frequency$summary
frequency$coverage
frequency$provenance$formula_parameters
```

The conditional means answer: "Among terms found in this TUBELEX resource,
what is their average frequency or prevalence?" Coverage answers the separate
question: "How much of this text did the resource measure?" Neither should be
reported without the other.

The Unicode tokenizer preserves contractions; the bundled TUBELEX resource
uses Treebank segmentation. A `lexdiv_tokenization` object therefore requires
an explicit `tokenization_mismatch = "allow"` override and is marked as using
a different tokenizer. Plain term vectors carry unverified alignment rather
than a claim of compatibility. Inspect contraction and punctuation handling
in the caller's preparation pipeline even when lookup coverage is 100%.
The optional `normalization = "tubelex_apostrophe"` only harmonizes term-level
apostrophe typography; it is not a segmentation adapter. For several documents,
`tubelex_profile_batch()` shares the verified resource snapshot within the call.

The lookup values are deliberately inspectable. `count`, `videos`, and
`channels` are raw resource counts. Derived scores use the following base-10
formulas. The four normalization constants used to form the three denominators
are stored in `frequency$provenance$formula_parameters`:

```text
zipf = log10(1e9 * (count + 1) /
             (token_total + source_vocabulary_size))
video_prevalence = log10((videos + 1) / (video_total + 2))
channel_prevalence = log10((channels + 1) / (channel_total + 2))
```

Video and channel prevalence are log proportions, so negative values are
expected; values closer to zero denote wider prevalence. Summary means use
matched rows only and must not be interpreted as though unmatched terms had a
frequency of zero.

## Frequency bands are profiles, not mean corpus frequencies

New JACET 8000 was published in 2016.
The [Ishikawa Laboratory vocabulary page](https://language.sakura.ne.jp/s/voc.html)
links the official `j8_2016.xlsx` workbook and explains that compilation used
BNC and COCA evidence plus Japanese educational materials. The workbook labels
the entries as representative lemmas. This is a ranked pedagogical list, not
the same estimand as a TUBELEX mean frequency.

`nj8_profile()` uses the bundled `jacet2016-8000-v1` table by default.
JACET has permitted redistribution with source attribution. Cite:
JACET Basic Word Revision Committee (Ed.). (2016).
*The New JACET List of 8000 Basic Words*. Tokyo: Kirihara Shoten.
The installed `licenses/nj8/NOTICE.md` contains the full attribution and the
three verified corrections to the supplied snapshot. All 8,000 pairs match
the official workbook. The permission does not relicense the table as MIT.

An external copy can still be passed explicitly, for example
`nj8_profile(annotated, "j8_2016.xlsx", unit = "lemma")` with `readxl` installed.
The example below uses a small synthetic table for hand-checkable results.

```{r}
synthetic_nj8 <- data.frame(
  NJ8 = c(1L, 1001L, 6001L, 8000L),
  Word = c("the", "develop", "rare", "extreme")
)
level_profile <- nj8_profile(
  c("The", "develop", "rare", "outside", "the"),
  synthetic_nj8,
  unit = "surface"
)
level_profile$summary
level_profile$coverage
```

The corpus adapter keeps document identity explicit and processes the external
list once. The following synthetic example uses surface units so both documents
share the same input contract without inventing lemma annotations.

```{r}
level_batch <- nj8_profile_batch(
  list(
    document_a = c("The", "develop", "rare"),
    document_b = c("extreme", "outside")
  ),
  synthetic_nj8,
  unit = "surface"
)
level_batch$coverage
level_batch$document_diagnostics
```

The returned `lookup` remains token-lossless and receives a `document_id`.
`max_rows` bounds the combined lookup and summary size before the list is
processed. Shared list provenance is stored once; document-specific
preprocessing provenance remains separate.

```{r, fig.width=7, fig.height=4.5}
plot(level_profile, weighting = "token")
```

The bars are the proportions occurring in each band. The line is the
cumulative proportion at or below each level. Both use all eligible items as
the denominator, including off-list items; therefore the Level 8 cumulative
rate equals list coverage. The off-list bar is deliberately not appended to
the cumulative line. Token and type plots answer different questions, and
surface, lemma, and flemma input can materially change both. With flemma input,
the selected conflict policy and alternative surface-headword rank/level should
also be reported.

Type-weighted off-list rates are especially preprocessing-sensitive: inflected
forms, spelling variants, and learner errors are distinct surface types but may
be collapsed or retained differently by a lemma/flemma backend. A token- and a
type-weighted profile are therefore parallel estimands, not interchangeable
replications of one score.

## Recommended reporting unit

For reproducible research, retain at least:

1. the core metric rows;
2. preprocessing provenance and token-level exclusions;
3. exact method and parameter identities;
4. resource identity and normalization method for frequency profiles;
5. exact/cumulative level-profile denominator and surface/lemma/flemma unit;
6. token/type coverage and unmatched counts.

Factor analysis, PCA, dominance analysis, proficiency prediction, and other
population-dependent models belong in downstream analysis or a research
vignette. They do not change the meaning of the package's versioned measurement
records.
