---
title: "Preprocessing, parameter sensitivity, and frequency coverage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preprocessing, parameter sensitivity, and frequency coverage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
library(ldfreq)
```

## What `ldfreq` adds to the R ecosystem

R already has mature lexical-diversity implementations. In particular,
`quanteda.textstats::textstat_lexdiv()` integrates with `quanteda` tokens and
document-feature matrices, while `koRpus::lex.div()` includes raw-character,
lemma, HD-D, MTLD, and MTLD-MA workflows. `ldfreq` does not claim that these
metrics were absent from R.

The narrower contribution of `ldfreq` is an auditable measurement boundary:

| Question | `ldfreq` representation |
|---|---|
| Which formula and boundary variant ran? | `method_id` and metric contract |
| Which window, segment, or sample was requested? | requested and effective parameter list-columns |
| Was a short-text request computable? | structured `status` and `missing_reason` |
| Did a screen change or censor a value? | no; screens are separate rows |
| Which raw-text decisions created the units? | preprocessing contract and token audit |
| Which resource produced a frequency value? | version, manifest hash, formula, and lookup identity |
| How much of the text matched that resource? | adjacent token/type coverage |

This makes `ldfreq` complementary to broad text-processing ecosystems rather
than a replacement for them.

## Why preprocessing is a measurement decision

Caltabellotta, Van Steendam, Noreillie, and Peters (2026,
<https://doi.org/10.1016/j.asw.2026.101039>) found that lemmatization and the
inclusion of all words versus content words materially changed MATTR and
frequency-based sophistication results. Their empirical best-performing choice
in one English/French study should not be converted into a universal software
default. `ldfreq` therefore exposes the alternatives and their coverage.

```{r}
tokenization <- lexdiv_tokenize("Cats and cat ran run.")
annotated <- lexdiv_lemmatize(
  tokenization,
  lemmas = c("cat", "and", "cat", "run", "run"),
  upos = c("NOUN", "CCONJ", "NOUN", "VERB", "VERB"),
  backend_id = "vignette-fixture",
  backend_version = "1"
)

surface <- lexdiv_metrics_text(tokenization, metrics = "ttr")
lemma_all <- lexdiv_metrics_text(
  annotated,
  unit = "lemma",
  metrics = "ttr"
)
lemma_content <- lexdiv_metrics_text(
  annotated,
  unit = "lemma",
  word_inclusion = "content",
  metrics = "ttr"
)

data.frame(
  operationalization = c("surface/all", "lemma/all", "lemma/content"),
  N = c(surface$results$N, lemma_all$results$N, lemma_content$results$N),
  V = c(surface$results$V, lemma_all$results$V, lemma_content$results$V),
  TTR = c(
    surface$results$value,
    lemma_all$results$value,
    lemma_content$results$value
  )
)
```

The point is not that TTR is preferred. The example demonstrates that the
lexical unit and denominator change the estimand before any diversity formula
is selected.

### Flemma as a caller-supplied resource transformation

For NWLC-oriented sensitivity analysis, `lexdiv_flemmatize()` reads a local
AntBNC lemma-list file supplied by the analyst. The payload is never bundled or
downloaded. Unknown forms use normalized-surface identity fallback so they
remain visible as off-list items, and explicit overrides take precedence over
the raw AntBNC mapping.

```{r, eval=FALSE}
flemma_annotations <- lexdiv_flemmatize(
  tokenization,
  "/path/to/antbnc_lemmas_ver_004.txt",
  overrides = data.frame(
    form = c("interesting", "interested"),
    flemma = c("interesting", "interested")
  ),
  resource_version = "004"
)
lexdiv_metrics_text(flemma_annotations, unit = "flemma", metrics = "ttr")
```

This is an explicit approximation boundary. NWLC reports using a manually
modified AntBNC mapping aligned to its selected word lists. Raw AntBNC can map
a surface form that is an independent New JACET headword to a different family
lemma. `new_jacet8000_profile()` records these conflicts and lets an analysis
retain AntBNC, prefer the word-list headword, or stop for explicit review.

## Length and parameter sensitivity

Bestgen (2024, <https://doi.org/10.1111/lang.12630>) distinguishes ordinary
text-length dependence from sensitivity to the reduction or window parameter.
Bestgen (2025, <https://doi.org/10.1016/j.rmal.2024.100168>) further shows why
MATTR is useful for within-text fluctuation but should not be treated as a
parameter-free universal score.

`lexdiv_plan()` and `lexdiv_profile()` preserve parameter variants as different
requests:

```{r}
long_tokens <- rep(
  c("one", "two", "three", "one", "four", "five", "one", "six"),
  20
)
plan <- lexdiv_plan(presets = "length_50_100")
profile <- lexdiv_profile(long_tokens, plan)
profile[profile$metric_id %in% c("mattr", "msttr"), c(
  "request_id", "metric_id", "value", "status", "N", "V"
)]
```

A downstream study may combine several indices, fit a PCA, or optimize a
window for a target population. Such sample-dependent modeling is not embedded
as a universal package score.

## Maas and MTLD are variant families, not single formulas

The version-pinned [TAALED 0.32 documentation](https://pypi.org/project/taaled/0.32/)
exposes distinctions that a shared label can hide. Its Maas calculation uses
the base-10, a-squared form, whereas the frozen `ldfreq` core uses natural logs
and a-squared. The base-10 value equals the natural-log value multiplied by
`log(10)`: rankings may be unchanged while coefficients, cutoffs, and raw-scale
comparisons are not.

For sequential MTLD, the frozen core reports the arithmetic mean of forward
and reverse tokens-per-factor scores. TAALED 0.32 exposes a pooled adjusted
mean-factor-length value, a mean of the two directional mean-factor-length
values, and an Original-style directional score mean. The final-token rule and
zero-denominator policy also matter. A label such as `MTLD` therefore does not
identify a method on its own.

`lexdiv_variant_metrics()` expands these choices in a separate contract:

```{r}
variant_tokens <- rep(
  c("one", "two", "three", "one", "four", "five"),
  12
)
variant_results <- lexdiv_variant_metrics(
  variant_tokens,
  mtld_thresholds = c(0.72, 0.92)
)
variant_results[, c(
  "family", "method_id", "reference_label", "value", "status",
  "comparison_scope"
)]
```

This is a sensitivity crosswalk, not a translation of TAALED code or an
official compatibility layer. The TAALED-relevant rows fix only the stated
formula, sequential factorization, and aggregation. Exact end-to-end comparison
still requires the same token sequence and a separately version-pinned TAALED
run; preprocessing, short-input, and missing-value policies must be compared
independently.

## Frequency is relative to a corpus

Bottini and Le Foll (2025, <https://doi.org/10.1075/ijlcr.23029.bot>) show that
mean-frequency measures can be influenced more by the reference corpus's mode
and register than by learner proficiency. The TUBELEX output is therefore named
as frequency and prevalence, not as an unqualified sophistication score.

```{r}
frequency <- tubelex_frequency_profile(tokenization)
frequency$summary
frequency$coverage
```

The conditional means answer: "Among terms found in this TUBELEX resource,
what is their average frequency or prevalence?" Coverage answers the separate
question: "How much of this text did the resource measure?" Neither should be
reported without the other.

## Frequency bands are profiles, not mean corpus frequencies

[JACET announced New JACET 8000 as a 2016 publication](https://www.jacet.org/info/%E2%96%A0%E3%80%8E%E5%A4%A7%E5%AD%A6%E8%8B%B1%E8%AA%9E%E6%95%99%E8%82%B2%E5%AD%A6%E4%BC%9A%E5%9F%BA%E6%9C%AC%E8%AA%9E%E3%83%AA%E3%82%B9%E3%83%88-%E6%96%B0jacet8000%E3%80%8F-%E5%88%8A%E8%A1%8C%EF%BC%81/).
The [Ishikawa Laboratory vocabulary page](https://language.sakura.ne.jp/s/voc.html)
links the official `j8_2016.xlsx` workbook and explains that compilation used
BNC and COCA evidence plus Japanese educational materials. The workbook labels
the entries as representative lemmas. This is a ranked pedagogical list, not
the same estimand as a TUBELEX mean frequency.

`new_jacet8000_profile()` accepts a list supplied by the user; `ldfreq` does
not bundle or download it. The example below uses a small synthetic fixture,
not the JACET list.

The downloaded workbook can be passed directly as
`new_jacet8000_profile(annotated, "j8_2016.xlsx", unit = "lemma")` when
`readxl` is installed. Its embedded notice permits personal research use
with citation and asks commercial users to consult JACET in advance. Public
download availability does not itself grant a downstream redistribution
license, so the file stays outside the package.

```{r}
synthetic_nj8 <- data.frame(
  NJ8 = c(1L, 1001L, 6001L, 8000L),
  Word = c("the", "develop", "rare", "extreme")
)
level_profile <- new_jacet8000_profile(
  c("The", "develop", "rare", "outside", "the"),
  synthetic_nj8
)
level_profile$summary
level_profile$coverage
```

The corpus adapter keeps document identity explicit and processes the external
list once. The following synthetic example uses surface units so both documents
share the same input contract without inventing lemma annotations.

```{r}
level_batch <- new_jacet8000_profile_batch(
  list(
    document_a = c("The", "develop", "rare"),
    document_b = c("extreme", "outside")
  ),
  synthetic_nj8,
  unit = "surface"
)
level_batch$coverage
level_batch$document_diagnostics
```

The returned `lookup` remains token-lossless and receives a `document_id`.
`max_rows` bounds the combined lookup and summary size before the list is
processed. Shared list provenance is stored once; document-specific
preprocessing provenance remains separate.

```{r, fig.width=7, fig.height=4.5}
plot(level_profile, weighting = "token")
```

The bars are the proportions occurring in each band. The line is the
cumulative proportion at or below each level. Both use all eligible items as
the denominator, including off-list items; therefore the Level 8 cumulative
rate equals list coverage. The off-list bar is deliberately not appended to
the cumulative line. Token and type plots answer different questions, and
surface, lemma, and flemma input can materially change both. With flemma input,
the selected conflict policy and alternative surface-headword rank/level should
also be reported.

Type-weighted off-list rates are especially preprocessing-sensitive: inflected
forms, spelling variants, and learner errors are distinct surface types but may
be collapsed or retained differently by a lemma/flemma backend. A token- and a
type-weighted profile are therefore parallel estimands, not interchangeable
replications of one score.

## Recommended reporting unit

For reproducible research, retain at least:

1. the core metric rows;
2. preprocessing provenance and token-level exclusions;
3. exact method and parameter identities;
4. resource identity and normalization method for frequency profiles;
5. exact/cumulative level-profile denominator and surface/lemma/flemma unit;
6. token/type coverage and unmatched counts.

Factor analysis, PCA, dominance analysis, proficiency prediction, and other
population-dependent models belong in downstream analysis or a research
vignette. They do not change the meaning of the package's frozen measurement
records.
