---
title: "Connecting corpus frequency with evidence of vocabulary use"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Connecting corpus frequency with evidence of vocabulary use}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ldfreq)
```

This guide develops one learner-focused application. The package also supports
native-speaker corpora, register comparisons and other descriptive corpus
research; see [comparison design](designing-comparisons.html). Those analyses
do not require recognition or recall responses. The same response-comparison
function can also describe scored L1 participant data: supply the appropriate
ID columns through `keys`, such as `c("participant_id", "item_id")`.

## Begin with what a learner can do

A learner selects the right definition in a vocabulary test. Can that learner
recall the meaning when reading, recognize a related derivative, or use the
word appropriately in a sentence? These are separate questions.
[Kremmel and Schmitt (2016)](https://doi.org/10.1080/15434303.2016.1237516)
use *lexical employability* for the ability to employ words in communication.
Their reading-focused study operationalized a prerequisite through meaning
recall, rather than observing fluent reading directly. Recognition, meaning
recall, form recall, derivatives, collocations, and contextual use must not be
collapsed into a single undifferentiated "known word" label.

TUBELEX supplies another kind of evidence: properties of word forms in a
reference corpus. Its original evaluation concerned lexical decision time,
familiarity and lexical complexity, not this learner's ability to use words
([Nohejl et al., 2025](https://aclanthology.org/2025.coling-main.641/)).
The frequency and video/channel counts can help describe or select target words
and serve as candidate predictors. They are not learner response scores.

Native-speaker corpora can supply attested contexts, register-specific use
patterns and candidate items for this research. With suitable contextual
annotation, an observed use can provide evidence of performance on that
occasion. Aggregate frequency cannot show which senses an individual can
retrieve, and an absent form may simply reflect the topic or opportunity to
use it. TUBELEX's English subtitle frequencies do not certify each speaker's
L1 background. Treat its documented source domain separately from participant
language background.

| Observation | What it can describe | What it does not establish alone |
|---|---|---|
| TUBELEX count / Zipf | Frequency of a form in the reference corpus | This learner's exposure or knowledge |
| Video / channel prevalence | How widely a form occurs across source parts | Diversity of word senses or automatic recall |
| Meaning recognition | A scored response with supplied options | Unaided meaning recall |
| Meaning recall | Recall under an explicit prompt and scoring rubric | Speed, comprehension of a passage, or productive use |
| Contextual use / comprehension | Performance in a specified task | Transfer to every task, sense or skill |
| An occurrence in an essay | A form used in that essay | Appropriateness, independent retrieval, or knowledge of absent words |

## Attach corpus features to items, not to inferred ability labels

These four items and all responses below are authored demonstrations, not
observations from learners or a reproduction of the published study. Two items
use the same form, `bank`, for different intended senses. `ldfreq` illustrates
a term outside this fixed reference corpus. Keep item and sense identities even
when the lookup form is identical.

```{r}
items <- data.frame(
  item_id = c("i1", "i2", "i3", "i4"),
  target_form = c("bank", "bank", "accurate", "ldfreq"),
  sense_id = c("financial-institution", "river-edge", "precise", "R-package"),
  context_id = c("authored-context-1", "authored-context-2",
                 "authored-context-3", "authored-context-4")
)
stopifnot(all(vapply(items, function(x)
  is.character(x) && !anyNA(x) && all(nzchar(trimws(x))), logical(1))),
  !anyDuplicated(items$item_id))
item_profile <- tubelex_profile(items$target_form, normalization = "identity")
stopifnot(item_profile$status == "ok",
  identical(item_profile$lookup$query_index, seq_len(nrow(items))),
  identical(item_profile$lookup$term, items$target_form))
item_features <- cbind(items, item_profile$lookup[, c(
  "lookup_term", "matched", "count", "videos", "channels",
  "zipf", "video_prevalence", "channel_prevalence")])
item_features[, c("item_id", "target_form", "sense_id", "matched", "zipf")]
```

The two `bank` items receive the same corpus values. This resource does not
separate their senses. `ldfreq` has missing frequency, which means no match,
not zero knowledge or maximal difficulty. Matching a target form directly is
useful for item features; it does not certify a tokenizer for connected text.
A lemma or word-family score must not be substituted for all of its forms'
frequencies. Summing derivative video/channel counts would double-count source
parts shared by the forms; the aggregate table cannot recover their union.

## Keep the learner-by-item evidence intact

This example uses already scored binary meaning-recognition and meaning-recall
responses. It does not score free text. Real data also need the prompt,
scoring rubric/version, rater or adjudication record, administration order and
missingness reason. Use separate rows/IDs for repeated occasions or senses.

```{r}
responses <- expand.grid(item_id = items$item_id,
  learner_id = c("learner_a", "learner_b", "learner_c"),
  stringsAsFactors = FALSE, KEEP.OUT.ATTRS = FALSE)
responses$recognition <- c(1, 1, 1, 0, 1, 1, 0, 0, 1, 0, 1, NA)
responses$meaning_recall <- c(1, 0, 1, 0, 1, 0, 0, 1, 1, 0, NA, 0)
responses$occasion_id <- "demo"
responses$test_format <- "meaning-recognition"
responses$test_rubric <- "authored-recognition-v1"
responses$criterion_rubric <- "authored-recall-v1"
responses$rater_id <- "authored-scores"
responses$test_missing_reason <- ifelse(is.na(responses$recognition),
  "not-administered", NA_character_)
responses$criterion_missing_reason <- ifelse(is.na(responses$meaning_recall),
  "not-administered", NA_character_)
index <- match(responses$item_id, item_features$item_id)
stopifnot(!anyNA(index),
  !any(setdiff(names(item_features), "item_id") %in% names(responses)))
analysis <- cbind(responses,
  item_features[index, setdiff(names(item_features), "item_id"), drop = FALSE])
rownames(analysis) <- NULL
stopifnot(nrow(analysis) == nrow(responses))
```

Matching by the unique item ID prevents a many-to-many join on `bank` from
silently multiplying observations. Repeated item features do not create more
independent words. Learners are also repeated across items.

## Inspect agreement and its direction before fitting a model

The same overall agreement can conceal different discrepancies. A recognized
word whose meaning was not recalled differs from a recalled word missed in the
recognition test. Following the comparison logic of Kremmel and Schmitt, count
both directions against the stated criterion, without treating the criterion
as an infallible observation of real-world ability.

```{r}
comparison <- lexdiv_compare_responses(analysis,
  test = "recognition", criterion = "meaning_recall")
comparison$counts
comparison$summary
stopifnot(identical(comparison$counts$n, c(4L, 2L, 1L, 3L, 1L, 1L, 0L)))
```

There are 10 complete pairs and two missing pairs. Recognition-only responses
are 2/10 of complete pairs, but 2/6 of recognition-correct complete pairs.
Those denominators answer different questions. `test_only` means recognition
correct and recall incorrect in this call; reversing `test` and `criterion`
reverses that interpretation. Each supplied row contributes once, so these are
pair-weighted proportions, not averages of learner-level proportions.
Missing responses stay missing;
a failure to observe a word in an essay likewise must not be coded as a failed
elicited response. The observed-case proportions may be biased if missingness
is related to knowledge or task difficulty.

Do not select a Zipf cutoff that labels words "employable", equate an NJ8 match
with learner knowledge, or multiply mean frequency and diversity into a new
ability score. Such rules would need independent outcome data and validation.

## Read scored CSV data without losing IDs or missingness

Use two tables: an item dictionary with unique `item_id`, `target_form`,
`sense_id` and `context_id`; and a response table with `learner_id`, `item_id`
and the two scores. Include occasion/task/format IDs, scoring versions, rater
records, presentation order and separate missingness reasons as appropriate to
your design. These columns remain in `$responses`; the function does not verify
that their labels accurately describe the administration or scoring.

The runnable round trip below writes the authored data to a temporary folder.
For your own data, set `input_dir` to the directory containing your two CSVs
and begin at `read.csv()`; do not write the demonstration over real files.
Read IDs as text so, for example, `001` stays distinct from `1`. This convention
uses an empty CSV field for a missing value, and preserves the literal text
`NA` as a possible ID. Scores must be exactly `0`, `1`, or missing.

```{r}
input_dir <- tempfile("vocabulary-example-")
dir.create(input_dir)
write.csv(items, file.path(input_dir, "items.csv"), row.names = FALSE,
  na = "", fileEncoding = "UTF-8")
write.csv(responses, file.path(input_dir, "responses.csv"),
  row.names = FALSE, na = "", fileEncoding = "UTF-8")

items_csv <- read.csv(file.path(input_dir, "items.csv"),
  colClasses = "character", na.strings = "", check.names = FALSE,
  fileEncoding = "UTF-8")
responses_csv <- read.csv(file.path(input_dir, "responses.csv"),
  colClasses = "character", na.strings = "", check.names = FALSE,
  fileEncoding = "UTF-8")
stopifnot(!anyDuplicated(names(items_csv)),
  all(c("item_id", "target_form", "sense_id", "context_id") %in% names(items_csv)),
  !anyDuplicated(names(responses_csv)),
  all(c("learner_id", "item_id", "recognition", "meaning_recall") %in%
        names(responses_csv)))
for (column in c("recognition", "meaning_recall")) {
  score <- responses_csv[[column]]
  stopifnot(all(is.na(score) | score %in% c("0", "1")))
  responses_csv[[column]] <- as.numeric(score)
}
stopifnot(!anyNA(items_csv$item_id), all(nzchar(trimws(items_csv$item_id))),
  !anyDuplicated(items_csv$item_id),
  all(responses_csv$item_id %in% items_csv$item_id))
from_csv <- lexdiv_compare_responses(responses_csv,
  test = "recognition", criterion = "meaning_recall")
stopifnot(identical(from_csv$counts, comparison$counts),
  identical(from_csv$summary, comparison$summary),
  identical(from_csv$responses[names(responses)], responses))
unlink(input_dir, recursive = TRUE)
```

For a real analysis, run the earlier item-profile and ID-matching code with
`items <- items_csv` and `responses <- responses_csv` before comparing the joined
rows. The item checks and `match()` prevent unknown IDs and many-to-many joins.
Do not drop unmatched TUBELEX items or replace their missing frequencies with
zero. Missing item/sense/context identifiers need resolution before making a
sense-specific claim; they are not inferable from a frequency lookup.

Blank scores here mean missing, not incorrect. Whether an administered but
unanswered item receives zero depends on your prespecified rubric; retain its
raw response and reason. Fractional scores are rejected instead of silently
rounded. Scheduled but wholly absent response rows cannot be detected by this
function: reconcile with the administration roster and add explicit missing
rows before reporting a completion denominator.

## Separate formats and repeated occasions

For repeated measurements, extend the composite key and choose the strata
explicitly. Merely adding `by` does not permit duplicate learner/item keys.
This duplicated demonstration is a software example, not evidence of change.

```{r}
repeated <- rbind(transform(analysis, occasion_id = "pre"),
                  transform(analysis, occasion_id = "post"))
by_occasion <- lexdiv_compare_responses(repeated,
  test = "recognition", criterion = "meaning_recall",
  keys = c("learner_id", "item_id", "occasion_id"),
  by = c("occasion_id", "test_format", "test_rubric", "criterion_rubric"))
by_occasion$summary[, c("occasion_id", "quantity", "numerator", "denominator")]
```

If a learner encounters several formats for the same item on one occasion,
include the format ID in `keys` as well. Keep different rubric versions or
criteria separate unless their comparability is justified. An `NA` group label
is retained as a missing-label stratum; it does not identify a common known
condition. Zero cells are retained, zero denominators give `NA`, and there are
no automatic significance tests or confidence intervals that would ignore
repeated learners and items. See `?lexdiv_compare_responses` for the full return
structure.

## Keep frequency, prevalence and coverage together

For document profiles, the diagnostic function produces reusable tables while
retaining failed and empty documents. The following tokens are authored inputs;
their preparation is explicit, and the package still records vector alignment
as unverified. See the [TUBELEX input guide](tubelex-input.html) for connected text.

```{r}
profiles <- tubelex_profile_batch(list(
  a = c("the", "bank", "bank", "ldfreq"),
  b = c("an", "accurate", "description"), blank = character(),
  invalid = c("bank", NA_character_)))
audit <- tubelex_diagnostics(profiles)
audit$summary[, c("document_id", "status", "weighting", "eligible_items",
                  "matched_items", "coverage", "mean_zipf")]
audit$unmatched_terms
audit$documents[, c("document_id", "tokenization_alignment", "query_normalization")]
```

The means remain conditional on matches. `unmatched_terms` contains confirmed
non-matches only; a resource failure is unresolved, not evidence of absence.
Token and type weighting remain separate. When input/query settings differ,
`condition_id` keeps aggregate counts apart; matching settings do not establish
validity or tokenizer equivalence. Do not pool unlike conditions as one corpus.

## Design the empirical validation around the claim

An initial study can ask whether frequency and channel prevalence help explain
meaning-recall performance beyond a recognition response on the same target
sense. This tests an association with the chosen recall criterion. A further
study needs performance in an appropriate reading, listening, speaking or
writing task, with accuracy and, where relevant, processing time. Derivative
and collocation knowledge require their own evidence.

Retain learner, item, target sense, task, occasion and rater IDs. Balance item
format and administration order to limit cueing between recognition and recall;
do not reuse this toy example as a test design. Predefine acceptable answers,
partial credit, missingness and any time criterion, then check independent
rating and adjudication. Start with a limited model such as a binomial model
with learner and item effects, provided the sample supports it. Compare a
frequency-only model with an explicitly justified addition of channel range;
frequency, video range and channel range can be strongly related. Evaluate
calibration and held-out prediction, with splits that match the intended claim
about new learners, new words, or both. Randomly splitting response rows leaks
both learner and item information across sets.

The ICNALE essay data used in package development do not include the paired
learner-by-word recognition/recall and contextual-use criteria needed for this
claim. Its general writing ratings can support a different question, but they
cannot establish employability of particular words. The package currently
supports the descriptive inputs and audit trail, not a validated employability
measure or an automatic response scorer.

```{r}
saved <- tempfile(fileext = ".rds")
saveRDS(list(items = items, responses = responses, item_profile = item_profile,
  analysis = analysis, comparison = comparison, profiles = profiles, audit = audit,
  scoring = list(example = "authored-not-empirical", version = "1"),
  session = sessionInfo()), saved)
restored <- readRDS(saved)
stopifnot(identical(restored$analysis, analysis),
  identical(lexdiv_compare_responses(restored$analysis,
    "recognition", "meaning_recall"), restored$comparison),
  identical(tubelex_diagnostics(restored$profiles), restored$audit))
unlink(saved)
```

Use a persistent destination for actual research. Save full records before
exporting display tables to CSV; the CSV alone cannot retain the corpus and
scoring provenance.
