---
title: "Word families, roots, and affixes: from texts to interpretable profiles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Word families, roots, and affixes: from texts to interpretable profiles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

An essay can repeat one family through several forms, or contain many words
with the same suffix. These observations answer different questions. This
tutorial shows how to count them while retaining the words, reference entries,
unresolved cases and decisions behind each result.

Work through the examples in order in a fresh R session. Use the
[documented installation](../index.html#installation). The counting examples
run offline in R; their displayed tables use `knitr` (install it once with
`install.packages("knitr")` if needed). The final KWIC exercise also requires the optional
`quanteda` package and a UTF-8 R session. The text, token boundaries, lemmas and
decisions below are authored teaching material, not learner observations or
independently validated annotations. Small tables make the arithmetic visible;
their TTR values are not recommendations for comparing short and long essays.

```{r morphology-setup}
library(ldfreq)
recipes <- new.env(parent = baseenv())
sys.source(system.file("examples", "family-count-comparison.R",
  package = "ldfreq", mustWork = TRUE), recipes)
sys.source(system.file("examples", "word-parts.R",
  package = "ldfreq", mustWork = TRUE), recipes)
sys.source(system.file("examples", "morpholex-word-parts.R",
  package = "ldfreq", mustWork = TRUE), recipes)
has_review <- requireNamespace("quanteda", quietly = TRUE) &&
  isTRUE(l10n_info()[["UTF-8"]])
linked <- NULL
```

The functions loaded into `recipes` are explicitly sourced example helpers,
not exported package APIs. They reuse complete imported annotations and
declared reference tables; they do not run a morphological analyzer.

## What changes when words are counted as families?

Start with two short word lists and an empty document. The first list makes
the counting units easy to compare. The second includes a word absent from
the chosen family inventory. The supplied lemmas leave `colour` and `color`
distinct so that spelling normalization is not silently mixed with lemmatization.

```{r family-input}
segments <- data.frame(
  document_id = c("listed", "unlisted", "empty"), segment_id = "s1",
  text = c("use uses colour color", "use reusability", ""))
tokens <- data.frame(
  document_id = c(rep("listed", 4), rep("unlisted", 2)),
  segment_id = "s1", token_index = c(1:4, 1:2),
  surface = c("use", "uses", "colour", "color", "use", "reusability"),
  lemma = c("use", "use", "colour", "color", "use", "reusability"))
annotations <- lexdiv_import_annotations(tokens, segments,
  list(language = "en", analyzer = "authored", analyzer_version = "1",
    dictionary = "authored lemmas", dictionary_version = "1",
    unit = "word", normalization = "none"))
nation <- bnccoca_data()
families <- lexdiv_family_profile(annotations, nation$dictionary,
  nation$resource, normalization = "nfkc_lower")
counts <- recipes$compare_family_counts(families)
knitr::kable(counts$comparison[counts$comparison$scope == "all_selected",
  c("document_id", "unit", "N", "V", "ttr", "status")], digits = 2,
  row.names = FALSE, col.names = c("Document", "Unit", "N", "V", "TTR", "Status"))
```

For `use uses colour color`, **four occurrences remain four tokens**. There
are four surface types, three supplied lemma types and two Nation families.
Grouping changes the type count, not the number or order of source occurrences.
A family can contain derivational relatives; a lemma or flemma is a different
counting unit. An inventory specifies which relatives are included.

The bundled Nation BNC/COCA Level 6, Version 1.0.0 inventory does **not** list
`REUSABILITY`. Therefore, the second document has an unavailable whole-document
family total and TTR, even though a researcher could define another inventory
that includes it in USE. Empty documents have zero types and undefined TTR.

### Check coverage before interpreting a smaller count

```{r family-coverage}
knitr::kable(counts$documents[c("document_id", "selected_tokens",
  "matched_tokens", "unlisted_tokens", "common_tokens", "token_coverage")],
  digits = 2, row.names = FALSE,
  col.names = c("Document", "Tokens", "Matched", "Unlisted", "Common subset", "Coverage"))
knitr::kable(counts$comparison[
  counts$comparison$document_id == "unlisted" &
    counts$comparison$scope == "common_resolved",
  c("unit", "N", "V", "ttr")], row.names = FALSE)
o <- families$occurrences
knitr::kable(o[o$status == "unlisted",
  c("document_id", "pre", "keyword", "post", "status")], row.names = FALSE)
```

The common resolved subset contains only `use`: all three units have N = 1,
V = 1 and TTR = 1. That is a result **conditional on one retained occurrence**,
not the family TTR of `use reusability`. Report coverage of the original
selection beside the comparison. Unlisted does not mean misspelled or unknown
to the writer. Do not remove unresolved words and close the resulting gaps for
MATTR or n-grams; the common subset here supports type counts and TTR only.

Nation's inclusion **Level 6** is not frequency **band 6**. Nor does family
membership establish that a learner knows every member. The inventory and its
CC BY-SA 4.0 attribution/ShareAlike terms are retained in `nation`; see the
[resource details](preprocessing-and-frequency.html#use-nations-bundled-bnccoca-inventory).

## Which roots and affixes occur in the text?

Use a separate, explicitly authored analysis to learn how part counts work.
Its five lexical tokens are `teachers smaller reuse transmit transmission`.
Punctuation remains in the imported source and is excluded from the count.

```{r authored-parts}
sys.source(system.file("examples", "word-parts-demo.R",
  package = "ldfreq", mustWork = TRUE), recipes)
parts <- recipes$word_parts_example$profile
knitr::kable(parts$occurrences[parts$occurrences$document_id == "parts" &
  parts$occurrences$selected %in% TRUE,
  c("surface", "observed_root_occurrences", "observed_prefix_occurrences",
    "observed_suffix_occurrences", "observed_inflection_occurrences",
    "observed_derivation_occurrences")],
  col.names = c("Form", "Roots", "Prefixes", "Suffixes", "Inflection", "Derivation"),
  row.names = FALSE)
```

Here, `teachers` has the root `teach`, derivational suffix `-er` and
inflectional suffix `-s`. The comparative `-er` in `smaller` has a different
part ID and function. Prefix/suffix describes **position**; inflection/derivation
describes **process**. They are two classifications of the same affix, not
four counts to add together.

```{r part-evidence}
knitr::kable(parts$frequency[parts$frequency$document_id == "parts",
  c("part_id", "canonical", "role", "process", "observed_occurrences")],
  col.names = c("Part ID", "Form", "Position", "Process", "Occurrences"),
  row.names = FALSE)
knitr::kable(parts$documents[c("document_id", "selected_tokens",
  "complete_coverage", "observed_affix_occurrences", "affix_occurrences",
  "affixes_per_token")], digits = 2, row.names = FALSE,
  col.names = c("Document", "Tokens", "Complete coverage", "Observed affixes",
    "Total affixes", "Affixes/token"))
# Follow a count back to its original token and supplied analysis.
knitr::kable(parts$part_occurrences[
  parts$part_occurrences$surface == "teachers",
  c("document_id", "token_index", "surface", "part_id", "canonical",
    "role", "process")], row.names = FALSE,
  col.names = c("Document", "Token", "Word", "Part ID", "Form", "Position", "Process"))
```

The five tokens contain seven affix occurrences: 7/5 = 1.4 affixes per token.
All five tokens contain an affix, giving a different statistic, 5/5 = 100%.
The repeated `trans-` counts at each occurrence. A part type is a distinct
declared part ID, whereas an occurrence is an instance at an original token.
The uncertain document contains ambiguous, unanalysed and unlisted forms;
its complete total is `NA`, not zero.

This example explicitly groups `mit`/`miss` as realizations of one bound root.
Its alternative analysis leaves `transmit` undivided, producing six affix
occurrences rather than seven on the **same five tokens**. These are declared
analytical alternatives, not evidence of how a learner mentally represents
the words. The helper counts supplied parts; it does not reconstruct a
derivation tree, infer semantic transparency or diagnose irregular inflection.

## What does each reference resource contribute?

| Resource | Question it can address | What a row does not establish |
|---|---|---|
| Nation BNC/COCA | Which listed forms belong to an educational family? | A complete segmentation or knowledge of every member |
| MorphoLex | What derivational segmentation and morphological variables does this reference record? | All inflections or the writer's mental decomposition |
| MorphyNet | Which source-to-target formation relations are recorded? | A unique full decomposition or Nation family membership |

### Inspect recorded segmentation before counting it

```{r morpholex-reference}
forms <- c("transmit", "transport", "transmission", "teacher", "teachers")
morpholex <- recipes$read_morpholex_parts(
  sheets = c("0-1-0", "0-1-1", "1-1-0"), words = forms)
knitr::kable(morpholex$analyses[c("form", "completeness")], row.names = FALSE)
source_words <- do.call(rbind, lapply(morpholex_data(
  c("0-1-0", "0-1-1", "1-1-0"))$sheets,
  function(x) x[c("Word", "MorphoLexSegm")]))
knitr::kable(source_words[source_words$Word %in% forms,
  c("Word", "MorphoLexSegm")], row.names = FALSE)
```

In this reference, `transmit` is undivided, `transport` is `trans- + port`,
and `transmission` uses canonical `transmit + -ion`. Both `teacher` and
`teachers` use `teach + -er`: the reference describes derivation and omits
the plural inflection. Do not interpret that omission as a zero count of
all inflectional morphology. A complete analysis is complete only within
its declared scope. The MorphoLex adapter preserves unsupported entries as
unanalysed and records requested forms missing from the selected sheets.

The bundled MorphoLex-en data use **CC BY-NC-SA 4.0**. Retain their citation
and noncommercial/ShareAlike conditions when using or adapting them. See the
[MorphoLex guide](preprocessing-and-frequency.html#use-the-bundled-morpholex-reference)
for source variables, provenance and an optional local-workbook route.

### Keep alternative formation relations separate

```{r morphynet-reference}
morphynet <- morphynet_read_derivations(system.file("extdata",
  "morphynet-example", "eng.derivational.example.tsv", package = "ldfreq",
  mustWork = TRUE), language = "en", resource_version = "Nine-row teaching excerpt")
r <- morphynet$relations
knitr::kable(r[r$target_word == "reusability",
  c("source_word", "target_word", "morpheme", "affix_position")], row.names = FALSE)
```

The incoming relations use `reuse + ability`, `reusable + ity`, and
`usability + re`. Three relations are not three affixes to sum within one
occurrence. Relations can coexist; even a full sentence need not determine
a uniquely correct derivation. A study must state what selecting a relation
means. The nine-row, CC BY-SA 3.0 teaching excerpt is not suitable for corpus
coverage estimates. For research, explicitly obtain and record the
[complete reference](preprocessing-and-frequency.html#inspect-morphynet-formation-relations).

## How do I record a judgment without changing the source?

This exercise uses two authored sentences, an unlisted form and an empty
document. Their lowercase spelling is intentional. Original surfaces are
retained: the morphology recipe uses exact surface matching, while Nation
lookup has its separately declared normalization. No POS or lemma is inferred.

The following three chunks require `quanteda` in a UTF-8 session. The core
counting and resource inspection above do not. If `has_review` is false,
install the optional package as needed and check `l10n_info()` before this
exercise; there is no automatic environment change.

```{r link-input, eval=has_review}
sys.source(system.file("examples", "morphology-link.R",
  package = "ldfreq", mustWork = TRUE), recipes)
review_segments <- data.frame(document_id = c("a", "b", "unlisted", "empty"),
  segment_id = "s1", text = c("the teacher discussed reusability.",
    "the teachers questioned reusability.", "quux.", ""))
review_tokens <- data.frame(document_id = c(rep("a", 5), rep("b", 5), rep("unlisted", 2)),
  segment_id = "s1", token_index = c(1:5, 1:5, 1:2),
  surface = c("the", "teacher", "discussed", "reusability", ".",
    "the", "teachers", "questioned", "reusability", ".", "quux", "."),
  upos = c(rep(c("DET", "NOUN", "VERB", "NOUN", "PUNCT"), 2), "X", "PUNCT"))
review_annotations <- lexdiv_import_annotations(review_tokens, review_segments,
  list(language = "en", analyzer = "authored", analyzer_version = "1",
    dictionary = "authored POS", dictionary_version = "1", unit = "word",
    normalization = "none"))
review_family <- lexdiv_family_profile(review_annotations, nation$dictionary,
  nation$resource, normalization = "nfkc_lower", exclude_pos = "PUNCT")
sheets <- grep("^[0-9]+-[0-9]+-[0-9]+$",
  names(morpholex_data()$sheets), value = TRUE)
review_morpholex <- recipes$read_morpholex_parts(sheets = sheets,
  words = unique(review_tokens$surface[review_tokens$upos != "PUNCT"]))
initial <- recipes$link_morphology_candidates(review_family,
  review_morpholex, morphynet)
knitr::kable(initial$occurrences[initial$occurrences$surface == "reusability",
  c("document_id", "pre", "keyword", "post", "morphynet_candidate_count")],
  row.names = FALSE, col.names = c("Document", "Left", "Word", "Right", "Candidates"))
```

Record one selected MorphyNet relation and one explicit unresolved decision.
Here selection is an **authored reporting convention**, not a claim that
the sentence proves a unique derivation. Candidate IDs come from the reference;
occurrence IDs distinguish the two uses of the same spelling.

```{r link-decisions, eval=has_review}
review <- initial$reviews$morphynet
decisions <- review$occurrences[review$occurrences$surface == "reusability",
  c("review_id", "occurrence_id")]
decisions$status <- c("selected", "unresolved")
chosen <- r$relation_id[r$target_word == "reusability" &
  r$source_word == "reusable" & r$morpheme == "ity"]
stopifnot(length(chosen) == 1L)
decisions$candidate_id <- c(chosen, NA_character_)
decisions$reviewer <- "authored-tutorial"
decisions$reason <- c("Report the reusable + ity relation in this example",
  "Keep alternative relations without selecting one")
decision_file <- tempfile(fileext = ".csv")
write.csv(decisions, decision_file, row.names = FALSE, na = "", fileEncoding = "UTF-8")
submitted <- read.csv(decision_file, colClasses = "character", na.strings = "",
  fileEncoding = "UTF-8", check.names = FALSE)
linked <- recipes$link_morphology_candidates(review_family,
  review_morpholex, morphynet, morphynet_decisions = submitted)
knitr::kable(linked$reviewed_relations[c("document_id", "surface",
  "source_word", "target_word", "morpheme", "affix_position")], row.names = FALSE,
  col.names = c("Document", "Word", "Source", "Target", "Morpheme", "Position"))
knitr::kable(linked$documents[c("document_id", "eligible_N",
  "morphynet_selected_N", "morphynet_unresolved_N", "selected_suffix_relations")],
  row.names = FALSE,
  col.names = c("Document", "Eligible tokens", "Selected", "Unresolved", "Suffix relations"))
```

Only the selected occurrence contributes one suffix **relation**. The second
occurrence stays unresolved; all other candidates stay unreviewed. Nation
membership is unchanged. MorphoLex has not been reviewed in this exercise,
so no complete-segmentation judgment or whole-document affix total is claimed.
Lookup and review counts remain separate in `linked$documents`.

The short CSV round trip above writes only submitted decisions. For editing
an entire worksheet in a spreadsheet, use the
[complete worksheet tutorial](ambiguity-review.html#edit-decision-worksheet):
it checks fixed IDs, missing/duplicate rows and partly filled decisions, then
returns only submitted rows. Start that workflow from
`initial$reviews$morphynet` (or `$morpholex`) and pass its returned decisions to
the corresponding `morphynet_decisions` (or `morpholex_decisions`) argument.
Decisions replace the supplied decision set; resume from the latest saved set
to retain earlier choices. Changed sources or references require a new review.

```{r link-identity, eval=has_review}
stopifnot(identical(initial$occurrences[c("document_id", "segment_id",
  "token_index", "surface", "family_id")],
  linked$occurrences[c("document_id", "segment_id", "token_index", "surface", "family_id")]))
```

## How do I use my own texts and report the results?

Use the [file-input tutorial](english-tokenization.html#import-text-files)
to obtain a document-ID/text table. Family and morphology recipes take
**complete imported annotations**, not a `lexdiv_tokenization` object or
filtered vocabulary list. Prepare a segment table containing `document_id`,
`segment_id`, `text`, and a token table containing those IDs, `token_index`,
`surface`, plus the annotations you actually obtained. Retain punctuation and
excluded tokens as context; choose exclusions during analysis. Include empty
documents in the segment table. Check alignment with the
[annotation import guide](annotation-alignment.html).

The input below names files you supply; it is not executed as part of the
teaching example. Declare the actual analyzer/dictionary versions. Lemma-based
comparisons require a supplied lemma column; missing values are not guessed.

```{r own-annotations, eval=FALSE, purl=FALSE}
segments <- read.csv("segments.csv", colClasses = "character",
  na.strings = character(), fileEncoding = "UTF-8", check.names = FALSE)
tokens <- read.csv("tokens.csv", colClasses = "character",
  na.strings = "", fileEncoding = "UTF-8", check.names = FALSE)
tokens$token_index <- as.numeric(tokens$token_index)
# Blank source text stays an empty string; blank token annotations are missing.
# Replace these labels with the actual tool and reference used for these files.
provenance <- list(language = "en", analyzer = "YOUR ANALYZER",
  analyzer_version = "YOUR VERSION", dictionary = "YOUR DICTIONARY",
  dictionary_version = "YOUR VERSION", unit = "word", normalization = "none")
annotations <- lexdiv_import_annotations(tokens, segments, provenance)
families <- lexdiv_family_profile(annotations, nation$dictionary,
  nation$resource, normalization = "nfkc_lower", exclude_pos = c("PUNCT", "SYM"))
counts <- recipes$compare_family_counts(families)
```

Keep form/lemma/family comparisons on explicitly shared occurrences, and keep
the original denominator and coverage beside them. For larger samples, show
the document values for each unit on the same scale; do not interpret totals
from different coverage subsets as a lexical-unit effect. Affix counts are
discrete: show counts and source forms before considering distribution plots.
Short-text type counts do not by themselves measure morphological productivity,
and neither reference membership nor observed use establishes learner knowledge.

Save the complete objects as well as any convenient CSV table. This executable
example uses a temporary directory; choose a persistent project directory for
your research. If the optional review was skipped, `linked` is `NULL`.

```{r save-morphology}
output_dir <- file.path(tempdir(), "ldfreq-morphology-tutorial")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
bundle <- list(families = families, counts = counts, nation = nation,
  authored_parts = parts, morpholex = morpholex, morphynet = morphynet,
  linked = linked, session = sessionInfo())
saveRDS(bundle, file.path(output_dir, "analysis.rds"))
write.csv(counts$comparison, file.path(output_dir, "family-counts.csv"),
  row.names = FALSE, na = "", fileEncoding = "UTF-8")
saved <- readRDS(file.path(output_dir, "analysis.rds"))
stopifnot(identical(recipes$compare_family_counts(saved$families), saved$counts))
stopifnot(identical(do.call(recipes$word_parts_profile,
  saved$authored_parts$inputs), saved$authored_parts))
if (!is.null(saved$linked)) {
  stopifnot(identical(do.call(recipes$link_morphology_candidates,
    saved$linked$inputs), saved$linked))
}
```

For reporting, identify the source population and document selection, token
boundaries, counting unit, normalization/exclusions, reference version and
license, original denominator, coverage, unresolved states and any review
criteria. Keep resource frequency bands, part occurrences, selected formation
relations and participant knowledge measures in separate columns. The saved
object retains inputs and provenance; CSV output is a convenient view, not
a replacement for that record.

Source descriptions and attribution are available in `?bnccoca_data`,
`?morpholex_data`, `?morphynet_read_derivations` and the installed resource
notices. For an alternative family inventory, partial analyses, repeated roots,
POS restrictions and the complete table contracts, see the
[detailed family and morphology reference](preprocessing-and-frequency.html#resource-defined-word-families).
